### 2012-01-20 notes.tex
initial commit of notes

\pagestyle{empty}
\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx, color}
\usepackage{amsmath}
%\usepackage[small,compact]{titlesec} 
%\usepackage[small,it]{caption}

\textwidth 6.5in
\headheight .1in
\topmargin -.2 in
\textheight 8.7in
\oddsidemargin 0in
\evensidemargin 1in

\begin{document}

 \centerline{\Large \bf Blockmodels for Dynamic Network Data} 
 \medskip
\centerline{\bf Christopher DuBois}
 \bigskip

\section{Introduction}

Viewing interactions among entities as a network is an increasingly useful approach for studying  phenomenon ranging from people communicating online to protein interactions. One class of statistical models for network data use latent variables for each node to help capture unobserved heterogeneity.  For example, the stochastic blockmodel \cite{Nowicki2001} and its nonparametric extension \cite{Kemp2006} assume each node in the network belongs to some block (or cluster) and models the probability of edges between blocks.   Such approaches are especially useful for large-scale network data where higher order dependencies can cause models such as exponential random graph models to be too complex to fit.%\cite{Schweinberger2011}.

Network data, however, is often collected as a sequence of events occurring over time.   Recently several works have adapted models from survival analysis and event history analysis to provide continuous time models of network-based event data \cite{Butts2008,Brandes2009,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011}.  These models are often semiparametric and allow one to specify the dependence of the process on the previous history of events.  In this way we can incorporate theories about the underlying processes and make predictions about future data conditioned on the past.

The above continuous time network models have two major drawbacks: 1) the entire network is assumed to have identical dynamics, and 2) the occurrence of any event can affect the rate at which any other event occurs.  Both aspects may be unrealistic.  Consider email communication in a small organization comprised of several teams.  Each team may have a different pattern for collaborating via email, and in many cases emails within one team will not affect emailing behavior within a separate team.

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction with the rest of the network.  This combines a latent variable model with a local dependence model for modeling the event sequence among nearby nodes, a view which reveals its relationship to hierarchical extensions to latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.

\section{Model}

%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.  We can then write the likelihood of $M$ events as
\begin{align}
\mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \\
&= \prod_{m=1}^M \lambda(t_m|\cdot) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}
\end{align}
\noindent where the $m$th event occurs at time $t_m$ and the intensity function $\lambda(t)$ is right continuous.
% \begin{align}
% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
% \end{align}

In general, we may have different types of events that can occur.  Let  the \emph{risk set}  $\mathcal{R}$ be the set of possible events.   In the context of relational events occurring among $N$ nodes, for example, we let $\mathcal{R}$ include all dyads.\footnote{We exclude self-loops.}  Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ which is piecewise constant.

For network data, we capture the notion that some events have no effect on the intensity of other events by assuming 1) each node $i$ belongs to some latent cluster $z_i$, and 2) each intensity function $\lambda_{ij}(t)$ may only change following an event involving a member of either $z_i$ or $z_j$.  See Figure \ref{fig:example}.

 In this case, the likelihood becomes

\begin{align}
\mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R_{k_1,k_2}}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
\end{align}

\begin{figure}
 \def\svgwidth{6in}
  \input{example.pdf_tex}
\caption{Illustration of the event data and the model assumptions.  Left: An event sequence among four nodes belonging to two clusters (circles and squares): (4,3) occurs at time $t_1$ and (2,1) occurs at time $t_2$.  Right: The intensity functions $\lambda_{21}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ only can change when events occur involving a node in its cluster (square).}
\label{fig:example}
%\includegraphics[width=6in]{example}
\end{figure}

For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $s_{ij}(t|A_t,\mathbf{z})$ that is a function of the previous history of events $A_t$ and the latent cluster assignments $z$.  We model the intensity functions via a log linear model conditioned on this information
\begin{align}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}_{ij}(t|\mathcal{A}_t,\mathbf{z}).
\end{align}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\beta_{z_i,z_j}$ that corresponds to the covariate vector $s_{ij}(t|\cdot)$.  This is reminiscent of the stochastic blockmodel \cite{Nowicki2001, Kemp2006} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.  Here, however, we use a blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

TODO: Discuss possible variations on the prior structure of $\beta$: 1. Enforce $\beta_{k,l} = \beta_{l,k}$. 2. Assume $\beta_{k,l,p} \sim \mbox{Normal}(\beta_p,\sigma_p^2)$. 3. Place other priors that provide different kinds of regularization.

TODO: Discuss decision to condition on $z$ for covariate vector $s_{ij}$.

TODO: Discuss choice of $\log \lambda$ instead of just $\lambda$.

TODO: Discuss choice against time-varying effects.

TODO: Discuss if the semiparametric partial likelihood approach can be used in this framework.

\subsection{Model specification}

We include a similar set of statistics in $s_{ij}$ as used in \cite{Butts2008,Vuy2011}.
Note that the model assumptions imply $s_{ij}(t|\mathcal{A}_t,\mathbf{z}) = f(\mathcal{A}$.

Let $r$ be the last event from a $z_i$ node to a $z_j$ node, that is $r = \mbox{argmax}_{r \in \{1, \ldots, m\}}I(t_r < t, z_{i_r} = z_i, z_{j_r} = z_j)$

\begin{itemize}
\item In and out degree of sender and receiver
\item Reciprocity
\item Transitivity, triangle closure
\item Shared contacters, contactees
\end{itemize}

TODO: Would like things that combine shared contacts with recency. (i.e. Facebook-like terms.)

\section{Inference}

We use Markov chain Monte Carlo to get samples from the posterior distribution of our parameters.

\begin{enumerate}
\item For $i \in [1,N]$: Sample $z_i$ via Gibbs sampling.
\item For each pair $(k,k')$: Sample $\beta_{k,k'}$ via MH or HMC.
\end{enumerate}

To perform Gibbs sampling for a given cluster assignment $z_i$, we need to compute the likelihood for each possible assignment of $z_i$.  This involves recomputing each $\boldsymbol{\tau}$ and possibly each $\mathbf{s}_{ij}$ if it also depends on $\mathbf{z}$.

TODO: Decide how to handle label switching.  Schweinberger has an interesting approach I hadn't seen before for his hierarchical ERGMS.  Should look into that more closely...

\subsection*{Bookkeeping considerations}
%For each dyad, we need to keep track of the set of piecewise constant intervals $d_{ij}$.  For a given set of assignments $z$ we can compute these intervals $(a_v,b_v)$.

Recall we let intensities change whenever an event occurs involving a fellow cluster member.  Thus the vector $\boldsymbol{\tau}_{ij}$ of changepoints (or knots) is given by
$$\boldsymbol{\tau}_{ij} = \left\{ t_m: z_{i_m} = z_i \ \mbox{or} \ z_{j_m} = z_j, m=1,\ldots,M\right\}$$
Any time we update we change node $k$'s cluster assignment from $a$ to $b$, the change point vectors for any nodes in $a$ and $b$ must be updated.  For now we assume we have to recompute  $\boldsymbol{\tau}_{ij}$  entirely.  Similarly, if the covariate vectors $\mathbf{s}_{ij}(t)$ depend on $\mathbf{z}$ then we need to recompute those as well prior to computing the likelihood.%More specifically, for a $D_{ij}$ such that $z_i, z_j \in \{a,b\}$, we must add knots $t_m$ for each.

\section{Simulation}

We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters.  The first cluster has an increased tendency for reciprocity; members of the second cluster have an increased tendency to continue speaking; any contact from the other group instigates discussion within the group.  Then the specification of each covariate vector is $s_{ij}(t) = [r_1(t,i,j), r_3(t,i,j), r_5(t,i,j)]$.  The parameter vectors $\beta_{1,1} = (1,0,0)$,  $\beta_{1,2} = \beta_{2,1} = (0,0,1)$, and $\beta_{1,1} = (0,1,0)$.

Simulation from the model follows the typical racing exponentials idea with a few subtleties.

\begin{enumerate}
\item $t_0 = 0$, all $s_{ij} = 1$
\item For $i = 1, \ldots, N$, draw $z_i \sim \mbox{Categorical}(\theta)$
\item For $m = 1, \ldots, M$:
  \begin{enumerate}
  \item For all $(i,j)$ such that $z_i,z_j \in \{z_{i_{m-1}},z_{j_{m-1}}\}$:
    \begin{enumerate}
    \item Compute $s_{ij}^*$ using events $1$ through $m-1$
    \item Set $\lambda_{ij}^* \leftarrow \exp\{ \beta_{z_i,z_j} s_{ij}^*\}$ 
    \end{enumerate}
  \item Draw $t_m \sim \mbox{Exponential}\left(\sum_{ij} \lambda_{ij}^*\right)$
  \item Draw $(i_m,j_m) \sim \mbox{Categorical}\left(\lambda_{ij}^* / \sum_{ij}\lambda_{ij}^*\right)$
  \item Save $s_{ij}(t_m) \leftarrow s_{ij}^*$
  \item Save $\lambda_{ij}(t_m) \leftarrow \lambda_{ij}^*$
  \end{enumerate}
\end{enumerate}

TODO: Show statistics relevant to these three effects for interactions within each cluster and interactions between the clusters.  This will show that the simulated data are reasonable given the model.

TODO: Show loglikelihood as a function of iteration to confirm it increases.

TODO: Show cluster assignments reasonable: proportion of times true cluster members co-occur.

TODO: Show parameter bias as a function of iteration; perhaps also as a function of the number of events in the dataset.

\section{Model checking and experiments}

\subsection*{Data}

Each of the following data sets are sequences of dyadic events, where each event has a time associated with it, a sender, and a recipient.

\begin{itemize}
\item Classroom:  Perhaps within a couple of the classrooms we can infer subgroups who tend to obey different relational event dynamics.
\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
\item Citation networks: Perhaps some groups of papers obey some types of rich-get-richer effects more than others.
\item Other
\end{itemize}

\subsection*{Prediction experiment}
TODO: Prediction experiment for the predicting the next event conditioned on the past.  We want to measure recall.  Compare to fitting the relational event model without the blockmodel structure.  Compare to relational event blockmodel without covariates: $$p(y_{ij}) = \exp\{\beta_{z_i,z_j}\} / \sum_{(i',j') \in \mathcal{R}}\exp\{\beta_{z_{i'},z_{j'}}\} $$.
Or more simply, just use $s_{ij}(t) = 1$.

TODO: Need to see how well the time-portion of the model is specified.

\subsection*{Posterior predictive simulations}

\section{Discussion}
Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches \cite{Meek2011} aim to learn the regions where an intensity is constant.

Possible future work could extend the present framework to a mixed-membership model where for each effect there is a latent cluster assignment.  This way a group of nodes might contact high in-degree nodes at a similar rate, though they vary in the degree to which they adhere to reciprocity.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}

### 2012-01-21 notes.tex
Change model to be have more independence assumptions.  Sketch out some c++ code to get a sense of how long likelihood computation will take.

@@
 
-Viewing interactions among entities as a network is an increasingly useful approach for studying  phenomenon ranging from people communicating online to protein interactions.
-One class of statistical models for network data use latent variables for each node to help capture unobserved heterogeneity.
+Methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions.
+One class of statistical models for network data uses latent variables for each node to help capture unobserved heterogeneity.
 For example, the stochastic blockmodel \cite{Nowicki2001} and its nonparametric extension \cite{Kemp2006} assume each node in the network belongs to some block (or cluster) and models the probability of edges between blocks.
-Such approaches are especially useful for large-scale network data where higher order dependencies can cause models such as exponential random graph models to be too complex to fit.%\cite{Schweinberger2011}.
+Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
 
@@
 Recently several works have adapted models from survival analysis and event history analysis to provide continuous time models of network-based event data \cite{Butts2008,Brandes2009,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011}.
-These models are often semiparametric and allow one to specify the dependence of the process on the previous history of events.
-In this way we can incorporate theories about the underlying processes and make predictions about future data conditioned on the past.
+These models allow one to specify the dependence of the process on the previous history of events.
+In this way one may incorporate theories about the underlying processes and make predictions about future data conditioned on the past.
 
@@
 
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction with the rest of the network.
-This combines a latent variable model with a local dependence model for modeling the event sequence among nearby nodes, a view which reveals its relationship to hierarchical extensions to latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
+Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.
+This combines a latent variable with a local dependence model for modeling the event sequence among nearby nodes.
+
+TODO: this view which reveals relationships between our approach and hierarchical extensions to latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
 
@@
 \begin{align}
-\mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \\
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \\
 &= \prod_{m=1}^M \lambda(t_m|\cdot) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}
 \end{align}
-\noindent where the $m$th event occurs at time $t_m$ and the intensity function $\lambda(t)$ is right continuous.
+\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is right continuous, and each $\tau_k \in [0,t_m]$.
 % \begin{align}
@@
 In general, we may have different types of events that can occur.
-Let  the \emph{risk set}  $\mathcal{R}$ be the set of possible events.
-In the context of relational events occurring among $N$ nodes, for example, we let $\mathcal{R}$ include all dyads.\footnote{We exclude self-loops.}  Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ which is piecewise constant.
-
-For network data, we capture the notion that some events have no effect on the intensity of other events by assuming 1) each node $i$ belongs to some latent cluster $z_i$, and 2) each intensity function $\lambda_{ij}(t)$ may only change following an event involving a member of either $z_i$ or $z_j$.
-See Figure \ref{fig:example}.
-
- In this case, the likelihood becomes
+In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ include all allowed dyads.
+Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ which is piecewise constant.
+As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
+Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
 
 \begin{align}
-\mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
-&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R_{k_1,k_2}}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
+\mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
+&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \} \\
 \end{align}
@@
 
-For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $s_{ij}(t|A_t,\mathbf{z})$ that is a function of the previous history of events $A_t$ and the latent cluster assignments $z$.
+Define the event history $\mathcal{A}_t = \{e_m: t_m \in [0,t)\}$.
+For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $s_{ij}(t|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous history of events $\mathcal{A}_t$ and the latent cluster assignments $z$.
+Note the statistics $s_{ij}$ also do not change until an event involving either $z_i$ or $z_j$ occurs.
 We model the intensity functions via a log linear model conditioned on this information

### 2012-01-21 build.r

@@ -1,0 +1,130 @@
+# Deprecated
+build.brem <- function(A,N) {
+  implementedEffects <- c("RSndSnd","RRecSnd","PSAB-BA","PSAB-BY","PSAB-XA","PSAB-XB","PSAB-XY","PSAB-AY")
+  M <- nrow(A)
+  P <- length(implementedEffects)
+  s <- array(0,c(M,N,N,P))
+  dimnames(s) <- list(NULL,NULL,NULL,implementedEffects)
+  rrs <- rss <- matrix(0,N,N)
+  for (m in 1:(M-1)) {
+   
+    a <- A[m,2]
+    b <- A[m,3]
+    
+    # P-shift effects
+    s[m,b,a,"PSAB-BA"] <- 1
+    s[m,b,-a,"PSAB-BY"] <- 1
+    s[m,-b,a,"PSAB-XA"] <- 1
+    s[m,-a,b,"PSAB-XB"] <- 1
+    s[m,-c(a,b),-c(a,b),"PSAB-XY"] <- 1
+    s[m,a,-b,"PSAB-AY"] <- 1
+    
+    # Recency effects
+    ix <- which(rss[a,] > 0)
+    if (length(ix) > 0) {
+      s[m,a,ix,"RSndSnd"] <- 1/(rss[a,ix]) 
+    }
+    ix <- which(rrs[a,] > 0)
+    if (length(ix) > 0) {
+      s[m,a,ix,"RRecSnd"] <- 1/(rrs[a,ix])
+    }
+
+    
+    # Update recency statistics: a just sent to b
+    # rrs[a,b]: rank of b in a's list of people recently received from
+    jx <- which(rrs[b,] > 0)  # b just received from a
+    rrs[b,jx] <- rrs[b,jx] + 1  # src to sent to rec
+    rrs[b,a] <- 1
+    # rss[a,b]: rank of b in a's list of people recently sent to
+    jx <- which(rss[a,] > 0)  # a just sent to b
+    rss[a,jx] <- rss[a,jx] + 1
+    rss[a,b] <- 1
+  }
+  return(s)
+}
+block.ps <- function(A,z) {
+  s <- A[,2]
+  r <- A[,3]
+  zs <- sort(unique(z))
+  setup <- expand.grid(i=zs,j=zs)
+  ds <- lapply(1:nrow(setup),function(k) {
+    ix <- which(z[s] == setup$i[k] & z[r]==setup$j[k])
[…]

### 2012-01-21 sim.r

@@ -1,0 +1,28 @@
+
+beta <- c(1,0)
+effects <- names(beta) <- c("PSAB-BA","PSAB-BY")
+A <- simulate.rem(1000,10,effects,beta)
+s <- build.brem(A,N)
+
+N <- 10
+K <- 2
+z <- c(rep(1,N/2),rep(2,N/2))
+P <- 9
+beta <- array(0,c(K,K,P))
+beta[1,1,] <- c(1,0,0,4,0,0,0,0,0)
+beta[2,1,] <- c(1,0,0,0,4,0,0,0,0)
+beta[1,2,] <- c(1,0,0,0,4,0,0,0,0)  # AB-
+beta[2,2,] <- c(1,0,0,0,0,0,0,0,2)  # AB-AY
+
+
+M <- 1000
+A <- simulate.brem(M,N,z,beta)
+
+df <- block.ps(A,z)
+qplot(pshift,value,data=df,geom="bar") + facet_grid(i~j) + theme_bw()
+
+precomp <- rem.dyad.setup(A,10,effects=effects)
+x <- rem.dyad(A,10,effects,ordinal=FALSE)
+
+value <- list(ms=precomp$modstats,pv=c(1,0))
+rem.llk(value)

### 2012-01-21 test.hash.r

@@ -1,0 +1,13 @@
+library(hash)
+
+ts <- sort(runif(10000,0,100))
+x <- rnorm(10000)
+a <- list()
+for (i in 1:length(ts)) {
+  a[as.character(ts[i])] <- x[i]
+}
+b <- hash(key=ts,values=x)
+
+head(a)
+system.time(a[8000:9000])
+system.time(b[ts[8000:9000]])

### 2012-01-22 notes.tex
Change example image.  First version of likelihood function with new independence assumptions.

@@
 As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
-Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
-
+Let $\tau_{i,j,m}$ be the time that event $(i,j)$ occurred before the $m$th event.
+% Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
+% \begin{align}
+% \mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
+% &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \}
+% \end{align}
+% \begin{align}
+% \mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{r=1}^N \exp\{ - (t_m - \tau_{i,r,m}) \lambda_{ir}(t_m | z_i,z_r)  - (t_m - \tau_{r,j,m}) \lambda_{rj}(t_m | z_r,z_j) \}
+% \end{align}
 \begin{align}
-\mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
-&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \} \\
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{s=1}^N\exp\{ - (t_m - \tau_{s,j,m}) \lambda_{sj}(t_m | z_s,z_j) \} \prod_{r=1}^N \exp\{ - (t_m - \tau_{i,r,m}) \lambda_{ir}(t_m | z_i,z_r) \}
 \end{align}
@@
  \def\svgwidth{6in}
-  \input{example.pdf_tex}
-\caption{Illustration of the event data and the model assumptions.
-Left: An event sequence among four nodes belonging to two clusters (circles and squares): (4,3) occurs at time $t_1$ and (2,1) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{21}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ only can change when events occur involving a node in its cluster (square).}
+  \input{example1.pdf_tex}
+\caption{Illustration of event data and the assumptions of the model.
+Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
+Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
+Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
+Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
 \label{fig:example}

### 2012-01-22 rem.cpp.r

@@ -1,102 +1,63 @@
+library(inline)
+library(Rcpp)
 
 fx <- cxxfunction(,"",includes=
   '
 int threeDIndex(int j, int k, int l, int J, int K, int L) { 
   return j*K*L + k*L + l;
 }
-// beta currently: AB-BA,AB-BY,AB-XA,
+double computeLambda(int i, int j, int a, int b, Rcpp::NumericVector beta) {
+  double lam = beta[0];
+  if (i==b & j==a) { // ab-ba
+    lam += beta[1];  
+  }
+  if (i==b & j!=a) { // ab-by
+    lam += beta[2]; 
+  }
+  if (i!=b & j==a) { // ab-xa
+    lam += beta[3];
+  }
+  if (i!=a & j==b) { // ab-xb
+    lam += beta[4];
+  }
+  if (i==b & j!=b) { // ab-ay
+    lam += beta[5];
+  }
+  return exp(lam);
+}
+// All senders, receivers must be 0-indexed.
+// Current "weirdness": Assumes all events "occur" at time 0.
 Rcpp::List llk(Rcpp::NumericVector beta, Rcpp::NumericVector times, Rcpp::IntegerVector sen, Rcpp::IntegerVector rec, int M, int N1, int N2, int P){
-  Rcpp::NumericVector lrm = Rcpp::NumericVector( Dimension(M,N1,N1) );
+
+  Rcpp::NumericMatrix mp = Rcpp::NumericMatrix(N1,N2);  // last event id that lam_ij changed
+
   double llk = 0;
-  int a,b;
+  int a,b,i,j;
+  double lam_ij, lam_ir, lam_rj;
   for (int m = 1; m<M; m++) {
-    a = sen[m-1];
-    b = rec[m-1];
-    for (int i = 0; i<N1; i++) {
-      for (int j = 0; j<N2; j++) {
-        // TODO: Make interface for using different effects cleaner
-        lrm[threeDIndex(m,i,j,M,N1,N2)] = beta[0];
 
-        // PSHIFTS
-        if (i==b & j==a) {
-          lrm[threeDIndex(m,i,j,M,N1,N2)] += beta[1]; //ab-ba
-        }
-        if (i==b & j!=a) {
-          lrm[threeDIndex(m,i,j,M,N1,N2)] += beta[2]; //ab-by
-        }
-        if (i!=b & j==a) {
[…]

### 2012-01-22 notes.tex
Update paper with description of statistics.  Add TODO.

@@
 This is reminiscent of the stochastic blockmodel \cite{Nowicki2001, Kemp2006} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.
-Here, however, we use a blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
+Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
 
@@
 We include a similar set of statistics in $s_{ij}$ as used in \cite{Butts2008,Vuy2011}.
-Note that the model assumptions imply $s_{ij}(t|\mathcal{A}_t,\mathbf{z}) = f(\mathcal{A}$.
+Let $v_{ijm}$ be the index of the last event where $(i,j)$ occurred; if $(i,j)$ has never occurred, we set $v_{ijm}=0$.
+%Note that the model assumptions imply $s_{ij}(t|\mathcal{A}_t,\mathbf{z}) = f(\mathcal{A}$.
 
-Let $r$ be the last event from a $z_i$ node to a $z_j$ node, that is $r = \mbox{argmax}_{r \in \{1, \ldots, m\}}I(t_r < t, z_{i_r} = z_i, z_{j_r} = z_j)$
+%Let $r$ be the last event from a $z_i$ node to a $z_j$ node, that is $r = \mbox{argmax}_{r \in \{1, \ldots, m\}}I(t_r < t, z_{i_r} = z_i, z_{j_r} = z_j)$
 
-\begin{itemize}
-\item In and out degree of sender and receiver
-\item Reciprocity
-\item Transitivity, triangle closure
-\item Shared contacters, contactees
-\end{itemize}
-
-TODO: Would like things that combine shared contacts with recency. (i.e. Facebook-like terms.)
+\begin{enumerate}
+\item Sender out-degree: $s_{ij1}(t) = \sum_{m:t_m<t} I(i_m=i)$
+\item Sender in-degree: $s_{ij2}(t) = \sum_{m:t_m<t} I(j_m=i)$
+\item Receiver out-degree: $s_{ij3}(t) = \sum_{m:t_m<t} I(i_m=j)$
+\item Receiver in-degree: $s_{ij4}(t) = \sum_{m:t_m<t} I(j_m=j)$
+\item Reciprocity (AB-BA): $s_{ij5}(t) = \sum_{m:t_m<t} I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item Turn-taking (AB-BY): $s_{ij6}(t) = \sum_{m:t_m<t} I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item Turn* (AB-XA): $s_{ij7}(t) = \sum_{m:t_m<t} I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item  (AB-XB): $s_{ij8}(t) = \sum_{m:t_m<t} I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item  (AB-AY): $s_{ij9}(t) = \sum_{m:t_m<t} I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item TODO: Shared contacters, contactees (also Facebook-like terms.) 
+\item TODO: Recency or functions of time-windows
+\end{enumerate}
 
@@
 \item For $i = 1, \ldots, N$, draw $z_i \sim \mbox{Categorical}(\theta)$
+\item Initialize $v = 0$
 \item For $m = 1, \ldots, M$:
   \begin{enumerate}
-  \item For all $(i,j)$ such that $z_i,z_j \in \{z_{i_{m-1}},z_{j_{m-1}}\}$:
+  \item For all $r \ne i$:
     \begin{enumerate}
-    \item Compute $s_{ij}^*$ using events $1$ through $m-1$
-    \item Set $\lambda_{ij}^* \leftarrow \exp\{ \beta_{z_i,z_j} s_{ij}^*\}$ 
+    \item Compute $s_{rj}^*$ using event $0$ through $v_{rj}$
+    \item Set $\lambda_{rj}^* \leftarrow \exp\{ \beta_{z_r,z_j} s_{rj}^*\}$ 
+    \end{enumerate}
+  \item For all $r \ne j$:
+    \begin{enumerate}
+    \item Compute $s_{ir}^*$ using event $0$ through $v_{ir}$
+    \item Set $\lambda_{ir}^* \leftarrow \exp\{ \beta_{z_i,z_r} s_{ir}^*\}$ 
     \end{enumerate}    

### 2012-01-22 fns.r

@@ -5,7 +5,7 @@ build.brem <- function(A,N) {
   P <- length(implementedEffects)
   s <- array(0,c(M,N,N,P))
   dimnames(s) <- list(NULL,NULL,NULL,implementedEffects)
-  rrs <- rss <- matrix(0,N,N)
+  rrs <- rss <- mp <- matrix(0,N,N)
   for (m in 1:(M-1)) {
    
     a <- A[m,2]
@@ -16,33 +16,13 @@ build.brem <- function(A,N) {
     s[m,b,-a,"PSAB-BY"] <- 1
     s[m,-b,a,"PSAB-XA"] <- 1
     s[m,-a,b,"PSAB-XB"] <- 1
-    s[m,-c(a,b),-c(a,b),"PSAB-XY"] <- 1
     s[m,a,-b,"PSAB-AY"] <- 1
     
-    # Recency effects
-    ix <- which(rss[a,] > 0)
-    if (length(ix) > 0) {
-      s[m,a,ix,"RSndSnd"] <- 1/(rss[a,ix]) 
-    }
-    ix <- which(rrs[a,] > 0)
-    if (length(ix) > 0) {
-      s[m,a,ix,"RRecSnd"] <- 1/(rrs[a,ix])
-    }
-
-    
-    # Update recency statistics: a just sent to b
-    # rrs[a,b]: rank of b in a's list of people recently received from
-    jx <- which(rrs[b,] > 0)  # b just received from a
-    rrs[b,jx] <- rrs[b,jx] + 1  # src to sent to rec
-    rrs[b,a] <- 1
-    # rss[a,b]: rank of b in a's list of people recently sent to
-    jx <- which(rss[a,] > 0)  # a just sent to b
-    rss[a,jx] <- rss[a,jx] + 1
-    rss[a,b] <- 1
   }
   return(s)
 }
 block.ps <- function(A,z) {
+  require(relevent)
   s <- A[,2]
   r <- A[,3]
   zs <- sort(unique(z))
@@ -57,74 +37,56 @@ block.ps <- function(A,z) {
   return(ds)
 }
 simulate.brem <- function(M,N,z,beta) {
-  implementedEffects <- c("Intercept","RSndSnd","RRecSnd","PSAB-BA","PSAB-BY","PSAB-XA","PSAB-XB","PSAB-XY","PSAB-AY")
+  implementedEffects <- c("Intercept","RSndSnd","RRecSnd","PSAB-BA","PSAB-BY","PSAB-XA","PSAB-XB","PSAB-AY")
[…]

### 2012-01-22 sim.r

@@ -7,12 +7,12 @@ s <- build.brem(A,N)
 N <- 10
 K <- 2
 z <- c(rep(1,N/2),rep(2,N/2))
-P <- 9
+P <- 8
 beta <- array(0,c(K,K,P))
-beta[1,1,] <- c(1,0,0,4,0,0,0,0,0)
-beta[2,1,] <- c(1,0,0,0,4,0,0,0,0)
-beta[1,2,] <- c(1,0,0,0,4,0,0,0,0)  # AB-
-beta[2,2,] <- c(1,0,0,0,0,0,0,0,2)  # AB-AY
+beta[1,1,] <- c(3,0,0,5,0,0,0,0)
+beta[2,1,] <- c(1,0,0,0,1,0,0,0)
+beta[1,2,] <- c(1,0,0,0,1,0,0,0)  # AB-
+beta[2,2,] <- c(2,0,0,0,0,0,0,2)  # AB-AY
 
 
 M <- 1000

### 2012-01-24 icml/draft.tex
Add ICML style files.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ICML 2012 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use the following line _only_ if you're still using LaTeX 2.09.
%\documentstyle[icml2012,epsf,natbib]{article}
% If you rely on Latex2e packages, like most moden people use this:
\documentclass{article}

% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}

% For citations
\usepackage{natbib}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx, color}
\usepackage{amsmath}

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2012} with
% \usepackage[nohyperref]{icml2012} above.
\usepackage{hyperref}

% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{icml2012} 
% Employ this version of the ``usepackage'' statement after the paper has
% been accepted, when creating the final version.  This will set the
% note in the first column to ``Appearing in''
% \usepackage[accepted]{icml2012}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Relational Event Blockmodels}

\begin{document}

\twocolumn[
\icmltitle{Relational Event Blockmodels}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2012
% package.
\icmlauthor{Your Name}{email@yourdomain.edu}
\icmladdress{Your Fantastic Institute,
            314159 Pi St., Palo Alto, CA 94306 USA}
\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
\icmladdress{Their Fantastic Institute,
            27182 Exp St., Toronto, ON M6H 2T1 CANADA}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{boring formatting information, machine learning, ICML}

\vskip 0.3in
]

\begin{abstract} 
ICML 2012 full paper submissions are due February 24, 2012. Reviewing will
be blind to the identities of the authors, and therefore identifying
information must not appear in any way in papers submitted for review. Submissions must be in PDF, 8 page length limit.
\end{abstract}

\pagestyle{empty}

\section{Introduction}

Methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions. One class of statistical models for network data uses latent variables for each node to help capture unobserved heterogeneity.  For example, the stochastic blockmodel \cite{Nowicki2001, Kemp} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between blocks.   Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.

Network data, however, is often collected as a sequence of events occurring over time.   Recently several works have adapted models from survival analysis and event history analysis to provide continuous time models of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.  These models allow one to specify the dependence of the process on the previous history of events.  In this way one may incorporate theories about the underlying processes and make predictions about future data conditioned on the past.

The above continuous time network models have two major drawbacks: 1) the entire network is assumed to have identical dynamics, and 2) the occurrence of any event can affect the rate at which any other event occurs.  Both aspects may be unrealistic.  Consider email communication in a small organization comprised of several teams.  Each team may have a different pattern for collaborating via email, and in many cases emails within one team will not affect emailing behavior within a separate team.

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.  [TODO: Explain how our experiments show that this helps.]  By combining a latent variable framework with a local dependence model for event sequences, we provide a hierarchical approach for modeling event-based network data, accompanying recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
\begin{figure*}
\center
 \def\svgwidth{5.5in}
  \input{example1.pdf_tex}
\caption{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.  Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
\label{fig:example}
\end{figure*}  
\section{Model}

%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.  We can then write the likelihood of $M$ events as
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \\
&= \prod_{m=1}^M \lambda(t_m|\cdot) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}
\end{align}
\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is right continuous, and each $\tau_k \in [0,t_m]$.
% \begin{align}
% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
% \end{align}

In general, we may have different types of events that can occur.   In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ include all allowed dyads.  Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ that is piecewise constant.  As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.  Let $\tau_{i,j,m}$ be the time that $\lambda_{i,j}$ changed before the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.  % Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
% &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \} 
% \end{align}
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{r=1}^N \exp\{ - (t_m - \tau_{i,r,m}) \lambda_{ir}(t_m | z_i,z_r)  - (t_m - \tau_{r,j,m}) \lambda_{rj}(t_m | z_r,z_j) \} 
% \end{align}
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) =& \prod_{m=1}^M 
\lambda_{i_m,j_m}(t_m|\cdot) \times \\
&\prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m})
\lambda_{ij}(t_m | z_i,z_j) \}
\label{eqn:llk}
\end{align}

For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}_{ij}(t|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$  and the latent cluster assignments $\mathbf{z}$.  Note the statistics $\mathbf{s}_{ij}$ also do not change until an event involving either $i$ or $j$ occurs.  We model the intensity functions via a log linear model
\begin{align}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}_{ij}(t|\mathcal{A}_t,\mathbf{z}).
\end{align}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}_{ij}(t|\cdot)$.  This is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.  Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

\subsection{Model specification}

We construct $\mathbf{s}_{ij}$ using a similar set of statistics as used in \cite{Butts2008,Vu2011}. Let $v_{ijm} \in [0,M]$ be the index of the last event where $(i,j)$ occurred (n.b. $t_{v_{ijm}} = \tau_{ijm}$). If $(i,j)$ has never occurred, we set $v_{ijm}=0$.  
\begin{enumerate}
\item Sender out-degree: $s_{ij1}(t) = \sum_{m:t_m<t} I(i_m=i)$
\item Sender in-degree: $s_{ij2}(t) = \sum_{m:t_m<t} I(j_m=i)$
\item Receiver out-degree: $s_{ij3}(t) = \sum_{m:t_m<t} I(i_m=j)$
\item Receiver in-degree: $s_{ij4}(t) = \sum_{m:t_m<t} I(j_m=j)$
\item Reciprocity (AB-BA): $s_{ij5}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-taking (AB-BY): $s_{ij6}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-usurping (AB-XA): $s_{ij7}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-usurping (AB-XB): $s_{ij8}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Turn-continuing (AB-AY): $s_{ij9}(t) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
\item Number of shared two-paths: $s_{ij10}(t) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
% \item Shared contacts (since last occurrence): $s_{ij10}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
% \item Triadic closure (since last occurrence): $s_{ij11}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
% \item Shared contacts (since last occurrence): $s_{ij12}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
\end{enumerate}

We include degree effects to model possible rich-get-richer phenomenon.  The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.  For example, an AB-BA effect indicates an increased propensity for reciprocity.  The final set of effects allow one to model phenomenon such as triadic closure.  Though we only use the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.  The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}

\section{Inference}

We use Markov chain Monte Carlo to get samples from the posterior distribution of our parameters.

\begin{enumerate}
\item For $i \in [1,N]$: Sample $z_i$ via Gibbs sampling: 
$z_i | A, z_{-i} \propto p(A|z,\beta) p(z_i)$
\item For each pair $(k,k')$: Sample $\beta_{k,k'}$ via MH or HMC.
\end{enumerate}

To perform Gibbs sampling for a given cluster assignment $z_i$, we need to compute the likelihood for each possible assignment of $z_i$. 
This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.  Computing the likelihood is then $O(T \cdot P \cdot 4N \cdot K)$.

% This involves recomputing each $\boldsymbol{\tau}$ and possibly each $\mathbf{s}_{ij}$ if it also depends on $\mathbf{z}$.

%TODO: Decide how to handle label switching.  Schweinberger has an interesting approach I hadn't seen before for his hierarchical ERGMS.  Should look into that more closely...

% \subsection*{Bookkeeping considerations}
% %For each dyad, we need to keep track of the set of piecewise constant intervals $d_{ij}$.  For a given set of assignments $z$ we can compute these intervals $(a_v,b_v)$.

% Recall we let intensities change whenever an event occurs involving a fellow cluster member.  Thus the vector $\boldsymbol{\tau}_{ij}$ of changepoints (or knots) is given by
% $$\boldsymbol{\tau}_{ij} = \left\{ t_m: z_{i_m} = z_i \ \mbox{or} \ z_{j_m} = z_j, m=1,\ldots,M\right\}$$
% Any time we update we change node $k$'s cluster assignment from $a$ to $b$, the change point vectors for any nodes in $a$ and $b$ must be updated.  For now we assume we have to recompute  $\boldsymbol{\tau}_{ij}$  entirely.  Similarly, if the covariate vectors $\mathbf{s}_{ij}(t)$ depend on $\mathbf{z}$ then we need to recompute those as well prior to computing the likelihood.%More specifically, for a $D_{ij}$ such that $z_i, z_j \in \{a,b\}$, we must add knots $t_m$ for each.

\section{Simulation}
\begin{figure*}
\center
\includegraphics[width=1.75in]{../figs/syn-mat.pdf}
\includegraphics[width=3in]{../figs/syn-counts.pdf}
\includegraphics[width=1.5in]{../figs/syn-llk.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2. Right: Loglikelihood vs. iteration during MCMC.}
\label{fig:syncounts}
\end{figure*}

We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to continue speaking, and 3) interactions between groups are more likely to be repeated.  The specification of $\textbf{s}$ is therefore $s_{ij}(t) = [s_0, s_{ij5}(t), s_{ij9}(t), s_{ij11}(t)]$.  For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (2,3,0,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (1,0,0,1)$, and $\boldsymbol{\beta}_{2,2} = (2,0,2,0)$.  See Appendix for details on simulating from the model.  In Figure \ref{fig:syncounts} we illustrate the simulated data and show the loglikelihood during the first iterations of MCMC [after being initialized with $\beta \sim N(\beta_{true},.5^2)$].

TODO: Show parameter bias as a function of iteration; perhaps also as a function of the number of events in the dataset.

\section{Model checking and experiments}

\subsection*{Data}

Each of the following data sets are sequences of dyadic events, where each event has a time associated with it, a sender, and a recipient.

\begin{itemize}
\item Classroom:  Perhaps within a couple of the classrooms we can infer subgroups who tend to obey different relational event dynamics.
\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
\item Citation networks: Perhaps some groups of papers obey some types of rich-get-richer effects more than others.
\item Other
\end{itemize}

\subsection*{Prediction experiment}
We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.  Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the posterior mean given the training data.  Recall at a given cutoff $k$ is found by first sorting the predicted intensity functions at each time, then computing the mean rank of the observed events in that list.

TODO: Compare to fitting the relational event model without the blockmodel structure.  Compare to relational event blockmodel without covariates: $$p(y_{ij}) = \exp\{\beta_{z_i,z_j}\} / \sum_{(a,b) \in \mathcal{R}}\exp\{\beta_{z_{a},z_{b}}\} $$
Or more simply, just use $s_{ij}(t) = 1$.

TODO: Need to see how well the time-portion of the model is specified.

\section{Discussion}
Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches \cite{Gunawardana2011} aim to learn the regions where an intensity is constant and allow for a nonlinear relationship between statistics and the intensity function.

Possible future work could extend the present framework to a mixed-membership model where for each effect there is a latent cluster assignment.  This way a group of nodes might contact high in-degree nodes at a similar rate, though they vary in the degree to which they adhere to reciprocity.

% TODO: Discuss possible variations on the prior structure of $\beta$: 1. Enforce $\beta_{k,l} = \beta_{l,k}$. 2. Assume $\beta_{k,l,p} \sim \mbox{Normal}(\beta_p,\sigma_p^2)$. 3. Place other priors that provide different kinds of regularization.

% TODO: Discuss decision to condition on $z$ for covariate vector $s_{ij}$.

% TODO: Discuss choice of $\log \lambda$ instead of just $\lambda$.

% TODO: Discuss choice against time-varying effects.

% TODO: Discuss if the semiparametric partial likelihood approach can be used in this framework.

% \begin{table}[t]
% \caption{Classification accuracies for naive Bayes and flexible 
% Bayes on various data sets.}
% \label{sample-table}
% \vskip 0.15in
% \begin{center}
% \begin{small}
% \begin{sc}
% \begin{tabular}{lcccr}
% \hline
% \abovespace\belowspace
% Data set & Naive & Flexible & Better? \\
% \hline
% \abovespace
% Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
% Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
% Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
% Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
% Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
% Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
% Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
% \belowspace
% Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
% \hline
% \end{tabular}
% \end{sc}
% \end{small}
% \end{center}
% \vskip -0.1in
% \end{table}

% Acknowledgements should only appear in the accepted version. 
\section*{Acknowledgments}

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\appendix
\section{Simulation}

Simulation from the model follows the typical racing exponentials idea with a few subtleties.

\begin{enumerate}
\item $t_0 = 0$, all $s_{ij} = 1$, Initialize $v = 0$
\item For $i = 1, \ldots, N$: Draw $z_i \sim \mbox{Categorical}(\theta)$
\item For $m = 1, \ldots, M$:
  \begin{enumerate}
  \item Draw $t_m \sim \mbox{Exponential}\left(\sum_{ij} \lambda_{ij}^*\right)$
  \item Draw $(i_m,j_m) \sim \mbox{Categorical}\left(\lambda_{ij}^* / \sum_{ij}\lambda_{ij}^*\right)$
  % \item Save $s_{ij}(t_m) \leftarrow s_{ij}^*$
  % \item Save $\lambda_{ij}(t_m) \leftarrow \lambda_{ij}^*$
  \item For all $(i,j)$ where $i$ or $j$ is in $\{i_m,j_m\}$:
    \begin{enumerate}
    \item Compute $s_{ij}^*$ using event $0$ through $v_{ij}$
    \item Set $\lambda_{ij}^* \leftarrow \exp\{ \beta_{z_i,z_j} s_{ij}^*\}$ 
    \end{enumerate}
  \end{enumerate}
\end{enumerate}

\bibliography{refs}
\bibliographystyle{icml2012}

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was
% created by Lise Getoor and Tobias Scheffer, it was slightly modified  
% from the 2010 version by Thorsten Joachims & Johannes Fuernkranz, 
% slightly modified from the 2009 version by Kiri Wagstaff and 
% Sam Roweis's 2008 version, which is slightly modified from 
% Prasad Tadepalli's 2007 version which is a lightly 
% changed version of the previous year's version by Andrew Moore, 
% which was in turn edited from those of Kristian Kersting and 
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2012-01-25 docs/notes.tex
Update TODO.  Add notes and eckmann analysis.

--- 2012-01-24 icml/draft.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% ICML 2012 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Use the following line _only_ if you're still using LaTeX 2.09.
-%\documentstyle[icml2012,epsf,natbib]{article}
-% If you rely on Latex2e packages, like most moden people use this:
-\documentclass{article}
-
-% For figures
-\usepackage{graphicx} % more modern
-%\usepackage{epsfig} % less modern
-\usepackage{subfigure}
-
-% For citations
-\usepackage{natbib}
-
-% For algorithms
-\usepackage{algorithm}
-\usepackage{algorithmic}
+\pagestyle{empty}
+\documentclass[11pt]{article}
 \usepackage{graphicx}
@@
 \usepackage{amsmath}
+%\usepackage[small,compact]{titlesec} 
+%\usepackage[small,it]{caption}
 
-% As of 2011, we use the hyperref package to produce hyperlinks in the
-% resulting PDF.  If this breaks your system, please commend out the
-% following usepackage line and replace \usepackage{icml2012} with
-% \usepackage[nohyperref]{icml2012} above.
-\usepackage{hyperref}
-
-% Packages hyperref and algorithmic misbehave sometimes.  We can fix
-% this with the following command.
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Employ the following version of the ``usepackage'' statement for
-% submitting the draft version of the paper for review.  This will set
-% the note in the first column to ``Under review.  Do not distribute.''
-\usepackage{icml2012} 
-% Employ this version of the ``usepackage'' statement after the paper has
-% been accepted, when creating the final version.  This will set the
-% note in the first column to ``Appearing in''
-% \usepackage[accepted]{icml2012}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Relational Event Blockmodels}
+\textwidth 6.5in
+\headheight .1in
+\topmargin -.2 in
+\textheight 8.7in
+\oddsidemargin 0in
+\evensidemargin 1in
 
@@
 
-\twocolumn[
-\icmltitle{Relational Event Blockmodels}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2012
-% package.
-\icmlauthor{Your Name}{email@yourdomain.edu}
-\icmladdress{Your Fantastic Institute,
-            314159 Pi St., Palo Alto, CA 94306 USA}
-\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
-\icmladdress{Their Fantastic Institute,
-            27182 Exp St., Toronto, ON M6H 2T1 CANADA}
-
-% You may provide any keywords that you 
-% find helpful for describing your paper; these are used to populate 
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{boring formatting information, machine learning, ICML}
-
-\vskip 0.3in
-]
-
-\begin{abstract} 
-ICML 2012 full paper submissions are due February 24, 2012. Reviewing will
-be blind to the identities of the authors, and therefore identifying
-information must not appear in any way in papers submitted for review. Submissions must be in PDF, 8 page length limit.
-\end{abstract}
-
-\pagestyle{empty}
+ \centerline{\Large \bf Blockmodels for Dynamic Network Data} 
+ \medskip
+\centerline{\bf Christopher DuBois}
+ \bigskip
 
@@
 [TODO: Explain how our experiments show that this helps.]  By combining a latent variable framework with a local dependence model for event sequences, we provide a hierarchical approach for modeling event-based network data, accompanying recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-\begin{figure*}
-\center
- \def\svgwidth{5.5in}
-  \input{example1.pdf_tex}
-\caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-\label{fig:example}
-\end{figure*}  
+
 \section{Model}
@@
 \begin{align}
-\mathcal{L}(\mathcal{A}|\theta) =& \prod_{m=1}^M 
-\lambda_{i_m,j_m}(t_m|\cdot) \times \\
-&\prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m})
-\lambda_{ij}(t_m | z_i,z_j) \}
-\label{eqn:llk}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | z_i,z_j) \}
 \end{align}
+
+\begin{figure}
+ \def\svgwidth{6in}
+  \input{example1.pdf_tex}
+\caption{Illustration of event data and the assumptions of the model.
+Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
+Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
+Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
+Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
+\label{fig:example}
+%\includegraphics[width=6in]{example}
+\end{figure}
 
@@
 
+\subsection{Submodels}
+
+The above framework generalizes several important special cases.
+The first is one where events within each block share the rate at which they occur.
+This can be implemented by only having an intercept statistic, $s_0(t,i,j) = 1$.
+As these intensities do not change under this specification, the likelihood simplifies to
+$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
+where $ \lambda_{ij} = \exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
+Each block in this model is homogeneous Poisson process where the dyad associated with each event is draw uniformly from the dyads in that block.
+
+Another important model is one where within group interactions are governed via a set of parameters while all between group interactions are governed by a different set.
+In this case $\boldsymbol{\beta}_{k1,k2} = \boldsymbol{\gamma}$ for all $k1 \ne k2$.
+
 \section{Inference}
@@
 This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.
-Computing the likelihood is then $O(T \cdot P \cdot 4N \cdot K)$.
+Computing the likelihood is then $O(T \cdot P \cdot 4N)$, though we essentially split this computation into $K^2$ pieces.
 
@@
 \section{Simulation}
-\begin{figure*}
+\begin{figure}
 \center
-\includegraphics[width=1.75in]{../figs/syn-mat.pdf}
-\includegraphics[width=3in]{../figs/syn-counts.pdf}
-\includegraphics[width=1.5in]{../figs/syn-llk.pdf}
+%\includegraphics[width=1.75in]{../figs/syn-mat.pdf}
+%\includegraphics[width=3in]{../figs/syn-counts.pdf}
+%\includegraphics[width=1.5in]{../figs/syn-llk.pdf}
 \caption{Illustration of 1000 simulated events, as described in text.
@@
 \label{fig:syncounts}
-\end{figure*}
+\end{figure}
 
@@
 TODO: Compare to fitting the relational event model without the blockmodel structure.
-Compare to relational event blockmodel without covariates: $$p(y_{ij}) = \exp\{\beta_{z_i,z_j}\} / \sum_{(a,b) \in \mathcal{R}}\exp\{\beta_{z_{a},z_{b}}\} $$
+Compare to relational event blockmodel without covariates: $p(y_{ij}) = \exp\{\beta_{z_i,z_j}\} / \sum_{(a,b) \in \mathcal{R}}\exp\{\beta_{z_{a},z_{b}}\} $
 Or more simply, just use $s_{ij}(t) = 1$.
@@
 
-% TODO: Discuss possible variations on the prior structure of $\beta$: 1.
+TODO: Discuss possible variations on the prior structure of $\beta$: 1.
 Enforce $\beta_{k,l} = \beta_{l,k}$.
@@
 
-% TODO: Discuss decision to condition on $z$ for covariate vector $s_{ij}$.
+TODO: Discuss if the semiparametric partial likelihood approach can be used in this framework.
 
-% TODO: Discuss choice of $\log \lambda$ instead of just $\lambda$.
-
-% TODO: Discuss choice against time-varying effects.
-
-% TODO: Discuss if the semiparametric partial likelihood approach can be used in this framework.
-
-% \begin{table}[t]
-% \caption{Classification accuracies for naive Bayes and flexible 
-% Bayes on various data sets.}
-% \label{sample-table}
-% \vskip 0.15in
-% \begin{center}
-% \begin{small}
-% \begin{sc}
-% \begin{tabular}{lcccr}
-% \hline
-% \abovespace\belowspace
-% Data set & Naive & Flexible & Better? \\
-% \hline
-% \abovespace
-% Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
-% Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
-% Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
-% Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
-% Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
-% Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
-% Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
-% \belowspace
-% Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
-% \hline
-% \end{tabular}
-% \end{sc}
-% \end{small}
-% \end{center}
-% \vskip -0.1in
-% \end{table}
-
-% Acknowledgements should only appear in the accepted version. 
-\section*{Acknowledgments}
-
-% In the unusual situation where you want a paper to appear in the
-% references without citing it in the main text, use \nocite
 \appendix
@@
 
+\bibliographystyle{plain}
 \bibliography{refs}
-\bibliographystyle{icml2012}
 
 \end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was
-% created by Lise Getoor and Tobias Scheffer, it was slightly modified  
-% from the 2010 version by Thorsten Joachims & Johannes Fuernkranz, 
-% slightly modified from the 2009 version by Kiri Wagstaff and 
-% Sam Roweis's 2008 version, which is slightly modified from 
-% Prasad Tadepalli's 2007 version which is a lightly 
-% changed version of the previous year's version by Andrew Moore, 
-% which was in turn edited from those of Kristian Kersting and 
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2012-01-25 data/eckmann.R

@@ -1,0 +1,53 @@
+d <- read.table("eckmann/eckmann.txt",header=FALSE)
+colnames(d) <- c("from","to","size","timestamp","chron",
+                 "weekday","multiplicity")
+d <- subset(d,multiplicity==1)
+times <- d$chron - d$chron[1]
+d$ntime <- times/max(times)
+all <- sort(unique(c(d$from,d$to)))
+min.count <- 50
+max.count <- 1000
+counts <- table(c(d$from,d$to))
+ix <- which(counts > min.count & counts < max.count)
+length(ix)
+keep <- as.numeric(names(counts[ix]))
+ix <- which(d$from %in% keep & d$to %in% keep)
+d <- d[ix,c("ntime","from","to")]
+              plot(as.numeric(table(table(c(d$from,d$to)))),type="l")
+
+nodes <- sort(unique(c(d$from,d$to)))
+d$s <- as.numeric(factor(d$from,nodes))
+d$r <- as.numeric(factor(d$to,nodes))
+         
+A <- d[,c("ntime","s","r")]
+
+# Visualize
+plotspmat <- function(A) {
+  mat <- table(A[,2],A[,3])
+  mat <- melt(as.matrix(mat))
+  mat <- subset(mat,value!=0)
+  plot(mat[,1],mat[,2],pch=".")
+}
+plotspmat(A)
+
+# Create even smaller subset
+ix <- which(A[,2] %in% 40:130 & A[,3] %in% 40:130)
+A <- A[ix,]
+A[,2] <- as.numeric(factor(A[,2],40:130))
+A[,3] <- as.numeric(factor(A[,3],40:130))
+
+source("rem.cpp.r")
+source("fns.r")
+N <- length(unique(c(A[,2],A[,3])))
+K <- 2
+P <- 7
+M <- nrow(A)
+
+z <- sample(1:K,N,replace=TRUE)
+beta <- array(rnorm(K^2*P),c(K,K,P))
+system.time(brem.llk(A,N,z,beta,px))
+
+set.seed(4)
+niter <- 100
+px <- c(1,rep(0,6))
+fit0 <- brem.mcmc(A,N,K,P,px,model.type="baserates",niter=3)

### 2012-01-25 sim.r

@@ -37,10 +37,11 @@ niter <- 100
 beta.init <- beta + rnorm(length(beta),0,1)
 fit0 <- brem.mcmc(sim$A,N,K,P,px,model.type="baserates",niter=niter,init=beta.init)
 fit1 <- brem.mcmc(sim$A,N,K,P,px,model.type="diag.rem",niter=niter,init=beta.init)
+px[1] <- 0
 fit2 <- brem.mcmc(sim$A,N,K,P,px,model.type="full",niter=niter,init=beta.init)
 
 # Compare llk and lpost of true and fit
-fit <- fit1
+fit <- fit2
 brem.llk(sim$A,N,z,beta,px)
 brem.llk(sim$A,N,fit$z,fit$beta,px)
 brem.lpost(sim$A,N,z,beta,px)

### 2012-01-26 docs/notes.tex
Add discussion about extensions of the model.

@@
 
+\subsection{Extensions}
+TODO: Explain difficulties in doing a mixed membership model.
+As in the mixed-membership stochastic blockmodel \cite{Airoldi2008}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.
+In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.
+One could extend our current framework by assuming that after the $m$th event, $i_m$ and $j_m$ redraw their latent class.
+
+TODO: Discuss other variants, e.g. HMM that guides which latent class over time?
+
+TODO: What about different clusters for senders and receivers (cite our KDD paper)?
 \section{Inference}
@@
 This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.
-Computing the likelihood is then $O(T \cdot P \cdot 4N)$, though we essentially split this computation into $K^2$ pieces.
+Computing the likelihood is then $O(T \cdot P \cdot 4N)$ (though this computation is split into $K^2$ pieces) and must computed $N \cdot K$ times per Gibbs iteration.
+
+\subsection{Approximate likelihood methods}
+
+TODO: Discuss how one might sample the likelihood.  Cite Raftery paper.
+
+TODO: Look at performance of this with respect to bias and performance measures.
 

### 2012-01-26 docs/notes.tex
reorg and precision plot

@@
 
-Methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions.
+Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.
 One class of statistical models for network data uses latent variables for each node to help capture unobserved heterogeneity.
@@
 
-In general, we may have different types of events that can occur.
-In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ include all allowed dyads.
+It is straightforward extend this to marked point processes where each event contains additional information.
+In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ of possible events include all allowed dyads.
 Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ that is piecewise constant.
@@
 \begin{align}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}_{ij}(t|\mathcal{A}_t,\mathbf{z}).
+\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z}).
 \end{align}
@@
 \begin{enumerate}
-\item Sender out-degree: $s_{ij1}(t) = \sum_{m:t_m<t} I(i_m=i)$
-\item Sender in-degree: $s_{ij2}(t) = \sum_{m:t_m<t} I(j_m=i)$
-\item Receiver out-degree: $s_{ij3}(t) = \sum_{m:t_m<t} I(i_m=j)$
-\item Receiver in-degree: $s_{ij4}(t) = \sum_{m:t_m<t} I(j_m=j)$
-\item Reciprocity (AB-BA): $s_{ij5}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Turn-taking (AB-BY): $s_{ij6}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Turn-usurping (AB-XA): $s_{ij7}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Turn-usurping (AB-XB): $s_{ij8}(t) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Turn-continuing (AB-AY): $s_{ij9}(t) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Number of shared two-paths: $s_{ij10}(t) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
-% \item Shared contacts (since last occurrence): $s_{ij10}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
-% \item Triadic closure (since last occurrence): $s_{ij11}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
-% \item Shared contacts (since last occurrence): $s_{ij12}(t) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
+\item Sender out-degree: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i)$
+\item Sender in-degree: $s_{2}(t,i,j) = \sum_{m:t_m<t} I(j_m=i)$
+\item Receiver out-degree: $s_{3}(t,i,j) = \sum_{m:t_m<t} I(i_m=j)$
+\item Receiver in-degree: $s_{4}(t,i,j) = \sum_{m:t_m<t} I(j_m=j)$
+\item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
+% \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
+% \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
+% \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
 \end{enumerate}
@@
 TODO: Explain difficulties in doing a mixed membership model.
-As in the mixed-membership stochastic blockmodel \cite{Airoldi2008}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.
+As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.
 In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.
@@
 
-TODO: Discuss other variants, e.g. HMM that guides which latent class over time?
+TODO: Discuss other variants, e.g. HMM that guides which latent class over time?  Or a DP instead of fixed K?
 
 TODO: What about different clusters for senders and receivers (cite our KDD paper)?
+
+TODO: What about doing a similar thing with topic model statistics in s(t,i,j)?
+
 \section{Inference}
@@
 \center
-%\includegraphics[width=1.75in]{../figs/syn-mat.pdf}
-%\includegraphics[width=3in]{../figs/syn-counts.pdf}
-%\includegraphics[width=1.5in]{../figs/syn-llk.pdf}
+\includegraphics[width=1.5in]{../figs/syn/mat.pdf}
+\includegraphics[width=2.5in]{../figs/syn/counts.pdf}
+\includegraphics[width=2in]{../figs/syn/bias.pdf}
 \caption{Illustration of 1000 simulated events, as described in text.
@@
 \subsection*{Prediction experiment}
+\begin{figure}
+\center
+\includegraphics[width=3.5in]{../figs/syn/test-recall.pdf}
+\caption{Predictive performance on test data.  Left: Recall on synthetic example shown in Figure \ref{fig:syncounts}.}
+\label{fig:recall}
+\end{figure}
 We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.

### 2012-01-26 array.example.r

@@ -1,32 +1,0 @@
-library(inline)
-library(Rcpp)
-fx <- cxxfunction(,"",includes=
-  '
-int threeDIndex(int j, int k, int l, int J, int K, int L) { 
-  return l*J*K + k*J + j;
-}
-Rcpp::IntegerVector putval(int v, int i, int j, int k, int I, int J, int K){
-  Rcpp::IntegerVector lrm = Rcpp::IntegerVector(Dimension(I,J,K));
-  lrm[threeDIndex(i-1,j-1,k-1,I,J,K)] = v;
-  return lrm;
-}
-RCPP_MODULE(foo){
-  function( "putval", &putval ) ;
-}
-', plugin="Rcpp")
-
-foo <- Module("foo",getDynLib(fx))
-
-foo$putval(5,1,2,3,2,3,4)[1,2,3] # 5
-
-
-m <- matrix(0,3,4)
-mind <- function(j,k,J,K) (k-1)*J + j
-m[mind(1,2,3,4)] <- 100
-m
-a <- array(0,c(2,3,4))
-aind <- function(i,j,k,I,J,K) k*I*J + j*I + i
-a[aind(1-1,2-1,3-1,2,3,4) + 1] <- 1
-a[1,2,3]
-
-tmp <- foo$fn(100,1,2,3,100,10,10)

### 2012-01-26 fns.r

@@ -1,283 +1,0 @@
-
-dyad.ps <- function(A,N) {
-  require(relevent)
-  x <- array(0,c(N,N,6))
-  mp <- matrix(0,N,N)
-  M <- nrow(A)
-  for (m in 1:M) {
-    i <- A[m,2]
-    j <- A[m,3]
-    # Find last time lambda_ij changed
-    if (mp[i,j]!=0) {
-      a <- A[mp[i,j],2]
-      b <- A[mp[i,j],3]
-      x[i,j,] <- x[i,j,] + pshift(i,j,a,b)
-    }
-    mp[i,] <- m
-    mp[,i] <- m
-    mp[j,] <- m
-    mp[,j] <- m
-  }
-  dimnames(x)[[3]] <- c("AB-BA","AB-BY","AB-XA","AB-XB","AB-AY","AB-AB")
-  return(x)
-}
-
-node.ps <- function(A,N) {
-  require(relevent)
-  x <- matrix(0,N,2)
-  p <- matrix(0,N,5)
-  M <- nrow(A)
-  for (m in 1:M) {
-    i <- A[m,2]
-    j <- A[m,3]
-    a <- x[i,1]
-    b <- x[i,2]
-    p[i,] <- p[i,] + pshift(i,j,a,b)
-    a <- x[j,1]
-    b <- x[j,2]
-    p[j,] <- p[j,] + pshift(i,j,a,b)
-    
-    # Update last event for i and j
-    x[i,] <- x[j,] <- c(i,j)
-  }
-  colnames(p) <- c("AB-BA","AB-BY","AB-XA","AB-XB","AB-AY")
-  return(p)
-}
-
-computeLambda <- function(i,j,a,b,beta) {
-  lam <- beta[1] + sum(beta[-1] * pshift(i,j,a,b))
-  return(lam)
-}
-pshift <- function(i,j,a,b) {
-  x <- rep(0,6)
-  if (i!=a & i==b & j==a & j!=b) { # ab-ba
-    x[1] <- 1
-  }
-  if (i!=a & i==b & j!=a & j!=b) { # ab-by
-    x[2] <- 1
-  }
-  if (i!=a & i!=b & j==a & j!=b) { # ab-xa
-    x[3] <- 1
-  }
-  if (i!=a & i!=b & j!=a & j==b) { # ab-xb
-    x[4] <- 1
-  }
-  if (i==a & i!=b & j!=a & j!=b) { # ab-ay
[…]

### 2012-01-26 rem.cpp.r

@@ -1,156 +1,0 @@
-library(inline)
-library(Rcpp)
-
-fx <- cxxfunction(,"",includes=
-  '
-int threeDIndex(int j, int k, int l, int J, int K, int L) { 
-  return l*J*K + k*J + j;
-}
-int threeDIndexOld(int j, int k, int l, int J, int K, int L) { 
-    return j*K*L + k*L + l;
-}
-Rcpp::IntegerVector testThreeDIndex() {
-  Rcpp::IntegerMatrix mp = Rcpp::IntegerMatrix(Dimension(3,4,5));
-  mp[threeDIndex(1,1,1,3,4,5)] = 1;
-  return mp;
-}
-
-int testIntConv(Rcpp::IntegerVector ps) {
-  int r;
-  if (ps[0] == 1) {
-    r = 1;
-  } else {
-    r = 2;
-  }
-  int k = ps[ps[0]];
-  return k;
-}
-double computeLambdaOld(int i, int j, int a, int b, Rcpp::NumericVector beta) {
-  double lam = 0;
-    lam += beta[0];
-    if (i!=a & i==b & j==a & j!=b) { // ab-ba
-      lam += beta[1];
-    }
-    if (i!=a & i==b & j!=a & j!=b) { // ab-by
-      lam += beta[2];
-    }
-    if (i!=a & i!=b & j==a & j!=b) { // ab-xa
-      lam += beta[3];
-    }
-    if (i!=a & i!=b & j!=a & j==b) { // ab-xb
-      lam += beta[4];
-    }
-    if (i==a & i!=b & j!=a & j!=b) { // ab-ay
-      lam += beta[5];
-    }
-  return lam;
-}
-double computeLambda(int i, int j, int a, int b, Rcpp::NumericVector beta,Rcpp::IntegerVector ps) {
-  double lam = 0;
-  //for (int p = 0; p < ps.size(); p++) {
-    //ps[p] = 1;
-    if (ps[0] == 1) {
-      lam += beta[0];
-    }
-    if (ps[1] == 1 & i!=a & i==b & j==a & j!=b) { // ab-ba
-      lam += beta[1];
-    }
-    if (ps[2] == 1 & i!=a & i==b & j!=a & j!=b) { // ab-by
-      lam += beta[2];
-    }
-    if (ps[3] == 1 & i!=a & i!=b & j==a & j!=b) { // ab-xa
-      lam += beta[3];
-    }
[…]

### 2012-01-26 sbm.r

@@ -1,72 +1,0 @@
-sbm.llk <- function(A,N,K,z,beta) {
-#   s <- factor(A[,2],1:N)
-#   r <- factor(A[,3],1:N)
-  zs <- factor(z[A[,2]],1:K)
-  zr <- factor(z[A[,3]],1:K)
-  mat <- table(zs,zr)
-  t_M <- A[nrow(A),1]
-  R <- table(factor(z,1:K))
-  R <- outer(R,R)
-  mat * log(beta) - t_M * R * beta
-  #dmultinom(mat,size=sum(mat),prob=eta,log=TRUE)  
-}
-sbm.lpost <- function(A,N,K,z,beta,alpha=1) {
-  beta <- exp(beta)
-  sum(sbm.llk(A,N,K,z,beta)) + sum(dgamma(beta,1,1,log=TRUE))
-}
-sbm.lrm <- function(A,N,z,beta) {
-  a <- array(0,c(nrow(A),N,N))
-  for (i in 1:nrow(A)) {
-    a[i,,] <- beta[z,z]
-  }
-  return(a)
-}
-sbm.mcmc <- function(A,N,K,niter=100,z=NULL,mcmc.sd=.1) {
-  if (is.null(z)) z <- sample(1:K,N,replace=TRUE)
-  beta <- matrix(rnorm(K^2),K,K)
-  llks <- rep(0,niter)
-  current <- beta
-  for (iter in 1:niter) {
-    # For each effect sample via MH
-    olp <- sbm.lpost(A,N,K,z,current)
-    for (p in 1:5) {
-      cand <- current
-      cand  <- cand + rnorm(K^2,0,mcmc.sd)
-      clp <- sbm.lpost(A,N,K,z,cand)
-      if (clp - olp > log(runif(1))) {
-        current <- cand
-        olp <- clp
-      }
-    }
-    
-    for (i in 1:N) {
-      ps <- rep(0,K)
-      for (k in 1:K) {
-        z[i] <- k
-        ps[k] <- sbm.lpost(A,N,K,z,current)
-      }
-      ps <- exp(ps - max(ps))
-      z[i] <- sample(1:K,size=1,prob=ps)
-    }
-
-    llks[iter] <- olp
-    cat("iter",iter,":",llks[iter],"z:",table(z),"\n")
-  }
-  return(list(beta=current,z=z))
-}
-mult.dir <- function(dataset,dims,prior=1/prod(dims)) {
-  if (ncol(dataset)==2)
[…]

### 2012-01-26 sim.r

@@ -1,95 +1,0 @@
-library(releventhier)
-library(ggplot2)
-source("rem.cpp.r")
-source("fns.r")
-source("sbm.r")
-N <- 10
-K <- 2
-P <- 7
-beta <- array(0,c(K,K,P))
-beta[1,1,] <- c(2,3,0,0,0,0,0)
-beta[2,1,] <- c(1,0,0,0,0,0,1)
-beta[1,2,] <- c(1,0,0,0,0,0,1) 
-beta[2,2,] <- c(2,0,0,0,0,2,0)  # AB-AY
-
-
-M <- 2000
-set.seed(1)
-z <- c(rep(1,5),rep(2,5))
-sim <- simulate.brem(M,N,z,beta)
-mat <- table(sim$A[,2],sim$A[,3])
-mat <- melt(as.matrix(mat))
-colnames(mat) <- c("X1","X2","value")
-require(releventhier)
-plotmat(mat)
-#ggsave("figs/syn-mat.pdf",width=3,height=3)
-
-
-px <- rep(1,7)
-brem.llk(sim$A,N,z,beta,px)
-brem.lpost(sim$A,N,z,beta,px)
-
-set.seed(4)
-niter <- 100
-beta.init <- beta + rnorm(length(beta),0,1)
-px0 <- px1 <- rep(1,7)
-px2 <- c(0,1,1,1,1,1,1)
-fit0 <- sbm.mcmc(sim$A,N,K)
-fit1 <- brem.mcmc(sim$A,N,K,P,px,model.type="diag.rem",niter=niter,init=beta.init)
-fit2 <- brem.mcmc(sim$A,N,K,P,px,model.type="full",niter=niter,z=z,beta=beta)
-
-# Compare llk and lpost of true and fit
-fit <- fit2
-brem.llk(sim$A,N,z,beta,px)
-brem.llk(sim$A,N,fit$z,fit$beta,px)
-brem.lpost(sim$A,N,z,beta,px)
-brem.lpost(sim$A,N,fit$z,fit$beta,px)
-
-pdf("figs/syn-llk.pdf",width=4,height=4)
-plot(fit$llk[1:100],type="l",ylab="loglikelihood",xlab="iteration")
-dev.off()
-
-# Compute dyad counts for each pshift
-df <- dyad.ps(sim$A,N)
-df <- melt(df)
-df$i <- z[df$X1]
-df$j <- z[df$X2]
[…]

### 2012-01-26 test.rem.cpp.r

@@ -1,38 +1,0 @@
-source("rem.cpp.r")
-library(testthat)
-
-set.seed(1)
-M <- 100
-N <- 10
-P <- 6
-times <- sort(runif(M,0,100))
-sen <- sample(1:N,M,replace=TRUE) - 1
-rec <- sample(1:N,M,replace=TRUE) - 1
-beta <- rnorm(P)
-ix <- 0:(N-1)
-px <- rep(1,7)
-
-test_that("llk works on different subsets of dyads",{
-  
-  llk <- drem$llk(beta,times,sen,rec,ix,ix,px,N,M,P)
-  llk
-  N <- 10
-  ix <- 1:5 - 1
-  jx <- 5:10 - 1
-  sen <- sample(ix,M,replace=TRUE)
-  rec <- sample(jx,M,replace=TRUE)
-  llk <- drem$llk(beta,times,sen,rec,ix,jx,px,N,M,P)
-})
-
-test_that("log rate matrix computation correct",{
-  lrm <- drem$lrm(beta,times,sen,rec,ix,ix,px,N,M,P)
-  expect_that(lrm[2,1,1],equals(beta[1]))
-  expect_that(lrm[2,rec[1]+1,sen[1]+1],equals(beta[1] + beta[2]))  # AB-BA
-})
-
-test_that("calculating lambda runs",{
-  drem$testIntConv(c(3,2,3))
-  beta <- rnorm(7)
-  expect_that(drem$computeLambdaOld(3,2,2,3,beta),
-              equals(drem$computeLambda(3,2,2,3,beta,rep(1L,7)))
-})

### 2012-01-27 R/brem.r
Implement approximate likelihood

@@ -147,7 +147,7 @@ drem.llk <- function(A,N,beta,ix,jx,px) {
   rec <- A[,3]-1
   ix <- ix-1
   jx <- jx-1
-  drem$llk(beta,times,sen,rec,ix,jx,px,N,M,length(P))$llk
+  drem$llk(beta,times,sen,rec,ix,jx,px,N,M,length(P))
 }
 drem.mle <- function(A,N,beta,ix,jx,px) {
   fn <- function(par) {

### 2012-01-27 R/rem.cpp.r

@@ -6,44 +6,30 @@ fx <- cxxfunction(,"",includes=
 int threeDIndex(int j, int k, int l, int J, int K, int L) { 
   return l*J*K + k*J + j;
 }
-int threeDIndexOld(int j, int k, int l, int J, int K, int L) { 
-    return j*K*L + k*L + l;
-}
-Rcpp::IntegerVector testThreeDIndex() {
-  Rcpp::IntegerMatrix mp = Rcpp::IntegerMatrix(Dimension(3,4,5));
-  mp[threeDIndex(1,1,1,3,4,5)] = 1;
-  return mp;
-}
-
-int testIntConv(Rcpp::IntegerVector ps) {
-  int r;
-  if (ps[0] == 1) {
-    r = 1;
-  } else {
-    r = 2;
+// http://stackoverflow.com/questions/311703/algorithm-for-sampling-without-replacement
+Rcpp::IntegerVector sampleWithoutReplacement
+(
+  Rcpp::IntegerVector y,    // size of set sampling from
+  int sampleSize
+  )
+{
+  RNGScope scope;
+  // Use Knuths variable names
+  int& n = sampleSize;
+  int N = y.size();
+  Rcpp::IntegerVector samples(n);
+  int t = 0; // total input records dealt with
+  int m = 0; // number of items selected so far
+  while (m < n) {
+    Rcpp::NumericVector u = runif(1);
+    if ( (N - t)*u[0] >= n - m ) {
+      t++;
+    } else {
+      samples[m] = y[t];
+      t++; m++;
+    }
   }
-  int k = ps[ps[0]];
-  return k;
-}
-double computeLambdaOld(int i, int j, int a, int b, Rcpp::NumericVector beta) {
-  double lam = 0;
-    lam += beta[0];
-    if (i!=a & i==b & j==a & j!=b) { // ab-ba
-      lam += beta[1];
-    }
-    if (i!=a & i==b & j!=a & j!=b) { // ab-by
-      lam += beta[2];
-    }
-    if (i!=a & i!=b & j==a & j!=b) { // ab-xa
-      lam += beta[3];
-    }
-    if (i!=a & i!=b & j!=a & j==b) { // ab-xb
-      lam += beta[4];
-    }
-    if (i==a & i!=b & j!=a & j!=b) { // ab-ay
-      lam += beta[5];
-    }
-  return lam;
+  return samples;
 }
[…]

### 2012-01-27 R/sbm.r

@@ -23,7 +23,7 @@ sbm.lrm <- function(A,N,z,beta) {
 }
 sbm.mcmc <- function(A,N,K,niter=100,z=NULL,mcmc.sd=.1) {
   if (is.null(z)) z <- sample(1:K,N,replace=TRUE)
-  beta <- matrix(rnorm(K^2),K,K)
+  beta <- matrix(rnorm(K^2,-5,1),K,K)
   llks <- rep(0,niter)
   current <- beta
   for (iter in 1:niter) {
@@ -39,15 +39,15 @@ sbm.mcmc <- function(A,N,K,niter=100,z=NULL,mcmc.sd=.1) {
       }
     }
     
-#     for (i in 1:N) {
-#       ps <- rep(0,K)
-#       for (k in 1:K) {
-#         z[i] <- k
-#         ps[k] <- sbm.lpost(A,N,K,z,current)
-#       }
-#       ps <- exp(ps - max(ps))
-#       z[i] <- sample(1:K,size=1,prob=ps)
-#     }
+    for (i in 1:N) {
+      ps <- rep(0,K)
+      for (k in 1:K) {
+        z[i] <- k
+        ps[k] <- sbm.lpost(A,N,K,z,current)
+      }
+      ps <- exp(ps - max(ps))
+      z[i] <- sample(1:K,size=1,prob=ps)
+    }
 
     llks[iter] <- olp
     cat("iter",iter,":",llks[iter],"z:",table(z),"\n")

### 2012-01-27 R/sim.r

@@ -1,8 +1,8 @@
 library(releventhier)
 library(ggplot2)
-source("rem.cpp.r")
-source("fns.r")
-source("sbm.r")
+source("R/rem.cpp.r")
+source("R/fns.r")
+source("R/sbm.r")
 N <- 10
 K <- 2
 P <- 7
@@ -20,7 +20,6 @@ sim <- simulate.brem(M,N,z,beta)
 mat <- table(sim$A[,2],sim$A[,3])
 mat <- melt(as.matrix(mat))
 colnames(mat) <- c("X1","X2","value")
-require(releventhier)
 plotmat(mat)
 ggsave("figs/syn/mat.pdf",width=3,height=3)
 
@@ -30,13 +29,14 @@ brem.llk(sim$A,N,z,beta,px)
 brem.lpost(sim$A,N,z,beta,px)
 
 set.seed(4)
-niter <- 1000
+niter <- 200
 px0 <- px1 <- rep(1,7)
 px2 <- c(0,1,1,1,1,1,1)
 fit0 <- sbm.mcmc(sim$A,N,K,niter=niter,z=z)
 fit1 <- brem.mcmc(sim$A,N,K,P,px,model.type="diag.rem",niter=niter,z=z,gibbs=FALSE)
 fit2 <- brem.mcmc(sim$A,N,K,P,px,model.type="full",niter=niter,z=z,gibbs=FALSE)
 fit3 <- brem.mcmc(sim$A,N,1,P,px,model.type="full",niter=niter,gibbs=FALSE)
+save(fit0,fit1,fit2,fit3,file="data/syn-fits.rdata")
 
 # Compare llk and lpost of true and fit
 fit <- fit2
@@ -73,7 +73,7 @@ ggsave("figs/syn/bias.pdf",width=5,height=4)
 library(releventhier)
 test <- simulate.brem(M,N,z,beta)
 lrms <- list(unif = array(1,c(M,N,N)),
-             true = brem.lrm(test$A,N,z,beta,px0),
+             true = brem.lrm(test$A,N,z,beta,px2),
              base = sbm.lrm(test$A,N,fit0$z,fit0$beta),
              diag = brem.lrm(test$A,N,fit1$z,fit1$beta,px1),
              full = brem.lrm(test$A,N,fit2$z,fit2$beta,px2),

### 2012-01-27 R/test.rem.cpp.r

@@ -1,4 +1,4 @@
-source("rem.cpp.r")
+source("R/rem.cpp.r")
 library(testthat)
 
 set.seed(1)
@@ -16,6 +16,8 @@ test_that("llk works on different subsets of dyads",{
   
   llk <- drem$llk(beta,times,sen,rec,ix,ix,px,N,M,P)
   llk
+})
+test_that("llk works on small example",{
   N <- 10
   ix <- 1:5 - 1
   jx <- 5:10 - 1
@@ -31,8 +33,33 @@ test_that("log rate matrix computation correct",{
 })
 
 test_that("calculating lambda runs",{
-  drem$testIntConv(c(3,2,3))
   beta <- rnorm(7)
-  expect_that(drem$computeLambdaOld(3,2,2,3,beta),
-              equals(drem$computeLambda(3,2,2,3,beta,rep(1L,7)))
+  drem$computeLambda(3,2,2,3,beta,rep(1L,7))
+})
+
+test_that("sampling without replacement examples are correct",{
+  x <- drem$sampleWithoutReplacement(1:20,10)
+  expect_that(length(x),equals(10))
+  expect_that(max(x) <= 20,is_true())
+  
+  x <- drem$sampleWithoutReplacement(6:10,5)
+  expect_that(x,equals(6:10))
+  
+  x <- drem$sampleWithoutReplacement(50:100,5)
+  expect_that(all(x <=100 & x>=50),is_true())
+})
+
+test_that("approximate likelihood correct",{
+  N <- 100
+  ix <- 1:N - 1
+  sen <- sample(ix,M,replace=TRUE)
+  rec <- sample(ix,M,replace=TRUE)
+  llk <- drem$llk(beta,times,sen,rec,ix,ix,px,N,M,P)
+  allk <- drem$allk(beta,times,sen,rec,ix,ix,px,N,M,P,N)
+  expect_that(llk,equals(allk))
+  
+  allks <- sapply((N/2):N,function(i) {
+    drem$allk(beta,times,sen,rec,ix,ix,px,N,M,P,i)
+  })
+  plot(allks,type="l")
 })

### 2012-01-27 data/eckmann.R

@@ -1,108 +1,0 @@
-d <- read.table("eckmann/eckmann.txt",header=FALSE)
-colnames(d) <- c("from","to","size","timestamp","chron",
-                 "weekday","multiplicity")
-d <- subset(d,multiplicity==1)
-times <- d$chron - d$chron[1]
-d$ntime <- times/max(times)
-all <- sort(unique(c(d$from,d$to)))
-min.count <- 50
-max.count <- 1000
-counts <- table(c(d$from,d$to))
-ix <- which(counts > min.count & counts < max.count)
-length(ix)
-keep <- as.numeric(names(counts[ix]))
-ix <- which(d$from %in% keep & d$to %in% keep)
-d <- d[ix,c("ntime","from","to")]
-              plot(as.numeric(table(table(c(d$from,d$to)))),type="l")
-
-nodes <- sort(unique(c(d$from,d$to)))
-d$s <- as.numeric(factor(d$from,nodes))
-d$r <- as.numeric(factor(d$to,nodes))
-
-A <- d[,c("ntime","s","r")]
-
-
-
-# Visualize
-plotspmat <- function(A) {
-  mat <- table(A[,2],A[,3])
-  mat <- melt(as.matrix(mat))
-  mat <- subset(mat,value!=0)
-  plot(mat[,1],mat[,2],pch=".")
-}
-plotspmat(A)
-
-# Create even smaller subset
-ix <- which(A[,2] %in% 40:130 & A[,3] %in% 40:130)
-A <- A[ix,]
-A[,2] <- as.numeric(factor(A[,2],40:130))
-A[,3] <- as.numeric(factor(A[,3],40:130))
-
-# Reorder names
-B <- A
-all <- sort(unique(c(A[,2],A[,3])))
-all <- sample(all)
-A[,2] <- match(A[,2],all)
-A[,3] <- match(A[,3],all)
-plotspmat(A)
-
-source("rem.cpp.r")
-source("fns.r")
-N <- length(unique(c(A[,2],A[,3])))
-K <- 2
-P <- 7
-M <- nrow(A)
-
-z <- sample(1:K,N,replace=TRUE)
-beta <- array(rnorm(K^2*P),c(K,K,P))
-system.time(brem.llk(A,N,z,beta,px))
-
-set.seed(4)
-niter <- 100
[…]

### 2012-01-27 docs/notes.tex
Implement approximate likelihood
Add section in notes on EM.

@@
 Computing the likelihood is then $O(T \cdot P \cdot 4N)$ (though this computation is split into $K^2$ pieces) and must computed $N \cdot K$ times per Gibbs iteration.
+The most straightforward implementation has space complexity $O(N^2)$ since we must keep track of $\tau_{i,j,m}$ for each $(i,j)$ at a given $m$.
 
@@
 
+\section{EM for baseline}
+
+In our baseline model, each block $(a,b) \in [1,K] \times [1,K]$ simply has a rate $\gamma_{a,b}$.
+The likelihood is
+\begin{align*}
+\mathcal{L}(A|\lambda) &= \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M\lambda_{ij}\} \\
+& =\prod_{a=1}^K \prod_{b=1}^K \prod_{m=1}^M \gamma_{ab}^{I(z_{i_m}=a,z_{j_m}=b)} \prod_{(i,j): z_i=a,z_j=b} \exp\{-t_M\gamma_{ab}\} \\
+& =\prod_{a=1}^K\prod_{b=1}^K \gamma_{ab}^{n_{ab}} \exp\{-t_Mn_an_b\gamma_{ab}\}
+\end{align*}
+where $n_{ab} = \sum_m I(z_{i_m}=a,z_{j_m}=b)$, $n_a = \sum_{i=1}^N I(z_i=a)$.
+
+The loglikelihood is $$\ell(A|\lambda) = \sum_{a=1}^K \sum_{b=1}^K n_{ab} \log \gamma_{ab} - t_M n_a n_b \gamma_{ab}$$
+and so $\hat{\gamma}_{ab} = \frac{n_{ab}}{t_Mn_an_b}$.
+
 \bibliographystyle{plain}

### 2012-01-30 docs/notes.tex
Small edits to notes.

@@
 Let $\tau_{i,j,m}$ be the time that $\lambda_{i,j}$ changed before the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.
+[TODO: We further assume that all intensity functions change at times $t=0$ and $t=t_M$.
+Also, assume the first event is drawn uniformly from the risk set.]
 % Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
@@
 One could extend our current framework by assuming that after the $m$th event, $i_m$ and $j_m$ redraw their latent class.
+
+Another idea: $\log \lambda_{ij} = (\pi_i \cdot \pi_j) \eta s(t,i,j)$
 
@@
 
+I do not believe it is possible to perform EM for a model where we have the more general linear predictor for the intensity function.
+Since the assignment of one node depends on the assignments of the other nodes, we suffer combinatorial explosion.
 \bibliographystyle{plain}

### 2012-01-31 docs/notes.tex
Organizing and cleaning

@@
 \center
-\includegraphics[width=1.5in]{../figs/syn/mat.pdf}
-\includegraphics[width=2.5in]{../figs/syn/counts.pdf}
-\includegraphics[width=2in]{../figs/syn/bias.pdf}
+\includegraphics[width=2in]{../figs/syn/mat.pdf}
+\includegraphics[width=3in]{../figs/syn/counts.pdf}
 \caption{Illustration of 1000 simulated events, as described in text.
 Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Loglikelihood vs.
-iteration during MCMC.}
+Right: Boxplot of distribution of participation counts across dyads.
+The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.}
 \label{fig:syncounts}
@@
 \subsection*{Prediction experiment}
+
 \begin{figure}
 \center
-\includegraphics[width=3.5in]{../figs/syn/test-recall.pdf}
-\caption{Predictive performance on test data.  Left: Recall on synthetic example shown in Figure \ref{fig:syncounts}.}
-\label{fig:recall}
+\includegraphics[width=3in]{../figs/syn/logposterior.pdf}
+\includegraphics[width=3in]{../figs/syn/test-recall.pdf}
+\caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
+\label{fig:synresults}
 \end{figure}

### 2012-01-31 R/build.r

@@ -1,130 +1,0 @@
-# Deprecated
-build.brem <- function(A,N) {
-  implementedEffects <- c("RSndSnd","RRecSnd","PSAB-BA","PSAB-BY","PSAB-XA","PSAB-XB","PSAB-XY","PSAB-AY")
-  M <- nrow(A)
-  P <- length(implementedEffects)
-  s <- array(0,c(M,N,N,P))
-  dimnames(s) <- list(NULL,NULL,NULL,implementedEffects)
-  rrs <- rss <- matrix(0,N,N)
-  for (m in 1:(M-1)) {
-   
-    a <- A[m,2]
-    b <- A[m,3]
-    
-    # P-shift effects
-    s[m,b,a,"PSAB-BA"] <- 1
-    s[m,b,-a,"PSAB-BY"] <- 1
-    s[m,-b,a,"PSAB-XA"] <- 1
-    s[m,-a,b,"PSAB-XB"] <- 1
-    s[m,-c(a,b),-c(a,b),"PSAB-XY"] <- 1
-    s[m,a,-b,"PSAB-AY"] <- 1
-    
-    # Recency effects
-    ix <- which(rss[a,] > 0)
-    if (length(ix) > 0) {
-      s[m,a,ix,"RSndSnd"] <- 1/(rss[a,ix]) 
-    }
-    ix <- which(rrs[a,] > 0)
-    if (length(ix) > 0) {
-      s[m,a,ix,"RRecSnd"] <- 1/(rrs[a,ix])
-    }
-
-    
-    # Update recency statistics: a just sent to b
-    # rrs[a,b]: rank of b in a's list of people recently received from
-    jx <- which(rrs[b,] > 0)  # b just received from a
-    rrs[b,jx] <- rrs[b,jx] + 1  # src to sent to rec
-    rrs[b,a] <- 1
-    # rss[a,b]: rank of b in a's list of people recently sent to
-    jx <- which(rss[a,] > 0)  # a just sent to b
-    rss[a,jx] <- rss[a,jx] + 1
-    rss[a,b] <- 1
-  }
-  return(s)
-}
-block.ps <- function(A,z) {
-  s <- A[,2]
-  r <- A[,3]
-  zs <- sort(unique(z))
-  setup <- expand.grid(i=zs,j=zs)
-  ds <- lapply(1:nrow(setup),function(k) {
-    ix <- which(z[s] == setup$i[k] & z[r]==setup$j[k])
[…]

### 2012-01-31 R/process/twitter.r

@@ -11,7 +11,7 @@ require(ROAuth)
 require(RCurl)
 source("pull.fns.r")
 options(cores=4)
-load("heroic.oauth.rdata")
+load("rstats/heroic.oauth.rdata")
 
 # xml files obtained from google reader by hand by motifying the continuation code repeatedly
 # http://www.google.com/reader/atom/feed%2Fhttp%3A%2F%2Fsearch.twitter.com%2Fsearch.atom%3Fq%3D%2523rstats?n=1000&c=CPHpgbSjipwC
@@ -41,13 +41,13 @@ grab.tweets.from.xml <- function(f) {
   return(df)
 }
 
-fs <- paste("data/rstats.",0:33,".xml",sep="")
+fs <- paste("data/rstats/rstats.",0:33,".xml",sep="")
 df <- lapply(fs,function(f) grab.tweets.from.xml(f))
 df <- do.call(rbind,df)
-save(df,file="data/rstats.tweets.rdata")
+save(df,file="data/rstats/rstats.tweets.rdata")
 
 # Select users and download their timelines
-load("data/rstats.tweets.rdata")
+load("data/rstats/rstats.tweets.rdata")
 K <- 1000
 tb <- sort(table(df$user),decreasing=TRUE)[1:K]
 plot(log(tb),type="l")
@@ -57,7 +57,7 @@ setup <- expand.grid(page=1:16,user=chosen)
 urls <- sapply(1:nrow(setup), function(i) {
     paste("https://api.twitter.com/1/statuses/user_timeline.json?include_entities=true&include_rts=true&count=200&screen_name=",setup$user[i],"&page=",setup$page[i],sep="")
 })
-save(urls,file="data/rstats.urls.rdata")
+save(urls,file="data/rstats/rstats.urls.rdata")
 
 
 # Try grabbing directed messages among #rstats tweets
@@ -83,7 +83,7 @@ get.recipient <- function(tweet) {
 }
 
 mentions <- lapply(df$text,get.mentions)
-save(mentions,file="data/rstats.mentions.rdata")
+save(mentions,file="data/rstats/rstats.mentions.rdata")
 
 rec <- lapply(df$text,get.recipient)
 length(rec)
@@ -94,51 +94,4 @@ s <- as.character(df$user[ix])
[…]

### 2012-01-31 R/rem.cpp.old.r

@@ -1,136 +1,0 @@
-
-fx <- cxxfunction(,"",includes=
-  '
-int threeDIndex(int j, int k, int l, int J, int K, int L) { 
-  return j*K*L + k*L + l;
-}
-// beta currently: AB-BA,AB-BY,AB-XA,
-Rcpp::List llk(Rcpp::NumericVector beta, Rcpp::NumericVector times, Rcpp::IntegerVector sen, Rcpp::IntegerVector rec, int M, int N1, int N2, int P){
-  Rcpp::NumericVector lrm = Rcpp::NumericVector( Dimension(M,N1,N1) );
-  double llk = 0;
-  int a,b;
-  for (int m = 1; m<M; m++) {
-    a = sen[m-1];
-    b = rec[m-1];
-    for (int i = 0; i<N1; i++) {
-      for (int j = 0; j<N2; j++) {
-        // TODO: Make interface for using different effects cleaner
-        lrm[threeDIndex(m,i,j,M,N1,N2)] = beta[0];
-
-        // PSHIFTS
-        if (i==b & j==a) {
-          lrm[threeDIndex(m,i,j,M,N1,N2)] += beta[1]; //ab-ba
-        }
-        if (i==b & j!=a) {
-          lrm[threeDIndex(m,i,j,M,N1,N2)] += beta[2]; //ab-by
-        }
-        if (i!=b & j==a) {
-          lrm[threeDIndex(m,i,j,M,N1,N2)] += beta[3]; //ab-xa
-        }
-        if (i!=a & j==b) {
-          lrm[threeDIndex(m,i,j,M,N1,N2)] += beta[4]; //ab-xb
-        }
-        if (i!=a & i!=b & j!=a & j!=b) {
-          lrm[threeDIndex(m,i,j,M,N1,N2)] += beta[5]; //ab-xy
-        }
-        if (i==b & j!=b) {
-          lrm[threeDIndex(m,i,j,M,N1,N2)] += beta[6]; //ab-ay
-        }
-
-        // Add to loglikelihood
-        double lambda = lrm[threeDIndex(m,i,j,M,N1,N2)];
-        llk += exp( - (times[m] - times[m-1]) * lambda);
-      }
-    }
-  }
-  return Rcpp::List::create(Rcpp::Named( "lrm" ) = lrm,
-                            Rcpp::Named( "llk" ) = llk);
-}
[…]

### 2012-01-31 R/rem.cpp.r

@@ -1,233 +1,0 @@
-library(inline)
-library(Rcpp)
-
-fx <- cxxfunction(,"",includes=
-  '
-int threeDIndex(int j, int k, int l, int J, int K, int L) { 
-  return l*J*K + k*J + j;
-}
-// http://stackoverflow.com/questions/311703/algorithm-for-sampling-without-replacement
-Rcpp::IntegerVector sampleWithoutReplacement
-(
-  Rcpp::IntegerVector y,    // size of set sampling from
-  int sampleSize
-  )
-{
-  RNGScope scope;
-  // Use Knuths variable names
-  int& n = sampleSize;
-  int N = y.size();
-  Rcpp::IntegerVector samples(n);
-  int t = 0; // total input records dealt with
-  int m = 0; // number of items selected so far
-  while (m < n) {
-    Rcpp::NumericVector u = runif(1);
-    if ( (N - t)*u[0] >= n - m ) {
-      t++;
-    } else {
-      samples[m] = y[t];
-      t++; m++;
-    }
-  }
-  return samples;
-}
-double computeLambda(int i, int j, int a, int b, Rcpp::NumericVector beta,Rcpp::IntegerVector ps) {
-  double lam = 0;
-  //for (int p = 0; p < ps.size(); p++) {
-    //ps[p] = 1;
-    if (ps[0] == 1) {
-      lam += beta[0];
-    }
-    if (ps[1] == 1 & i!=a & i==b & j==a & j!=b) { // ab-ba
-      lam += beta[1];
-    }
-    if (ps[2] == 1 & i!=a & i==b & j!=a & j!=b) { // ab-by
-      lam += beta[2];
-    }
-    if (ps[3] == 1 & i!=a & i!=b & j==a & j!=b) { // ab-xa
-      lam += beta[3];
-    }
-    if (ps[4] == 1 & i!=a & i!=b & j!=a & j==b) { // ab-xb
-      lam += beta[4];
-    }
-    if (ps[5] == 1 & i==a & i!=b & j!=a & j!=b) { // ab-ay
-      lam += beta[5];
-    }
-  // if (ps[6] == 6 & i==a & i!=b & j!=a & j==b) { // ab-ab
-  //    lam += beta[6];
-  //  }
-  //}
-  return lam;
-}
-// All senders, receivers (ix,jx) must be 0-indexed.
-// Current "weirdness": Assumes all events "occur" at time 0.
[…]

### 2012-01-31 R/sim.r

@@ -67,7 +67,7 @@ fit0 <- sbm.mcmc(sim$A,N,K,niter=niter,z=z,gibbs=FALSE)
 fit1 <- brem.mcmc(sim$A,N,K,model.type="diag.rem",niter=niter,z=z,gibbs=FALSE)
 fit2 <- brem.mcmc(sim$A,N,K,model.type="full",niter=niter,z=z,gibbs=FALSE)
 fit3 <- brem.mcmc(sim$A,N,1,model.type="full",niter=niter,gibbs=FALSE,mcmc.sd=.05)
-save(sim,true.lpost,fit0,fit1,fit2,fit3,file="data/syn-fits.rdata")
+save(sim,true.lpost,fit0,fit1,fit2,fit3,file="data/syn/fits.rdata")
 
 llks <- melt(list(base=fit0$llks,diag=fit1$llks,full=fit2$llks,sing=fit3$llks))
 llks$iter <- 1:niter
@@ -124,7 +124,7 @@ lposts <- list(true = brem.lpost(test$A,N,K,z,beta),
                full = brem.lpost(test$A,N,K,fit2$z,fit2$beta),
                sing = brem.lpost(test$A,N,K,fit3$z,fit3$beta))
 unlist(lposts)
-
+save(lposts,file="data/syn/lpost.rdata")
 
 # Look at bias of estimates
 true <- brem.lrm(sim$A,N,z,beta)

### 2012-01-31 R/test.fit.r

@@ -1,15 +1,0 @@
-library(ggplot2)
-source("rem.cpp.r")
-source("fns.r")
-N <- 10
-K <- 1
-z <- rep(1,N)
-P <- 7
-beta <- array(0,c(K,K,P))
-beta[1,1,] <- c(0,3,0,0,0,2,0)
-
-M <- 5000
-set.seed(1)
-sim <- simulate.brem(M,N,z,beta)
-
-drem.mle(sim$A,beta,1:N,1:N)

### 2012-01-31 R/test.rem.cpp.r

@@ -1,149 +1,0 @@
-source("R/brem.r")
-source("R/rem.cpp.r")
-library(testthat)
-
-test_that("lrm and llk functions work on small example",{
-  # Set up example
-  set.seed(1)
-  M <- 4
-  N <- 4
-  P <- 7
-  times <- c(1,2,3,4)
-  sen <- c(1,3,3,1)
-  rec <- c(3,1,1,3)
-  beta <- c(1,1,0,0,0,0,0)
-  ix <- 1:N
-  px <- rep(1,7)
-  
-  # Constract log rate matrix by hand and compare to drem$lrm
-  a <- array(1,c(M,N,N))
-  a[1,,] <- matrix(0,N,N)
-  a[2,3,1] <- 2
-  a[3,1,3] <- 2
-  a[4,1,3] <- 2
-  lrm <- drem$lrm(beta,times,sen-1,rec-1,ix-1,ix-1,px,N,M)
-  expect_that(lrm,equals(a))
-  
-  # Compute log likelihood by hand.  
-  diag(a[1,,]) <- diag(a[2,,]) <- diag(a[3,,]) <- diag(a[4,,]) <- -Inf
-  llks <- c(a[1,sen[1],rec[1]],
-            a[2,sen[2],rec[2]] - (times[2]-times[2-1]) * sum(exp(a[2,,])),
-            a[3,sen[3],rec[3]] - (times[3]-times[3-1]) * sum(exp(a[3,,])),
-            a[4,sen[4],rec[4]] - (times[4]-times[4-1]) * sum(exp(a[4,,])) )
-  sum(llks)
-  
-  # Compare to drem$llk2
-  llk2 <- drem$llk2(lrm,times,sen-1,rec-1,N,M)
-  expect_that(sum(llks),equals(llk2))
-  
-  # Compare to drem$llk
-  px <- c(1,1,0,0,0,0,0)
-  llk3 <- drem$llk(beta,times,sen-1,rec-1,ix-1,ix-1,px,N,M)
-  expect_that(sum(llks),equals(llk3))
-  
-  # Compare to drem$allk
-  px <- c(1,1,0,0,0,0,0)  
-  allk <- drem$allk(beta,times,sen-1,rec-1,ix-1,ix-1,px,N,M,N)
-  expect_that(sum(llks),equals(allk))
-})
-
-N <- 10
-M <- 100
-ix <- 1:5 - 1
-jx <- 5:10 - 1
[…]

### 2012-01-31 R/test.stat.r

@@ -1,44 +1,0 @@
-require(inline)
-require(Rcpp)
-require(RcppEigen)
-require(testthat)
-
-fx <- cxxfunction(,"",includes=
-'
-int llk(Rcpp::NumericVector times, Rcpp::IntegerVector sen, Rcpp::IntegerVector rec, int N1, int N2, int P ){
-  int llk = 0;
-  for (int m = 0; m<M; m++) {
-    for (int i = 0; i<N1; i++) {
-      for (int j = 0; j<N2; j++) {
-        for (int p = 0; p<P; p++) {
-          llk += 1;
-        }
-      }
-    }
-  }
-  return llk;
-}
-RCPP_MODULE(foo){
-  function( "llk", &llk ) ;
-}
-', plugin="Rcpp")
-
-foo <- Module("foo",getDynLib(fx))
-# norm <- function(M,N1,N2,P) M*N1*N2*P
-
-system.time(foo$norm(10000,200,200,4))
-
-transCpp <- '
-using Eigen::Map;
-using Eigen::MatrixXi;
-// Map the integer matrix AA from R
-const Map<MatrixXi> A(as<Map<MatrixXi> >(AA));
-// evaluate and return the transpose of A
-const MatrixXi At(A.transpose());
-return wrap(At);
-'
-ftrans <- cxxfunction(signature(AA="matrix"), transCpp, plugin="RcppEigen") 
-a <- matrix(1:6,nr=2)   
-all.equal(ftrans(a),t(a))
-
-

### 2012-02-11 docs/notes.tex
Add Gibbs sampling to notes.
small changes

@@
 Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.
-[TODO: Explain how our experiments show that this helps.]  By combining a latent variable framework with a local dependence model for event sequences, we provide a hierarchical approach for modeling event-based network data, accompanying recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
+[TODO: Explain how our experiments show that this helps.]   [TODO: perhaps use ``role'' terminology.] By combining a latent variable framework with a local dependence model for event sequences, we provide a hierarchical approach for modeling event-based network data, accompanying recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
 
@@
 
+\section{Gibbs sampling}
+
+Let $\mathcal{U}_a = \{a: a \in (i_m,j_m), m \in [1,M]\}$ be the set of events where actor is involved:
+
+$$p(A|z,\beta,a \ \mbox{involved}) = \prod_{a \in \mathcal{U}_a} \lambda_{i_a,j_a}(t_a|\cdot)
+\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{ija}) \lambda_{ij}(t_a|cdot)\}$$
+
 \section{EM for baseline}

### 2012-02-11 process/synthetic.r

@@ -1,5 +1,6 @@
-library(brem)
+
 library(ggplot2)
+library(brem)
 
 N <- 10
 K <- 2

### 2012-02-11 run.sh

@@ -13,4 +13,14 @@ parallel --sshlogin 8/d9,8/d10,8/d11 -j 4 'cd /extra/duboisc0/blockrem; ./R/brem
 ./R/brem.Rscript -f "data/synthetic.rdata" -k 1 -n 2000 -m "full" -d "results/synthetic"
 
 # Use this now:
-./brem -f "data/synthetic.rdata" -k 2 -n 500 -m "full" -d "results/synthetic" ::: 
+./brem.r -f "data/synthetic.rdata" -k 2 -n 1000 -m {} -d "results/synthetic" -g "slow" ::: "full" "slow"
+
+seq 10|./parallel --sshlogin 8/d10 -j +0 'cd /extra/duboisc0/blockrem/; ./brem.r'
+seq 5|./parallel --sshlogin 8/d9,8/d10 'module load R/2.13.1; cd /extra/duboisc0/blockrem/; ./tmp.r'
+ls | parallel --sshlogin 8/d8,8/d9,8/d10 'echo -n {}" "; ls {}|wc -l'
+
+./parallel --sshlogin 8/d8,8/d9,8/d10 'cd /extra/duboisc0/blockrem;./brem.r -f "data/synthetic.rdata" -k 2 -n 1000 -m {} -d "results/synthetic" -g "slow"'  ::: "full" "shared" "baserates"
+
+cd ~/Documents/blockrem/
+rsync -auvz . duboisc@d1:/extra/duboisc0/blockrem/
+rsync -auvz duboisc@d1:/extra/duboisc0/blockrem/results blockrem/

### 2012-02-16 docs/notes.tex
Edit notes.

@@
 Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.
-One class of statistical models for network data uses latent variables for each node to help capture unobserved heterogeneity.
-For example, the stochastic blockmodel \cite{Nowicki2001, Kemp} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between blocks.
-Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
+One class of statistical models for network data uses latent variables to help model unobserved heterogeneity.
+For example, the stochastic blockmodel \cite{Nowicki2001, Kemp} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between nodes of each block.
+% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
 
 Network data, however, is often collected as a sequence of events occurring over time.
-Recently several works have adapted models from survival analysis and event history analysis to provide continuous time models of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify the dependence of the process on the previous history of events.
+Recently several works have adapted models from event history analysis \cite{} to provide continuous time models of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
+These models allow one to specify how the process depends on the previous history of events.
 In this way one may incorporate theories about the underlying processes and make predictions about future data conditioned on the past.
@@
 Consider email communication in a small organization comprised of several teams.
-Each team may have a different pattern for collaborating via email, and in many cases emails within one team will not affect emailing behavior within a separate team.
+Each team may have a different pattern for collaborating via email, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
 
 Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.
-[TODO: Explain how our experiments show that this helps.]   [TODO: perhaps use ``role'' terminology.] By combining a latent variable framework with a local dependence model for event sequences, we provide a hierarchical approach for modeling event-based network data, accompanying recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
+[TODO: Explain how our experiments show that this helps.]   [TODO: perhaps use ``role'' terminology.]
 
@@
 \mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \\
-&= \prod_{m=1}^M \lambda(t_m|\cdot) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}
+&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}
 \end{align}
-\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is right continuous, and each $\tau_k \in [0,t_m]$.
+\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
 % \begin{align}
@@
 
-It is straightforward extend this to marked point processes where each event contains additional information.
+One can extend the above to marked point processes where each event contains additional information.
 In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ of possible events include all allowed dyads.
-Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ that is piecewise constant.
-As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
-Let $\tau_{i,j,m}$ be the time that $\lambda_{i,j}$ changed before the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.
-[TODO: We further assume that all intensity functions change at times $t=0$ and $t=t_M$.
-Also, assume the first event is drawn uniformly from the risk set.]
-% Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
-% \begin{align}
-% \mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
-% &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \}
-% \end{align}
-% \begin{align}
-% \mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{r=1}^N \exp\{ - (t_m - \tau_{i,r,m}) \lambda_{ir}(t_m | z_i,z_r)  - (t_m - \tau_{r,j,m}) \lambda_{rj}(t_m | z_r,z_j) \}
-% \end{align}
+Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t)$ that is piecewise constant.
+Assume each node belongs to a cluster $z_i$.
+Let $\tau_{i,j,m}$ be the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.
+\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
+
 \begin{align}
@@
 
-For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}_{ij}(t|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$  and the latent cluster assignments $\mathbf{z}$.
-Note the statistics $\mathbf{s}_{ij}$ also do not change until an event involving either $i$ or $j$ occurs.
+For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$  and the latent cluster assignments $\mathbf{z}$.
+As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
+Note the statistics $\mathbf{s}(t,i,j)$ also do not change until an event involving either $i$ or $j$ occurs.
 We model the intensity functions via a log linear model
@@
 where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}_{ij}(t|\cdot)$.
-This is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.
-Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
 
@@
 
-We construct $\mathbf{s}_{ij}$ using a similar set of statistics as used in \cite{Butts2008,Vu2011}.
+We construct $\mathbf{s}(t,i,j)$ using a similar set of statistics as used in \cite{Butts2008,Vu2011}.
 Let $v_{ijm} \in [0,M]$ be the index of the last event where $(i,j)$ occurred (n.b.
@@
 \begin{enumerate}
-\item Sender out-degree: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i)$
-\item Sender in-degree: $s_{2}(t,i,j) = \sum_{m:t_m<t} I(j_m=i)$
-\item Receiver out-degree: $s_{3}(t,i,j) = \sum_{m:t_m<t} I(i_m=j)$
-\item Receiver in-degree: $s_{4}(t,i,j) = \sum_{m:t_m<t} I(j_m=j)$
+\item Dyad count: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i) / m$
+\item Sender out-degree: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i) / m$
+\item Sender in-degree: $s_{2}(t,i,j) = \sum_{m:t_m<t} I(j_m=i) / m$
+\item Receiver out-degree: $s_{3}(t,i,j) = \sum_{m:t_m<t} I(i_m=j)/m$
+\item Receiver in-degree: $s_{4}(t,i,j) = \sum_{m:t_m<t} I(j_m=j)/m$
 \item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
@@
 \item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
+%\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
 % \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
@@
 
-\subsection{Submodels}
+\subsection{Relation to other models}
+
+This is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.
+Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
 
@@
 where $ \lambda_{ij} = \exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
-Each block in this model is homogeneous Poisson process where the dyad associated with each event is draw uniformly from the dyads in that block.
-
-Another important model is one where within group interactions are governed via a set of parameters while all between group interactions are governed by a different set.
-In this case $\boldsymbol{\beta}_{k1,k2} = \boldsymbol{\gamma}$ for all $k1 \ne k2$.
-
-\subsection{Extensions}
-TODO: Explain difficulties in doing a mixed membership model.
-As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.
-In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.
-One could extend our current framework by assuming that after the $m$th event, $i_m$ and $j_m$ redraw their latent class.
-
-Another idea: $\log \lambda_{ij} = (\pi_i \cdot \pi_j) \eta s(t,i,j)$
-
-TODO: Discuss other variants, e.g. HMM that guides which latent class over time?  Or a DP instead of fixed K?
-
-TODO: What about different clusters for senders and receivers (cite our KDD paper)?
-
-TODO: What about doing a similar thing with topic model statistics in s(t,i,j)?
+Analogous to the stochastic block model for static networks, this model posits each dyad is a homogeneous Poisson process and all dyads within block $(z_i,z_j)$ have the intensity $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
 
@@
 
-\subsection{Approximate likelihood methods}
+% \subsection{Approximate likelihood methods}
 
-TODO: Discuss how one might sample the likelihood.  Cite Raftery paper.
+% TODO: Discuss how one might sample the likelihood.  Cite Raftery paper.
 
-TODO: Look at performance of this with respect to bias and performance measures.
-
-% This involves recomputing each $\boldsymbol{\tau}$ and possibly each $\mathbf{s}_{ij}$ if it also depends on $\mathbf{z}$.
-
-%TODO: Decide how to handle label switching.  Schweinberger has an interesting approach I hadn't seen before for his hierarchical ERGMS.  Should look into that more closely...
-
-% \subsection*{Bookkeeping considerations}
-% %For each dyad, we need to keep track of the set of piecewise constant intervals $d_{ij}$.  For a given set of assignments $z$ we can compute these intervals $(a_v,b_v)$.
-
-% Recall we let intensities change whenever an event occurs involving a fellow cluster member.  Thus the vector $\boldsymbol{\tau}_{ij}$ of changepoints (or knots) is given by
-% $$\boldsymbol{\tau}_{ij} = \left\{ t_m: z_{i_m} = z_i \ \mbox{or} \ z_{j_m} = z_j, m=1,\ldots,M\right\}$$
-% Any time we update we change node $k$'s cluster assignment from $a$ to $b$, the change point vectors for any nodes in $a$ and $b$ must be updated.
-For now we assume we have to recompute  $\boldsymbol{\tau}_{ij}$  entirely.
-Similarly, if the covariate vectors $\mathbf{s}_{ij}(t)$ depend on $\mathbf{z}$ then we need to recompute those as well prior to computing the likelihood.%More specifically, for a $D_{ij}$ such that $z_i, z_j \in \{a,b\}$, we must add knots $t_m$ for each.
+% TODO: Look at performance of this with respect to bias and performance measure
 
@@
 We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to continue speaking, and 3) interactions between groups are more likely to be repeated.
-The specification of $\textbf{s}$ is therefore $s_{ij}(t) = [s_0, s_{ij5}(t), s_{ij9}(t), s_{ij11}(t)]$.
+The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{5}(t,i,j), s_{9}(t,i,j), s_{11}(t,i,j)]$.
 For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (2,3,0,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (1,0,0,1)$, and $\boldsymbol{\beta}_{2,2} = (2,0,2,0)$.
@@
 \begin{itemize}
-\item Classroom:  Perhaps within a couple of the classrooms we can infer subgroups who tend to obey different relational event dynamics.
+%\item Classroom:  Perhaps within a couple of the classrooms we can infer subgroups who tend to obey different relational event dynamics.
 \item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
-\item Citation networks: Perhaps some groups of papers obey some types of rich-get-richer effects more than others.
-\item Other
+%\item Citation networks: Perhaps some groups of papers obey some types of rich-get-richer effects more than others.
+\item We collected tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{#rstats}.
+This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
+We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
+Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
+We use the subset of 487 users who participated in more than one event.
 \end{itemize}
@@
 TODO: Discuss if the semiparametric partial likelihood approach can be used in this framework.
+
+\subsection{Extensions}
+TODO: Explain difficulties in doing a mixed membership model.
+As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.
+In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.
+One could extend our current framework by assuming that after the $m$th event, $i_m$ and $j_m$ redraw their latent class.
+
+Another idea: $\log \lambda_{ij} = (\pi_i \cdot \pi_j) \eta s(t,i,j)$
+
+TODO: Discuss other variants, e.g. HMM that guides which latent class over time?  Or a DP instead of fixed K?
+
+TODO: What about different clusters for senders and receivers (cite our KDD paper)?
+
+TODO: What about doing a similar thing with topic model statistics in s(t,i,j)?
+
+\section{Conclusion}
+ We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
+The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models) to provide detailed models of event dynamics among subsets of nodes while simultaneously allowing for heterogeneity in dynamics of the network as a whole.
 

### 2012-02-16 docs/notes.tex
Edits to paper and add refs to git.

@@
  \bigskip
+
+\abstract{TODO}
 
@@
 Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t)$ that is piecewise constant.
-Assume each node belongs to a cluster $z_i$.
-Let $\tau_{i,j,m}$ be the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
+We can then write the likelihood as
 
@@
 \end{align}
+\noindent where event $m$ is $(i_m,j_m)$, $\tau_{i,j,m}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads who have a changepoint in their intensity when $(i_m,j_m)$ occurs.
+\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
 
@@
 
-For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$  and the latent cluster assignments $\mathbf{z}$.
-As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
-Note the statistics $\mathbf{s}(t,i,j)$ also do not change until an event involving either $i$ or $j$ occurs.
-We model the intensity functions via a log linear model
+For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$.
+Though this will be useful, we will constrain the types of dependencies so that an interaction between one dyad does not affect the rate of an entirely separate dyad.
+As shown in Figure \ref{fig:example}, we do this by assuming each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
+This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
+
+We hope to learn about the dynamics between subsets of nodes while allowing for variety among disjoint subsets.
+To facilitate this, we assume each node belongs to a cluster $z_i \sim \mbox{Categorical}(\alpha)$ where $\sum_k \alpha_k = 1$ and  model the intensity functions via a log linear model
 \begin{align}
@@
 where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}_{ij}(t|\cdot)$.
+Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
 
@@
 \begin{enumerate}
+\item Intercept: $s_{0}(t,i,j) = 1$
 \item Dyad count: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i) / m$
@@
 We include degree effects to model possible rich-get-richer phenomenon.
+For example, a particular node sending often may indicate they will continue to be a sender in the future.
+We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
+TODO: Develop some theory about these sorts of statistics.}
+
 The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
@@
 The final set of effects allow one to model phenomenon such as triadic closure.
-Though we only use the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.
+Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.
 The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
@@
 
-This is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.
+Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.
 Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
@@
 \item For $i \in [1,N]$: Sample $z_i$ via Gibbs sampling: 
-$z_i | A, z_{-i} \propto p(A|z,\beta) p(z_i)$
+$z_i | A, z_{-i} \propto p(A|z,\beta) p(z_i)$.  (See Appendix.)
 \item For each pair $(k,k')$: Sample $\beta_{k,k'}$ via MH or HMC.
@@
 
+% TODO: Update the following discussion about sampling.
+
  To perform Gibbs sampling for a given cluster assignment $z_i$, we need to compute the likelihood for each possible assignment of $z_i$.
-This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.
+By limiting the changepoints to times when either node is involved, computing the likelihood is $O(T \cdot P \cdot 4N)$ where 4 comes from considering all dyads where either $i$ or $j$ is involved.
+We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.
+Note that
+% This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.
 Computing the likelihood is then $O(T \cdot P \cdot 4N)$ (though this computation is split into $K^2$ pieces) and must computed $N \cdot K$ times per Gibbs iteration.
@@
 In Figure \ref{fig:syncounts} we illustrate the simulated data and show the loglikelihood during the first iterations of MCMC [after being initialized with $\beta \sim N(\beta_{true},.5^2)$].
+Note that we recover the same cluster assignments and the full model converges to the true loglikelihood.
 
@@
 %\item Citation networks: Perhaps some groups of papers obey some types of rich-get-richer effects more than others.
-\item We collected tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{#rstats}.
+\item We collected tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.
 This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
@@
 Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
-We use the subset of 487 users who participated in more than one event.
+We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
 \end{itemize}
@@
 We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the posterior mean given the training data.
-Recall at a given cutoff $k$ is found by first sorting the predicted intensity functions at each time, then computing the mean rank of the observed events in that list.
+Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.
 
-TODO: Compare to fitting the relational event model without the blockmodel structure.
-Compare to relational event blockmodel without covariates: $p(y_{ij}) = \exp\{\beta_{z_i,z_j}\} / \sum_{(a,b) \in \mathcal{R}}\exp\{\beta_{z_{a},z_{b}}\} $
-Or more simply, just use $s_{ij}(t) = 1$.
+In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.
+At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
+
+Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $\sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $\sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
+Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
+
+We consider several variations of our model to investigate its properties.
+The \texttt{baserates} model only is an intercept only model (and mimics the behavior of a stochastic block model).
+The \texttt{full} model uses all the statistics discussed in Section \ref{sec:modeladditional}.
+For the \texttt{shared} model we force all diagonal blocks to use the same vector ($\boldsymbol{\beta}_{k_1,k_2} = \boldsymbol{\gamma}_1$ where $k_1=k_2$) and all non-diagonal blocks to use the same vector to be the same  ($\boldsymbol{\beta}_{k_1,k_2} = \boldsymbol{\gamma}_2$ where $k_1 \ne k_2$).
+For each of these we may vary the number of clusters $K$.
+The \texttt{single} model uses all statistics but assumes a single cluster $K=1$.
+
+TODO: In Figure \ref{fig:} we show the recall for each dataset.
+
+TODO: In Table \ref{tab:} we show the likelihood on the test dataset.
 
@@
 \section{Discussion}
+
+TODO: Discuss experimental results.  Describe qualitative behavior of each model's recall line.
+
+TODO: Discuss any interpretation of the parameters and clusters for both datasets.
+
 Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches \cite{Gunawardana2011} aim to learn the regions where an intensity is constant and allow for a nonlinear relationship between statistics and the intensity function.
 
+\subsection{Extensions}
 Possible future work could extend the present framework to a mixed-membership model where for each effect there is a latent cluster assignment.
 This way a group of nodes might contact high in-degree nodes at a similar rate, though they vary in the degree to which they adhere to reciprocity.
+As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.
+In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.
+One could extend our current framework by assuming that for each event $m$, any node  $i,j \in \mathcal{R}_{i_m,j_m}$ redraws their latent class.
 
-TODO: Discuss possible variations on the prior structure of $\beta$: 1.
+Possible variants:
+\begin{itemize}
+\item Discuss possible variations on the prior structure of $\beta$: 1.
 Enforce $\beta_{k,l} = \beta_{l,k}$.
@@
 Place other priors that provide different kinds of regularization.
-
-TODO: Discuss if the semiparametric partial likelihood approach can be used in this framework.
-
-\subsection{Extensions}
-TODO: Explain difficulties in doing a mixed membership model.
-As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.
-In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.
-One could extend our current framework by assuming that after the $m$th event, $i_m$ and $j_m$ redraw their latent class.
-
-Another idea: $\log \lambda_{ij} = (\pi_i \cdot \pi_j) \eta s(t,i,j)$
-
-TODO: Discuss other variants, e.g. HMM that guides which latent class over time?  Or a DP instead of fixed K?
-
-TODO: What about different clusters for senders and receivers (cite our KDD paper)?
-
-TODO: What about doing a similar thing with topic model statistics in s(t,i,j)?
+\item $\log \lambda_{ij} = (\pi_i \cdot \pi_j) \eta s(t,i,j)$
+\item  HMM that guides which latent class over time?
+\item  HDP instead of fixed K: Hierarchical, nonparametric model that allows parameters to be shared across blocks.
+\item different clusters for senders and receivers (cite our KDD paper)?
+\item More statistics, e.g. topic model similarity
+\end{itemize}
 
@@
 $$p(A|z,\beta,a \ \mbox{involved}) = \prod_{a \in \mathcal{U}_a} \lambda_{i_a,j_a}(t_a|\cdot)
-\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{ija}) \lambda_{ij}(t_a|cdot)\}$$
+\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{ija}) \lambda_{ij}(t_a|\cdot)\}$$
 

### 2012-02-16 dashboard.r
Implement likelihood that can work on subset of events.  Add test.

@@ -118,11 +118,16 @@ lposts <- lapply(fits,function(f) {
 # unlist(lposts)
 
 
+# trace plots
+library(coda)
+r <- melt(res$param)
+q7 <- qplot(X1,value,data=r, colour=factor(X4),geom="line") + labs(colour="parameters for\n 1x1 block",x="iteration") + theme_bw() + facet_grid(X2~X3)
+
 cat("Creating dashboard.\n")
 library(gridExtra)
 pdf(paste("figs/",opts$dataset,"/dashboard.pdf",sep=""),width=20,height=10)
 blankPanel <- grid.rect(gp=gpar(col="white"))
-grid.arrange(q1, q2, q3, q4, q5, q6, ncol=3)
+grid.arrange(q1, q2, q3, q4, q5, q6, q7, ncol=4)
 dev.off()
 
 #ggsave(paste("figs/",opts$dataset,"/dashboard.pdf",sep=""),width=20,height=10)

### 2012-02-16 pkg/R/utils.r

@@ -223,6 +223,9 @@ get.indices <- function(A,N) {
   return(xs)
 }
 
+#' Compute the rank of the each event in the observed edgelist given an array of intensities
+#' @edgelist Mx3 matrix of event times, senders and recipients
+#' @ratemats MxNxN array of intensities.  
 ranks <- function(edgelist,ratemats,...) {
   M <- nrow(edgelist)
   n <- dim(ratemats)[2]
@@ -235,6 +238,10 @@ ranks <- function(edgelist,ratemats,...) {
   }
   return(r)
 }
+
+#' Compute the recall metric using a vector of ranks
+#' @rs vector of observed ranks
+#' @top vector of cutpoints to compute
 recall <- function(rs,top=c(1:20)) {
   data.frame(k=top,recall=sapply(top,function(k) mean(rs <= k)))
 }
@@ -374,4 +381,56 @@ ratemat.from.marginals <- function(train,test,N) {
   lrm <- array(0,c(M,N,N))
   for (i in 1:M) lrm[i,,] <- r
   return(lrm)
+}
+
+#' Compute likelihood using the log intensity array.
+#' @lrm log intensity array
+#' @times 
+#' @sen
+#' @rec
+#' @z
+#' @N
+#' @M
+brem.llk.slow <- function(lrm,times,sen,rec,z,N,M) {
+  mp <- matrix(1,N,N)
+  llk <- 0
+  llks <- rep(0,4)
+  for (m in 2:(M-1)) {
+    i = sen[m];
+    j = rec[m];
+    zi = z[i];
+    zj = z[j];
+    llk = llk + lrm[m,i,j]
+    
+    for (r in 1:N) {
+      zr = z[r];
+      if (r != i) {
+        lam  = lrm[m,i,r]
+        llk  = llk - (times[m] - times[mp[i,r]]) * exp(lam);
+        lam  = lrm[m,r,i]
+        llk  = llk - (times[m] - times[mp[r,i]]) * exp(lam);
+        mp[i,r] = m;
+        mp[r,i] = m;
+      }
+      if (r != j) {
+        lam  = lrm[m,j,r]
+        llk  = llk - (times[m] - times[mp[j,r]]) * exp(lam);
+        lam  = lrm[m,r,j]
+        llk  = llk - (times[m] - times[mp[r,j]]) * exp(lam);
+        mp[j,r] = m;
+        mp[r,j] = m;
+      }
+    }
[…]

### 2012-02-16 pkg/inst/tests/test.brem.cpp.r

@@ -177,6 +177,10 @@ test_that("lrm and llk functions work on small example for K=2",{
   s$precompute()
   llk3 <- loglikelihood_fast(beta,z-1,s$ptr(),K)
   
+  # Use new likelihood function that can work with a subset of the events
+  llk5 <- loglikelihood_fast_subset(beta,z-1,s$ptr(),K,1:(M-2))
+  expect_that(sum(llk5),equals(sum(llk3)))
+  
   llk4 <- llk_fast(lrm,times,sen-1,rec-1,M,N)
   expect_that(sum(llk3),equals(sum(llk4)))
   

### 2012-02-16 pkg/src/brem.cpp

@@ -447,8 +447,13 @@ Rcpp::List gibbs(Rcpp::IntegerVector ix, Rcpp::NumericVector beta, Rcpp::Integer
 }
 
 
-// Use the precomputed statistics s to compute the likelihood
-Rcpp::NumericVector llkfast(Rcpp::NumericVector beta, Rcpp::IntegerVector z, SEXP statptr_, int K) {
+
+/*
+ Use the precomputed statistics s to compute the likelihood
+ mx: vector of events to compute by default be a vector of 1 to M-2 (inclusive)
+ */
+
+Rcpp::NumericVector llkfast_subset(Rcpp::NumericVector beta, Rcpp::IntegerVector z, SEXP statptr_, int K, Rcpp::IntegerVector mx) {
   RemStat *s = XPtr<RemStat>(statptr_);
   int N = s->N;
   int P = s->P;
@@ -470,8 +475,10 @@ Rcpp::NumericVector llkfast(Rcpp::NumericVector beta, Rcpp::IntegerVector z, SEX
   zi = z[i];
   zj = z[j];
   llks[0] = computeLambdaFast(i,j,zi,zj,s->get_s(0,i,j),beta,N,K,P);
-
-  for (int m = 1; m < (M-1); m++) {
+  
+  for (int v = 0; v < mx.size(); v++) {
+    int m = mx[v];
+  //for (int m = 1; m < (M-1); m++) {
     i = sen[m];
     j = rec[m];
     zi = z[i];
@@ -510,7 +517,7 @@ Rcpp::NumericVector llkfast(Rcpp::NumericVector beta, Rcpp::IntegerVector z, SEX
     llks[m] = llk;
   }
 
-  //All intensities assumed to change at the last event
+  // All intensities assumed to change at the last event
   int m = M-1;
   llk = 0;
   for (int i = 0; i < N; i++) {
@@ -525,6 +532,12 @@ Rcpp::NumericVector llkfast(Rcpp::NumericVector beta, Rcpp::IntegerVector z, SEX
   }
   llks[M-1] = llk;
   return llks;
+}
+Rcpp::NumericVector llkfast(Rcpp::NumericVector beta, Rcpp::IntegerVector z, SEXP statptr_, int K) {
+  RemStat *s = XPtr<RemStat>(statptr_);
+  int M = s->M;
+  Rcpp::IntegerVector mx = Rcpp::seq(1,M-2);
[…]

### 2012-02-19 brem.r
Debuggin slice sampler drop offs.

@@ -39,6 +39,10 @@ if (K > 1 & file.exists(f)) {
   beta <- NULL
 }
 
+px <- rep(1,13)
+px[13] <- 0
+if (K==1) px[1] <- 0 # identifiability
+
 fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice,
-		             niter=opts$numiterations,gibbs=opts$gibbs,beta=beta,px=px,
+                 niter=opts$numiterations,gibbs=opts$gibbs,beta=beta,px=px,
                  outdir=paste("results/",opts$dataset,"/",sep=""))

### 2012-02-19 dashboard.r

@@ -37,12 +37,10 @@ fnames <- unlist(strsplit(dir(results.dir),".rdata"))
 
 
 cat("Plotting log posterior during MCMC.\n")
-niter <- 500
-llks <- lapply(fits,function(f) f$llks)#cbind(llk=f$llks,iter=1:f$niter))
-names(llks) <- fnames
-llks <- melt(llks)
-llks$iter <- 1:niter
-llks <- subset(llks,value > mean(value))
+llks <- lapply(1:length(fits),function(i) {
+  data.frame(model=fnames[i],llk=fits[[i]]$llks,iter=1:fits[[i]]$niter)
+})
+llks <- do.call(rbind,llks)
 
 if (opts$dataset == "synthetic") {
   q1 <- qplot(iter,value,data=subset(llks,iter>10),geom="line",colour=factor(L1)) + labs(x="iteration",y="log posterior",colour="model") + theme_bw() + geom_abline(intercept=true.lpost,slope=0)

### 2012-02-19 pkg/R/slice.r

@@ -52,8 +52,8 @@ uni.slice.evals <- 0	# Number of density evaluations done in these calls
 
 uni.slice <- function (x0, g, w=1, m=Inf, lower=-Inf, upper=+Inf, gx0=NULL)
 {
-  uni.slice.calls <- 0  # Number of calls of the slice sampling function
-  uni.slice.evals <- 0
+  uni.slice.calls <<- 0  # Number of calls of the slice sampling function
+  uni.slice.evals <<- 0
   # Check the validity of the arguments.
   
   if (!is.numeric(x0) || length(x0)!=1
@@ -70,14 +70,17 @@ uni.slice <- function (x0, g, w=1, m=Inf, lower=-Inf, upper=+Inf, gx0=NULL)
   
   # Keep track of the number of calls made to this function.
   
-  uni.slice.calls <- uni.slice.calls + 1
+  uni.slice.calls <<- uni.slice.calls + 1
   
   # Find the log density at the initial point, if not already known.
   
   if (is.null(gx0)) 
-  { uni.slice.evals <- uni.slice.evals + 1
+  { uni.slice.evals <<- uni.slice.evals + 1
     gx0 <- g(x0)
   }
+
+  # TEMP: Debugging
+  if (gx0 != g(x0)) stop("incorrect gx0 supplied")
   
   # Determine the slice level, in log terms.
   
@@ -96,14 +99,16 @@ uni.slice <- function (x0, g, w=1, m=Inf, lower=-Inf, upper=+Inf, gx0=NULL)
   { 
     repeat
     { if (L<=lower) break
-      uni.slice.evals <- uni.slice.evals + 1
+      uni.slice.evals <<- uni.slice.evals + 1
+      if (uni.slice.evals > 50) browser()
       if (g(L)<=logy) break
       L <- L - w
     }
     
     repeat
     { if (R>=upper) break
-      uni.slice.evals <- uni.slice.evals + 1
+      uni.slice.evals <<- uni.slice.evals + 1
+      if (uni.slice.evals > 50) browser()
       if (g(R)<=logy) break
       R <- R + w
     }
@@ -116,7 +121,9 @@ uni.slice <- function (x0, g, w=1, m=Inf, lower=-Inf, upper=+Inf, gx0=NULL)
     
     while (J>0)
     { if (L<=lower) break
-      uni.slice.evals <- uni.slice.evals + 1
[…]

### 2012-02-19 run.sh

@@ -10,16 +10,14 @@
 ./brem.r -f "data/eckmann-small.rdata" -k 2 -n 500 -m "full" -d "results/eckmann-small" -g "fast"
 ./brem.r -f "data/eckmann-small.rdata" -k 1 -n 500 -m "full" -d "results/eckmann-small" -g "fast"
 
-opts=list(dataset="synthetic",numclusters=1,model.type="full",niter=10,gibbs="fast",numiterations=10,slice=TRUE)
+opts=list(dataset="eckmann-small",numclusters=1,model.type="full",niter=10,gibbs="fast",numiterations=100,slice=TRUE)
 
 ./parallel --sshlogin 8/d6 'cd /extra/duboisc0/blockrem;./brem.r -d {} -k 1 -n 500 -m "full"' ::: "synthetic" "eckmann-small" "twitter-small"
 
 ./parallel --sshlogin 8/d5 'cd /extra/duboisc0/blockrem;./brem.r -d {} -k 2 -n 500 -m "shared"' ::: "synthetic" "eckmann-small" 
 
 # Fit models
-
-./parallel --sshlogin 8/d5 'cd /extra/duboisc0/blockrem;./brem.r -d {} -k 1 -n 500 -m "full"' ::: "synthetic" "eckmann-small"
-
+cd /extra/duboisc0/blockrem;./brem.r -d "twitter-small" -k 1 -n 500 -m "full" 
 ./parallel --sshlogin 8/d5 'cd /extra/duboisc0/blockrem;./brem.r -d "synthetic" -k 2 -n 500 -m {}' ::: "baserates" "shared" "full"
 
 ./parallel --sshlogin 8/d5 'cd /extra/duboisc0/blockrem;./brem.r -d "eckmann-small" -k 2 -n 500 -m {}' ::: "baserates" "shared" "full"

### 2012-02-19 synthetic.example.r

@@ -1,7 +1,6 @@
 opts <- list(dataset="synthetic",numclusters=1,model.type="full",gibbs="fast",numiterations=20,slice=TRUE)
 
 load(paste("data/",opts$dataset,".rdata",sep=""))
-
 library(brem)
 
 # Precompute data structures
@@ -13,6 +12,8 @@ s <- new(RemStat,train[,1],as.integer(train[,2])-1,as.integer(train[,3])-1,N,M,P
 s$precompute()
 
 # Degree vs. no degree effects, slice sampling, K=1 full
+source("pkg/R/brem.r")
+source("pkg/R/slice.r")
 px <- rep(1,13)
 px[8:13] <- 0
 fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice,

### 2012-02-21 docs/notes.tex
Implement likelihood that can work on subset of events.  Add test.
Debuggin slice sampler drop offs.
Update draft with some tikz

@@
 \documentclass[11pt]{article}
-\usepackage{graphicx}
+\usepackage{tikz}
+\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
+
+%\usepackage{tikz,fullpage}
+%\usetikzlibrary{arrows,petri,topaths}%
+%\usepackage{tkz-berge}
 \usepackage{verbatim}
@@
 \usepackage{amsmath}
+\usepackage{subfigure}
 %\usepackage[small,compact]{titlesec} 
@@
 
-\abstract{TODO}
+\abstract{
+
+\begin{itemize}
+\item Blockmodels and clustering algorithms often applied to static networks for exploratory data analysis
+\item continuous time methods allow one to predict future events given the previous history
+\item we combine these two ideas with a hierarchical model such that each pair of blocks has a set of parameters guiding the event dynamics for those dyads
+\item prediction experiments on real datasets
+\item show that the parameter estimates are interpretable
+\end{itemize}
+
+}
 
@@
 Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.
-[TODO: Explain how our experiments show that this helps.]   [TODO: perhaps use ``role'' terminology.]
+The behavior of the model is illustrated with data simulated from the model.
+Using  data sets of dyadic communication among people, we demonstrate the ability to improve upon simple baselines for predicting the next event.
+Finally we show that the parameter estimates provide an interpretable structure to the event dynamics among blocks of individuals, sometimes identifying particular \emph{roles} of communication that exist within the network.
 
@@
 \begin{figure}
- \def\svgwidth{6in}
-  \input{example1.pdf_tex}
+
+\centering 
+\subfigure[Dynamic network data]
+{
+\begin{tikzpicture}[scale=.8,transform shape,thick]
+  \node[vertex] (n1) at (0,0)  {1};
+  \node[vertex] (n2) at (2,2)  {2};
+  \node[vertex] (n3) at (4,0) {3};
+  \node[vertex] (n4) at (2,-2) {4};
+\draw[-stealth] (.4,.4) -- (1.6,1.6);
+\draw[-stealth] (2.4,1.6) -- (3.6,.4);
+\node at (.7,1.1) {$t_1$};
+\node at (3.3,1.1) {$t_2$};
+\end{tikzpicture}
+}
+\subfigure[Intensities for two dyads]
+{
+\begin{tikzpicture}[scale=1,transform shape,thick]
+\draw[-stealth] (0,0) -- (0,3);
+\draw[-stealth] (0,0) -- (5,0);
+
+% First timeline
+\draw (0,2) -- (1,2);
+\draw[fill=black] (1,2) circle (0.75mm);
+\draw (1,1.7) circle (0.75mm);
+\draw (1.08,1.7) -- (3,1.7);
+\draw[fill=black] (3,1.7) circle (0.75mm);
+\draw (3,1.7) circle (0.75mm);
+\draw[fill=black] (3,2.2) circle (0.75mm);
+\draw[-stealth] (3,2.2) -- (5,2.2);
+
+% second timeline
+\draw[dashed] (0,1) -- (3,1);
+\draw[fill=black] (3,1) circle (0.75mm);
+\draw (3,.75) circle (0.75mm);
+\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
+
+% labels
+\node at (-.5,2) {$\lambda_{1,2}$};
+\node at (-.5,1) {$\lambda_{3,4}$};
+\node at (1,-.3) {$t_1$};
+\node at (3,-.3) {$t_2$};
+\end{tikzpicture}
+}
 \caption{Illustration of event data and the assumptions of the model.
@@
 Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
+[TODO?
+Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.]}
+\label{fig:example}
+\end{figure}
+
+% \begin{figure}
+%  \def\svgwidth{6in}
+%   \input{example1.pdf_tex}
+% \caption{Illustration of event data and the assumptions of the model.
+Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
+Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
+Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
 Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-\label{fig:example}
-%\includegraphics[width=6in]{example}
-\end{figure}
+% \label{fig:example}
+% %\includegraphics[width=6in]{example}
+% \end{figure}
 
@@
 \begin{itemize}
-%\item Classroom:  Perhaps within a couple of the classrooms we can infer subgroups who tend to obey different relational event dynamics.
-\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
-%\item Citation networks: Perhaps some groups of papers obey some types of rich-get-richer effects more than others.
-\item We collected tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.
-This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
+%\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
+\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
 We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
@@
 We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
+\item The Irvine data set
 \end{itemize}
@@
 
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $\sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $\sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
+Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg} = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
 Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
+
+Our method jointly models both which dyads occur and when they occur.
+To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j} = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
 
@@
 
+TODO: discuss non-specification of rate and just predicting next
+event, could still use our approach?
+
 Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches \cite{Gunawardana2011} aim to learn the regions where an intensity is constant and allow for a nonlinear relationship between statistics and the intensity function.
@@
 One could extend our current framework by assuming that for each event $m$, any node  $i,j \in \mathcal{R}_{i_m,j_m}$ redraws their latent class.
+TODO: Discuss why we *didn't* do this.
 
@@
 \end{itemize}
+
+\section{Related work}
+\cite{Tillman2011} explores the idea of having nonpoisson processes at each edge and solves the master equation for the dynamical system.
 

### 2012-02-22 brem.r
Using slice sampling (with m=20 to avoid getting stuck in infinite loops) samples the betas better and moves around from initial solution.

@@ -1,5 +1,7 @@
 #!/usr/bin/env Rscript
 suppressPackageStartupMessages(library("optparse"))
+suppressPackageStartupMessages(library("brem"))
+source("pkg/R/brem.r")
 
 option_list <- list(
   make_option(c("-d", "--dataset"), 
@@ -13,14 +15,17 @@ option_list <- list(
   make_option(c("-g","--gibbs"), default="fast",
               help="Slow or fast version of gibbs."),
   make_option(c("-s","--slice"), default=FALSE,
-              help="Slice sample instead of MH.")
+              help="Slice sample instead of MH."),
+  make_option(c("-i","--initialize"), default=FALSE,
+              help="initialize with K=1 solution"),
+  make_option(c("-f","--fixz"), default=FALSE,
+              help="if twitter, fix z to predetermined values")
   )
 parser <- OptionParser(usage = "%prog [options] file", option_list=option_list)
 opts   <- parse_args(OptionParser(option_list=option_list))
 
 load(paste("data/",opts$dataset,".rdata",sep=""))
 
-library(brem)
 
 # Precompute data structures
 # N should be loaded by dataset
@@ -31,19 +36,33 @@ s <- new(RemStat,train[,1],as.integer(train[,2])-1,as.integer(train[,3])-1,N,M,P
 s$precompute()
 
 # Initialize with K=1 solution, if available
-#f <- paste("results/",opts$dataset,"/full.1.rdata",sep="")
-#if (K > 1 & file.exists(f)) {
-#  load(f)
-#  beta <- array(res$beta,c(P,K,K))
-#} else {
-#  beta <- NULL
-#}
-beta = NULL
+f <- paste("results/",opts$dataset,"/full.1.rdata",sep="")
+if (K > 1 & file.exists(f) & opts$initialize) {
+ load(f)
+ beta <- array(res$beta,c(P,K,K))
+} else {
+ beta <- NULL
+}
 
 px <- rep(1,13)
 px[13] <- 0
+px[7]  <- 0
 if (K==1) px[1] <- 0 # identifiability
 
+if (opts$dataset=="twitter-small" & opts$fixz) {
+  tb <- table(factor(c(train[,2],train[,3]),1:N))
+  chosen <- names(tb)[which(tb > 20)]
+  chosen <- as.numeric(chosen)
+  z <- rep(1,N)
+  z[chosen] <- 2
+  opts$gibbs <- FALSE
+} else {
[…]

### 2012-02-22 pkg/R/brem.r

@@ -112,7 +112,7 @@ brem.lpost.fast.block <- function(A,N,K,z,s,beta,k1,k2,priors=list(beta=list(mu=
   }    
 }
 
-brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,beta=NULL,z=NULL,gibbs="fast",mh=FALSE,outdir=getwd(),priors=list(beta=list(mu=0,sigma=1)),verbose=FALSE,px=NULL) {
+brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,m=20,beta=NULL,z=NULL,gibbs="fast",mh=FALSE,outfile=NULL,priors=list(beta=list(mu=0,sigma=1)),verbose=FALSE,px=NULL) {
   llks <- rep(0,niter)
   M <- nrow(A)
   P <- 13
@@ -131,8 +131,9 @@ brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,beta=NULL,z=N
   
   olp <- brem.lpost.fast(A,N,K,z,s,current)
   
+  deltas <- matrix(0,niter,5)
   for (iter in 1:niter) {
-    
+    stime <- proc.time()
     # For each effect sample via MH
     if (mh) {
       for (i in 1:2) {
@@ -141,7 +142,7 @@ brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,beta=NULL,z=N
         olp <- res$olp
       }
     } else {
-      res <- brem.slice(A,N,K,P,z,s,current,px,model.type,priors,olp)
+      res <- brem.slice(A,N,K,P,z,s,current,px,model.type,priors,olp,m=m)
       current <- res$current
       olp <- res$olp
     }
@@ -159,6 +160,7 @@ brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,beta=NULL,z=N
       }
     }
     if (gibbs=="fast") {
+      cat("gibbs")
       z <- gibbs(1:N-1,current,z-1,s$ptr(),K)$z + 1
     }
     
@@ -166,11 +168,16 @@ brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,beta=NULL,z=N
     param[iter,,,] <- current
     llks[iter] <- brem.lpost.fast(A,N,K,z,s,current,priors)
     
[…]

### 2012-02-22 synthetic.example.r

@@ -2,6 +2,7 @@ opts <- list(dataset="synthetic",numclusters=1,model.type="full",gibbs="fast",nu
 
 load(paste("data/",opts$dataset,".rdata",sep=""))
 library(brem)
+source("pkg/R/brem.r")
 
 # Precompute data structures
 N <- max(c(train[,2],train[,3]))
@@ -18,7 +19,7 @@ px[8:13] <- 0
 px[7] <- 0
 fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice,
                  niter=opts$numiterations,gibbs=opts$gibbs,px=px,
-                 outdir=NULL)
+                 outfile=NULL)
 fit1 <- fit
 
 px <- rep(1,13)
@@ -27,7 +28,7 @@ px[7] <- 0
 #px[8]  <- 0
 fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice,
                  niter=opts$numiterations,gibbs=opts$gibbs,px=px,
-                 outdir=NULL)
+                 outfile=NULL)
 fit2 <- fit
 
 # look at mixing for both
@@ -45,7 +46,7 @@ px[8:13] <- 0
 px[7] <- 0
 fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice,
                  niter=opts$numiterations,gibbs=opts$gibbs,px=px,
-                 outdir=NULL)
+                 outfile=NULL)
 fit1 <- fit
 
 px <- rep(1,13)
@@ -53,7 +54,7 @@ px[13] <- 0
 px[7] <- 0
 fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice,
                  niter=opts$numiterations,gibbs=opts$gibbs,px=px,#beta=beta,
-                 outdir=NULL)
+                 outfile=NULL)
 fit2 <- fit
 
 # Use shared model

### 2012-02-22 twitter.example.r

@@ -1,17 +1,15 @@
 # Get estimates for full 2 when we manually set the groups
-
+library(brem)
 load("data/twitter-small.rdata")
+source("pkg/R/brem.r")
 
 tb <- table(factor(c(train[,2],train[,3]),1:N))
 chosen <- names(tb)[which(tb > 20)]
 chosen <- as.numeric(chosen)
-
 z <- rep(1,N)
 z[chosen] <- 2
 
-
-opts <- list(slice=FALSE,gibbs=FALSE,model.type="full",dataset="twitter-small",numiterations=500,numclusters=2)
-
+opts <- list(slice=FALSE,gibbs=FALSE,model.type="full",dataset="twitter-small",numiterations=500,numclusters=2,initialize=TRUE)
 
 # Precompute data structures
 M <- nrow(train)
@@ -22,27 +20,70 @@ s$precompute()
 
 px <- rep(1,13)
 px[13] <- 0
-beta <- NULL
+f <- paste("results/",opts$dataset,"/full.1.rdata",sep="")
+if (K > 1 & file.exists(f) & opts$initialize) {
+  load(f)
+  beta <- array(res$beta,c(P,K,K))
+} else {
+  beta <- NULL
+}
+
+# fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice, mcmc.sd=.05,m=5,
+#                  niter=opts$numiterations, gibbs=opts$gibbs,  beta=beta, px=px, z=z,   
+#                  outfile="tmp.slice.m5.rdata")
+# 
+
+opts$gibbs <- "fast"
+opts$slice <- TRUE
+# fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice, mcmc.sd=.05,m=20,
+#                  niter=opts$numiterations, gibbs=opts$gibbs,  beta=beta, px=px, z=z,   
+#                  outfile="tmp.slice.gibbs.rdata")
+
+
+fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice, mcmc.sd=.05,m=20,
+                 niter=opts$numiterations, gibbs=opts$gibbs,  beta=beta, px=px,  
+                 outfile="tmp.slice.gibbs.learn2.rdata")
+
+# Visualzie model paramters
+
+load("tmp.slice.rdata")
+r <- melt(res$param)
+r <- subset(r,X1<10)
[…]

### 2012-03-07 docs/notes.tex
Using slice sampling (with m=20 to avoid getting stuck in infinite loops) samples the betas better and moves around from initial solution.
Modify example for online debugging.

@@
 
- \centerline{\Large \bf Blockmodels for Dynamic Network Data} 
+ \centerline{\Large \bf Stochastic Blockmodels for Relational Events} 
  \medskip
@@
 
-\begin{itemize}
-\item Blockmodels and clustering algorithms often applied to static networks for exploratory data analysis
-\item continuous time methods allow one to predict future events given the previous history
-\item we combine these two ideas with a hierarchical model such that each pair of blocks has a set of parameters guiding the event dynamics for those dyads
-\item prediction experiments on real datasets
-\item show that the parameter estimates are interpretable
-\end{itemize}
+Stochastic blockmodels are a tool for exploring the structure of static social networks by learning latent groups of individuals who have similar tendencies in their  group-wise interactions.
+Continuous-time methods, such as the relational event framework (Butts 2008), model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.
+We propose a hierarchical model combining these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.
+The  method is illustrated using  dyadic interaction data (e.g.
+email, Twitter) among hundreds of actors and thousands of events.
+We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than a single relational event model for all actors in the network.
+Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
+
+% \begin{itemize}
+% \item Blockmodels and clustering algorithms often applied to static networks for exploratory data analysis
+% \item continuous time methods allow one to predict future events given the previous history
+% \item we combine these two ideas with a hierarchical model such that each pair of blocks has a set of parameters guiding the event dynamics for those dyads
+% \item prediction experiments on real datasets
+% \item show that the parameter estimates are interpretable
+% \end{itemize}
 
@@
 \draw (3,1.7) circle (0.75mm);
-\draw[fill=black] (3,2.2) circle (0.75mm);
-\draw[-stealth] (3,2.2) -- (5,2.2);
+\draw (3,2.2) circle (0.75mm);
+\draw[-stealth] (3.05,2.2) -- (5,2.2);
 
@@
 % TODO: Update the following discussion about sampling.
+
+Slice sampling with $m=20$.
 

### 2012-03-07 online.debug.r

@@ -55,6 +55,14 @@ get.pred <- function(train,A,test.ix,fit) {
 #   }
   list(lrm=lrm,m=list(train=m.train,test=m.test))
 }
+multinomial.score <- function(m,x) {
+  M <- nrow(x)
+  r <- rep(0, M)
+  for (i in 1:M) {
+    r[i] <- m[i,x[i,2],x[i,3]]
+  }
+  return(r)
+}
 
 # Compare fit of abba, abby, dyad count to baseline on eckmann
 opts <- list(dataset="eckmann-small",numclusters=K,model.type="full",gibbs="none",numiterations=10,slice=TRUE)
@@ -67,12 +75,16 @@ K <- 1#opts$numclusters
 s <- new(RemStat,train[1:M,1],as.integer(train[1:M,2])-1,as.integer(train[1:M,3])-1,N,M,P)
 s$precompute()
 px <- rep(0,13)
-px[c(2,3,12)] <- 1
+#px[c(2,3,12)] <- 1
+px[c(12)] <- 1
 fit <- brem.mcmc(train[1:M,],N,K,s,model.type=opts$model.type,mh=!opts$slice,
                  niter=opts$numiterations,gibbs=opts$gibbs,beta=NULL,px=px,
                  outfile=NULL)
 save(fit,file="tmp.rdata")
 
+load("tmp.rdata")
+fit$beta[2:3] <- 0
+fit$beta[12] <- 20
 preds <- get.pred(train,A,test.ix,fit)
 mllk.train <- log(multinomial.score(preds$m$train,train))
 llk.train <- loglikelihood_vec_from_lrm(preds$lrm$train,train[1:M,1],as.integer(train[1:M,2])-1,as.integer(train[1:M,3])-1,N,M)
@@ -88,6 +100,103 @@ abline(0,1)
 plot(llk.train,llk.train.online,asp=1,pch=".")
 abline(0,1)
 dev.off()
+
+############################################
+# After meeting with Smyth
+
+# Compute p_ij(m)
+m.train <- ratemat.online(train,N)
+m.train[which(m.train==-Inf)] <- 0
+eps <- 1
+for (i in 1:nrow(train)) {
+  lam <- m.train[i,,] + eps
+  m.train[i,,] <- lam/sum(lam)
+}
+m.train.online <- m.train
+
+# Compute log lambda_ij(m)
+fit$beta[2:3] <- 0
+fit$beta[12] <- 1#exp(1)
[…]

### 2012-03-07 pkg/src/brem.cpp

@@ -302,7 +302,8 @@ double computeLambdaFast(int i, int j, int zi, int zj, vector<int> s, Rcpp::Nume
   }
   double numEvents = double(s[12]);
   for (int p = 7; p < 12; p++) {
-    lam += s[p]/(numEvents + 1) * beta[threeDIndex(p,zi,zj,P,K,K)];
+    //    lam += s[p]/(numEvents + 1) * beta[threeDIndex(p,zi,zj,P,K,K)];
+    lam += log((s[p]+1)/(numEvents + N*(N-1))) * beta[threeDIndex(p,zi,zj,P,K,K)];
   }
   return lam;
 }
@@ -658,7 +659,8 @@ double computeLambda(int i, int j, int zi, int zj, Rcpp::NumericVector s, Rcpp::
   }
   double numEvents = double(s[threeDIndex(12,i,j,P,N,N)]);//s[12];
   for (int p = 7; p < 12; p++) {
-    lam += s[threeDIndex(p,i,j,P,N,N)]/(numEvents+1) * beta[threeDIndex(p,zi,zj,P,K,K)];
+    //    lam += s[threeDIndex(p,i,j,P,N,N)]/(numEvents+1) * beta[threeDIndex(p,zi,zj,P,K,K)];
+    lam += log((s[threeDIndex(p,i,j,P,N,N)] + 1)/(numEvents + N*(N-1))) * beta[threeDIndex(p,zi,zj,P,K,K)];
   }
 
   return lam;

### 2012-03-12 docs/slides/slides.tex
Start on slides for Sunbelt.

\documentclass{beamer}
\usetheme{default}
\begin{document}

\begin{frame}{Stochastic blockmodels}

Idea: 
\begin{itemize}
\item Each actor belongs to a latent class. 
\item Model block-wise interactions (mixing rates).
\end{itemize}

\includegraphics[scale=.5]{../../figs/synthetic/mat}
\includegraphics[scale=.5]{../../figs/synthetic/mat}
\includegraphics[scale=.5]{../../figs/synthetic/bm}

$$\boldsymbol{\eta} = \left[
\begin {array}{cc}
 \eta_{11}& \eta_{12} \\
 \eta_{11}& \eta_{12} 
\end {array}
\right]$$

% \begin{align*}
%$$ \eta = \left[ 
% \begin{tabular}{cc}
% \eta_{11}& \eta_{12} \\
% \eta_{11}& \eta_{12} 
% \end{tabular} 
%\right]$$
% \end{align*}

\begin{align*}
y_{ij} \sim& F(\eta_{z_i,z_j}) \\
z_i \sim& \mbox{Categorical}(\pi)
\end{align*}

\end{frame}

\end{document}

### 2012-03-12 docs/slides/slides.tex
Add material to slides.

@@
 \usetheme{default}
+\usepackage{tikz}
+\usepackage{subfigure}
+\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
+\title[Blockmodels for relational events]%(optional, only for long titles)
+{Stochastic blockmodels for relational events}
+\subtitle{}
+\author[DuBois, Christopher] % (optional, for multiple authors)
+{Christopher~DuBois\inst{1} \and Carter~Butts\inst{2} \and Padhraic~Smyth\inst{3}}
+\institute[] % (optional)
+{
+  \inst{1}%
+Department of Statistics \\
+University of California, Irvine
+  \and
+  \inst{2}%
+Department of Sociology \\
+University of California, Irvine
+  \and
+  \inst{2}%
+Department of Computer Science \\
+University of California, Irvine
+}
+\date[Sunbelt 2012] % (optional)
+{Sunbelt 2012}
+\subject{Informatik}
 \begin{document}
+\frame{\titlepage}
+
+\begin{frame}{Dyadic event data}
+\begin{figure}
+{
+\begin{tikzpicture}[scale=1,transform shape,thick]
+  \node[vertex] (n1) at (0,0)  {1};
+  \node[vertex] (n2) at (2,2)  {2};
+  \node[vertex] (n3) at (4,0) {3};
+  \node[vertex] (n4) at (2,-2) {4};
+\draw[-stealth] (.4,.4) -- (1.6,1.6);
+\draw[-stealth] (2.4,1.6) -- (3.6,.4);
+\node at (.7,1.1) {$t_1$};
+\node at (3.3,1.1) {$t_2$};
+\end{tikzpicture}
+}
+\end{figure}
+
+\begin{itemize}
+\item Above example: 2 dyadic events occurring among 4 nodes
+\item Nodes may have covariates
+\item Interested in inferences about the rate of particular events, conditioned on what we have observed
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Relational event models}
+
+Multiplicative models for the hazard:
+$$\log \lambda_{ij}(t | \cdot) = \boldsymbol{\beta}' \mathbf{s}(t,i,j)$$
+
+Likelihood:
+
+\begin{align}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | .) \}
+\end{align}
+
+\begin{itemize}
+\item $t_m$: time of event $m$
+\item $\tau_{i,j,m}$: changepoint time $\lambda_{ij}(t)$ prior to $t_m$
+\item $\mathcal{R}_{i,j}$: set of dyads involving either $i$ or $j$
+\end{itemize}
+
+\end{frame}
 
@@
 
-Idea: 
 \begin{itemize}
@@
 
-\includegraphics[scale=.5]{../../figs/synthetic/mat}
+\includegraphics[scale=.5]{../../figs/synthetic/unsorted}
 \includegraphics[scale=.5]{../../figs/synthetic/mat}
@@
 
-$$\boldsymbol{\eta} = \left[
+\vspace{-.8cm}
+\begin{align*}
+\mbox{Mixing rates} & & \boldsymbol{\eta} =& \left[
 \begin {array}{cc}
  \eta_{11}& \eta_{12} \\
- \eta_{11}& \eta_{12} 
+ \eta_{21}& \eta_{22} 
 \end {array}
-\right]$$
-
-% \begin{align*}
-%$$ \eta = \left[ 
-% \begin{tabular}{cc}
-% \eta_{11}& \eta_{12} \\
-% \eta_{11}& \eta_{12} 
-% \end{tabular} 
-%\right]$$
-% \end{align*}
-
-\begin{align*}
-y_{ij} \sim& F(\eta_{z_i,z_j}) \\
-z_i \sim& \mbox{Categorical}(\pi)
+\right]
+= \left[
+\begin {array}{cc}
+.8 & .2\\.2 & .8
+\end {array}
+\right] \\
+\mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
+\mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi)
 \end{align*}
@@
 
+\begin{frame}{Blockmodels for relational event data}
+Idea:
+  \begin{itemize}
+  \item Assume each actor $i$ belongs to a latent class $z_i$.
+  \item Model event dynamics between blocks.
+  \end{itemize}
+
+\begin{align}
+\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z}).
+\end{align}
+
+Notation:
+\begin{itemize}
+\item $\boldsymbol{\beta}_{k,l}$: parameter vector for relational events from an actor in block $k$ to an actor in block $l$
+\item  $\mathcal{A}_t$: history prior to time $t$
+\item $\mathbf{z}$: vector of class assignments
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Model specification for $X_{ij}(t)$}
+\textbf{Participation shifts} adjust for the next event
+\begin{itemize}
+\item AB-BA
+\item AB-BY
+\item AB-AY
+\item AB-XA
+\item AB-XB
+\end{itemize}
+
+\textbf{Degree effects} count the number of previous events
+\begin{itemize}
+\item sent by sender
+\item received by sender
+\item sent be receiver
+\item received by receiver
+\item previous occurrences
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Specifying degree effects (and explosion)}
+
+\end{frame}
+
+\begin{frame}{Scalability}
+Only use covariates for $(i,j)$ that involve either $i$ or $j$.
+
+\begin{figure}
+
+\centering 
+\subfigure[Dynamic network data]
+{
+\begin{tikzpicture}[scale=.8,transform shape,thick]
+  \node[vertex] (n1) at (0,0)  {1};
+  \node[vertex] (n2) at (2,2)  {2};
+  \node[vertex] (n3) at (4,0) {3};
+  \node[vertex] (n4) at (2,-2) {4};
+\draw[-stealth] (.4,.4) -- (1.6,1.6);
+\draw[-stealth] (2.4,1.6) -- (3.6,.4);
+\node at (.7,1.1) {$t_1$};
+\node at (3.3,1.1) {$t_2$};
+\end{tikzpicture}
+}
+\subfigure[Intensities for two dyads]
+{
+\begin{tikzpicture}[scale=1,transform shape,thick]
+\draw[-stealth] (0,0) -- (0,3);
+\draw[-stealth] (0,0) -- (5,0);
+
+% First timeline
+\draw (0,2) -- (1,2);
+\draw[fill=black] (1,2) circle (0.75mm);
+\draw (1,1.7) circle (0.75mm);
+\draw (1.08,1.7) -- (3,1.7);
+\draw[fill=black] (3,1.7) circle (0.75mm);
+\draw (3,1.7) circle (0.75mm);
+\draw (3,2.2) circle (0.75mm);
+\draw[-stealth] (3.05,2.2) -- (5,2.2);
+
+% second timeline
+\draw[dashed] (0,1) -- (3,1);
+\draw[fill=black] (3,1) circle (0.75mm);
+\draw (3,.75) circle (0.75mm);
+\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
+
+% labels
+\node at (-.5,2) {$\lambda_{1,2}$};
+\node at (-.5,1) {$\lambda_{3,4}$};
+\node at (1,-.3) {$t_1$};
+\node at (3,-.3) {$t_2$};
+\end{tikzpicture}
+}
+\end{figure}
+
+\begin{itemize}
+\item Restricts the class of effects one can use
+\item Complexity of likelihood computation reduced ($N^2 \rightarrow N$)
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Synthetic example: 2000 events among 10 actors}
+\includegraphics[scale=.12]{../../figs/synthetic/network}
+\includegraphics[scale=.7]{../../figs/synthetic/mat}
+\end{frame}
+
+\begin{frame}{Synthetic example}
+Parameter estimates
+
+\end{frame}
+
 \end{document}

### 2012-03-12 process/synthetic.r

@@ -72,6 +72,20 @@ mat$X2 <- factor(mat$X2,1:10)
 plotmat(mat,limits=c(-10,3))
 ggsave("figs/synthetic/bm.pdf",width=3,height=3)
 
+library(sna)
+edgelist <- melt(table(sim$A[,2],sim$A[,3]))
+colnames(edgelist) <- c("sen","rec","val")
+mypal <- brewer.pal(9,"Greys")
+edge.colors <- col2rgb(mypal[as.numeric(cut(edgelist$val,9))])
+edge.colors <- rbind(edge.colors,100)
+edgelist <- as.matrix(edgelist)
+attr(edgelist,"n") <- N
+net <- as.edgelist.sna(edgelist)
+pdf("figs/synthetic/network.pdf",width=20,height=20)
+par(mar=rep(0,4))
+set.seed(4)
+gplot(net,pad=0,thresh=10,vertex.cex=2,label.cex=4,vertex.sides=30,label=1:10)#,edge.col=1:5)#t(edge.colors))
+dev.off()
 
 test_that("simulated lrm agrees with brem.lrm",{
   tmp <- brem.lrm(sim$A,N,z,beta)
@@ -79,6 +93,9 @@ test_that("simulated lrm agrees with brem.lrm",{
   diag(sim$lrm[1,,]) <- -Inf
   expect_that(all.equal(tmp[-1,,],sim$lrm[-1,,]),is_true())
 })
+
+
+
 
 
 ## OLD CODE

### 2012-03-12 docs/slides/slides.tex
Work on getting dashboard and predict scripts running again.
Add material to slides.

@@
 \end{itemize}
-
 \end{frame}
@@
 \begin{frame}{Synthetic example}
-Parameter estimates
+True parameters
+
+Observed counts for each block.
+
+Parameter estimates (using Gibbs sampling for $\mathbf{z}$ and slice sampling for $\boldsymbol{\beta}$).
+\end{frame}
+
+\begin{frame}{Application: Dyadic interactions on Twitter}
+  \begin{itemize}
+  \item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 with hashtag \texttt{\#rstats}.
+  \item Selecting tweets beginning with the \texttt{@} symbol, and mark the first mentioned user as the recipient.
+  \item Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
+  \item We consider a subset 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Application: Dyadic interactions on Twitter}
+Prediction experiments and model comparison
+\end{frame}
+
+\begin{frame}{Application: Dyadic interactions on Twitter}
+Plot parameter estimates for winning model
+Compare to fitting a single relational event model
+\end{frame}
+
+\begin{frame}{Discussion}
+Interpretation of parameter estimates
+  \begin{itemize}
+  \item Method clusters with respect to the relational event dynamics for a set of actors
+  \end{itemize}
+
+Main takeaways
+\begin{itemize}
+\item Detailed, highly dependent model for local structure\\ (using relational event models)
+\item Latent variable model to capture meso-level structure \\(using stochastic block model)
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Future directions}
+  \begin{itemize}
+  \item Benefits of mixed-membership ideas
+  \item Dirichlet process instead of fixed $K$ (with split-merge sampling?)
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{References}
 

### 2012-03-12 brem.r

@@ -12,7 +12,7 @@ option_list <- list(
               help="Number of MCMC iterations [default %default]"),
   make_option(c("-m","--model.type"), default="full",
               help="Type of model to fit.  Options: \"baserate\", \"shared\", \"full\"."),
-  make_option(c("-g","--gibbs"), default="fast",
+  make_option(c("-g","--gibbs"), default=TRUE,
               help="Slow or fast version of gibbs."),
   make_option(c("-s","--slice"), default=FALSE,
               help="Slice sample instead of MH."),
@@ -37,7 +37,7 @@ s$precompute()
 
 # Initialize with K=1 solution, if available
 f <- paste("results/",opts$dataset,"/full.1.rdata",sep="")
-if (K > 1 & file.exists(f) & opts$initialize) {
+if ((K > 1) & file.exists(f) & opts$initialize) {
  load(f)
  beta <- array(res$beta,c(P,K,K))
 } else {
@@ -47,7 +47,7 @@ if (K > 1 & file.exists(f) & opts$initialize) {
 px <- rep(1,13)
 px[13] <- 0
 px[7]  <- 0
-if (K==1) px[1] <- 0 # identifiability
+#if (K==1) px[1] <- 0 # identifiability?
 
 if (opts$dataset=="twitter-small" & opts$fixz) {
   tb <- table(factor(c(train[,2],train[,3]),1:N))
@@ -60,9 +60,10 @@ if (opts$dataset=="twitter-small" & opts$fixz) {
   z <- NULL
 }
 
-#outfile <- paste("results/",opts$dataset,"/",opts$model.type,".",K,".",opts$slice,".",opts$fixz,".rdata",sep="")
 outfile <- paste("results/",opts$dataset,"/",opts$model.type,".",K,".rdata",sep="")
+
+priors <- list(beta=list(mu=0,sigma=3))
 
 fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice,
                  niter=opts$numiterations,gibbs=opts$gibbs,beta=beta,px=px,z=z,
-                 outfile=outfile)
+                 outfile=outfile,priors=priors,skip.intercept=TRUE)

### 2012-03-12 dashboard.r

@@ -58,7 +58,7 @@ if (opts$dataset == "synthetic") {
 }
 
 library(coda)
-load(paste("results/",opts$dataset,"/full.3.rdata",sep=""))
+load(paste("results/",opts$dataset,"/full.2.rdata",sep=""))
 r <- melt(res$param)
 r <- subset(r,X1 < 60)
 q1a <- qplot(X1,value,data=r, colour=factor(X2),geom="line") + labs(colour="parameters for\n each block",x="iteration") + theme_bw() + facet_grid(X3~X4,scales="free")
@@ -74,7 +74,7 @@ q2 <- qplot(X1,X2,data=zs,geom="tile",fill=factor(value)) + facet_wrap(~L1) + la
 
 cat("Compute distribution of activity using results from the full model.\n")
 tb <- table(factor(c(train[,2],train[,3]),1:N))
-fx <- grep("full.3",names(fits))
+fx <- grep("full.2",names(fits))
 z <- fits[[fx]]$z
 df <- data.frame(group=z,count=tb)
 q3a <- qplot(log(count.Freq),data=df,geom="histogram")+facet_grid(group~.,scales="free")+labs(y="number of users",x="log(number of events)") + theme_bw()
@@ -151,7 +151,7 @@ betas <- lapply(fits,function(f) f$beta)
 names(betas) <- names(fits)
 b <- melt(betas)
 q6 <- qplot(X1,value,data=b,geom="point",colour=factor(L1)) + facet_grid(X2~X3) + theme_bw() #+ labs(colour="model")
-qplot(X1,value,data=subset(b,L1=="full.3"),geom="point") + facet_grid(X2~X3) + theme_bw() #+ labs(colour="model")
+#qplot(X1,value,data=subset(b,L1=="full.3"),geom="point") + facet_grid(X2~X3) + theme_bw() #+ labs(colour="model")
 
 if (opts$dataset=="twitter-small") {
   lapply(1:3,function(k) nmap[which(z==k)])
@@ -194,15 +194,13 @@ if (opts$predictions) {
   mllks.test$event <- 1:nrow(test)
   
   # Examine log likelihood of observations
-  qplot(event,value,data=llks.train,geom="point",colour=factor(L1))
-  qplot(event,value,data=llks.test,geom="point",colour=factor(L1))
[…]

### 2012-03-12 pkg/R/brem.r

@@ -119,7 +119,7 @@ brem.lpost.fast.block <- function(A,N,K,z,s,beta,k1,k2,priors=list(beta=list(mu=
   }    
 }
 
-brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,m=20,beta=NULL,z=NULL,gibbs="fast",mh=FALSE,outfile=NULL,priors=list(beta=list(mu=0,sigma=1)),verbose=FALSE,px=NULL,skip.intercept=TRUE) {
+brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,m=20,beta=NULL,z=NULL,gibbs=TRUE,mh=FALSE,outfile=NULL,priors=list(beta=list(mu=0,sigma=1)),verbose=FALSE,px=NULL,skip.intercept=TRUE) {
   llks <- rep(0,niter)
   M <- nrow(A)
   P <- 13
@@ -154,21 +154,31 @@ brem.mcmc <- function(A,N,K,s,niter=5,model.type="full",mcmc.sd=.1,m=20,beta=NUL
       olp <- res$olp
     }
     
-    if (gibbs=="slow") {
-      # Gibbs sample assignments
-      for (i in 1:N) {
-        ps <- rep(0,K)
-        for (k in 1:K) {
-          z[i] <- k
-          ps[k] <- brem.lpost.fast(A,N,K,z,s,current,priors)
-        }
-        ps <- exp(ps - max(ps))
-        z[i] <- sample(1:K,size=1,prob=ps)
-      }
-    }
-    if (gibbs=="fast") {
+#     if (gibbs=="slow") {
+#       # Gibbs sample assignments
+#       for (i in 1:N) {
+#         ps <- rep(0,K)
+#         for (k in 1:K) {
+#           z[i] <- k
+#           ps[k] <- brem.lpost.fast(A,N,K,z,s,current,priors)
+#         }
+#         ps <- exp(ps - max(ps))
+#         z[i] <- sample(1:K,size=1,prob=ps)
+#       }
+#     }
+    if (gibbs) {
       cat("gibbs")
-      z <- gibbs(1:N-1,current,z-1,s$ptr(),K)$z + 1
+      z <- RemGibbsPc(1:N-1,current,z-1,s$ptr(),K)$z + 1
+      
+      # Sample empty cluster parameters from prior.  Only sample baserates.
+      #     for (k in 1:K) {
+      #       if (length(which(z==k))==0) {
[…]

### 2012-03-12 predict.r

@@ -18,19 +18,25 @@ load(paste("data/",opts$dataset,".rdata",sep=""))
 
 test.ix <- (1:nrow(A))[-(1:nrow(train))]
 
+if (opts$model %in% c("online","uniform","marg")) {
+  pred <- get.pred.baseline(train,A,test.ix,opts$model)
+} else {
+  load(paste("results/",opts$dataset,"/",opts$model,".rdata",sep=""))
+  pred <- get.pred(train,A,test.ix,res)
+}
 
-llkm.train <- log(multinomial.score(m.train,train))
-llkm.test  <- log(multinomial.score(m.test, test))
+llkm.train <- log(multinomial.score(pred$m$train,train))
+llkm.test  <- log(multinomial.score(pred$m$test, test))
 
 # Compute loglikelihood of each observation
-llk.train <- loglikelihood_vec_from_lrm(lrm.train,train[,1],as.integer(train[,2])-1,as.integer(train[,3])-1,N,nrow(train))
-llk.test <- loglikelihood_vec_from_lrm(lrm.test,test[,1],as.integer(test[,2])-1,as.integer(test[,3])-1,N,nrow(test))
+llk.train <- RemLogLikelihoodVecFromArray(pred$lrm$train,train[,1],as.integer(train[,2])-1,as.integer(train[,3])-1,N,nrow(train))
+llk.test <- RemLogLikelihoodVecFromArray(pred$lrm$test,test[,1],as.integer(test[,2])-1,as.integer(test[,3])-1,N,nrow(test))
 
 # Compute rank of each observation
 cat("ranks (train)\n")
-rk.train <- ranks(train,-lrm.train,ties.method="random")
+rk.train <- ranks(train,-pred$lrm$train,ties.method="random")
 cat("ranks (test)\n")
-rk.test  <- ranks(test, -lrm.test, ties.method="random")
+rk.test  <- ranks(test, -pred$lrm$test, ties.method="random")
 
 # Save
 dir.create(paste("results/",opts$dataset,"/llks/",sep=""))

### 2012-03-12 run.sh

@@ -1,9 +1,11 @@
-opts=list(dataset="synthetic",numclusters=1,model.type="full",niter=10,gibbs="fast",numiterations=100,slice=TRUE)
+opts=list(dataset="synthetic",numclusters=1,model.type="full",niter=10,gibbs="fast",numiterations=100,slice=TRUE,initialize=FALSE,fixz=FALSE)
 
 # Fit models
 ./parallel --sshlogin 4/d6 'cd /extra/duboisc0/blockrem;./brem.r -d {1} -k {2} -n 500 -m {3} -s TRUE' ::: "irvine" ::: 2 4 ::: "shared" "full"
 
 ./parallel --sshlogin 6/d5,6/m 'cd /extra/duboisc0/blockrem;./brem.r -d {1} -k {2} -n 200 -m {3} -s TRUE' ::: "synthetic" "eckmann-small" ::: 1 2 3 ::: "shared" "full"
+
+./parallel --sshlogin 6/d12,6/m 'cd /extra/duboisc0/blockrem;./brem.r -d {1} -k {2} -n 200 -m {3} -s TRUE' ::: "synthetic"  ::: 1 2 :::  "full"
 
 
 ./parallel --sshlogin 3/d7,3/d8 'cd /extra/duboisc0/blockrem;./brem.r -d {1} -k {2} -n 200 -m {3} -s TRUE' ::: "twitter-small" ::: 2 3 ::: "shared" "full"

### 2012-03-13 docs/notes.tex
Add gradient derivation.  Add material/figs to slides.

@@
 
- \centerline{\Large \bf Stochastic Blockmodels for Relational Events} 
+ \centerline{\Large \bf Stochastic Blockmodels for Relational Event Data} 
  \medskip
@@
 \abstract{
+Stochastic blockmodels assume latent groups of individuals have similar tendencies in their  group-wise interactions.
+Many find these methods useful for exploratory analysis of static networks.
+On the other hand, continuous-time methods such as the relational event framework (Butts 2008) model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.
+We propose to combine these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.
+The method is illustrated using dyadic interaction data (e.g.
+email, Twitter direct messages) among hundreds of actors and thousands of events.
+We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than using a single relational event model.
+Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
 
-Stochastic blockmodels are a tool for exploring the structure of static social networks by learning latent groups of individuals who have similar tendencies in their  group-wise interactions.
+%Stochastic blockmodels are a tool for exploring the structure of static social networks by learning latent groups of individuals who have similar tendencies in their  group-wise interactions.
 Continuous-time methods, such as the relational event framework (Butts 2008), model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.
@@
 \center
-\includegraphics[width=2in]{../figs/syn/mat.pdf}
-\includegraphics[width=3in]{../figs/syn/counts.pdf}
+\includegraphics[width=2in]{../figs/synthetic/mat.pdf}
+\includegraphics[width=3in]{../figs/synthetic/counts.pdf}
 \caption{Illustration of 1000 simulated events, as described in text.
@@
 \center
-\includegraphics[width=3in]{../figs/syn/logposterior.pdf}
-\includegraphics[width=3in]{../figs/syn/test-recall.pdf}
+%\includegraphics[width=3in]{../figs/synthetic/logposterior.pdf}
+\includegraphics[width=3in]{../figs/synthetic/test-recall.pdf}
 \caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
@@
 Since the assignment of one node depends on the assignments of the other nodes, we suffer combinatorial explosion.
+
+\section{Gradient}
+
+\begin{align*}
+\frac{\partial \ell}{\partial \beta_{p,k,l}} = \frac{\partial}{\partial \beta_{p,k,l}} \sum_{m = 1}^M \left[ \log \lambda_{i_m,j_m}(t_m | \cdot) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) \lambda_{i,j}(t_m|\cdot) \right] \\
+  =\sum_{m \in V} \left[ s_p(t_m,i_m,j_m) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) s_p(t_m,i,j) \lambda_{i,j}(t_m|\cdot) \right]
+\end{align*}
+
+where $V$ is the set of $m$ where either $z_i \in \{k,l\}$ or $z_j \in \{k,l\}$.
+
 \bibliographystyle{plain}

### 2012-03-13 docs/slides/slides.tex
Add gradient derivation.  Add material/figs to slides.

@@
 \usepackage{tikz}
-\usepackage{subfigure}
+\usepackage{subfig}
+\setlength{\unitlength}{\textwidth} 
 \tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
@@
 \begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | .) \}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | .) \}
 \end{align}
@@
 \begin{itemize}
+\item $M$: number of events
 \item $t_m$: time of event $m$
-\item $\tau_{i,j,m}$: changepoint time $\lambda_{ij}(t)$ prior to $t_m$
-\item $\mathcal{R}_{i,j}$: set of dyads involving either $i$ or $j$
+\item $\mathcal{R}$: set of dyadic events
 \end{itemize}
@@
 
-\includegraphics[scale=.5]{../../figs/synthetic/unsorted}
-\includegraphics[scale=.5]{../../figs/synthetic/mat}
-\includegraphics[scale=.5]{../../figs/synthetic/bm}
+\includegraphics[scale=.45]{../../figs/synthetic/unsorted}
+\uncover<2->{\includegraphics[scale=.45]{../../figs/synthetic/mat}}
+\uncover<3->{\includegraphics[scale=.45]{../../figs/synthetic/bm} }
 
@@
 \begin{align*}
+\uncover<4->{
 \mbox{Mixing rates} & & \boldsymbol{\eta} =& \left[
@@
 \right] \\
+}
+\uncover<5->{
 \mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
+}
+\uncover<6->{
 \mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi)
+}
 \end{align*}
@@
 
-\begin{frame}{Blockmodels for relational event data}
+\begin{frame}{Stochastic blockmodels for relational event data}
 Idea:
@@
 
-Notation:
+\uncover<2>{Notation:
 \begin{itemize}
@@
 \end{itemize}
+}
+
+\uncover<3>{
+\vspace{-3cm}
+Benefits:  
+\begin{itemize}
+\item Adjust for unobserved heterogeneity.
+\item Obtain clusters of nodes who have similar interaction dynamic with other parts of the network.
+\end{itemize}
+
+}
 
@@
 \begin{figure}
-
 \centering 
-\subfigure[Dynamic network data]
-{
+%\subfigure[Dynamic network data]
+%{
 \begin{tikzpicture}[scale=.8,transform shape,thick]
@@
 \end{tikzpicture}
-}
-\subfigure[Intensities for two dyads]
-{
+% }
+% \subfigure[Intensities for two dyads]
+% {
 \begin{tikzpicture}[scale=1,transform shape,thick]
@@
 \end{tikzpicture}
-}
+%}
 \end{figure}
@@
 
-\begin{frame}{Synthetic example}
-True parameters
+\begin{frame}{Synthetic example: $\beta$ and observed counts}
+\begin{picture}(0.0,0.0)
+   \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-true}}
+   \put(.6,-0.3){\includegraphics[width=0.5\textwidth]{../../figs/synthetic/counts}}
+\end{picture}
+\end{frame}
 
-Observed counts for each block.
+\begin{frame}{Synthetic example: slice sampling for $\hat{\beta}$ and Gibbs on
+    $\mathbf{z}$}
+\begin{picture}(0.0,0.0)
+   \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-estimates}}
+   \put(.6,-0.3){\includegraphics[width=0.5\textwidth]{../../figs/synthetic/counts}}
+\end{picture}
 
-Parameter estimates (using Gibbs sampling for $\mathbf{z}$ and slice sampling for $\boldsymbol{\beta}$).
+%\vspace{1in}
+%From 100 samples after 50 burnin using Gibbs sampling for $\mathbf{z}$ and slice sampling for $\boldsymbol{\beta}$).
 \end{frame}

### 2012-03-13 brem.r

@@ -62,8 +62,11 @@ if (opts$dataset=="twitter-small" & opts$fixz) {
 
 outfile <- paste("results/",opts$dataset,"/",opts$model.type,".",K,".rdata",sep="")
 
-priors <- list(beta=list(mu=0,sigma=3))
+priors <- list(beta=list(mu=0,sigma=1))
 
+M <- nrow(train)
+s <- new(RemStat,train[,1],as.integer(train[,2])-1,as.integer(train[,3])-1,N,M,P)
+s$precompute()
 fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice,
                  niter=opts$numiterations,gibbs=opts$gibbs,beta=beta,px=px,z=z,
-                 outfile=outfile,priors=priors,skip.intercept=TRUE)
+                 outfile=outfile,priors=priors,skip.intercept=FALSE)

### 2012-03-13 dashboard.r

@@ -87,7 +87,10 @@ if (length(fx) > 0) {
   df <- melt(df)
   df$i <- z[df$X1]
   df$j <- z[df$X2]
-  q3 <- qplot(X3,value,data=df,geom="boxplot",outlier.size=0.1) + facet_grid(i ~ j) + theme_bw() + labs(x="shift type",y="count for a given dyad") + opts(axis.text.x=theme_text(angle=90))
+  levels(df$X3) <- rev(list("AB-BA"="ab-ba","AB-BY"="ab-by","AB-XA"="ab-xa","AB-AY"="ab-ay","AB-XB"="ab-xb","AB-AB"="ab-ab"))
+  q3 <- qplot(X3,value,data=df,geom="boxplot",outlier.size=0.1) + facet_grid(i ~ j) + theme_bw() + labs(x="",y="count for a given dyad") + opts(axis.text.x=theme_text(angle=90)) + coord_flip()
+  q3
+  ggsave("figs/synthetic/counts.pdf",width=4,height=5)
 } else {
   q3 <- qplot(0,0,label="not available",geom="text")
 }
@@ -236,6 +239,37 @@ blankPanel <- grid.rect(gp=gpar(col="white"))
 grid.arrange(q1, q1a, q2, q3, q3a, q4, q5, q6,ncol=4)
 dev.off()
 
+if (opts$dataset=="synthetic") {
+  load("data/synthetic.rdata")
+  load("results/synthetic/full.2.rdata")
+  b <- melt(res$param)
+  colnames(b) <- c ("iter","p","k1","k2","value")
+  pnames <- c("Intercept","ab-ba","ab-by","ab-xa","ab-xb","ab-ay","ab-ab","sen. outdeg","rec. outdegree","sen. indegree","rec. indegree","dyad count","total")
+  dimnames(beta)[[1]] <- pnames
+  tb <- melt(beta)
+  colnames(tb) <- c("p","k1","k2","value")
+ # tb <- subset(tb,!(p %in% c("ab-ab","total")))
+  d <- ddply(b,.(p,k1,k2),function(x) c(mean=mean(x$value),quantile(x$value,c(.025,.2,.8,.975))))
+  
+  # Fix group assignments
+  d$k1 <- c(2,1)[d$k1]
+  d$k2 <- c(2,1)[d$k2]
+  
+  # Fix parameter names
+  d$p <- pnames[d$p]
+  d <- subset(d,! p %in% c("ab-ab","total"))
[…]

### 2012-03-13 process/synthetic.r

@@ -77,6 +77,7 @@ plotmat(mat,limits=c(-10,3))
 ggsave("figs/synthetic/bm.pdf",width=3,height=3)
 
 library(sna)
+library(RColorBrewer)
 edgelist <- melt(table(sim$A[,2],sim$A[,3]))
 colnames(edgelist) <- c("sen","rec","val")
 mypal <- brewer.pal(9,"Greys")
@@ -87,8 +88,8 @@ attr(edgelist,"n") <- N
 net <- as.edgelist.sna(edgelist)
 pdf("figs/synthetic/network.pdf",width=20,height=20)
 par(mar=rep(0,4))
-set.seed(4)
-gplot(net,pad=0,thresh=10,vertex.cex=2,label.cex=4,vertex.sides=30,label=1:10)#,edge.col=1:5)#t(edge.colors))
+set.seed(5)
+gplot(net,pad=0,thresh=30,vertex.cex=2,label.cex=4,vertex.sides=30,label=1:10)#,edge.col=1:5)#t(edge.colors))
 dev.off()
 
 test_that("simulated lrm agrees with brem.lrm",{

### 2012-03-14 docs/slides/slides.tex
Add material to slides

@@
 \usepackage{tikz}
-\usepackage{subfig}
+
 \setlength{\unitlength}{\textwidth} 
 \tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-\title[Blockmodels for relational events]%(optional, only for long titles)
-{Stochastic blockmodels for relational events}
-\subtitle{}
+\title{Stochastic blockmodels for relational events}
+% \subtitle{}
 \author[DuBois, Christopher] % (optional, for multiple authors)
@@
 {Sunbelt 2012}
-\subject{Informatik}
 \begin{document}
@@
 \item Above example: 2 dyadic events occurring among 4 nodes
-\item Nodes may have covariates
+\item Nodes and edges may have covariates
 \item Interested in inferences about the rate of particular events, conditioned on what we have observed
@@
 
-\begin{frame}{Relational event models}
+\begin{frame}{Relational event models: Modeling the \emph{hazard}}
+Multiplicative models for the hazard of a dyad $(i,j)$:
+$$\log \lambda_{ij}(t | \cdot) = \boldsymbol{\beta}' \mathbf{s}(t,i,j)$$
+\begin{itemize}
+\item $\lambda_{ij}(t | \cdot)$: ``rate'' of interaction from $i$ to
+  $j$ at time $t$
+\item $\boldsymbol{\beta}$: vector of model parameters
+\item $\mathbf{s}(t,i,j)$: vector of sufficient statistics for $(i,j)$
+  given the events before time $t$
+\end{itemize}
+\end{frame}
 
-Multiplicative models for the hazard:
-$$\log \lambda_{ij}(t | \cdot) = \boldsymbol{\beta}' \mathbf{s}(t,i,j)$$
-
-Likelihood:
+\begin{frame}{Relational event models: Likelihood}
 
@@
 
-\includegraphics[scale=.45]{../../figs/synthetic/unsorted}
-\uncover<2->{\includegraphics[scale=.45]{../../figs/synthetic/mat}}
-\uncover<3->{\includegraphics[scale=.45]{../../figs/synthetic/bm} }
+\uncover<2->{\includegraphics[scale=.45]{../../figs/synthetic/unsorted}}
+\uncover<3->{\includegraphics[scale=.45]{../../figs/synthetic/mat}}
+\uncover<4->{\includegraphics[scale=.45]{../../figs/synthetic/bm} }
 
@@
 \begin{align*}
-\uncover<4->{
+\uncover<5->{
 \mbox{Mixing rates} & & \boldsymbol{\eta} =& \left[
@@
 }
-\uncover<5->{
+\uncover<6->{
 \mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
 }
-\uncover<6->{
+\uncover<7->{
 \mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi)
@@
 
+\begin{frame}{Model adequacy}
+Plot parameter estimates for winning model
+Compare to fitting a single relational event model
+\end{frame}
+
 \begin{frame}{Discussion}

### 2012-03-14 docs/slides/slides.tex
Synthetic prediction experiment good (again).  Whole pipeline working.

@@
 \uncover<6->{
-\mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
+\mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi) \\
 }
 \uncover<7->{
-\mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi)
+\mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
 }
@@
 
+\end{frame}
+
+\begin{frame}{Stochastic blockmodels for relational event data}
+Subsets of interactions may have distinct dynamics
+\begin{itemize}
+  \item Core-periphery structures \\(e.g. periphery only sends, core reciprocates immediately)
+  \item<2-> Cohesive subgroups with differing behavior \\(e.g. students and professors)
+  \end{itemize}
 \end{frame}
@@
 \end{itemize}
-
 }
@@
 
-\begin{frame}{Model specification for $X_{ij}(t)$}
+\begin{frame}{Model specification for $\mathbf{s}(t,i,j)$}
 \textbf{Participation shifts} adjust for the next event
@@
    \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-true}}
-   \put(.6,-0.3){\includegraphics[width=0.5\textwidth]{../../figs/synthetic/counts}}
+   \put(.55,-0.25){\includegraphics[width=0.55\textwidth]{../../figs/synthetic/counts}}
 \end{picture}
@@
 
-\begin{frame}{Synthetic example: slice sampling for $\hat{\beta}$ and Gibbs on
+\begin{frame}{Synthetic example: slice sampling for $\beta$ and Gibbs on
     $\mathbf{z}$}
@@
    \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-estimates}}
-   \put(.6,-0.3){\includegraphics[width=0.5\textwidth]{../../figs/synthetic/counts}}
+   \put(.65,-0.25){\includegraphics[width=0.55\textwidth]{../../figs/synthetic/counts}}
 \end{picture}

### 2012-03-14 dashboard.r

@@ -97,18 +97,22 @@ z <- fits[[fx]]$z
 df <- data.frame(group=z,count=tb)
 q3a <- qplot(log(count.Freq),data=df,geom="histogram")+facet_grid(group~.,scales="free")+labs(y="number of users",x="log(number of events)") + theme_bw()
 
-cat("Compute dyad counts for each pshift using results from the full model.\n")
+cat("Compute dyad counts for each pshift using results from the true model.\n")
 fx <- grep("full.2",fnames)
 if (length(fx) > 0) {
-  z <- fits[[fx]]$z
+  if (opts$dataset!="synthetic") z <- fits[[fx]]$z
+  else z <- c(1,1,1,1,1,2,2,2,2,2)
+  
   df <- dyad.ps(train,N)
   df <- melt(df)
   df$i <- z[df$X1]
   df$j <- z[df$X2]
-  levels(df$X3) <- rev(list("AB-BA"="ab-ba","AB-BY"="ab-by","AB-XA"="ab-xa","AB-AY"="ab-ay","AB-XB"="ab-xb","AB-AB"="ab-ab"))
-  q3 <- qplot(X3,value,data=df,geom="boxplot",outlier.size=0.1) + facet_grid(i ~ j) + theme_bw() + labs(x="",y="count for a given dyad") + opts(axis.text.x=theme_text(angle=90)) + coord_flip()
+  df <- subset(df,X3 != "AB-AB")
+  df$X3 <- factor(as.character(df$X3))
+  levels(df$X3) <- rev(list("AB-BA"="ab-ba","AB-BY"="ab-by","AB-XA"="ab-xa","AB-XB"="ab-xb","AB-AY"="ab-ay"))#,"AB-AB"="ab-ab"))
+  q3 <- qplot(X3,value,data=df,geom="boxplot",outlier.size=0.1) + facet_grid(i ~ j) + theme_bw() + labs(x="",y="count for a given dyad") + coord_flip()# +opts(axis.text.x=theme_text(angle=90)) 
   q3
-  ggsave("figs/synthetic/counts.pdf",width=4,height=5)
+  ggsave("figs/synthetic/counts.pdf",width=4,height=4)
 } else {
   q3 <- qplot(0,0,label="not available",geom="text")
 }
@@ -214,14 +218,14 @@ if (opts$predictions) {
   mllks.test <- melt(mllks.test)
   mllks.test$event <- 1:nrow(test)
   
[…]

### 2012-03-14 predict.r

@@ -23,7 +23,7 @@ if (opts$model %in% c("online","uniform","marg")) {
   pred <- get.pred.baseline(train,A,test.ix,opts$model)
 } else {
   if (opts$dataset == "synthetic" & opts$model == "truth") {
-    res <- list(beta=beta,z=z,llks=rep(200,true.lpost),param=beta)  # true values
+    res <- list(beta=beta,z=z,llks=rep(true.lpost,500),niter=500,zs=rep(list(z),500),param=beta)  # true values
     save(res,file=paste("results/",opts$dataset,"/",opts$model,".rdata",sep=""))
   }
   load(paste("results/",opts$dataset,"/",opts$model,".rdata",sep=""))
@@ -37,16 +37,21 @@ llkm.test  <- log(multinomial.score(pred$m$test, test))
 llk.train <- RemLogLikelihoodVecFromArray(pred$lrm$train,train[,1],as.integer(train[,2])-1,as.integer(train[,3])-1,N,nrow(train))
 llk.test <- RemLogLikelihoodVecFromArray(pred$lrm$test,test[,1],as.integer(test[,2])-1,as.integer(test[,3])-1,N,nrow(test))
 
+cat("\nllkm.train:",mean(llkm.train))
+cat("\nllkm.test:",mean(llkm.test))
+cat("\nllk.train:",mean(llk.train))
+cat("\nllk.test:",mean(llk.test))
+
 # Compute rank of each observation
-cat("ranks (train)\n")
+cat("\nranks (train)\n")
 rk.train <- ranks(train,-pred$lrm$train,ties.method="random")
 cat("ranks (test)\n")
 rk.test  <- ranks(test, -pred$lrm$test, ties.method="random")
 
 # Save
-dir.create(paste("results/",opts$dataset,"/llks/",sep=""))
+dir.create(paste("results/",opts$dataset,"/llks/",sep=""),showWarnings=FALSE)
 save(llkm.train,llkm.test,llk.train,llk.test,opts,file=paste("results/",opts$dataset,"/llks/",opts$model,".rdata",sep=""))
-dir.create(paste("results/",opts$dataset,"/ranks/",sep=""))
+dir.create(paste("results/",opts$dataset,"/ranks/",sep=""),showWarnings=FALSE)
 save(rk.train,rk.test,opts,file=paste("results/",opts$dataset,"/ranks/",opts$model,".rdata",sep=""))
 
 

### 2012-03-14 docs/slides/slides.tex
Sunbelt talk submitted to Carter.

@@
 {Sunbelt 2012}
+
+\renewcommand\footnotemark{}
+\renewcommand*\footnoterule{}
+
 \begin{document}
@@
 
-\begin{frame}{Relational event models: Modeling the \emph{hazard}}
+\begin{frame}{Relational event models}
 Multiplicative models for the hazard of a dyad $(i,j)$:
@@
 \end{itemize}
+\vspace{2cm}
+\footnotesize{See \cite{Butts2008, Vu2011,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011}.}
 \end{frame}
@@
 \end{align*}
-
+\uncover<8->{
+\footnotesize{See \cite{Nowicki2001, Kemp}.}
+}
 \end{frame}
@@
 \item received by receiver
-\item previous occurrences
+\item previous occurrences of this dyad
 \end{itemize}
@@
 
-\begin{frame}{Specifying degree effects (and explosion)}
+% \begin{frame}{Specifying degree effects (and explosion)}
 
-\end{frame}
+% \end{frame}
 
@@
    \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-estimates}}
-   \put(.65,-0.25){\includegraphics[width=0.55\textwidth]{../../figs/synthetic/counts}}
+   \put(.55,-0.25){\includegraphics[width=0.55\textwidth]{../../figs/synthetic/counts}}
 \end{picture}
@@
 
-\begin{frame}{Application: Dyadic interactions on Twitter}
-  \begin{itemize}
-  \item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 with hashtag \texttt{\#rstats}.
-  \item Selecting tweets beginning with the \texttt{@} symbol, and mark the first mentioned user as the recipient.
-  \item Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
-  \item We consider a subset 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-  \end{itemize}
+\begin{frame}{Prediction experiment: Synthetic data}
+Mean log likelihood on observed events and held out events.
+\input{../../figs/synthetic/results.tex}
 \end{frame}
 
-\begin{frame}{Application: Dyadic interactions on Twitter}
-Prediction experiments and model comparison
+\begin{frame}{Application: Modeling email (88 people, 2000 events)}
+
+\begin{picture}(0.0,0.0)
+   \put(-.05,-0.25){\includegraphics[scale=.4]{../../figs/eckmann-small/countsa}}
+ \uncover<2->{  \put(.5,-0.25){\includegraphics[scale=.5]{../../figs/eckmann-small/params-estimates}}}
+\end{picture}
 \end{frame}
 
-\begin{frame}{Application: Dyadic interactions on Twitter}
-Plot parameter estimates for winning model
-Compare to fitting a single relational event model
+\begin{frame}{Prediction experiment: Eckmann subset}
+Mean log likelihood on observed events and held out events.
+\input{../../figs/eckmann-small/results.tex}
 \end{frame}
 
-\begin{frame}{Model adequacy}
-Plot parameter estimates for winning model
-Compare to fitting a single relational event model
-\end{frame}
+% \begin{frame}{Application: Dyadic interactions on Twitter}
+%   \begin{itemize}
+%   \item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 with hashtag \texttt{\#rstats}.
+%   \item Selecting tweets beginning with the \texttt{@} symbol, and mark the first mentioned user as the recipient.
+%   \item Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
+%   \item We consider a subset 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
+%   \end{itemize}
+% \end{frame}
+
+% \begin{frame}{Application: Dyadic interactions on Twitter}
+% Prediction experiments and model comparison
+% \end{frame}
+
+% \begin{frame}{Application: Dyadic interactions on Twitter}
+% Plot parameter estimates for winning model
+% Compare to fitting a single relational event model
+% \end{frame}
+
+% \begin{frame}{Model adequacy}
+% Plot parameter estimates for winning model
+% Compare to fitting a single relational event model
+% \end{frame}
 
@@
   \begin{itemize}
-  \item Method clusters with respect to the relational event dynamics for a set of actors
+  \item The proposed method clusters individuals with respect to
+    shared relational event dynamics with other clusters
   \end{itemize}
 
+\uncover<2->{
 Main takeaways
@@
 \end{itemize}
+}
+
+\uncover<3->{
+Future directions
+  \begin{itemize}
+  \item Consider a mixed-membership approach
+  \item Dirichlet process instead of fixed $K$
+  \end{itemize}
+}
 \end{frame}
 
-\begin{frame}{Future directions}
-  \begin{itemize}
-  \item Benefits of mixed-membership ideas
-  \item Dirichlet process instead of fixed $K$ (with split-merge sampling?)
-  \end{itemize}
+\begin{frame}[allowframebreaks]{References}
+\tiny
+\bibliographystyle{apalike}
+\bibliography{../refs}
 \end{frame}
 
-\begin{frame}{References}
+\begin{frame}{Thanks}
+Work supported by National Defense Science and Engineering Graduate
+Fellowship.
+
+\vspace{2cm}
+Special thanks to Carter Butts and Padhraic Smyth.
 

### 2012-03-14 brem.r

@@ -36,6 +36,8 @@ P <- 13
 K <- opts$numclusters
 
 # Initialize with K=1 solution, if available
+if (K==1) opts$gibbs <- FALSE
+
 f <- paste("results/",opts$dataset,"/full.1.rdata",sep="")
 if ((K > 1) & file.exists(f) & opts$initialize) {
  load(f)
@@ -65,8 +67,8 @@ if (opts$dataset=="twitter-small" & opts$fixz) {
 
 outfile <- paste("results/",opts$dataset,"/",opts$model.type,".",K,".rdata",sep="")
 
-priors <- list(beta=list(mu=0,sigma=3))
+priors <- list(beta=list(mu=0,sigma=1))
 
 fit <- brem.mcmc(train,N,K,model.type=opts$model.type,slice=opts$slice,gibbs=opts$gibbs,mh=opts$mh,
-                 niter=opts$numiterations,beta=beta,px=px,z=z,
+                 niter=opts$numiterations,beta=beta,px=px,z=z,m=10,
                  outfile=outfile,priors=priors,skip.intercept=FALSE)

### 2012-03-14 dashboard.r

@@ -39,7 +39,7 @@ fnames <- unlist(strsplit(fs,".rdata"))
 fitnames <- sapply(fnames,function(f) strsplit(f,paste(results.dir,"/",sep=""))[[1]][2])
 fits <- lapply(fs,function(f) {
   load(f)
-  return(res)
+  return(fit)
 })
 names(fits) <- fitnames
 
@@ -59,7 +59,7 @@ if (opts$dataset == "synthetic") {
 
 library(coda)
 load(paste("results/",opts$dataset,"/full.2.rdata",sep=""))
-r <- melt(res$param)
+r <- melt(fit$param)
 r <- subset(r,X1 < 60)
 q1a <- qplot(X1,value,data=r, colour=factor(X2),geom="line") + labs(colour="parameters for\n each block",x="iteration") + theme_bw() + facet_grid(X3~X4,scales="free")
 
@@ -68,7 +68,7 @@ cat("Plot overall rate\n")
 if (FALSE) {
   s <- new(RemStat,A[,1],as.integer(A[,2])-1,as.integer(A[,3])-1,N,nrow(A),P)
   s$precompute()
-  lrm <- LogIntensityArrayPc(res$beta,res$z-1,s$ptr(),K)
+  lrm <- LogIntensityArrayPc(fit$beta,fit$z-1,s$ptr(),K)
   for (m in 1:nrow(A)) diag(lrm[m,,]) <- -Inf
   total.lambda  <- sapply(1:nrow(A),function(i) sum(exp(lrm[i,,])))
   pdf("figs/online-debug-time1.pdf",width=7,height=7)
@@ -246,12 +246,29 @@ blankPanel <- grid.rect(gp=gpar(col="white"))
 grid.arrange(q1, q1a, q2, q3, q3a, q4, q5, q6,ncol=4)
 dev.off()
 
+if (opts$dataset=="eckmann-small") {
+  load("data/eckmann-small.rdata")
+  load("results/eckmann-small-bk2/full.2.rdata")
+  b <- melt(fit$param)
+  colnames(b) <- c ("iter","p","k1","k2","value")
+  b <- subset(b, iter > 50 & value != 0)
+  pnames <- c("Intercept","ab-ba","ab-by","ab-xa","ab-xb","ab-ay","ab-ab","sen. outdegree","rec. outdegree","sen. indegree","rec. indegree","dyad count","total")
+  b$p <- pnames[b$p]
[…]

### 2012-03-14 pkg/R/brem.r

@@ -1,43 +1,62 @@
-#' Simulate data from the block relational event model
+#' Simulate data from the block relational event model.
 #' @M number of events
 #' @N number of actors
 #' @z latent class assignments
 #' @beta P x K x K array of parameters
-simulate.brem <- function(M,N,z,beta) {
+#' 
+generate.brem <- function(M,N,z,beta,nsim) {
+  param <- do.call(rbind)
+  fit <- list(param=)
+}
+
+simulate.brem <- function(fit,nsim=1,return.lrm = FALSE) {
+  z <- fit$z
+  beta <- fit$beta
+  M <- fit$M
+  N <- fit$N
+  K <- fit$K
+  P <- fit$P
   
-  # start with a event from 1 to 2
-  time <- 0
-  A <- matrix(c(time,1,2),1,3)
-  lambda <- beta[1,z,z]
-  lrm <- array(0,c(M,N,N))
-  s <- InitializeStatisticsArray(N,P);
-  z <- z-1
-  for (m in 2:M) {
+  sims <- mclapply(1:nsim,function(sim) { 
+    cat(".")
+    beta <- fit$param[fit$niter - sim,,,]
+    beta <- array(beta,c(P,K,K))
+    # start with a event from 1 to 2
+    time <- 0
+    A <- matrix(c(time,1,2),1,3)
+    lambda <- beta[1,z,z]
+    if (return.lrm) lrm <- array(0,c(M,N,N))
+    else lrm <- NULL
     
-    i <- A[m-1,2]
-    j <- A[m-1,3]
-    
-    s <- UpdateStatisticsArray(s,m-1,i-1,j-1,N,P)
-    # Compute changes to lambda
-    for (r in 1:N) {
-      lambda[r,j] <- LogLambda(r-1,j-1,z[r],z[j],s,beta,N,K,P)
-      lambda[j,r] <- LogLambda(j-1,r-1,z[j],z[r],s,beta,N,K,P)
-      lambda[i,r] <- LogLambda(i-1,r-1,z[i],z[r],s,beta,N,K,P)
-      lambda[r,i] <- LogLambda(r-1,i-1,z[r],z[i],s,beta,N,K,P)
+    s <- InitializeStatisticsArray(N,P);
+    for (m in 2:M) {
+      
+      i <- A[m-1,2]
+      j <- A[m-1,3]
+      
+      s <- UpdateStatisticsArray(s,m-1,i-1,j-1,N,P)
[…]

### 2012-03-14 post.pred.r

@@ -1,0 +1,68 @@
+library(brem)
+source("pkg/R/brem.r")
+require(ggplot2)
+post.pred.times <- function(sims) {
+  d <- melt(sapply(sims,function(s) s$edgelist[,1]))
+  colnames(d) <- c("event","sim","value")
+  return(d)
+}
+post.pred.pshift <- function(sims) {
+  require(relevent)
+  d <- mclapply(1:length(sims), function(i) {
+    M <- nrow(sims[[i]]$edgelist)
+    pshifts <- accum.ps(sims[[i]]$edgelist)[M,c(1,3,8:10,13)]
+    data.frame(sim=i,stat=names(pshifts),value=pshifts)
+  })
+  d <- do.call(rbind,d)
+  rownames(d) <- c()
+  return(d)
+}
+post.pred.degree <- function(sims) {
+  require(network)
+  require(sna)
+  degrees <- mclapply(1:length(sims), function(i) {
+    net <- network(sims[[i]]$edgelist[,2:3])
+    N.i <- sims[[i]]$N
+    rbind(
+      data.frame(sim=i,stat="indegree",actor=1:N.i,obs=degree(net,cmode="indegree")),
+      data.frame(sim=i,stat="outdegree",actor=1:N.i,obs=degree(net,cmode="outdegree")),
+      data.frame(sim=i,stat="degree",actor=1:N.i,obs=degree(net)))
+  })
+  do.call(rbind,degrees)
+}
+post.pred.stat <- function(sims,stat) {
+  switch(stat,
+         "global"=post.pred.times(sims),
+         "pshift"=post.pred.pshift(sims),
+         "degree"=post.pred.degree(sims))
+}
+
+# Eckmann
+load("data/eckmann-small.rdata")
+load("results/eckmann-small-bk2/full.2.rdata")
+#load("results/eckmann-small/full.2.rdata")
+A <- as.matrix(A)
+dimnames(A) <- NULL
+library(multicore)
+sims <- simulate(fit,nsim=50)
+obs <- list(list(edgelist=A[1:fit$M,],N=N))
+
+stats <- list()
+stats$sim <- stats$obs <- list()
+
+# Plot overall event vs. time
+stats$sim$time <- post.pred.stat(sims,"global")
+stats$obs$time <- post.pred.stat(obs,"global")
+stats$sim$pshift <- post.pred.stat(sims,"pshift")
[…]

### 2012-03-14 predict.r

@@ -23,11 +23,11 @@ if (opts$model %in% c("online","uniform","marg")) {
   pred <- get.pred.baseline(train,A,test.ix,opts$model)
 } else {
   if (opts$dataset == "synthetic" & opts$model == "truth") {
-    res <- list(beta=beta,z=z,llks=rep(true.lpost,500),niter=500,zs=rep(list(z),500),param=beta)  # true values
-    save(res,file=paste("results/",opts$dataset,"/",opts$model,".rdata",sep=""))
+    fit <- list(beta=beta,z=z,llks=rep(true.lpost,500),niter=500,zs=rep(list(z),500),param=beta)  # true values
+    save(fit,file=paste("results/",opts$dataset,"/",opts$model,".rdata",sep=""))
   }
   load(paste("results/",opts$dataset,"/",opts$model,".rdata",sep=""))
-  pred <- get.pred(train,A,test.ix,res)
+  pred <- get.pred(train,A,test.ix,fit)
 }
 
 llkm.train <- log(multinomial.score(pred$m$train,train))

### 2012-03-14 run.sh

@@ -5,6 +5,8 @@
 
 ./parallel --sshlogin 6/d5,6/m 'cd /extra/duboisc0/blockrem;./brem.r -d {1} -k {2} -n 200 -t {3} -s TRUE' ::: "eckmann-small" ::: 1 2 3 ::: "full"
 
+./parallel --sshlogin 6/d5,6/m 'cd /extra/duboisc0/blockrem;./brem.r -d {1} -k {2} -n 200 -t {3} -s TRUE --initialize TRUE' ::: "eckmann-small" ::: 2 3 ::: "full"
+
 ./parallel --sshlogin 6/d12,6/m 'cd /extra/duboisc0/blockrem;./brem.r -d {1} -k {2} -n 500 -t {3} -s TRUE' ::: "synthetic"  ::: 1 2 :::  "full"
 
 
@@ -22,7 +24,7 @@
 
 ./parallel --sshlogin  7/m 'cd /extra/duboisc0/blockrem;./predict.r -d {1} -t {2}' ::: "twitter-small" ::: "full.1" "full.2" "online" "full.3" "shared.2" "shared.3" "uniform" "online" "marg"
 
-./parallel --sshlogin 2/d4,2/d5,2/d6,2/d7 'cd /extra/duboisc0/blockrem;./predict.r -d {1} -t {2}' ::: "synthetic" ::: "full.2" "full.1" "uniform" "online" "marg"
+./parallel --sshlogin 2/d4,2/d5,2/d6,2/d7 'cd /extra/duboisc0/blockrem;./predict.r -d {1} -t {2}' ::: "synthetic" ::: "full.2" "full.1" "uniform" "online" "marg" "truth"
 
 ./parallel --sshlogin 2/d4,2/d5,2/d6,2/d7 'cd /extra/duboisc0/blockrem;./predict.r -d {1} -t {2}' ::: "eckmann-small" ::: "full.3" "full.2" "full.1" "uniform" "online" "marg"
 
@@ -33,9 +35,7 @@
 
 # Run dashboard
 rsync -auvz dashboard.r duboisc@d1:/extra/duboisc0/blockrem/
-./parallel --sshlogin 8/d10 'cd /extra/duboisc0/blockrem;./dashboard.r -d {} -s TRUE' ::: "synthetic" 
-./parallel --sshlogin 8/d10 'cd /extra/duboisc0/blockrem;./dashboard.r -d {} -s TRUE' ::: "twitter-small" 
[…]

### 2012-03-15 docs/slides/slides.tex
Update slides with posterior predictive checks material.

@@
   \end{itemize}
-
+\uncover<2->{
 \begin{align}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z}).
+\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t).
 \end{align}
-
-\uncover<2>{Notation:
+}
+\uncover<3>{Notation:
 \begin{itemize}
@@
 
-\uncover<3>{
+\uncover<4>{
 \vspace{-3cm}
@@
 \begin{itemize}
-\item AB-BA
-\item AB-BY
-\item AB-AY
-\item AB-XA
-\item AB-XB
+\item AB-BA: Reciprocity
+\item AB-BY:  Turn-taking 
+\item AB-AY: Turn-continuing
+\item AB-XA: Turn-usurping
+\item AB-XB: Turn-usurping
 \end{itemize}
@@
 \item received by receiver
-\item previous occurrences of this dyad
+\item involving this dyad
 \end{itemize}
@@
 
+\uncover<2->{
 \begin{figure}
@@
 \end{figure}
-
+}
+\uncover<3->{
 \begin{itemize}
@@
 \end{itemize}
+}
 \end{frame}
@@
 
+\begin{frame}{Posterior predictive checks}
+  \begin{itemize}
+  \item Investigate whether observed data is reasonable under model.
+  \item Interested in a particular statistic of a sequence $T(Y)$ \\
+    (e.g. total time, degree distribution, etc.)
+  \item Simulate $Y^{(i)} \sim \mbox{REM}(\beta^{(i)},\mathcal{A})$
+  \item Compare $T(Y)$ to distribution of $T(Y^{(i)})$
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Posterior predictive checks}
+\includegraphics[scale=.4]{../../figs/synthetic/postpred-time}
+\includegraphics[scale=.4]{../../figs/synthetic/postpred-pshift}
+\end{frame}
+
 \begin{frame}{Prediction experiment: Synthetic data}
-Mean log likelihood on observed events and held out events.
+Mean log likelihood for observed events and held out events.
 \input{../../figs/synthetic/results.tex}
@@
 \begin{frame}{Prediction experiment: Eckmann subset}
-Mean log likelihood on observed events and held out events.
+Mean log likelihood for observed events and held out events.
 \input{../../figs/eckmann-small/results.tex}
+\end{frame}
+
+\begin{frame}{Posterior predictive checks for $K=2$ model}
+\includegraphics[scale=.4]{../../figs/eckmann-small/postpred-time}
+\includegraphics[scale=.4]{../../figs/eckmann-small/postpred-pshift}
 \end{frame}
@@
   \item Consider a mixed-membership approach
-  \item Dirichlet process instead of fixed $K$
+  \item Intelligent way to share parameters as $K$ grows
   \end{itemize}

### 2012-03-20 pkg/inst/include/utils.h
Outline REM DP c++ and add tests.

@@ -1,0 +1,16 @@
+#include <iostream>
+#include <algorithm>
+#include <vector>
+#include <Rcpp.h>
+
+/*
+ Utilities
+*/
+
+// Return the (j,k,l) element of a (J,K,L) array represented as a vector.
+int threeDIndex(int j, int k, int l, int J, int K, int L);
+
+bool IsFiniteNumber(double x);
+
+// For a vector of log probabilities, return a draw from the categorical distribution
+int rcategorical (Rcpp::NumericVector lp);

### 2012-03-20 pkg/inst/tests/test.gradient.r

@@ -1,176 +1,180 @@
-context("gradient")
-library(brem)
-set.seed(2)
-M <- 2000
-N <- 500
-times <- sort(runif(M,0,1))
-sen <- sample(1:N,M,replace=TRUE)
-rec <- sample(1:N,M,replace=TRUE)
-ix <- which(sen==rec)
-times <- times[-ix]
-sen <- sen[-ix]
-rec <- rec[-ix]
-K <- 1
-beta <- list("intercept"=matrix(-1,K,K),
-             "abba" = matrix(1,K,K),
-             "abby"=matrix(0,K,K),
-             "abxa"=matrix(0,K,K),
-             "abxb"=matrix(0,K,K),
-             "abay"=matrix(1,K,K),
-             "abab"=matrix(0,K,K),
-             "sod"=matrix(0,K,K),
-             "rod"=matrix(0,K,K),
-             "sid"=matrix(0,K,K),
-             "rid"=matrix(0,K,K),
-             "dc"=matrix(1,K,K),
-             "cc"=matrix(0,K,K))
-z <- rep(1,N)#c(rep(1,N/2),rep(2,N/2))
-P <- length(beta)
-beta <- abind(beta,rev.along=3)
-
-
-brem.lpost.block <- function(A,N,K,z,s,beta,k1,k2,px,priors=list(beta=list(mu=0,sigma=1))) { 
-  beta[which(px==0),,] <- 0
-  M <- nrow(A)
-  zs <- z[A[,2]]
-  zr <- z[A[,3]]
-  ix <- which(zs==k1 | zr==k2) - 1  # 0 based indexing for c++
-  if (length(ix) == 0) return(0)
-  sum(dnorm(unlist(beta[which(px==1),k1,k2]),priors$beta$mu,priors$beta$sigma,log=TRUE)) +  sum(RemLogLikelihoodPcSubset(beta,z-1,s$ptr(),K,ix))
-}
-brem.lpost.block.grad <- function(A,N,K,z,s,beta,k1,k2,px,priors=list(beta=list(mu=0,sigma=1))) {   
-  M <- nrow(A)
-  zs <- z[A[,2]]
-  zr <- z[A[,3]]
-  ix <- which(zs==k1 | zr==k2) - 1  # 0 based indexing for c++
[…]

### 2012-03-20 pkg/src/brem.cpp

@@ -40,76 +40,10 @@ rcategorical is a helper function for gibbs.  It draws from a categorical distri
 #include <algorithm>
 #include <vector>
 #include <Rcpp.h>
+#include <utils.h>
 using namespace std;
 using namespace Rcpp;
 RNGScope scope;
-
-/*
- Utilities
-*/
-
-// Return the (j,k,l) element of a (J,K,L) array represented as a vector.
-int threeDIndex(int j, int k, int l, int J, int K, int L) { 
-  return l*J*K + k*J + j;
-}
-
-bool IsFiniteNumber(double x) {
-  return (x <= DBL_MAX && x >= -DBL_MAX); 
-} 
-
-// For a vector of log probabilities, return a draw from the categorical distribution
-int rcategorical (Rcpp::NumericVector lp) {
-  int max = 0;
-  int min = 0;
-  int K = lp.size();
-  for (int i = 0; i < K; i++) {
-    if (lp[i] > lp[max]) {
-      max = i;
-    }
-    if (lp[i] < lp[min]) {
-      min = i;
-    }
-  }
-  
-  // If all small/large values, subtract max/min value for numeric stability.
-  if (lp[max] < 0) {
-    double lmax = lp[max];
-    for (int i = 0; i < K; i++) {
-      lp[i] -= lmax;
-    }
-  }
-  if (lp[min] > 0) {
-    double lmin = lp[min];
-    for (int i = 0; i < K; i++) {
-      lp[i] -= lmin;
-    }
-  }
-
-  // Inverse CDF method
-  Rcpp::NumericVector p = exp(lp);
-
-  // Get sum.  If any Infinite values, return that index. 
-  // TODO: This probably introduces a bias towards smaller k.
-  int k;
-  double cuml = 0;
-  double sum = 0;
-  for (k=0;k<K;k++) {
-    if (!IsFiniteNumber(p[k])) {
-      return k;
-    }
-    sum += p[k];
-  }
-
-  double r = as<double>(Rcpp::runif(1));
-  for (k=0;k<K;k++) {
-    cuml += p[k]/sum;
-    if (r < cuml) {
-      break;
-    } 
-  }
-  return k;
-}
-
 
 /*
 Data structure for precomputing relational event statistics

### 2012-03-20 pkg/src/remdp.cpp

@@ -1,0 +1,373 @@
+#include <iostream>
+#include <algorithm>
+#include <vector>
+#include <Rcpp.h>
+using namespace std;
+using namespace Rcpp;
+
+class RemDP {
+public:
+
+  RemDP(Rcpp::List edgelist_,  int N_) : 
+    edgelist(edgelist_), N(N_) {
+    times = edgelist(0);
+    sen = edgelist(1);
+    rec = edgelist(2);
+    M = sen.size();
+    P = 13;
+    initialize();
+    //precompute();
+  }
+
+
+  RemDP(Rcpp::NumericVector times_, 
+        Rcpp::IntegerVector sen_,  
+        Rcpp::IntegerVector rec_, 
+        int N_, int M_, int P_) : 
+    times(times_),sen(sen_),rec(rec_),
+    N(N_),M(M_),P(P_) {
+    initialize();
+    //precompute();
+  }
+
+  void initialize() {
+
+    // Current vectors of statistics
+    for (int i = 0; i < N; i++) {
+      vector< vector<int> > w_i;
+      for (int j = 0; j < N; j++) {
+        vector<int> w_ij(M);
+        w_i.push_back(w_ij);
+      }
+      w.push_back(w_i);
+    }
+
+    // Current vectors of statistics
+    for (int i = 0; i < N; i++) {
+      vector< vector<int> > s_i;
+      for (int j = 0; j < N; j++) {
+        vector<int> s_ij(P);
+        s_i.push_back(s_ij);
+      }
+      s.push_back(s_i);
+    }
+
+    // Actor changepoitn indices
+    for (int i = 0; i < N; i++) {
+      vector<int> u_i;
+      u.push_back(u_i);
+    }
+
+    // Dyad changepoint indices
+    for (int i = 0; i < N; i++) {
+      vector< vector<int> > v_i;
+      for (int j = 0; j < N; j++) {
+        //        int arr[] = {0};
+        vector<int> v_ij(1);// (arr, arr + sizeof(arr) / sizeof(arr[0]) );
+        v_i.push_back(v_ij);
+      }
+      v.push_back(v_i);
+    }
+    
+    // Dyadic level statistics at each changepoing
+    for (int i = 0; i < N; i++) {
+      vector< vector< vector<int> > > x_i;
+      for (int j = 0; j < N; j++) {
+        vector< vector<int> > x_ij;
+        vector<int> x_ijp(P);
+        //        x_ijp[0] = 1;
+        x_ij.push_back(x_ijp);
[…]

### 2012-03-20 pkg/src/utils.cpp

@@ -1,0 +1,70 @@
+#include <iostream>
+#include <algorithm>
+#include <vector>
+#include <Rcpp.h>
+
+/*
+ Utilities
+*/
+
+// Return the (j,k,l) element of a (J,K,L) array represented as a vector.
+int threeDIndex(int j, int k, int l, int J, int K, int L) { 
+  return l*J*K + k*J + j;
+}
+
+bool IsFiniteNumber(double x) {
+  return (x <= DBL_MAX && x >= -DBL_MAX); 
+} 
+
+// For a vector of log probabilities, return a draw from the categorical distribution
+int rcategorical (Rcpp::NumericVector lp) {
+  int max = 0;
+  int min = 0;
+  int K = lp.size();
+  for (int i = 0; i < K; i++) {
+    if (lp[i] > lp[max]) {
+      max = i;
+    }
+    if (lp[i] < lp[min]) {
+      min = i;
+    }
+  }
+  
+  // If all small/large values, subtract max/min value for numeric stability.
+  if (lp[max] < 0) {
+    double lmax = lp[max];
+    for (int i = 0; i < K; i++) {
+      lp[i] -= lmax;
+    }
+  }
+  if (lp[min] > 0) {
+    double lmin = lp[min];
+    for (int i = 0; i < K; i++) {
+      lp[i] -= lmin;
+    }
+  }
+
+  // Inverse CDF method
+  Rcpp::NumericVector p = exp(lp);
+
+  // Get sum.  If any Infinite values, return that index. 
+  // TODO: This probably introduces a bias towards smaller k.
+  int k;
+  double cuml = 0;
+  double sum = 0;
+  for (k=0;k<K;k++) {
+    if (!IsFiniteNumber(p[k])) {
+      return k;
+    }
+    sum += p[k];
+  }
+
+  double r = Rcpp::as<double>(Rcpp::runif(1));
+  for (k=0;k<K;k++) {
+    cuml += p[k]/sum;
+    if (r < cuml) {
+      break;
+    } 
+  }
+  return k;
+}

### 2012-03-23 docs/notes.tex
Outline REM DP c++ and add tests.
Edit notes for final report.

@@
 \abstract{
-Stochastic blockmodels assume latent groups of individuals have similar tendencies in their  group-wise interactions.
-Many find these methods useful for exploratory analysis of static networks.
-On the other hand, continuous-time methods such as the relational event framework (Butts 2008) model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.
-We propose to combine these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.
-The method is illustrated using dyadic interaction data (e.g.
-email, Twitter direct messages) among hundreds of actors and thousands of events.
-We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than using a single relational event model.
+Observing network data as a stream of events provides an opportunity to understand differences in how nodes behave given the past.
+Recent methods for continuous-time network data model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.
+Methods for static networks, such as stochastic blockmodels, often account for actor-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
+We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
+The method is illustrated using dyadic interaction data -- e.g.
+email, Twitter direct messages.
+We show parameter estimates from the model reveal heterogeneity in the dynamics among groups of individuals.
+The fitted models have better predictive accuracy than both baselines and relational event models without the latent structure.
 Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
@@
 Network data, however, is often collected as a sequence of events occurring over time.
-Recently several works have adapted models from event history analysis \cite{} to provide continuous time models of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
+Recently work leverages  continuous time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
 These models allow one to specify how the process depends on the previous history of events.
-In this way one may incorporate theories about the underlying processes and make predictions about future data conditioned on the past.
+In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
 
-The above continuous time network models have two major drawbacks: 1) the entire network is assumed to have identical dynamics, and 2) the occurrence of any event can affect the rate at which any other event occurs.
-Both aspects may be unrealistic.
-Consider email communication in a small organization comprised of several teams.
-Each team may have a different pattern for collaborating via email, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
+The above continuous time network models assume all nodes in the network behave according to identical dynamics.
+In many cases this is unrealistic.
+For example, in a small organization comprised of several teams, each team may exhibit different patterns for collaborating via email --  individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
+%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
 
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.
-The behavior of the model is illustrated with data simulated from the model.
-Using  data sets of dyadic communication among people, we demonstrate the ability to improve upon simple baselines for predicting the next event.
-Finally we show that the parameter estimates provide an interpretable structure to the event dynamics among blocks of individuals, sometimes identifying particular \emph{roles} of communication that exist within the network.
+Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.
+The behavior of the model is illustrated with simulated data.
+We describe parameter estimation and the learning of the latent cluster assignments via MCMC.
+Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.
+Finally we show that the parameter estimates provide an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication that exist.
 
@@
 One can extend the above to marked point processes where each event contains additional information.
-In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ of possible events include all allowed dyads.
-Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t)$ that is piecewise constant.
+In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
+We assume each dyad $(i,j) \in \mathcal{R}$ has an intensity function $\lambda_{ij}(t)$ that is piecewise constant.
 We can then write the likelihood as
-
 \begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | z_i,z_j) \}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | z_i,z_j) \}
+\label{eqn:llk}
 \end{align}
-\noindent where event $m$ is $(i_m,j_m)$, $\tau_{i,j,m}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads who have a changepoint in their intensity when $(i_m,j_m)$ occurs.
+\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{i,j,m}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs.
 \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
@@
 }
-\caption{Illustration of event data and the assumptions of the model.
+\caption[]{Illustration of event data and the assumptions of the model.
 Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-[TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.]}
+Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
+Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.\footnotemark }
 \label{fig:example}
 \end{figure}
+
+\footnotetext{TODO?
+Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
 
@@
 We hope to learn about the dynamics between subsets of nodes while allowing for variety among disjoint subsets.
-To facilitate this, we assume each node belongs to a cluster $z_i \sim \mbox{Categorical}(\alpha)$ where $\sum_k \alpha_k = 1$ and  model the intensity functions via a log linear model
-\begin{align}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z}).
-\end{align}
-where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}_{ij}(t|\cdot)$.
+To facilitate this, we assume node $i$ is assigned a latent cluster $z_i$ and the intensity functions via a log linear model
+\begin{align*}
+\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) &= \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t) &
+z_i &\sim \mbox{CRP}(\alpha) &
+\beta_{k,l,p} &\sim \mbox{N}(\mu_p,\sigma_p^2)
+\end{align*}
+where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}(t,i,j|\cdot)$.
 Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
+To allow for the blocks to share information by placing a hierarchical prior on each $\boldsymbol{\beta}_{k.l}$.
 
@@
 
-We construct $\mathbf{s}(t,i,j)$ using a similar set of statistics as used in \cite{Butts2008,Vu2011}.
-Let $v_{ijm} \in [0,M]$ be the index of the last event where $(i,j)$ occurred (n.b.
-$t_{v_{ijm}} = \tau_{ijm}$).
-If $(i,j)$ has never occurred, we set $v_{ijm}=0$.
-\begin{enumerate}
-\item Intercept: $s_{0}(t,i,j) = 1$
-\item Dyad count: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i) / m$
-\item Sender out-degree: $s_{1}(t,i,j) = \sum_{m:t_m<t} I(i_m=i) / m$
-\item Sender in-degree: $s_{2}(t,i,j) = \sum_{m:t_m<t} I(j_m=i) / m$
-\item Receiver out-degree: $s_{3}(t,i,j) = \sum_{m:t_m<t} I(i_m=j)/m$
-\item Receiver in-degree: $s_{4}(t,i,j) = \sum_{m:t_m<t} I(j_m=j)/m$
-\item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-\item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-%\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
-% \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
-% \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
-% \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
-\end{enumerate}
+% \begin{enumerate}
+% \item Intercept: $s_{0}(t,i,j) = 1$
+% \item Dyad count: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
+% \item Sender out-degree: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
+% \item Sender in-degree: $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$
+% \item Receiver out-degree: $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$
+% \item Receiver in-degree: $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$
+% \item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+% \item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+% \item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+% \item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+% \item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+% %\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
+% % \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
+% % \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
+% % \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
+% \end{enumerate}
 
-We include degree effects to model possible rich-get-richer phenomenon.
+\begin{table}[t]
+\footnotesize
+\center
+\begin{tabular}{|l|l|}
+\hline
+Statistic & Formula \\
+\hline
+\hline
+Intercept:& $s_{0}(t,i,j) = 1$\\
+Dyad count:& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
+Sender out-degree:& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
+Sender in-degree:& $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
+Receiver out-degree:& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
+Receiver in-degree:& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
+Reciprocity (AB-BA):& $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$\\
+Turn-taking (AB-BY): &$s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}\ne i)$\\
+Turn-usurping (AB-XA):& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}} \ne j,j_{v_{ijm}}=i)$\\
+Turn-usurping (AB-XB): &$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}} \ne i,j_{v_{ijm}}=j)$\\
+Turn-continuing (AB-AY):& $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}\ne j)$\\
+\hline
+\end{tabular}
+\label{tab:stats}
+\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
+\end{table}
+
+In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.
+Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred (n.b.
+$t_{v_{mij}} = \tau_{mij}$).
+If $(i,j)$ has never occurred, we set $v_{mij}=0$.
+Here we include degree effects to model possible rich-get-richer phenomenon.
 For example, a particular node sending often may indicate they will continue to be a sender in the future.
 We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
-TODO: Develop some theory about these sorts of statistics.}
+TODO: Explore the theory about these sorts of statistics.
+Aalen has a few suggestions.} Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.
+\footnote{Describe the form of $f()$.
+Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
 
@@
 
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is essentially a mixing rate between group $z_i$ and group $z_j$.
-Here, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
+Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
+In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
 
 The above framework generalizes several important special cases.
-The first is one where events within each block share the rate at which they occur.
-This can be implemented by only having an intercept statistic, $s_0(t,i,j) = 1$.
+For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.
+Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
 As these intensities do not change under this specification, the likelihood simplifies to
 $$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-where $ \lambda_{ij} = \exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
-Analogous to the stochastic block model for static networks, this model posits each dyad is a homogeneous Poisson process and all dyads within block $(z_i,z_j)$ have the intensity $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
+
+Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood becomes a conditional logit
+$$\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.$$
+The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
+
+\subsection{Related work}
+
+Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant and allow for a nonlinear relationship between statistics and intensity functions.
+
+(TODO: More here.  For example: \cite{Tillman2011} explores the idea of having nonpoisson processes at each edge and solves the master equation for the dynamical system.)
 
@@
 
-We use Markov chain Monte Carlo to get samples from the posterior distribution of our parameters.
+We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.
 
-\begin{enumerate}
-\item For $i \in [1,N]$: Sample $z_i$ via Gibbs sampling: 
-$z_i | A, z_{-i} \propto p(A|z,\beta) p(z_i)$.  (See Appendix.)
-\item For each pair $(k,k')$: Sample $\beta_{k,k'}$ via MH or HMC.
-\end{enumerate}
+\subsection*{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
+
+We Gibbs sample the latent class assignments $\mathbf{z}$ from the conditional distribution
+\begin{align*}
+p(z_i | z_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto p(z_i | z_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}).% & z_i &= \mbox{new} = \alpha
+\end{align*}
+Letting $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ be the set of events where node $r$ is involved, the likelihood term above can be written as
+$$p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) = \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
+\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}$$
+
+\noindent Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$. 
+% \begin{align*}
+% p(z_i = k | z_{-i},\alpha) &= n_k & p(z_i = \mbox{new} |  z_{-i},\alpha) &= \alpha
+% \end{align*}
+
+ By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot 4N)$.
+The factor of 4 is due to the number of dyads involving a single node.
+We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.
+
+\subsection*{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
+
+For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
+\begin{align*}
+p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \mu, \sigma) &\propto p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\
+p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)
+\end{align*}
+the likelihood term above can be written as
+$$p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) = \prod_{a \in \mathcal{V}_{k,l}} \lambda_{i_a,j_a}(t_a|\cdot)
+\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}$$
+\noindent where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \mbox{or} z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.
+\footnote{TODO: If we end up using HMC, mention step size and step length used.}
+The above distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}
+
+\subsection*{Split-merge algorithm}
+
+Motivation: simply Gibbs sampling does not seem to mix well.  \footnote{This section is planned future work.}
+
+In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
+For these chains we do not propose split moves when the number of non-empty clusters is already $K$.
 
@@
 
-Slice sampling with $m=20$.
-
- To perform Gibbs sampling for a given cluster assignment $z_i$, we need to compute the likelihood for each possible assignment of $z_i$.
-By limiting the changepoints to times when either node is involved, computing the likelihood is $O(T \cdot P \cdot 4N)$ where 4 comes from considering all dyads where either $i$ or $j$ is involved.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.
-Note that
 % This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.
@@
 
-\section{Simulation}
 \begin{figure}
 \center
-\includegraphics[width=2in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=3in]{../figs/synthetic/counts.pdf}
+\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
+\includegraphics[width=2.25in]{../figs/synthetic/counts.pdf}
+\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
 \caption{Illustration of 1000 simulated events, as described in text.
 Left: Counts of each dyad.
-Right: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.}
+Center: Boxplot of distribution of participation counts across dyads.
+The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
+Right: Parameters (in red) and posterior credible intervals (in black).}
 \label{fig:syncounts}
@@
 
+\section{Simulation}
+
 We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to continue speaking, and 3) interactions between groups are more likely to be repeated.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{5}(t,i,j), s_{9}(t,i,j), s_{11}(t,i,j)]$.
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (2,3,0,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (1,0,0,1)$, and $\boldsymbol{\beta}_{2,2} = (2,0,2,0)$.
-See Appendix for details on simulating from the model.
-In Figure \ref{fig:syncounts} we illustrate the simulated data and show the loglikelihood during the first iterations of MCMC [after being initialized with $\beta \sim N(\beta_{true},.5^2)$].
-Note that we recover the same cluster assignments and the full model converges to the true loglikelihood.
+The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{5}(t,i,j), s_{9}(t,i,j)]$.
+For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,0,2)$.
+This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
 
-TODO: Show parameter bias as a function of iteration; perhaps also as a function of the number of events in the dataset.
+In Figure \ref{fig:syncounts} we illustrate the simulated data.
+While the dyad counts suggest a stochastic blockmodel, the center plot shows each block has empirical differences in their dynamics: intensities for reciprocal actions among actors in block 1 are $e^3$ times greater, intensities for turn-taking actions among actors in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
+Fitting the model with $K=2$ converges to the true loglikelihood (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.
 
 \section{Model checking and experiments}
+
+\small{
+\input{../figs/results.tex}
+}
 
@@
 %\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
+\item Eckmann email:
+\item Kiel email: Still need to hunt this down.
+\item Enron email:
 \item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
@@
 We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item The Irvine data set
+\item MIT Reality Mining: 
 \end{itemize}
@@
 
-\begin{figure}
-\center
+\begin{figure}[th]
+\caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
+\label{fig:recall}
+\end{figure}
+
+%\begin{figure}
+%\center
 %\includegraphics[width=3in]{../figs/synthetic/logposterior.pdf}
-\includegraphics[width=3in]{../figs/synthetic/test-recall.pdf}
-\caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
-\label{fig:synresults}
-\end{figure}
-We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.
+% \includegraphics[width=3in]{../figs/synthetic/test-recall.pdf}
+% \caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
+% \label{fig:synresults}
+% \end{figure}
+We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.
+Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
+I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
 
@@
 
-Our method jointly models both which dyads occur and when they occur.
+Our method jointly models \emph{which} dyads occur and \emph{when} they occur.
 To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j} = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
 
-We consider several variations of our model to investigate its properties.
+%We consider several variations of our model to investigate its properties.
 The \texttt{baserates} model only is an intercept only model (and mimics the behavior of a stochastic block model).
@@
 
-TODO: In Figure \ref{fig:} we show the recall for each dataset.
+ Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
+Two likelihoods are considered: 1) \texttt{rem} (relational event model) likelihood as given in Equation \ref{eqn:llk}, and \texttt{mult} (multinomial likelihood) given in Equation \ref{eqn:multllk}.
+The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
 
-TODO: In Table \ref{tab:} we show the likelihood on the test dataset.
+For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
+Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.
 
-TODO: Need to see how well the time-portion of the model is specified.
+For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
+This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
+\footnote{Will need to discuss results for other datasets once they are complete.}
+
+Figure \ref{fig:recall} provides recall curves for several of the datasets.
 
@@
 
+\begin{figure}[ht]
+\includegraphics[width=3in]{../figs/eckmann-small/params-estimates}
+\caption{Left: Posterior credible intervals for parameter estimates from fitting the $K=2$ model to the Eckmann subset.
+Right: Posterior distribution of $K$ when fitting the model and allowing for flexible $K$.}
+\label{fig:posteriorparams}
+\end{figure}
+
+ We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
+The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models) to provide detailed models of event dynamics among subsets of nodes while simultaneously allowing for heterogeneity in dynamics of the network as a whole.
+
+%By combining stochastic blockmodels with a relational event framework, the proposed method accounts for unobserved heterogeneity in the way dyadic events occur conditioned on the past.
+
+Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions.
+Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
+
+In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods with respect to ranking tasks and the likelihood of unobserved data.
+
 TODO: Discuss experimental results.  Describe qualitative behavior of each model's recall line.
 
-TODO: Discuss any interpretation of the parameters and clusters for both datasets.
+TODO: Describe and include posterior predictive checks.
 
-TODO: discuss non-specification of rate and just predicting next
-event, could still use our approach?
+Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j)$.
 
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches \cite{Gunawardana2011} aim to learn the regions where an intensity is constant and allow for a nonlinear relationship between statistics and the intensity function.
+Our method is able to reveal heterogeneity that exists among individuals.  
+The purpose of our method is to learn about event dynamics within latent groups of individuals.
+Other types of heterogeneity likely exist in some data sets.
+For example, dynamics might change over time \cite{Vu2011}.
+Alternatively, if nodes change groups.
+One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.
+In some contexts such extensions might be substantively important and warrant future work.
 
-\subsection{Extensions}
-Possible future work could extend the present framework to a mixed-membership model where for each effect there is a latent cluster assignment.
+% \subsection{Extensions}
+% Possible future work could extend the present framework to a mixed-membership model where for each effect there is a latent cluster assignment.
 This way a group of nodes might contact high in-degree nodes at a similar rate, though they vary in the degree to which they adhere to reciprocity.
@@
 
-Possible variants:
-\begin{itemize}
-\item Discuss possible variations on the prior structure of $\beta$: 1.
+% Possible variants:
+% \begin{itemize}
+% \item Discuss possible variations on the prior structure of $\beta$: 1.
 Enforce $\beta_{k,l} = \beta_{l,k}$.
@@
 Place other priors that provide different kinds of regularization.
-\item $\log \lambda_{ij} = (\pi_i \cdot \pi_j) \eta s(t,i,j)$
-\item  HMM that guides which latent class over time?
-\item  HDP instead of fixed K: Hierarchical, nonparametric model that allows parameters to be shared across blocks.
-\item different clusters for senders and receivers (cite our KDD paper)?
-\item More statistics, e.g. topic model similarity
-\end{itemize}
+% \item $\log \lambda_{ij} = (\pi_i \cdot \pi_j) \eta s(t,i,j)$
+% \item  HMM that guides which latent class over time?
+% \item  HDP instead of fixed K: Hierarchical, nonparametric model that allows parameters to be shared across blocks.
+% \item different clusters for senders and receivers (cite our KDD paper)?
+% \item More statistics, e.g. topic model similarity
+% \end{itemize}
 
-\section{Related work}
-\cite{Tillman2011} explores the idea of having nonpoisson processes at each edge and solves the master equation for the dynamical system.
+% \appendix
+% \section{Simulation}
 
-\section{Conclusion}
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models) to provide detailed models of event dynamics among subsets of nodes while simultaneously allowing for heterogeneity in dynamics of the network as a whole.
+% Simulation from the model follows the typical racing exponentials idea with a few subtleties.
 
-\appendix
-\section{Simulation}
+% \begin{enumerate}
+% \item $t_0 = 0$, all $s_{ij} = 1$, Initialize $v = 0$
+% \item For $i = 1, \ldots, N$: Draw $z_i \sim \mbox{Categorical}(\theta)$
+% \item For $m = 1, \ldots, M$:
+%   \begin{enumerate}
+%   \item Draw $t_m \sim \mbox{Exponential}\left(\sum_{ij} \lambda_{ij}^*\right)$
+%   \item Draw $(i_m,j_m) \sim \mbox{Categorical}\left(\lambda_{ij}^* / \sum_{ij}\lambda_{ij}^*\right)$
+%   % \item Save $s_{ij}(t_m) \leftarrow s_{ij}^*$
+%   % \item Save $\lambda_{ij}(t_m) \leftarrow \lambda_{ij}^*$
+%   \item For all $(i,j)$ where $i$ or $j$ is in $\{i_m,j_m\}$:
+%     \begin{enumerate}
+%     \item Compute $s_{ij}^*$ using event $0$ through $v_{ij}$
+%     \item Set $\lambda_{ij}^* \leftarrow \exp\{ \beta_{z_i,z_j} s_{ij}^*\}$ 
+%     \end{enumerate}
+%   \end{enumerate}
+% \end{enumerate}
 
-Simulation from the model follows the typical racing exponentials idea with a few subtleties.
+% \section{EM for baseline}
 
-\begin{enumerate}
-\item $t_0 = 0$, all $s_{ij} = 1$, Initialize $v = 0$
-\item For $i = 1, \ldots, N$: Draw $z_i \sim \mbox{Categorical}(\theta)$
-\item For $m = 1, \ldots, M$:
-  \begin{enumerate}
-  \item Draw $t_m \sim \mbox{Exponential}\left(\sum_{ij} \lambda_{ij}^*\right)$
-  \item Draw $(i_m,j_m) \sim \mbox{Categorical}\left(\lambda_{ij}^* / \sum_{ij}\lambda_{ij}^*\right)$
-  % \item Save $s_{ij}(t_m) \leftarrow s_{ij}^*$
-  % \item Save $\lambda_{ij}(t_m) \leftarrow \lambda_{ij}^*$
-  \item For all $(i,j)$ where $i$ or $j$ is in $\{i_m,j_m\}$:
-    \begin{enumerate}
-    \item Compute $s_{ij}^*$ using event $0$ through $v_{ij}$
-    \item Set $\lambda_{ij}^* \leftarrow \exp\{ \beta_{z_i,z_j} s_{ij}^*\}$ 
-    \end{enumerate}
-  \end{enumerate}
-\end{enumerate}
+% In our baseline model, each block $(a,b) \in [1,K] \times [1,K]$ simply has a rate $\gamma_{a,b}$.
+% The likelihood is
+% \begin{align*}
+% \mathcal{L}(A|\lambda) &= \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M\lambda_{ij}\} \\
+% & =\prod_{a=1}^K \prod_{b=1}^K \prod_{m=1}^M \gamma_{ab}^{I(z_{i_m}=a,z_{j_m}=b)} \prod_{(i,j): z_i=a,z_j=b} \exp\{-t_M\gamma_{ab}\} \\
+% & =\prod_{a=1}^K\prod_{b=1}^K \gamma_{ab}^{n_{ab}} \exp\{-t_Mn_an_b\gamma_{ab}\}
+% \end{align*}
+% where $n_{ab} = \sum_m I(z_{i_m}=a,z_{j_m}=b)$, $n_a = \sum_{i=1}^N I(z_i=a)$.
 
-\section{Gibbs sampling}
+% The loglikelihood is $$\ell(A|\lambda) = \sum_{a=1}^K \sum_{b=1}^K n_{ab} \log \gamma_{ab} - t_M n_a n_b \gamma_{ab}$$
+% and so $\hat{\gamma}_{ab} = \frac{n_{ab}}{t_Mn_an_b}$.
 
-Let $\mathcal{U}_a = \{a: a \in (i_m,j_m), m \in [1,M]\}$ be the set of events where actor is involved:
-
-$$p(A|z,\beta,a \ \mbox{involved}) = \prod_{a \in \mathcal{U}_a} \lambda_{i_a,j_a}(t_a|\cdot)
-\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{ija}) \lambda_{ij}(t_a|\cdot)\}$$
-
-\section{EM for baseline}
-
-In our baseline model, each block $(a,b) \in [1,K] \times [1,K]$ simply has a rate $\gamma_{a,b}$.
-The likelihood is
-\begin{align*}
-\mathcal{L}(A|\lambda) &= \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M\lambda_{ij}\} \\
-& =\prod_{a=1}^K \prod_{b=1}^K \prod_{m=1}^M \gamma_{ab}^{I(z_{i_m}=a,z_{j_m}=b)} \prod_{(i,j): z_i=a,z_j=b} \exp\{-t_M\gamma_{ab}\} \\
-& =\prod_{a=1}^K\prod_{b=1}^K \gamma_{ab}^{n_{ab}} \exp\{-t_Mn_an_b\gamma_{ab}\}
-\end{align*}
-where $n_{ab} = \sum_m I(z_{i_m}=a,z_{j_m}=b)$, $n_a = \sum_{i=1}^N I(z_i=a)$.
-
-The loglikelihood is $$\ell(A|\lambda) = \sum_{a=1}^K \sum_{b=1}^K n_{ab} \log \gamma_{ab} - t_M n_a n_b \gamma_{ab}$$
-and so $\hat{\gamma}_{ab} = \frac{n_{ab}}{t_Mn_an_b}$.
-
-I do not believe it is possible to perform EM for a model where we have the more general linear predictor for the intensity function.
+% I do not believe it is possible to perform EM for a model where we have the more general linear predictor for the intensity function.
 Since the assignment of one node depends on the assignments of the other nodes, we suffer combinatorial explosion.
 
-\section{Gradient}
+% \section{Gradient}
 
-\begin{align*}
-\frac{\partial \ell}{\partial \beta_{p,k,l}} = \frac{\partial}{\partial \beta_{p,k,l}} \sum_{m = 1}^M \left[ \log \lambda_{i_m,j_m}(t_m | \cdot) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) \lambda_{i,j}(t_m|\cdot) \right] \\
-  =\sum_{m \in V} \left[ s_p(t_m,i_m,j_m) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) s_p(t_m,i,j) \lambda_{i,j}(t_m|\cdot) \right]
-\end{align*}
+% \begin{align*}
+% \frac{\partial \ell}{\partial \beta_{p,k,l}} &= \frac{\partial}{\partial \beta_{p,k,l}} \sum_{m = 1}^M \left[ \log \lambda_{i_m,j_m}(t_m | \cdot) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) \lambda_{i,j}(t_m|\cdot) \right] \\
+%   &=\sum_{m \in V} \left[ \frac{\partial}{\partial \beta_{p,k,l}} \sum_p \beta_{p,k,l} s_p(t,i,j|\mathcal{A}_t)  - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) \frac{\partial}{\partial \beta_{p,k,l}} \exp \left\{\sum_p \beta_{p,k,l} s_p(t,i,j|\mathcal{A}_t) \right\} \right] \\
+%   &=\sum_{m \in V} \left[ s_p(t_m,i_m,j_m) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) s_p(t_m,i,j) \lambda_{i,j}(t_m|\cdot) \right]
+% \end{align*}
 
-where $V$ is the set of $m$ where either $z_i \in \{k,l\}$ or $z_j \in \{k,l\}$.
+% where $V$ is the set of $m$ where either $z_i \in \{k,l\}$ or $z_j \in \{k,l\}$.
 

### 2012-03-23 figs/results.tex
Edit notes for final report.

% latex table generated in R 2.14.0 by xtable 1.7-0 package
% Thu Mar 22 18:27:09 2012
\begin{table}[t]
\begin{center}
{\footnotesize
\begin{tabular}{lllrrrrrrrr}
  \hline
dataset & likelihood & type & uniform & marg & online & K=1 & K=2 & K=3 & dp & truth \\ 
  \hline
synthetic & rem & train & -0.928 & -0.916 & -0.689 & -0.577 & -0.106 &  &  & -0.107 \\ 
   &  & test & -0.919 & -0.919 & -0.636 & -0.589 & -0.130 &  &  & -0.122 \\ 
   & mult & train & -4.605 & -4.593 & -4.281 & -4.157 & -3.688 &  &  & -3.689 \\ 
   &  & test & -4.605 & -4.606 & -4.225 & -4.179 & -3.729 &  &  & -3.720 \\ 
  eckmann-small & rem & train & -8.800 & -7.722 & -7.309 & -6.552 & -6.350 & -6.430 &  &  \\ 
   &  & test & -8.761 & -7.727 & -6.659 & -6.151 & -6.106 & -6.455 &  &  \\ 
   & mult & train & -8.955 & -7.889 & -7.462 & -6.514 & -6.438 & -6.456 &  &  \\ 
   &  & test & -8.955 & -7.933 & -6.849 & -6.051 & -6.238 & -6.340 &  &  \\ 
   \hline
\end{tabular}
}
\caption{Comparing mean loglikelihood for each event across methods for each dataset.  Larger values are better.  See text for details. [TODO: More datasets as well as results for the DP version.  Could also remove training set results.]}
\label{tab:results}
\end{center}
\end{table}

### 2012-03-23 dashboard.r

@@ -33,6 +33,7 @@ results.dir <- paste("results/",opts$dataset,sep="")
 load(paste("data/",opts$dataset,".rdata",sep=""))
 fs <- list.files(results.dir,full.names=TRUE)
 fs <- fs[-grep("ranks",fs)]
+fs <- fs[-grep("final",fs)]
 fs <- fs[-grep("llks",fs)]
 fs <- fs[-grep("counts",fs)]
 fnames <- unlist(strsplit(fs,".rdata"))
@@ -228,15 +229,20 @@ if (opts$predictions) {
   }
   
   # Make results table
-  r <- cbind(ddply(llks.train,.(L1),summarise,llk=mean(value)),
-             ddply(llks.test,.(L1),summarise,llk=mean(value)),
-             ddply(mllks.train,.(L1),summarise,llk=mean(value)),
-             ddply(mllks.test,.(L1),summarise,llk=mean(value)))
-  r <- r[,c(1,2,4,6,8)]  # remove extraneous columns
-  colnames(r) <- c("method","brem.train","brem.test","multin.train","multin.test")
-  for (i in 2:5) r[,i] <- round(r[,i],2)
-  r
-  print(xtable(r),include.rownames=FALSE,file=paste("figs/",opts$dataset,"/results.tex",sep=""))
+  df <- rbind(
+        cbind(likelihood="rem",type="train",
+              ddply(llks.train,.(L1),summarise,value=mean(value))),
+        cbind(likelihood="rem",type="test",
+              ddply(llks.test,.(L1),summarise,value=mean(value))),
+        cbind(likelihood="mult",type="train",
+              ddply(mllks.train,.(L1),summarise,value=mean(value))),
+        cbind(likelihood="mult",type="test",
+              ddply(mllks.test,.(L1),summarise,value=mean(value)))
+              )
+  df$L1 <- factor(df$L1,c("uniform","marg","online","full.1","full.2","full.3","dp","truth"))
+  df$dataset <- opts$dataset
+  dir.create(paste("results/",opts$dataset,"/final",sep=""))
+  save(df,file=paste("results/",opts$dataset,"/final/results.rdata",sep=""))
 }
 
 cat("Creating dashboard.\n")

### 2012-03-23 eda/eda.r

@@ -1,0 +1,68 @@
+load("data/eckmann-small.rdata")
+P <- 13
+library(brem)
+s <- new(RemStat,train[,1],train[,2]-1,train[,3]-1,N,nrow(train),P)
+s$precompute()
+s$transform()
+
+# Arrange statistics vectors in large matrix with P columns
+x <- s$get_all_s()
+x[[1]][[2]][1:5]
+xs <- list()
+k <- 0
+for (i in 1:N) {
+  for (j in (1:N)[-i]) {
+    xsk <- do.call(rbind,x[[i]][[j]][-1])
+    ms <- xsk[,ncol(xsk)] + 1
+    timegap <- c(0,diff(train[ms,1]))
+    xs[[k <- k+1]] <- cbind(i,j,1:nrow(xsk),timegap,xsk)
+  }
+}
+ys <- do.call(rbind,xs)
+colnames(ys) <- c("i","j","v","timegap",paste("s",0:12,sep=""))
+
+# Whether or not that dyad was observed
+ms <- ys[,ncol(ys)]   # index of changepoints
+obs <- (train[ms+1,2] == ys[,1]) & (train[ms+1,3] == ys[,2])
+ys <- cbind(obs,ys)
+
+# Sanity check
+ix <- head(which(obs))
+ys[ix,1:5]
+train[ms[ix]+1,]
+
+# Remove event ids
+tmp <- ys[,-c(2,3,4,ncol(ys))]
+
+# Kmeans
+K <- 10
+km <- kmeans(tmp,K)
+
+clusters <- factor(km$cluster,1:K)
+sen <- factor(ys[,2],1:N)
+rec <- factor(ys[,3],1:N)
+plot.phi(table(clusters,sen) + table(clusters,rec))
+plot.phi(table(clusters,sen))
+plot.phi(table(clusters,rec))
+
+# Starting with most popular cluster, get list of actors with higher than average assignments.
+plot.phi <- function(phi) {
+    o.k <- order(rowSums(phi),decreasing=TRUE)
+    ixs <- lapply(1:K,function(i) {
+      o.a <- order(phi[o.k[i],],decreasing=TRUE)
+      ix <- which(phi[o.k[i],o.a] > mean(phi[o.k[i],]))
+      o.a[ix]
+    })
+    o <- unique(unlist(ixs))
+
+    phi <- apply(phi,1,function(p) p/colSums(phi))
+
+    library(ggplot2)
+    phi <- melt(phi)
[…]

### 2012-03-23 results.r

@@ -1,0 +1,20 @@
+
+res <- lapply(c("synthetic","eckmann-small"),function(x) {
+  load(paste("results/",x,"/final/results.rdata",sep=""))
+  return(df)
+})
+res <- do.call(rbind,res)
+res$dataset <- factor(res$dataset,c("synthetic","eckmann-small"))
+
+r <- cast(res,dataset + likelihood + type ~ L1,add.missing=TRUE)
+
+# Clean up
+colnames(r)[7:9] <- paste("K=",1:3,sep="")
+r$dataset    <- as.character(r$dataset)
+r$likelihood <- as.character(r$likelihood)
+r$dataset[-seq(1,nrow(r),by=4)] <- ""
+r$likelihood[-seq(1,nrow(r),by=2)] <- ""
+#for (i in 4:ncol(r)) r[,i] <- round(r[,i],3)
+
+ r <- xtable(r,caption="Comparing mean loglikelihood for each event across methods for each dataset.  Larger values are better.  See text for details. [TODO: More datasets as well as results for the DP version.  Could also remove training set results.]",label="tab:results",digits=3)
+print(r,include.rownames=FALSE,file=paste("figs/results.tex",sep=""),NA.string="",table.placement="t",size="footnotesize")

### 2012-03-23 docs/submission.tex
Put report for Smyth in NIPS format and submit.

--- 2012-03-23 docs/notes.tex
@@
-\pagestyle{empty}
-\documentclass[11pt]{article}
+\documentclass{article}
+\usepackage{nips12submit_e,times}
 \usepackage{tikz}
 \tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-
-%\usepackage{tikz,fullpage}
-%\usetikzlibrary{arrows,petri,topaths}%
-%\usepackage{tkz-berge}
 \usepackage{verbatim}
@@
 \usepackage{subfigure}
-%\usepackage[small,compact]{titlesec} 
-%\usepackage[small,it]{caption}
 
-\textwidth 6.5in
-\headheight .1in
-\topmargin -.2 in
-\textheight 8.7in
-\oddsidemargin 0in
-\evensidemargin 1in
+\title{Stochastic Blockmodels for Relational Event Data}
+
+\author{
+David S.~Hippocampus\thanks{ Use footnote for providing further information
+about author (webpage, alternative address)---\emph{not} for acknowledging
+funding agencies.} \\
+Department of Computer Science\\
+Cranberry-Lemon University\\
+Pittsburgh, PA 15213 \\
+\texttt{hippo@cs.cranberry-lemon.edu} \\
+\And
+Coauthor \\
+Affiliation \\
+Address \\
+\texttt{email} \\
+\AND
+Coauthor \\
+Affiliation \\
+Address \\
+}
 
@@
 
- \centerline{\Large \bf Stochastic Blockmodels for Relational Event Data} 
- \medskip
-\centerline{\bf Christopher DuBois}
- \bigskip
+\maketitle
 
-\abstract{
+\begin{abstract}
 Observing network data as a stream of events provides an opportunity to understand differences in how nodes behave given the past.
-Recent methods for continuous-time network data model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.
-Methods for static networks, such as stochastic blockmodels, often account for actor-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
+Recent methods for continuous-time network data model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
+Methods for static networks, such as stochastic blockmodels, often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
 We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
@@
 Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-
-%Stochastic blockmodels are a tool for exploring the structure of static social networks by learning latent groups of individuals who have similar tendencies in their  group-wise interactions.
-Continuous-time methods, such as the relational event framework (Butts 2008), model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.
-We propose a hierarchical model combining these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.
-The  method is illustrated using  dyadic interaction data (e.g.
-email, Twitter) among hundreds of actors and thousands of events.
-We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than a single relational event model for all actors in the network.
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-
-% \begin{itemize}
-% \item Blockmodels and clustering algorithms often applied to static networks for exploratory data analysis
-% \item continuous time methods allow one to predict future events given the previous history
-% \item we combine these two ideas with a hierarchical model such that each pair of blocks has a set of parameters guiding the event dynamics for those dyads
-% \item prediction experiments on real datasets
-% \item show that the parameter estimates are interpretable
-% \end{itemize}
-
-}
+\end{abstract}
 
@@
 Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.
-Finally we show that the parameter estimates provide an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication that exist.
+Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
 
@@
 In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ has an intensity function $\lambda_{ij}(t)$ that is piecewise constant.
+We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t)$ that is piecewise constant.
 We can then write the likelihood as
@@
 \end{align}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{i,j,m}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs.
+\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs.
 \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
@@
 Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.\footnotemark }
+Note $\lambda_{34}$ can only change when events occur that involve either node 3 or node 4.\footnotemark }
 \label{fig:example}
@@
 For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$.
-Though this will be useful, we will constrain the types of dependencies so that an interaction between one dyad does not affect the rate of an entirely separate dyad.
-As shown in Figure \ref{fig:example}, we do this by assuming each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
+In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
+As shown in Figure \ref{fig:example}, we incorporate this idea by assuming each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
 This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
 
-We hope to learn about the dynamics between subsets of nodes while allowing for variety among disjoint subsets.
-To facilitate this, we assume node $i$ is assigned a latent cluster $z_i$ and the intensity functions via a log linear model
+We hope to learn about the dynamics between subsets of nodes.   To facilitate this, we assume node $i$ is assigned a latent cluster $z_i$ and the intensity functions via a log linear model
 \begin{align*}
@@
 \end{align*}
-where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}(t,i,j|\cdot)$.
+where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the vector of statistics $\mathbf{s}(t,i,j|\cdot)$.
 Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-To allow for the blocks to share information by placing a hierarchical prior on each $\boldsymbol{\beta}_{k.l}$.
+We allow  the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$.
 
@@
 % \item Receiver in-degree: $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$
-% \item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-% \item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-% \item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-% \item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
-% \item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
+% \item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
+% \item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
+% \item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
+% \item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
+% \item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
 % %\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
-% % \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
-% % \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
-% % \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
+% % \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}=r)$
+% % \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=r,j_{v_{mij}}=j)$
+% % \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=r,j_{v_{mij}}=r)$
 % \end{enumerate}
@@
 \hline
-Intercept:& $s_{0}(t,i,j) = 1$\\
-Dyad count:& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender out-degree:& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree:& $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Receiver out-degree:& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-Receiver in-degree:& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-Reciprocity (AB-BA):& $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$\\
-Turn-taking (AB-BY): &$s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}\ne i)$\\
-Turn-usurping (AB-XA):& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}} \ne j,j_{v_{ijm}}=i)$\\
-Turn-usurping (AB-XB): &$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}} \ne i,j_{v_{ijm}}=j)$\\
-Turn-continuing (AB-AY):& $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}\ne j)$\\
+Intercept& $s_{0}(t,i,j) = 1$\\
+Dyad count& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
+Sender out-degree& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
+Sender in-degree& $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
+Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
+Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
+Reciprocity (AB-BA)& $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
+Turn-taking (AB-BY)&$s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
+Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
+Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
+Turn-continuing (AB-AY)& $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
 \hline
@@
 For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.
-Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
+Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
 As these intensities do not change under this specification, the likelihood simplifies to
@@
 
-Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood becomes a conditional logit
-$$\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.$$
+Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
+\begin{align}
+\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
+\label{eqn:multllk}
+\end{align}
 The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
-\subsection{Related work}
-
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant and allow for a nonlinear relationship between statistics and intensity functions.
-
-(TODO: More here.  For example: \cite{Tillman2011} explores the idea of having nonpoisson processes at each edge and solves the master equation for the dynamical system.)
 
@@
 \begin{align*}
-p(z_i | z_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto p(z_i | z_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}).% & z_i &= \mbox{new} = \alpha
+p(z_i | z_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) &\propto p(z_i | z_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}) \\
+p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
+\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
 \end{align*}
-Letting $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ be the set of events where node $r$ is involved, the likelihood term above can be written as
-$$p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) = \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
-\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}$$
-
-\noindent Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$. 
+where $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ is the set of events where node $r$ is involved.
+Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
 % \begin{align*}
@@
 
-\subsection*{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
+\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
 
@@
 p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \mu, \sigma) &\propto p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\
-p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)
+p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
+p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
+\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
 \end{align*}
-the likelihood term above can be written as
-$$p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) = \prod_{a \in \mathcal{V}_{k,l}} \lambda_{i_a,j_a}(t_a|\cdot)
-\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}$$
-\noindent where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \mbox{or} z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.
+where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \ \mbox{or} \ z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.
 \footnote{TODO: If we end up using HMC, mention step size and step length used.}
-The above distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}
+This distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}
 
-\subsection*{Split-merge algorithm}
+\subsection{Split-merge algorithm}
 
-Motivation: simply Gibbs sampling does not seem to mix well.  \footnote{This section is planned future work.}
-
+Motivation: simply Gibbs sampling does not seem to mix well.
+\footnote{This section is planned future work.
 In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.
+For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
 
@@
 \includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2.25in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 1000 simulated events, as described in text.
+\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
+\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
+\caption{Illustration of 2000 simulated events, as described in text.
 Left: Counts of each dyad.
@@
 
-In Figure \ref{fig:syncounts} we illustrate the simulated data.
-While the dyad counts suggest a stochastic blockmodel, the center plot shows each block has empirical differences in their dynamics: intensities for reciprocal actions among actors in block 1 are $e^3$ times greater, intensities for turn-taking actions among actors in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
+Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
+Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
 Fitting the model with $K=2$ converges to the true loglikelihood (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.
@@
 
-\small{
-\input{../figs/results.tex}
-}
-
-\subsection*{Data}
+\subsection{Data}
 
@@
 \begin{itemize}
-%\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
-\item Eckmann email:
-\item Kiel email: Still need to hunt this down.
+\item Kiel email: \cite{Ebel2002}
 \item Enron email:
@@
 
-\subsection*{Prediction experiment}
+\subsection{Prediction experiments}
 
-\begin{figure}[th]
-\caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-\label{fig:recall}
-\end{figure}
-
-%\begin{figure}
-%\center
-%\includegraphics[width=3in]{../figs/synthetic/logposterior.pdf}
-% \includegraphics[width=3in]{../figs/synthetic/test-recall.pdf}
-% \caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
-% \label{fig:synresults}
-% \end{figure}
 We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.
 Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
+I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}   We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk} .
+The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
 
@@
 At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
+
+\input{../figs/results.tex}
+
+\subsection{Baselines}
 
@@
 
-%We consider several variations of our model to investigate its properties.
-The \texttt{baserates} model only is an intercept only model (and mimics the behavior of a stochastic block model).
-The \texttt{full} model uses all the statistics discussed in Section \ref{sec:modeladditional}.
-For the \texttt{shared} model we force all diagonal blocks to use the same vector ($\boldsymbol{\beta}_{k_1,k_2} = \boldsymbol{\gamma}_1$ where $k_1=k_2$) and all non-diagonal blocks to use the same vector to be the same  ($\boldsymbol{\beta}_{k_1,k_2} = \boldsymbol{\gamma}_2$ where $k_1 \ne k_2$).
-For each of these we may vary the number of clusters $K$.
-The \texttt{single} model uses all statistics but assumes a single cluster $K=1$.
+\begin{figure}[t]
+\center
+\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
+\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
+\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
+\caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
+\label{fig:recall}
+\end{figure}
+
+\subsection{Results}
 
  Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
-Two likelihoods are considered: 1) \texttt{rem} (relational event model) likelihood as given in Equation \ref{eqn:llk}, and \texttt{mult} (multinomial likelihood) given in Equation \ref{eqn:multllk}.
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
-
 For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
@@
 
-Figure \ref{fig:recall} provides recall curves for several of the datasets.
+Figure \ref{fig:recall} will provide recall curves for several of the datasets.
 
@@
 
-\begin{figure}[ht]
+%% TODO: Discuss scalability.
+%% TODO: Describe and include posterior predictive checks.
+
+\begin{figure}[t]
 \includegraphics[width=3in]{../figs/eckmann-small/params-estimates}
+\fbox{\rule[.5cm]{0cm}{5cm} \rule[.5cm]{5cm}{0cm}}
 \caption{Left: Posterior credible intervals for parameter estimates from fitting the $K=2$ model to the Eckmann subset.
@@
  We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models) to provide detailed models of event dynamics among subsets of nodes while simultaneously allowing for heterogeneity in dynamics of the network as a whole.
-
-%By combining stochastic blockmodels with a relational event framework, the proposed method accounts for unobserved heterogeneity in the way dyadic events occur conditioned on the past.
+The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models) to provide detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in dynamics of the network as a whole.
 
@@
 Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
+For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
+Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
 
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods with respect to ranking tasks and the likelihood of unobserved data.
+This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share.
+Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j)$, and similarly one use our method to study how their role varies across nodes.
 
-TODO: Discuss experimental results.  Describe qualitative behavior of each model's recall line.
+In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
+We also note that we improve over the situation where we constrain $K=1$.
 
-TODO: Describe and include posterior predictive checks.
+Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
+By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
 
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j)$.
-
-Our method is able to reveal heterogeneity that exists among individuals.  
-The purpose of our method is to learn about event dynamics within latent groups of individuals.
-Other types of heterogeneity likely exist in some data sets.
+Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.
 For example, dynamics might change over time \cite{Vu2011}.
@@
 
-% \subsection{Extensions}
-% Possible future work could extend the present framework to a mixed-membership model where for each effect there is a latent cluster assignment.
-This way a group of nodes might contact high in-degree nodes at a similar rate, though they vary in the degree to which they adhere to reciprocity.
-As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.
-In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.
-One could extend our current framework by assuming that for each event $m$, any node  $i,j \in \mathcal{R}_{i_m,j_m}$ redraws their latent class.
-TODO: Discuss why we *didn't* do this.
+\subsubsection*{Acknowledgements}
 
-% Possible variants:
-% \begin{itemize}
-% \item Discuss possible variations on the prior structure of $\beta$: 1.
-Enforce $\beta_{k,l} = \beta_{l,k}$.
-2.
-Assume $\beta_{k,l,p} \sim \mbox{Normal}(\beta_p,\sigma_p^2)$.
-3.
-Place other priors that provide different kinds of regularization.
-% \item $\log \lambda_{ij} = (\pi_i \cdot \pi_j) \eta s(t,i,j)$
-% \item  HMM that guides which latent class over time?
-% \item  HDP instead of fixed K: Hierarchical, nonparametric model that allows parameters to be shared across blocks.
-% \item different clusters for senders and receivers (cite our KDD paper)?
-% \item More statistics, e.g. topic model similarity
-% \end{itemize}
-
-% \appendix
-% \section{Simulation}
-
-% Simulation from the model follows the typical racing exponentials idea with a few subtleties.
-
-% \begin{enumerate}
-% \item $t_0 = 0$, all $s_{ij} = 1$, Initialize $v = 0$
-% \item For $i = 1, \ldots, N$: Draw $z_i \sim \mbox{Categorical}(\theta)$
-% \item For $m = 1, \ldots, M$:
-%   \begin{enumerate}
-%   \item Draw $t_m \sim \mbox{Exponential}\left(\sum_{ij} \lambda_{ij}^*\right)$
-%   \item Draw $(i_m,j_m) \sim \mbox{Categorical}\left(\lambda_{ij}^* / \sum_{ij}\lambda_{ij}^*\right)$
-%   % \item Save $s_{ij}(t_m) \leftarrow s_{ij}^*$
-%   % \item Save $\lambda_{ij}(t_m) \leftarrow \lambda_{ij}^*$
-%   \item For all $(i,j)$ where $i$ or $j$ is in $\{i_m,j_m\}$:
-%     \begin{enumerate}
-%     \item Compute $s_{ij}^*$ using event $0$ through $v_{ij}$
-%     \item Set $\lambda_{ij}^* \leftarrow \exp\{ \beta_{z_i,z_j} s_{ij}^*\}$ 
-%     \end{enumerate}
-%   \end{enumerate}
-% \end{enumerate}
-
-% \section{EM for baseline}
-
-% In our baseline model, each block $(a,b) \in [1,K] \times [1,K]$ simply has a rate $\gamma_{a,b}$.
-% The likelihood is
-% \begin{align*}
-% \mathcal{L}(A|\lambda) &= \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M\lambda_{ij}\} \\
-% & =\prod_{a=1}^K \prod_{b=1}^K \prod_{m=1}^M \gamma_{ab}^{I(z_{i_m}=a,z_{j_m}=b)} \prod_{(i,j): z_i=a,z_j=b} \exp\{-t_M\gamma_{ab}\} \\
-% & =\prod_{a=1}^K\prod_{b=1}^K \gamma_{ab}^{n_{ab}} \exp\{-t_Mn_an_b\gamma_{ab}\}
-% \end{align*}
-% where $n_{ab} = \sum_m I(z_{i_m}=a,z_{j_m}=b)$, $n_a = \sum_{i=1}^N I(z_i=a)$.
-
-% The loglikelihood is $$\ell(A|\lambda) = \sum_{a=1}^K \sum_{b=1}^K n_{ab} \log \gamma_{ab} - t_M n_a n_b \gamma_{ab}$$
-% and so $\hat{\gamma}_{ab} = \frac{n_{ab}}{t_Mn_an_b}$.
-
-% I do not believe it is possible to perform EM for a model where we have the more general linear predictor for the intensity function.
-Since the assignment of one node depends on the assignments of the other nodes, we suffer combinatorial explosion.
-
-% \section{Gradient}
-
-% \begin{align*}
-% \frac{\partial \ell}{\partial \beta_{p,k,l}} &= \frac{\partial}{\partial \beta_{p,k,l}} \sum_{m = 1}^M \left[ \log \lambda_{i_m,j_m}(t_m | \cdot) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) \lambda_{i,j}(t_m|\cdot) \right] \\
-%   &=\sum_{m \in V} \left[ \frac{\partial}{\partial \beta_{p,k,l}} \sum_p \beta_{p,k,l} s_p(t,i,j|\mathcal{A}_t)  - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) \frac{\partial}{\partial \beta_{p,k,l}} \exp \left\{\sum_p \beta_{p,k,l} s_p(t,i,j|\mathcal{A}_t) \right\} \right] \\
-%   &=\sum_{m \in V} \left[ s_p(t_m,i_m,j_m) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) s_p(t_m,i,j) \lambda_{i,j}(t_m|\cdot) \right]
-% \end{align*}
-
-% where $V$ is the set of $m$ where either $z_i \in \{k,l\}$ or $z_j \in \{k,l\}$.
-
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 \bibliography{refs}

### 2012-03-30 docs/submission.tex
Add notes on split-merge sampler.

@@
 
+IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
+A proposed split has REM parameters that are only slightly different than those for the original block.
+Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
+By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
+
+This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
+We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg.
+8).
+
+First create an empty cluster whose parameters will be used as auxiliary variables.
+If we have $K$ unempty clusters, let $l = K+1$ and for $r \in [1,K]$ let $\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
+
+Next pick two nodes $i,j$ uniformly.
+If they belong to the same block $k$ (i.e.
+$z_i=z_j=k$), then we will call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as follows:
+\begin{itemize}
+\item For $p \in [1,P], r \in [1,K]$: $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
+\item Perform a sequential Gibbs scan for $\mathbf{z}^{split}$ for all $a : z_a = k$: $$p(z_{a}^{split}=k|\cdot)  \propto \sum_{-a}I(z_a=k) p(A|z,\phi)$$
+and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
+\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $\alpha^*$, defined below.
+\end{itemize}
+
+  If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
+$z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
+\begin{itemize}
+\item For each $p \in [1,P]$ draw $\phi_{kkp}^{merge}$ uniformly from $\{\phi_{kkp}^{split}, \phi_{llp}^{split}\}$ and for all other blocks $r \in [1,K] \setminus \{k,l\}$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$ and $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$.
+%\item Let $l$ be the index of an empty cluster, if one exists; otherwise, let $l=K+1$.
+For $r \in [1,K]$ let $\phi_{l,r}^{split} \sim G_0$, $\phi_{r,l}^{split} \sim G_0$, and $\phi_{l,l}^{split} \sim G_0$
+\item Accept $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $1/\alpha^*$, defined below.
+\end{itemize}
+
+The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})}\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}\frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
+
+The first two terms represent the prior.
+For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.
+We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.
+For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.
+$ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
+
 % TODO: Update the following discussion about sampling.

### 2012-03-30 docs/submission.tex
Update split merge description.

--- 2012-03-23 docs/submission.tex
@@
 
+IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
+A proposed split has REM parameters that are only slightly different than those for the original block.
+Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
+By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
+
+This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
+We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg.
+8).
+
+%$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
+
+Next pick two nodes $i,j$ uniformly.
+If they belong to the same block $k$ (i.e.
+$z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
+\begin{itemize}
+%\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+\item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
+\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
+\item Let $S$ be all nodes assigned to $k$.
+Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
+where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
+\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
+\end{itemize}
+
+  If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
+$z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
+\begin{itemize}
+\item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$
+\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+\item Let $S$ be the set of nodes assigned to $k$ or $l$.
+Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
+\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
+\end{itemize}
+
+The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})}\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}\frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
+
+The first two terms represent the prior.
+For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.
+We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.
+For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.
+$ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
+We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.
+One can return to the previous state with positive probability...
+
 % TODO: Update the following discussion about sampling.

### 2012-03-30 pkg/R/brem.r

@@ -4,32 +4,12 @@
 #' @z latent class assignments
 #' @beta P x K x K array of parameters
 #' 
-<<<<<<< HEAD
-generate.brem <- function(M,N,z,beta,nsim) {
-  param <- do.call(rbind)
-  fit <- list(param=0)
-=======
 generate.brem <- function(M,N,beta,z) {
   edgelist <- matrix(c(0,1,2),nr=1)
   simulate.brem.cond(beta,z,edgelist,M,N)
->>>>>>> fixgradient
 }
-
 simulate.brem <- function(fit,train,M=100,nsim=1) {
   mclapply(1:nsim,function(i) {
-<<<<<<< HEAD
-    beta <- fit$param[fit$niter - i,,,]
-    simulate.brem.cond(beta,train,M)
-  })
-}
-simulate.brem.cond <- function(beta,train,M=100) {
-  edgelist <- train[nrow(train),,drop=FALSE]
-  time <- edgelist[1]
-  s <- InitializeStatisticsArray(N,P);
-  for (m in 2:nrow(train)) {
-    i <- train[m-1,2]
-    j <- train[m-1,3]
-=======
     beta <- fit$param[fit$niter - i + 1,,,]
     z <- fit$zs[[fit$niter - i + 1]]
     simulate.brem.cond(beta,z,train,M,fit$N)
@@ -44,7 +24,6 @@ simulate.brem.cond <- function(beta,z,train,M=100,N) {
   for (m in 1:nrow(train)) {
     i <- train[m,2]
     j <- train[m,3]
->>>>>>> fixgradient
     s <- UpdateStatisticsArray(s,m-1,i-1,j-1,N,P)
   }
   lambda <- matrix(0,N,N)
@@ -67,10 +46,7 @@ simulate.brem.cond <- function(beta,z,train,M=100,N) {
     j <- cells[drawcell,2]
     time <- time + rexp(1,sum(cells[,3]))
     edgelist <- rbind(edgelist,c(time,i,j))
-<<<<<<< HEAD
-=======
 
->>>>>>> fixgradient
     s <- UpdateStatisticsArray(s,m-1,i-1,j-1,N,P)
   }
   dimnames(edgelist) <- NULL

### 2012-03-30 pkg/R/samplers.r

@@ -66,7 +66,6 @@ hmc <- function (current_q, lposterior, epsilon, L)
     }  
 }
 
-# Requires A, s, priors, z, px, k1, and k2 to be in environment
 lposterior <- function(A,s,beta,z,px,k1,k2,priors=list(beta=list(mu=0,sigma=1)),lp=TRUE,lgrad=FALSE) { 
   if (lp) {
     res <- block(A,s,beta,z,k1,k2,px,priors)

### 2012-03-31 docs/submission.tex
Typo in split-merge description.
debugging splitmerge

@@
 \begin{abstract}
-Observing network data as a stream of events provides an opportunity to understand differences in how nodes behave given the past.
-Recent methods for continuous-time network data model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
-Methods for static networks, such as stochastic blockmodels, often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
+Observing network data as a stream of relational events provides an
+opportunity to understand differences in how nodes interact given the
+past.
+For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
+For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
 We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
-The method is illustrated using dyadic interaction data -- e.g.
-email, Twitter direct messages.
-We show parameter estimates from the model reveal heterogeneity in the dynamics among groups of individuals.
-The fitted models have better predictive accuracy than both baselines and relational event models without the latent structure.
+The method is illustrated using dyadic interaction data such as email and Twitter direct messages.
+Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
+The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.
 Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
@@
 Network data, however, is often collected as a sequence of events occurring over time.
-Recently work leverages  continuous time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
+Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
 These models allow one to specify how the process depends on the previous history of events.
@@
 In many cases this is unrealistic.
-For example, in a small organization comprised of several teams, each team may exhibit different patterns for collaborating via email --  individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
+For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --  individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
 %, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
@@
 \end{align}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs.
+\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
 \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
@@
 In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
-As shown in Figure \ref{fig:example}, we incorporate this idea by assuming each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.
+As shown in Figure \ref{fig:example}, we incorporate this idea by assuming each intensity function $\lambda_{ij}(t)$ can only change following an event involving either $i$ or $j$.
 This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
 
-We hope to learn about the dynamics between subsets of nodes.   To facilitate this, we assume node $i$ is assigned a latent cluster $z_i$ and the intensity functions via a log linear model
+We hope to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
 \begin{align*}
@@
 z_i &\sim \mbox{CRP}(\alpha) &
-\beta_{k,l,p} &\sim \mbox{N}(\mu_p,\sigma_p^2)
+\boldsymbol{\beta}_{k,l} &\sim \mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
 \end{align*}
@@
 Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-We allow  the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$.
+We allow  the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p \sim \mbox{Gamma}(3,1)$ (TODO).
 
@@
 If $(i,j)$ has never occurred, we set $v_{mij}=0$.
-Here we include degree effects to model possible rich-get-richer phenomenon.
+We include degree effects to model possible rich-get-richer phenomenon.
 For example, a particular node sending often may indicate they will continue to be a sender in the future.
@@
 
-We Gibbs sample the latent class assignments $\mathbf{z}$ from the conditional distribution
+We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
 \begin{align*}
-p(z_i | z_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) &\propto p(z_i | z_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}) \\
+p(z_i | \mathbf{z}_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) &\propto p(z_i | \mathbf{z}_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}) \\
 p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
@@
 
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot 4N)$.
-The factor of 4 is due to the number of dyads involving a single node.
+ By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
 We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.
@@
 \begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \mu, \sigma) &\propto p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\
+p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\
 p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
@@
 We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.
-One can return to the previous state with positive probability...
+One can return to the previous state with positive probability.
 
@@
 Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ converges to the true loglikelihood (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.
+Fitting the model with $K=2$ makes the same predictions as the true model (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.
 
@@
 
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions.
+Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann data.
 Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
@@
 
+\appendix{Split-merge updates}
+
+Motivation: simply Gibbs sampling does not seem to mix well.
+\footnote{This section is planned future work.
+In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
+For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
+
+IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
+A proposed split has REM parameters that are only slightly different than those for the original block.
+Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
+By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
+
+This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
+We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg.
+8).
+
+%$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
+
+Next pick two nodes $i,j$ uniformly.
+If they belong to the same block $k$ (i.e.
+$z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
+\begin{itemize}
+%\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+\item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
+\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
+\item Let $S$ be all nodes assigned to $k$.
+Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
+where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
+\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
+\end{itemize}
+
+  If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
+$z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
+\begin{itemize}
+\item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$
+\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+\item Let $S$ be the set of nodes assigned to $k$ or $l$.
+Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
+\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
+\end{itemize}
+
+The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})}\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}\frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
+
+The first two terms represent the prior.
+For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.
+We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.
+For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.
+$ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
+We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.
+One can return to the previous state with positive probability.
+
 \end{document}

### 2012-03-31 pkg/R/splitmerge.r

@@ -60,6 +60,8 @@ lnormalize <- function(ps) {
   ps <- exp(ps)
   return(ps/sum(ps))
 }
+
+
 gibbs_restricted <- function(phi,z,S,k,l,llk_node,ztrue=NULL) {
 
   # Randomly assign nodes that are currently either in k or l
@@ -76,13 +78,13 @@ gibbs_restricted <- function(phi,z,S,k,l,llk_node,ztrue=NULL) {
       z[a] <- j
       ys[a,j] <- llk_node(a,phi,z) + log(counts[j] + priors$alpha)
     }
-    ys[a,clusters] <- lnormalize(ys[a,clusters])
-    z[a] <- sample(clusters,1,prob=ys[a,clusters]) #rcatlog(ys[a,])
+    z[a] <- sample(clusters,1,prob=lnormalize(ys[a,clusters])) #rcatlog(ys[a,])
     if (!is.null(ztrue)) z[a] <- ztrue[a]
     counts[z[a]] <- counts[z[a]] + 1
   }
-  q <- sum(log(ys[cbind(S,z[S])]))
 #  browser()
+  ys[S,clusters] <- t(apply(ys[S,clusters],1,lnormalize))
+  q <- sum(log(ys[cbind(S,z[S])]))
   list(z=z,transition=q,ys=ys,z.init=z.init)
 }
 
@@ -103,7 +105,7 @@ pm2ps <- function(phi.merge,phi.split,k,l,sigma) {
   return(sum(ps))
 }
 
-splitmerge <- function(phi,z,lposterior,llk_node,priors) {
+splitmerge <- function(phi,z,lposterior,llk_node,priors,verbose=TRUE) {
   N <- length(z)
   K <- dim(phi)[2]
   ij <- sample(1:N,2,replace=FALSE)
@@ -116,22 +118,22 @@ splitmerge <- function(phi,z,lposterior,llk_node,priors) {
     phi$merge <- add_cluster(phi$init)
     phi$merge <- sample_cluster_from_prior(phi$merge,K+1,priors)
     z$merge <- z$init
-    phi$split <- split_phi(phi$merge,z$init[i],K+1,sigma)
+    phi$split <- split_phi(phi$merge,z$init[i],K+1,priors$sigma)
     h <- gibbs_restricted(phi$split,z$merge,S,z$init[i],K+1,llk_node)
     z$split <- h$z
     q$zm2zs <- h$transition
[…]

### 2012-03-31 pkg/inst/tests/test.gradient.r

@@ -1,0 +1,323 @@
+
+## context("gradient")
+## library(brem)
+## library(multicore)
+## source("pkg/R/brem.r")
+
+## set.seed(2)
+## M <- 5000
+## N <- 10
+## K <- 1
+## beta <- list("intercept"=matrix(-1,K,K),
+##              "abba" = matrix(1,K,K),
+##              "abby"=matrix(0,K,K),
+##              "abxa"=matrix(0,K,K),
+##              "abxb"=matrix(0,K,K),
+##              "abay"=matrix(1,K,K),
+##              "abab"=matrix(0,K,K),
+##              "sod"=matrix(0,K,K),
+##              "rod"=matrix(0,K,K),
+##              "sid"=matrix(0,K,K),
+##              "rid"=matrix(0,K,K),
+##              "dc"=matrix(0,K,K),
+##              "cc"=matrix(0,K,K))
+## z <- rep(1,N)#c(rep(1,N/2),rep(2,N/2))
+## P <- length(beta)
+## beta <- abind(beta,rev.along=3)
+## <<<<<<< HEAD
+## set.seed(1)
+## z <- c(rep(1,10))#,rep(2,5))
+
+## sim <- generate.brem(M,N,beta,z)
+## A <- sim$edgelist
+## px <- rep(1,13)
+
+## #' General purpose Metropolis-Hastings sampler
+## #' @param current
+## #' @param likelihood
+## #' @param olp
+## #' @param sd
+
+## #' @return ...
+## mh <- function(current,likelihood,olp=NULL,sd=.1) {
+##   if (is.null(olp)) {
+##     olp <- attr(current,"log.density") <- likelihood(current)
+##   }
+##   cand <- current + rnorm(length(current),0,sd)
+##   clp <- likelihood(cand)
+##   if (clp - olp > log(runif(1))) {
+##     current <- cand
+##     attr(current,"log.density") <- clp
+##   }
+##   return(current)
+## }
+
+## slice <- function(current,likelihood,olp=NULL,m=20) {
+##   uni.slice.alt(current,likelihood,gx0=olp,m=m)
+## }
+
[…]

### 2012-03-31 pkg/inst/tests/test.splitmerge.r

@@ -68,7 +68,7 @@ phi[1,,] <- rnorm(9,0,1)
 sim <- generate.brem(M,N,phi,z)
 table(sim$edgelist[,2],sim$edgelist[,3])
 
-A <- sim$edgelist
+edgelist <- A <- sim$edgelist
 px <- rep(0,13)
 px[1] <- 1
 s <- new(RemStat,A[,1],A[,2]-1,A[,3]-1,N,nrow(A),length(px))
@@ -93,6 +93,8 @@ llk_node_test<- function(a,phi,z) {
 }
 
 test_that("Gibbs sampling on small example gives true assignments",{
+
+  a <- 1
   truellk <- llk_node(1,phi,z)
   get.ys <- function(phi,z,llk_node) {
     z2 <- z
@@ -113,8 +115,11 @@ test_that("Gibbs sampling on small example gives true assignments",{
   expect_that(apply(ys,1,which.max), equals(z))
 })
 
-g <- gibbs_restricted(phi,z,2,3,llk_node_test)
-g
+test_that("restricted gibbs scan finds correct assignments",{
+  S <- 4:9
+  g <- gibbs_restricted(phi,z,S,2,3,llk_node_test)
+  expect_that(all(g$z==z), is_true())
+})
 
 lposterior_test <- function(phi,z,priors) {
   llk <- llk_node_test(1,phi,z)
@@ -125,14 +130,15 @@ lposterior_test <- function(phi,z,priors) {
   return(llk + pr.phi + pr.z)
 }
 
-priors <- list(alpha=1,phi=list(mu=0,sigma=1))
+priors <- list(alpha=1,phi=list(mu=0,sigma=1),sigma=.1)
 lposterior_test(phi,z,priors)
 
 source("pkg/R/splitmerge.r")
-s <- splitmerge(phi,z,lposterior_test,llk_node_test,priors)
+s <- splitmerge(phi,z,lposterior_test,llk_node_test,priors,verbose=TRUE)
 
 truth <- list(phi=phi,z=z)
-
+phi <- truth$phi
+z <- truth$z
 
 # Still bugs in trying this...
 for (a in 1:3) {

### 2012-04-18 docs/submission.tex
Update split merge appendix.

@@
 One class of statistical models for network data uses latent variables to help model unobserved heterogeneity.
-For example, the stochastic blockmodel \cite{Nowicki2001, Kemp} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between nodes of each block.
+For example, the stochastic blockmodel \cite{Nowicki2001, Kemp, Ishiguro2010} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between nodes of each block.
 % Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
@@
 
-Motivation: simply Gibbs sampling does not seem to mix well.
+%Motivation: simply Gibbs sampling does not seem to mix well.
 \footnote{This section is planned future work.
@@
 For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
+
+Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
+\begin{align*}
+  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
+\end{align*}
 
@@
 
-This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
-We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg.
-8).
-
 %$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
-
-Next pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
-\begin{itemize}
-%\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
-\item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
-\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
-\item Let $S$ be all nodes assigned to $k$.
-Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
-where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
-  If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
-\begin{itemize}
-\item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$
-\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
-\item Let $S$ be the set of nodes assigned to $k$ or $l$.
-Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
-\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
-The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})}\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}\frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
-
-The first two terms represent the prior.
-For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.
-We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.
-For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.
-$ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
-We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.
-One can return to the previous state with positive probability.
 
@@
 Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ makes the same predictions as the true model (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.
+Fitting the model with $K=2$ makes the same predictions as the true model (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
 
@@
 
+Trivially extended to co-clustering applications and for directed data.  Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
+
 Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.
@@
 
-\appendix{Split-merge updates}
+\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
 
-Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
+This is a two-stage MH procedure.
+First, pick two nodes $i,j$ uniformly.
+If they belong to the same block $k$ (i.e.
+$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ with new cluster $l$:
+\begin{itemize}
+\item For blocks $r \ne l$, sample $\phi_{lr}^{split} \sim N(\phi_{kr}^{merge},\sigma^2)$ and $\phi_{rl}^{split} \sim N(\phi_{rk}^{merge},\sigma^2)$ and $\phi_{ll}^{split} \sim N(\phi_{kk}^{merge}, \sigma^2)$.
+\item For all $a$ such that $z_a^{merge} = k$, randomly initialize $z_a^{split}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{split}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{split}_{-a},\boldsymbol{\phi}^{split})$
+where $n_k=\sum_{i}I(z_i^{split}=k)$.% and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
+\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
+\end{itemize}
 
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
+ If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
+$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
+\begin{itemize}
+\item For all blocks $r \ne k,l$: sample $\phi_{kr}^{merge} \sim \mbox{Unif}(\{\phi_{kr}^{split}, \phi_{lr}^{split}\})$, and $\phi_{rk}^{merge} \sim \mbox{Unif}(\{\phi_{rk}^{split}, \phi_{rl}^{split}\})$.%, and sample $\phi_{lr}^{merge},\phi_{rl}^{merge}$, and $\phi_{ll}^{merge}$ independently from the prior.
+\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
+\end{itemize}
+
+ The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(Y|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
+
+The first term is the likelihood ratio of the two states.
+Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{split}$, the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan from  $\mathbf{z}^{merge}$ is computed as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}^{merge}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{split})$.
+Since there is only one way of combining two clusters we know $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$.
+The transition $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ is a product of proposal densities and
+$q(\boldsymbol{\phi}^{split} \rightarrow
+\boldsymbol{\phi}^{merge}) = \frac{1}{2}\prod_{r \ne k,l}\left(\frac{1}{2}\right)^2$.
+\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{merge})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
+
+ \appendix{\textbf{Split-merge Metropolis-Hastings Move for Relational Event Blockmodel}}
 
@@
 \item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
-\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
+%\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
 \item Let $S$ be all nodes assigned to $k$.
@@
 \begin{itemize}
-\item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$
-\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+\item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$, and sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
+%\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
 \item Let $S$ be the set of nodes assigned to $k$ or $l$.
@@
 
-The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})}\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}\frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
+The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
 
-The first two terms represent the prior.
+For the proposed merge step,
+$$q(\boldsymbol{\phi}^{split} \rightarrow
+\boldsymbol{\phi}^{merge}) = \prod_{q,r \ne l}\frac{1}{2} \prod_r p(\phi_{r,l,p}|\mu_p,\sigma_p^2)p(\phi_{l,r,p}|\mu_p,\sigma_p^2) p(\phi_{l,l,p}|\mu_p,\sigma_p^2)$$ though the latter product cancels with a corresponding term in $p(\boldsymbol{\phi}^{merge})$.
+We can compute the transition $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.
+For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.
+$ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
+
+% Motivation: simply Gibbs sampling does not seem to mix well.
+\footnote{This section is planned future work.
+In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
+For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
+
+% IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
+A proposed split has REM parameters that are only slightly different than those for the original block.
+Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
+By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
+
+% This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
+We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg.
+8).
+
+% %$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
+
+% Next pick two nodes $i,j$ uniformly.
+If they belong to the same block $k$ (i.e.
+$z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
+% \begin{itemize}
+% %\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+% \item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
+% \item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
+% \item Let $S$ be all nodes assigned to $k$.
+Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
+% where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
+% \item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
+% \end{itemize}
+
+%   If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
+$z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
+% \begin{itemize}
+% \item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$
+% %\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+% \item Let $S$ be the set of nodes assigned to $k$ or $l$.
+Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
+% \item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
+% \end{itemize}
+
+% The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})}\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}\frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
+
+% The first two terms represent the prior.
 For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.

### 2012-04-27 docs/submission.tex
Update split merge algorithm to new idea.

@@
 % Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
+% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
 
@@
 If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ with new cluster $l$:
+$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
 \begin{itemize}
-\item For blocks $r \ne l$, sample $\phi_{lr}^{split} \sim N(\phi_{kr}^{merge},\sigma^2)$ and $\phi_{rl}^{split} \sim N(\phi_{rk}^{merge},\sigma^2)$ and $\phi_{ll}^{split} \sim N(\phi_{kk}^{merge}, \sigma^2)$.
-\item For all $a$ such that $z_a^{merge} = k$, randomly initialize $z_a^{split}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{split}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{split}_{-a},\boldsymbol{\phi}^{split})$
-where $n_k=\sum_{i}I(z_i^{split}=k)$.% and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
+\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
+\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
+where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
+\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
 \end{itemize}
@@
  If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
+$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
 \begin{itemize}
-\item For all blocks $r \ne k,l$: sample $\phi_{kr}^{merge} \sim \mbox{Unif}(\{\phi_{kr}^{split}, \phi_{lr}^{split}\})$, and $\phi_{rk}^{merge} \sim \mbox{Unif}(\{\phi_{rk}^{split}, \phi_{rl}^{split}\})$.%, and sample $\phi_{lr}^{merge},\phi_{rl}^{merge}$, and $\phi_{ll}^{merge}$ independently from the prior.
-\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
+\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
+  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
+\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
+\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
 \end{itemize}
 
- The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(Y|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
+ The Metropolis-Hastings acceptance probabilities require us to compute 
+$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
 
 The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{split}$, the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan from  $\mathbf{z}^{merge}$ is computed as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}^{merge}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{split})$.
-Since there is only one way of combining two clusters we know $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$.
-The transition $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ is a product of proposal densities and
-$q(\boldsymbol{\phi}^{split} \rightarrow
-\boldsymbol{\phi}^{merge}) = \frac{1}{2}\prod_{r \ne k,l}\left(\frac{1}{2}\right)^2$.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{merge})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
+Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
+Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
+The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
+$q(\boldsymbol{\phi}^{s} \rightarrow
+\boldsymbol{\phi}^{m})$ are products of Normal densities.
+\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
 
- \appendix{\textbf{Split-merge Metropolis-Hastings Move for Relational Event Blockmodel}}
+%  \appendix{\textbf{Split-merge Metropolis-Hastings Move for Relational Event Blockmodel}}
 
-This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
+% This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
 We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg.
@@
 
-%$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
+% %$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
 
-Next pick two nodes $i,j$ uniformly.
+% Next pick two nodes $i,j$ uniformly.
 If they belong to the same block $k$ (i.e.
 $z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
-\begin{itemize}
-%\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
-\item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
-%\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
-\item Let $S$ be all nodes assigned to $k$.
+% \begin{itemize}
+% %\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+% \item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
+% %\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
+% \item Let $S$ be all nodes assigned to $k$.
 Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
-where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
+% where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
+% \item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
+% \end{itemize}
 
-  If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
+%   If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
 $z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
-\begin{itemize}
-\item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$, and sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
-%\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
-\item Let $S$ be the set of nodes assigned to $k$ or $l$.
+% \begin{itemize}
+% \item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$, and sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
+% %\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
+% \item Let $S$ be the set of nodes assigned to $k$ or $l$.
 Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
-\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
+% \item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
+% \end{itemize}
 
-The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
+% The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
 
-For the proposed merge step,
-$$q(\boldsymbol{\phi}^{split} \rightarrow
-\boldsymbol{\phi}^{merge}) = \prod_{q,r \ne l}\frac{1}{2} \prod_r p(\phi_{r,l,p}|\mu_p,\sigma_p^2)p(\phi_{l,r,p}|\mu_p,\sigma_p^2) p(\phi_{l,l,p}|\mu_p,\sigma_p^2)$$ though the latter product cancels with a corresponding term in $p(\boldsymbol{\phi}^{merge})$.
+% For the proposed merge step,
+% $$q(\boldsymbol{\phi}^{split} \rightarrow
+% \boldsymbol{\phi}^{merge}) = \prod_{q,r \ne l}\frac{1}{2} \prod_r p(\phi_{r,l,p}|\mu_p,\sigma_p^2)p(\phi_{l,r,p}|\mu_p,\sigma_p^2) p(\phi_{l,l,p}|\mu_p,\sigma_p^2)$$ though the latter product cancels with a corresponding term in $p(\boldsymbol{\phi}^{merge})$.
 We can compute the transition $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.

### 2012-05-02 docs/submission.tex
Edits to split/merge description.

@@
 
+Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
+This increases the acceptance probability for splits (and decreases the probability of merges).
+However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
+
 %  \appendix{\textbf{Split-merge Metropolis-Hastings Move for Relational Event Blockmodel}}

### 2012-05-24 docs/submission.tex
Small changes to slides and tex draft.
Work in Model section in draft.

@@
 \begin{abstract}
-Observing network data as a stream of relational events provides an
-opportunity to understand differences in how nodes interact given the
-past.
+Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.
 For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
@@
 %Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.  We can then write the likelihood of $M$ events as
+%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
+We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
+Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.
+If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
 \begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \\
-&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
+&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
 \end{align}
-\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
+%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
+
 % \begin{align}
@@
 In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t)$ that is piecewise constant.
-We can then write the likelihood as
+We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
+The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
 \begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | z_i,z_j) \}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
 \label{eqn:llk}
 \end{align}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
+%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
 
 \begin{figure}
-
 \centering 
@@
 
-For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$.
+%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
+
+We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
+\begin{align*}
+\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
+%z_i &\sim \mbox{CRP}(\alpha) &
+%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
+\end{align*}
+where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
+Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
+
  In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
-As shown in Figure \ref{fig:example}, we incorporate this idea by assuming each intensity function $\lambda_{ij}(t)$ can only change following an event involving either $i$ or $j$.
-This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
+As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event involving either $i$ or $j$.
+In the Appendix we show how this reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
 
-We hope to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
-\begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) &= \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t) &
-z_i &\sim \mbox{CRP}(\alpha) &
-\boldsymbol{\beta}_{k,l} &\sim \mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
-\end{align*}
-where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the vector of statistics $\mathbf{s}(t,i,j|\cdot)$.
-Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-We allow  the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p \sim \mbox{Gamma}(3,1)$ (TODO).
+We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Gamma}(3,1)$ (TODO).
 
@@
 In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred (n.b.
+Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.
+% (n.b.
 $t_{v_{mij}} = \tau_{mij}$).
@@
 We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
+% \begin{align*}
+% p(z_r | \mathbf{z}_{-r},\mathcal{A}_t,\alpha,\boldsymbol{\beta}) &\propto p(z_i | \mathbf{z}_{-i},\alpha) p(\mathcal{A}_t|\mathbf{z},\boldsymbol{\beta}) \\
+% p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{m=1}^M \lambda_{i_a,j_a}(t_a|\cdot)
+% \prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
+% \end{align*}
 \begin{align*}
-p(z_i | \mathbf{z}_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) &\propto p(z_i | \mathbf{z}_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}) \\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
-\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
+p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
+p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
+\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{I(r \in \{i_m,j_m\})}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
+%& \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t - t_{M}) \lambda_{ij}(t|\cdot)\}  
 \end{align*} 
-where $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ is the set of events where node $r$ is involved.
+where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
 Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
@@
 
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m,i,j$.
-
 \subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
@@
 \begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\
+p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\ %, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\
 p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
-\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
+p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)
+\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
+%p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
+%\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
 \end{align*}
-where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \ \mbox{or} \ z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.
-\footnote{TODO: If we end up using HMC, mention step size and step length used.}
-This distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}
+where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
+This distribution is sampled using slice sampling.
 
@@
 Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ makes the same predictions as the true model (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
+Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
 
@@
 
+\appendix{\textbf{Complexity}}
+
+We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood
+ Equation \ref{eqn:likelihood} as 
+\begin{align*}
+\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
+\end{align*}
+\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
+\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
+
+ By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
+We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
+
 \appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
@@
 
- The Metropolis-Hastings acceptance probabilities require us to compute 
+ The Metropolis-Hastings acceptance probabilities require the quantity
 $$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$

### 2012-05-27 docs/submission.tex
Minor changes.

@@
 Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.
-One class of statistical models for network data uses latent variables to help model unobserved heterogeneity.
-For example, the stochastic blockmodel \cite{Nowicki2001, Kemp, Ishiguro2010} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between nodes of each block.
+Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
 % Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
@@
 
-The above continuous time network models assume all nodes in the network behave according to identical dynamics.
+The above continuous-time network models assume all nodes in the network behave according to identical dynamics.
 In many cases this is unrealistic.
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --  individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
+For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
+
 %, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.

### 2012-05-27 dashboard.r

@@ -28,11 +28,13 @@ options(verbose=FALSE)
 library(brem)
 library(ggplot2)
 
+#opts <- list(dataset="synthetic-1",predictions=TRUE)
+
 # Pull data from each saved file and grab the name of the fit
 results.dir <- paste("results/",opts$dataset,sep="")
 load(paste("data/",opts$dataset,".rdata",sep=""))
 fs <- list.files(results.dir,full.names=TRUE)
-fs <- fs[-grep("ranks",fs)]
+fs <- fs[-grep("ranks",fs)]  # TODO: Bug if one of these not present
 fs <- fs[-grep("final",fs)]
 fs <- fs[-grep("llks",fs)]
 fs <- fs[-grep("counts",fs)]

### 2012-05-27 pkg/R/splitmerge.r

@@ -240,7 +240,7 @@ splitmerge <- function(phi,z,lposterior,llk_node,priors,sigma=.1,verbose=TRUE) {
 ##' @export
 ##' @return 
 ##' @author chris
-mcmc.blockmodel <- function(lposterior,llk_node,priors,N,P,K,niter=20,do.sm=TRUE,num.extra=0,verbose=FALSE,sigma=.1) {
+mcmc.blockmodel <- function(lposterior,llk_node,priors,N,P,K,niter=20,do.sm=TRUE,num.extra=0,verbose=FALSE,sigma.proposal=.1) {
   priors$sigma <- sigma
   phi <- array(0,c(P,K,K))
   phi[1,,] <- rnorm(K^2)

### 2012-05-27 run.sh

@@ -40,7 +40,8 @@
 
 # Run dashboard
 rsync -auvz dashboard.r duboisc@d1:/extra/duboisc0/blockrem/
-./parallel --sshlogin 8/d6 'cd /extra/duboisc0/blockrem;./dashboard.r -d {} -s TRUE --predictions TRUE' ::: "synthetic" "eckmann-small"
+./parallel --sshlogin 8/d6 'cd /extra/duboisc0/blockrem;./dashboard.r -d {} -s TRUE --predictions TRUE' ::: "synthetic-1"
+./parallel 'cd /extra/duboisc0/blockrem;./dashboard.r -d {} -s TRUE --predictions TRUE' ::: "synthetic-1"
 
 # Helper commands
 cd ~/Documents/blockrem/

### 2012-05-29 docs/submission.tex
Add Eckmann example figs to paper.

@@
 \usepackage{amsmath}
-\usepackage{subfigure}
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
 
-\begin{figure}
-\centering 
-\subfigure[Dynamic network data]
-{
-\begin{tikzpicture}[scale=.8,transform shape,thick]
-  \node[vertex] (n1) at (0,0)  {1};
-  \node[vertex] (n2) at (2,2)  {2};
-  \node[vertex] (n3) at (4,0) {3};
-  \node[vertex] (n4) at (2,-2) {4};
-\draw[-stealth] (.4,.4) -- (1.6,1.6);
-\draw[-stealth] (2.4,1.6) -- (3.6,.4);
-\node at (.7,1.1) {$t_1$};
-\node at (3.3,1.1) {$t_2$};
-\end{tikzpicture}
-}
-\subfigure[Intensities for two dyads]
-{
-\begin{tikzpicture}[scale=1,transform shape,thick]
-\draw[-stealth] (0,0) -- (0,3);
-\draw[-stealth] (0,0) -- (5,0);
+% \begin{figure}
+% \centering 
+% \begin{subfigure}
+% \begin{tikzpicture}[scale=.8,transform shape,thick]
+%   \node[vertex] (n1) at (0,0)  {1};
+%   \node[vertex] (n2) at (2,2)  {2};
+%   \node[vertex] (n3) at (4,0) {3};
+%   \node[vertex] (n4) at (2,-2) {4};
+% \draw[-stealth] (.4,.4) -- (1.6,1.6);
+% \draw[-stealth] (2.4,1.6) -- (3.6,.4);
+% \node at (.7,1.1) {$t_1$};
+% \node at (3.3,1.1) {$t_2$};
+% \end{tikzpicture}
+% \caption{Dynamic network data}
+% \end{subfigure}
+% \begin{subfigure}
+% \begin{tikzpicture}[scale=1,transform shape,thick]
+% \draw[-stealth] (0,0) -- (0,3);
+% \draw[-stealth] (0,0) -- (5,0);
 
-% First timeline
-\draw (0,2) -- (1,2);
-\draw[fill=black] (1,2) circle (0.75mm);
-\draw (1,1.7) circle (0.75mm);
-\draw (1.08,1.7) -- (3,1.7);
-\draw[fill=black] (3,1.7) circle (0.75mm);
-\draw (3,1.7) circle (0.75mm);
-\draw (3,2.2) circle (0.75mm);
-\draw[-stealth] (3.05,2.2) -- (5,2.2);
+% % First timeline
+% \draw (0,2) -- (1,2);
+% \draw[fill=black] (1,2) circle (0.75mm);
+% \draw (1,1.7) circle (0.75mm);
+% \draw (1.08,1.7) -- (3,1.7);
+% \draw[fill=black] (3,1.7) circle (0.75mm);
+% \draw (3,1.7) circle (0.75mm);
+% \draw (3,2.2) circle (0.75mm);
+% \draw[-stealth] (3.05,2.2) -- (5,2.2);
 
-% second timeline
-\draw[dashed] (0,1) -- (3,1);
-\draw[fill=black] (3,1) circle (0.75mm);
-\draw (3,.75) circle (0.75mm);
-\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
+% % second timeline
+% \draw[dashed] (0,1) -- (3,1);
+% \draw[fill=black] (3,1) circle (0.75mm);
+% \draw (3,.75) circle (0.75mm);
+% \draw[dashed, -stealth] (3.08,.75) -- (5,.75);
 
-% labels
-\node at (-.5,2) {$\lambda_{1,2}$};
-\node at (-.5,1) {$\lambda_{3,4}$};
-\node at (1,-.3) {$t_1$};
-\node at (3,-.3) {$t_2$};
-\end{tikzpicture}
-}
-\caption[]{Illustration of event data and the assumptions of the model.
+% % labels
+% \node at (-.5,2) {$\lambda_{1,2}$};
+% \node at (-.5,1) {$\lambda_{3,4}$};
+% \node at (1,-.3) {$t_1$};
+% \node at (3,-.3) {$t_2$};
+% \end{tikzpicture}
+% \caption{Intensities for two dyads}
+% \end{subfigure}
+% \caption[]{Illustration of event data and the assumptions of the model.
 Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
@@
 Note $\lambda_{34}$ can only change when events occur that involve either node 3 or node 4.\footnotemark }
-\label{fig:example}
-\end{figure}
+% \label{fig:example}
+% \end{figure}
 
@@
 \end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
+where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
 Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
@@
 \begin{figure}[t]
-\includegraphics[width=3in]{../figs/eckmann-small/params-estimates}
-\fbox{\rule[.5cm]{0cm}{5cm} \rule[.5cm]{5cm}{0cm}}
-\caption{Left: Posterior credible intervals for parameter estimates from fitting the $K=2$ model to the Eckmann subset.
-Right: Posterior distribution of $K$ when fitting the model and allowing for flexible $K$.}
+\begin{subfigure}[b]{0.5\textwidth}
+\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
+%\caption{Observed counts}
+\end{subfigure}
+\qquad
+\begin{subfigure}[b]{0.5\textwidth}
+\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
+%\caption{Samples of $\beta_{p,k,l}$}
+\end{subfigure}
+
+%\fbox{\rule[.5cm]{0cm}{5cm} \rule[.5cm]{5cm}{0cm}}
+\caption{%Observed counts versus model estimates from Eckmann email data.
+ Left: Observed statistics across all dyadic email interactions from members of group 1 to group 3 (178 total events in total) and from members of group 3 to group 1 (179 events in total).
+Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
+In addition to finding differences in mean activity for these dyads, the model learns that events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
 \label{fig:posteriorparams}

### 2012-05-29 docs/submission.tex
Add matrix plots and caption.

@@
 \begin{figure}[t]
-\begin{subfigure}[b]{0.5\textwidth}
+\centering
+\begin{subfigure}[b]{0.45\textwidth}
+\centering
 \includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
@@
 \qquad
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[b]{0.45\textwidth}
+\centering
 \includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
@@
 
-%\fbox{\rule[.5cm]{0cm}{5cm} \rule[.5cm]{5cm}{0cm}}
 \caption{%Observed counts versus model estimates from Eckmann email data.
@@
 
+\begin{figure}[t]
+\centering
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
+\caption{Observed counts}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
+\caption{Intercept estimates}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
+\caption{\texttt{ab-ba} estimates}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
+\caption{\texttt{ab-by} estimates}
+\end{subfigure}
+\caption{Comparing observed counts and parameter estimates.
+Darker values are larger.
+Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
+Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics as seen in (c) and (d).}
+\label{fig:parmats}
+\end{figure}
+
  We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
@@
 Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
+%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
 Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.

### 2012-05-29 docs/submission.tex
Small changes to paper and todos.

@@
 
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Gamma}(3,1)$ (TODO).
+We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$ (TODO).
 
@@
 \item Kiel email: \cite{Ebel2002}
-\item Enron email:
+\item Classrooms: \cite{McFarland2003}
+\item Eckmann email: dyadic emails  \cite{Eckmann}
+\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
+\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
 \item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
@@
 We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: 
+\item MIT Reality Mining:  \cite{MITreality}
 \end{itemize}

### 2012-05-29 docs/submission.tex
Paper edits and cleanup.

@@
 \usepackage{subcaption}
+\usepackage{footnote}
 
@@
 \maketitle
+
+\savenotes
 
@@
 
-% \begin{figure}
-% \centering 
-% \begin{subfigure}
-% \begin{tikzpicture}[scale=.8,transform shape,thick]
-%   \node[vertex] (n1) at (0,0)  {1};
-%   \node[vertex] (n2) at (2,2)  {2};
-%   \node[vertex] (n3) at (4,0) {3};
-%   \node[vertex] (n4) at (2,-2) {4};
-% \draw[-stealth] (.4,.4) -- (1.6,1.6);
-% \draw[-stealth] (2.4,1.6) -- (3.6,.4);
-% \node at (.7,1.1) {$t_1$};
-% \node at (3.3,1.1) {$t_2$};
-% \end{tikzpicture}
-% \caption{Dynamic network data}
-% \end{subfigure}
-% \begin{subfigure}
-% \begin{tikzpicture}[scale=1,transform shape,thick]
-% \draw[-stealth] (0,0) -- (0,3);
-% \draw[-stealth] (0,0) -- (5,0);
+\begin{figure}
+\centering 
+\begin{minipage}[b]{0.45\linewidth}
+  \begin{tikzpicture}[scale=.8,transform shape,thick]
+    \node[vertex] (n1) at (0,0)  {1};
+    \node[vertex] (n2) at (2,2)  {2};
+    \node[vertex] (n3) at (4,0) {3};
+    \node[vertex] (n4) at (2,-2) {4};
+    \draw[-stealth] (.4,.4) -- (1.6,1.6);
+    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
+    \node at (.7,1.1) {$t_1$};
+    \node at (3.3,1.1) {$t_2$};
+  \end{tikzpicture}
+  \caption{Dynamic network data}
+\end{minipage}
+\begin{minipage}[b]{0.45\linewidth}
+\begin{tikzpicture}[scale=1,transform shape,thick]
+\draw[-stealth] (0,0) -- (0,3);
+\draw[-stealth] (0,0) -- (5,0);
 
-% % First timeline
-% \draw (0,2) -- (1,2);
-% \draw[fill=black] (1,2) circle (0.75mm);
-% \draw (1,1.7) circle (0.75mm);
-% \draw (1.08,1.7) -- (3,1.7);
-% \draw[fill=black] (3,1.7) circle (0.75mm);
-% \draw (3,1.7) circle (0.75mm);
-% \draw (3,2.2) circle (0.75mm);
-% \draw[-stealth] (3.05,2.2) -- (5,2.2);
+% First timeline
+\draw (0,2) -- (1,2);
+\draw[fill=black] (1,2) circle (0.75mm);
+\draw (1,1.7) circle (0.75mm);
+\draw (1.08,1.7) -- (3,1.7);
+\draw[fill=black] (3,1.7) circle (0.75mm);
+\draw (3,1.7) circle (0.75mm);
+\draw (3,2.2) circle (0.75mm);
+\draw[-stealth] (3.05,2.2) -- (5,2.2);
 
-% % second timeline
-% \draw[dashed] (0,1) -- (3,1);
-% \draw[fill=black] (3,1) circle (0.75mm);
-% \draw (3,.75) circle (0.75mm);
-% \draw[dashed, -stealth] (3.08,.75) -- (5,.75);
+% second timeline
+\draw[dashed] (0,1) -- (3,1);
+\draw[fill=black] (3,1) circle (0.75mm);
+\draw (3,.75) circle (0.75mm);
+\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
 
-% % labels
-% \node at (-.5,2) {$\lambda_{1,2}$};
-% \node at (-.5,1) {$\lambda_{3,4}$};
-% \node at (1,-.3) {$t_1$};
-% \node at (3,-.3) {$t_2$};
-% \end{tikzpicture}
-% \caption{Intensities for two dyads}
-% \end{subfigure}
-% \caption[]{Illustration of event data and the assumptions of the model.
+% labels
+\node at (-.5,2) {$\lambda_{1,2}$};
+\node at (-.5,1) {$\lambda_{3,4}$};
+\node at (1,-.3) {$t_1$};
+\node at (3,-.3) {$t_2$};
+\end{tikzpicture}
+\caption{Intensities for two dyads}
+\end{minipage}
+\caption[]{Illustration of event data and the assumptions of the model.
 Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
 Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur that involve either node 3 or node 4.\footnotemark }
-% \label{fig:example}
-% \end{figure}
+We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
+\label{fig:example}
+\end{figure}
 
@@
  In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event involving either $i$ or $j$.
-In the Appendix we show how this reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
+As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.
+This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
+In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
 
@@
 
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.
+We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
 
@@
 p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{I(r \in \{i_m,j_m\})}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
+\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
 %& \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t - t_{M}) \lambda_{ij}(t|\cdot)\}  
@@
 Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
+In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
+This aims for similarly sized groups.}
 % \begin{align*}
@@
 p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)
+p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
 \prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
@@
 \centering
-\begin{subfigure}[b]{0.45\textwidth}
+\begin{subfigure}[b]{0.35\textwidth}
 \centering
@@
 \qquad
-\begin{subfigure}[b]{0.45\textwidth}
+\begin{subfigure}[b]{0.35\textwidth}
 \centering
@@
 Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics as seen in (c) and (d).}
+Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
 \label{fig:parmats}
@@
 
-%  \appendix{\textbf{Split-merge Metropolis-Hastings Move for Relational Event Blockmodel}}
-
-% This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
-We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg.
-8).
-
-% %$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
-
-% Next pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
-% \begin{itemize}
-% %\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
-% \item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
-% %\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
-% \item Let $S$ be all nodes assigned to $k$.
-Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
-% where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
-% \item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
-% \end{itemize}
-
-%   If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
-% \begin{itemize}
-% \item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$, and sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
-% %\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
-% \item Let $S$ be the set of nodes assigned to $k$ or $l$.
-Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
-% \item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
-% \end{itemize}
-
-% The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
-
-% For the proposed merge step,
-% $$q(\boldsymbol{\phi}^{split} \rightarrow
-% \boldsymbol{\phi}^{merge}) = \prod_{q,r \ne l}\frac{1}{2} \prod_r p(\phi_{r,l,p}|\mu_p,\sigma_p^2)p(\phi_{l,r,p}|\mu_p,\sigma_p^2) p(\phi_{l,l,p}|\mu_p,\sigma_p^2)$$ though the latter product cancels with a corresponding term in $p(\boldsymbol{\phi}^{merge})$.
-We can compute the transition $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.
-For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.
-$ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
-
-% Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-% IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-% This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.
-We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg.
-8).
-
-% %$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
-
-% Next pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
-% \begin{itemize}
-% %\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
-% \item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
-% \item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
-% \item Let $S$ be all nodes assigned to $k$.
-Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
-% where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
-% \item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
-% \end{itemize}
-
-%   If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
-% \begin{itemize}
-% \item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$
-% %\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
-% \item Let $S$ be the set of nodes assigned to $k$ or $l$.
-Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
-% \item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
-% \end{itemize}
-
-% The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})}\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}\frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$
-
-% The first two terms represent the prior.
-For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.
-We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.
-For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.
-$ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
-We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.
-One can return to the previous state with positive probability.
+\spewnotes
 

### 2012-05-29 docs/submission.tex
Add new results tables.

@@
 
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$ (TODO).
+We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.
+The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
 
 \subsection{Model specification}
-
-% \begin{enumerate}
-% \item Intercept: $s_{0}(t,i,j) = 1$
-% \item Dyad count: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
-% \item Sender out-degree: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
-% \item Sender in-degree: $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$
-% \item Receiver out-degree: $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$
-% \item Receiver in-degree: $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$
-% \item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
-% \item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
-% \item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
-% \item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
-% \item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
-% %\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
-% % \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}=r)$
-% % \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=r,j_{v_{mij}}=j)$
-% % \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=r,j_{v_{mij}}=r)$
-% \end{enumerate}
 
@@
 Intercept& $s_{0}(t,i,j) = 1$\\
-Dyad count& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-Sender out-degree& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-Reciprocity (AB-BA)& $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-taking (AB-BY)&$s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Turn-continuing (AB-AY)& $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
+Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
+Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
+Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
+%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
+%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
+Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
+Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
+Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
+%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
+%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
 \hline
@@
 
-\subsection{Split-merge algorithm}
+\subsection*{Sampling $\mu$  and $\sigma$ given $\beta$ }
 
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
+The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
+\begin{align}
+\label{eqn:gibbs.sigma}
+\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
+ \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
+\end{align}
+The posterior distribution of $\mu$ given $\sigma$ and the lower level parameters can in turn be written as
+\begin{align}
+\label{eqn:gibbs.mu}
+\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K}\sum_{k=1}^K \beta_{k,p},\frac{ \sigma_p^2}{\sqrt{K}}\right),
+\end{align}
 
@@
 
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to continue speaking, and 3) interactions between groups are more likely to be repeated.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{5}(t,i,j), s_{9}(t,i,j)]$.
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,0,2)$.
+We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
+The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.
+For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
 This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
@@
 \begin{itemize}
-\item Kiel email: \cite{Ebel2002}
-\item Classrooms: \cite{McFarland2003}
-\item Eckmann email: dyadic emails  \cite{Eckmann}
+%\item Kiel email: \cite{Ebel2002}
+\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
+\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
 \item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
@@
 We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining:  \cite{MITreality}
+\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
 \end{itemize}
@@
 
-\input{../figs/results.tex}
+\input{../figs/results-llk.tex}
 
@@
 
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg} = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
+Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
 Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
@@
 Our method jointly models \emph{which} dyads occur and \emph{when} they occur.
-To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j} = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
+To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
 
-\begin{figure}[t]
-\center
-\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-\caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-\label{fig:recall}
-\end{figure}
+\input{../figs/results-recall.tex}
+
+% \begin{figure}[t]
+% \center
+% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
+% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
+% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
+% \caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
+% \label{fig:recall}
+% \end{figure}
 
@@
 Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
+
 %For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
@@
 This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share.
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j)$, and similarly one use our method to study how their role varies across nodes.
+Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
 
@@
 \appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
+
+%Motivation: simply Gibbs sampling does not seem to mix well.
+\footnote{This section is planned future work.
+In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
+For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
+
+Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
+\begin{align*}
+  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
+\end{align*}
+
+IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
+A proposed split has REM parameters that are only slightly different than those for the original block.
+Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
+By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
 

### 2012-05-29 docs/submission.tex
Fix up inference section.
Rearrange results/discussion a bit.

@@
 
-\subsection*{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
+\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
 
 We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
-% \begin{align*}
-% p(z_r | \mathbf{z}_{-r},\mathcal{A}_t,\alpha,\boldsymbol{\beta}) &\propto p(z_i | \mathbf{z}_{-i},\alpha) p(\mathcal{A}_t|\mathbf{z},\boldsymbol{\beta}) \\
-% p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{m=1}^M \lambda_{i_a,j_a}(t_a|\cdot)
-% \prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
-% \end{align*}
 \begin{align*}
@@
 \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
-%& \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t - t_{M}) \lambda_{ij}(t|\cdot)\}  
 \end{align*} 
@@
 This aims for similarly sized groups.}
-% \begin{align*}
-% p(z_i = k | z_{-i},\alpha) &= n_k & p(z_i = \mbox{new} |  z_{-i},\alpha) &= \alpha
-% \end{align*}
 
@@
 \begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\ %, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\
+p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
 p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
@@
 \prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-%p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
-%\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
 \end{align*}
@@
 
-\subsection*{Sampling $\mu$  and $\sigma$ given $\beta$ }
+\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
 
@@
 \end{align}
-The posterior distribution of $\mu$ given $\sigma$ and the lower level parameters can in turn be written as
+Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$, 
 \begin{align}
 \label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K}\sum_{k=1}^K \beta_{k,p},\frac{ \sigma_p^2}{\sqrt{K}}\right),
+\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
 \end{align}
-
-%$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$
-
-% TODO: Update the following discussion about sampling.
-
-% This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.
-Computing the likelihood is then $O(T \cdot P \cdot 4N)$ (though this computation is split into $K^2$ pieces) and must computed $N \cdot K$ times per Gibbs iteration.
-The most straightforward implementation has space complexity $O(N^2)$ since we must keep track of $\tau_{i,j,m}$ for each $(i,j)$ at a given $m$.
-
-% \subsection{Approximate likelihood methods}
-
-% TODO: Discuss how one might sample the likelihood.  Cite Raftery paper.
-
-% TODO: Look at performance of this with respect to bias and performance measure
 
@@
 
-\section{Model checking and experiments}
+\section{Model fitting and experiments}
 
@@
 \end{itemize}
+
+\subsection{Exploratory analysis}
+
+Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we
 
@@
 
- Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
+TODO: Quickly describe main take-aways from the results tables.
+
+% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
 For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
@@
 
-For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
+%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
 This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
 \footnote{Will need to discuss results for other datasets once they are complete.}
-
-Figure \ref{fig:recall} will provide recall curves for several of the datasets.
 
@@
  We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models) to provide detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in dynamics of the network as a whole.
+The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}).
+This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in dynamics of the network as a whole.
 
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann data.
-Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
+%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
 
@@
 
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share.
+This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.
 Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.

### 2012-05-29 docs/submission.tex
Put sentences on their own lines.

@@
 In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.
-% (n.b.
-$t_{v_{mij}} = \tau_{mij}$).
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.
-We include degree effects to model possible rich-get-richer phenomenon.
+Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$). 
+If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.  
 For example, a particular node sending often may indicate they will continue to be a sender in the future.  
@@
 TODO: Explore the theory about these sorts of statistics.
-Aalen has a few suggestions.} Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.
-\footnote{Describe the form of $f()$.
-Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
+Aalen has a few suggestions.}
+Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
 
@@
 
-Each of the following data sets are sequences of dyadic events, where each event has a time associated with it, a sender, and a recipient.
+Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
 
@@
 Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}   We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk} .
+I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
+We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.  
 The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
@@
 The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}). 
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in dynamics of the network as a whole.
+This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
 
@@
 This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.  
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
 In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.  
 We also note that we improve over the situation where we constrain $K=1$.  
+Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
 
@@
 
-Trivially extended to co-clustering applications and for directed data.  Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
+Trivially extended to co-clustering applications and for directed data.  
+Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
 
@@
 
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood
- Equation \ref{eqn:likelihood} as 
+We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as 
 \begin{align*}

### 2012-05-29 predict.r
Fix bug in synthetic recall results.  Update tables.

@@ -29,7 +29,6 @@ save.pred <- function(pred,dataset,model) {
   llk.test  <- pred$llk$test
   rk.train  <- pred$rks$train
   rk.test <- pred$rks$test
-
   dir.create(paste("results/",dataset,"/final/",sep=""),showWarnings=FALSE)
   dir.create(paste("results/",dataset,"/llks/",sep=""),showWarnings=FALSE)
   save(llkm.train,llkm.test,llk.train,llk.test,opts,file=paste("results/",dataset,"/llks/",model,".rdata",sep=""))
@@ -60,7 +59,7 @@ for (model in models) {
 if (dataset == "synthetic-1") {
   niter <- 500
   fit <- list(params=list(beta=beta,z=z),llks=rep(true.lpost,niter),niter=niter,zs=rep(list(z),niter),param=beta,ego=1,transform=TRUE)  # true values
-  pred <- evaluate(A,N,train.ix,test.ix,fit)
+  pred <- evaluate(A,N,train.ix,test.ix,fit,ties.method="random")
   save.pred(pred,dataset,"truth")
 }
 
@@ -109,16 +108,17 @@ names(rks) <- models
 
 ds <- lapply(rks,function(r) {
   d <- list(train=list(recall.30  = recall(r$train,top=1:30),
-                       recall.200 = recall(r$train,top=1:100),
+                       #recall.200 = recall(r$train,top=1:100),
                        recall.all = recall(r$train,top=seq(1,N^2,length.out=100))),
             test =list(recall.30  = recall(r$test,top=1:30),
-                       recall.200 = recall(r$test,top=1:100),
+                       #recall.200 = recall(r$test,top=1:100),
                        recall.all = recall(r$test,top=seq(1,N^2,length.out=100))))
   return(melt(d,id.vars="k",measure.vars="recall"))
 })
 for (i in 1:length(ds)) ds[[i]]$model <- models[i]
 ds <- do.call(rbind,ds)
 rownames(ds) <- c()
+t(t(table(ds$model)))
 
 save(ds,file=paste("results/",dataset,"/final/recall.rdata",sep=""))
 
@@ -167,25 +167,11 @@ ks <- lapply(models,function(model) {
 })
 names(ks) <- models
 
-for (m in c("uniform","marg","online")) {
-  ks[[m]] <- data.frame(model=m,mean.K=NA,sd.K=NA)
-}
[…]

### 2012-05-29 results.r

@@ -3,7 +3,7 @@ datasets <- c("synthetic-1","eckmann-small", "realitymining-small", "classroom-1
 res <- lapply(datasets,function(x) {
   f <-paste("results/",x,"/final/results.rdata",sep="")
   if (file.exists(f)) load(f)
-  return(df)
+  return(df[,1:5])
 })
 names(res) <- datasets
 lapply(res,dim)
@@ -11,7 +11,7 @@ lapply(res,dim)
 res <- do.call(rbind,res)
 
 res$dataset <- factor(res$dataset,datasets)
-res <- res[,-c(6,7)]
+#res <- res[,-c(6,7)]
 library(reshape2)
 res <- res[,c(1:3,5,4)]
 
@@ -25,6 +25,10 @@ r$metric <- as.character(r$metric)
 r <- r[,c(1,2,6,4,5,3)]
 r$dataset[-seq(1,nrow(r),by=2)] <- ""
 
+# Add truth for synthetic data
+r$truth <- NA
+r$truth[1:4] <- subset(res,L1 == "truth" & type=="test")$value
+
 a <- subset(r, metric %in% c("rem","mult"))
 #r$likelihood[-seq(1,nrow(r),by=2)] <- ""
 library(xtable)
@@ -33,17 +37,6 @@ print(xr,include.rownames=FALSE,file=paste("figs/results-llk.tex",sep=""),NA.str
 
 b <- subset(r, metric %in% c("rank5","rank20"))
 b$metric <- c("@  5", "@ 20")
-xr <- xtable(r,caption="Predictive accuracy on a recall task.  Larger values are better.  See text for details.",label="tab:results",digits=3)
+xr <- xtable(b,caption="Predictive accuracy on a recall task.  Larger values are better.  See text for details.",label="tab:results",digits=3)
 print(xr,include.rownames=FALSE,file=paste("figs/results-recall.tex",sep=""),NA.string="",table.placement="t",size="footnotesize")
 
-
-# Clean up
-colnames(r)[7:9] <- paste("K=",1:3,sep="")
-r$dataset    <- as.character(r$dataset)
-r$likelihood <- as.character(r$likelihood)
-r$dataset[-seq(1,nrow(r),by=4)] <- ""
-r$likelihood[-seq(1,nrow(r),by=2)] <- ""
-#for (i in 4:ncol(r)) r[,i] <- round(r[,i],3)
-
[…]

### 2012-05-29 docs/submission.tex
Sketch out exploratory analysis section.

@@
 
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we
+Figure \ref{fig:posteriorparams} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.  
+We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
+For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.  
+The two panels on the left show the variation in $s_0(t_M,i,j)$, $s_1(t_M,i,j)$ and $s_3(t_M,i,j)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
+The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{p,1,3}$ and $\beta_{p,3,1}$. 
+Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.  
+Similarly, block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
+The model has learned that this structure exists in the data and our inferences from the parameter estimates are sensible given the observed statistics.
 
@@
 \caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions from members of group 1 to group 3 (178 total events in total) and from members of group 3 to group 1 (179 events in total).
+ Left: Observed statistics across all dyadic email interactions between  members of group 1 and group 3.
 Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-In addition to finding differences in mean activity for these dyads, the model learns that events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
+In addition to finding differences in mean activity for these dyads, estimates suggest events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
 \label{fig:posteriorparams}

### 2012-05-30 docs/submission.tex
Paper edits.
Edited the intro a bit (3rd para mainly)

@@
 \title{Stochastic Blockmodels for Relational Event Data}
+% Latent Variable Modeling of Relational Event Dynamics
+% Latent Variable Modeling of Relational Event Sequences
+% Latent Variable Modeling for Relational Event Sequences
 
@@
 Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.
-For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
-For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
+For continuous-time network data, a number of approaches have recently been proposed for modeling the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
+For static networks, methods such as stochastic blockmodels have traditionally been used for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
 We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
-The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
+The method is illustrated using dyadic interaction data such as email communication and Twitter direct messages.
 Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
@@
 
-The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
-In many cases this is unrealistic.  
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
+A limitation of all of these approaches is the assumption of a single common behavior for all individuals.
+In contrast, in real-world networks it is reasonable to expect {\it heterogeneity} in dynamic behavior.  
+In an organization, the manner in which an individual communicates with others may be a function of an individual's {\it role} in the organization.
+For example, in a university,  email communication patterns over time between professors, students, and staff will likely be quite different within and between the three groups.
+%PS: not totally happy with this example since the "roles" are completely determined by one's "job title" and I think we would prefer something
+% less explicit, e.g., that there are "leaders", "followers", etc, that go beyond just job titles. So feel free to edit this further!
+
+%
+%comprised of several teams, each team may uncover different patterns for collaborating via email---individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
 
@@
 
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
-The behavior of the model is illustrated with simulated data.  
-We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
-Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
+Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data to capture node-level heterogeneity, where we assume that latent clusters of nodes share similar patterns of interaction.
+We describe parameter estimation and learning of  latent cluster assignments via MCMC, and illustrate the behavior of the model  with simulated data.
+Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
+Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, allowing identification of particular \emph{roles} in communication.
 
@@
 
-\subsection{Exploratory analysis}
+\subsection{Model-based exploratory analysis}
 
@@
 
-\subsubsection*{Acknowledgements}
+%\subsubsection*{Acknowledgements}
 

### 2012-05-30 docs/abstract.tex
Split draft into different tex files.

--- 2012-05-29 docs/submission.tex
@@
-\documentclass{article}
-\usepackage{nips12submit_e,times}
-\usepackage{tikz}
-\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-\usepackage{verbatim}
-\usepackage{url}
-\usepackage{graphicx, color}
-\usepackage{amsmath}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{footnote}
-
-\title{Stochastic Blockmodels for Relational Event Data}
-
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-}
-
-\begin{document}
-
-\maketitle
-
-\savenotes
-
 \begin{abstract}
 Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.
-For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
-For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
+For continuous-time network data, a number of approaches have recently been proposed for modeling the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
+For static networks, methods such as stochastic blockmodels have traditionally been used for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
 We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
-The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
+The method is illustrated using dyadic interaction data such as email communication and Twitter direct messages.
 Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
@@
 \end{abstract}
-
-\section{Introduction}
-
-Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
-% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
-% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
-
-Network data, however, is often collected as a sequence of events occurring over time.   
-Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify how the process depends on the previous history of events.  
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
-
-The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
-In many cases this is unrealistic.  
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
-
-%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
-
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
-The behavior of the model is illustrated with simulated data.  
-We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
-Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
-
-\section{Model}
-
-%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
-We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
-Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.  
-If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
-&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
-\end{align}
-%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
-
-% \begin{align}
-% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
-% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
-% \end{align}
-
-One can extend the above to marked point processes where each event contains additional information. 
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
-The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
-\label{eqn:llk}
-\end{align}
-%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
-
-\begin{figure}
-\centering 
-\begin{minipage}[b]{0.45\linewidth}
-  \begin{tikzpicture}[scale=.8,transform shape,thick]
-    \node[vertex] (n1) at (0,0)  {1};
-    \node[vertex] (n2) at (2,2)  {2};
-    \node[vertex] (n3) at (4,0) {3};
-    \node[vertex] (n4) at (2,-2) {4};
-    \draw[-stealth] (.4,.4) -- (1.6,1.6);
-    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
-    \node at (.7,1.1) {$t_1$};
-    \node at (3.3,1.1) {$t_2$};
-  \end{tikzpicture}
-  \caption{Dynamic network data}
-\end{minipage}
-\begin{minipage}[b]{0.45\linewidth}
-\begin{tikzpicture}[scale=1,transform shape,thick]
-\draw[-stealth] (0,0) -- (0,3);
-\draw[-stealth] (0,0) -- (5,0);
-
-% First timeline
-\draw (0,2) -- (1,2);
-\draw[fill=black] (1,2) circle (0.75mm);
-\draw (1,1.7) circle (0.75mm);
-\draw (1.08,1.7) -- (3,1.7);
-\draw[fill=black] (3,1.7) circle (0.75mm);
-\draw (3,1.7) circle (0.75mm);
-\draw (3,2.2) circle (0.75mm);
-\draw[-stealth] (3.05,2.2) -- (5,2.2);
-
-% second timeline
-\draw[dashed] (0,1) -- (3,1);
-\draw[fill=black] (3,1) circle (0.75mm);
-\draw (3,.75) circle (0.75mm);
-\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
-
-% labels
-\node at (-.5,2) {$\lambda_{1,2}$};
-\node at (-.5,1) {$\lambda_{3,4}$};
-\node at (1,-.3) {$t_1$};
-\node at (3,-.3) {$t_2$};
-\end{tikzpicture}
-\caption{Intensities for two dyads}
-\end{minipage}
-\caption[]{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
-\label{fig:example}
-\end{figure}
-
-\footnotetext{TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-
-% \begin{figure}
-%  \def\svgwidth{6in}
-%   \input{example1.pdf_tex}
-% \caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-% \label{fig:example}
-% %\includegraphics[width=6in]{example}
-% \end{figure}
-
-%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
-
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
-\begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
-%z_i &\sim \mbox{CRP}(\alpha) &
-%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
-\end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-
- In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.  
-This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.  
-In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
-
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.  
-The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
-
-\subsection{Model specification}
-
-\begin{table}[t]
-\footnotesize
-\center
-\begin{tabular}{|l|l|}
-\hline
-Statistic & Formula \\
-\hline
-\hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-\hline
-\end{tabular}
-\label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
-\end{table}
-
-In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$). 
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.  
-For example, a particular node sending often may indicate they will continue to be a sender in the future.  
-We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
-TODO: Explore the theory about these sorts of statistics.
-Aalen has a few suggestions.}
-Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
-
-The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.  
-For example, an AB-BA effect indicates an increased propensity for reciprocity.  
-The final set of effects allow one to model phenomenon such as triadic closure.  
-Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.  
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
-
-\subsection{Relation to other models}
-
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
-In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
-
-The above framework generalizes several important special cases.  
-For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  
-Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  
-As these intensities do not change under this specification, the likelihood simplifies to 
-$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-
-Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
-\begin{align}
-\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
-\label{eqn:multllk}
-\end{align}
-The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
-\section{Inference}
-
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
-
-\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
-
-We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
-\begin{align*}
-p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
-\end{align*} 
-where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.  
-Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.  
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
-This aims for similarly sized groups.}
-
-\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
-
-For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
-\begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
-p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
-\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-\end{align*}
-where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
-This distribution is sampled using slice sampling.
-
-\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
-
-The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
-\begin{align}
-\label{eqn:gibbs.sigma}
-\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
- \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
-\end{align}
-Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$, 
-\begin{align}
-\label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
-\end{align}
-
-\begin{figure}
-\center
-\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
-\label{fig:syncounts}
-\end{figure}
-
-\section{Simulation}
-
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.  
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.  
-This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
-
-Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
-Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
-
-\section{Model fitting and experiments}
-
-\subsection{Data}
-
-Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
-
-\begin{itemize}
-%\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
-\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
-\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
-We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
-Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  
-We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
-\end{itemize}
-
-\subsection{Exploratory analysis}
-
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we
-
-\subsection{Prediction experiments}
-
-We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.  
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
-We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.  
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
-
-In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.  
-At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
-
-\input{../figs/results-llk.tex}
-
-\subsection{Baselines}
-
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
-Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
-
-Our method jointly models \emph{which} dyads occur and \emph{when} they occur.  
-To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
-
-\input{../figs/results-recall.tex}
-
-% \begin{figure}[t]
-% \center
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-% \label{fig:recall}
-% \end{figure}
-
-\subsection{Results}
-
-TODO: Quickly describe main take-aways from the results tables.
-
-% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
-For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
-Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.
-
-%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
-This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
-\footnote{Will need to discuss results for other datasets once they are complete.}
-
-\section{Discussion}
-
-%% TODO: Discuss scalability.
-%% TODO: Describe and include posterior predictive checks.
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
-%\caption{Observed counts}
-\end{subfigure}
-\qquad
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
-%\caption{Samples of $\beta_{p,k,l}$}
-\end{subfigure}
-
-\caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions from members of group 1 to group 3 (178 total events in total) and from members of group 3 to group 1 (179 events in total).
-Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-In addition to finding differences in mean activity for these dyads, the model learns that events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
-\label{fig:posteriorparams}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
-\caption{Observed counts}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
-\caption{Intercept estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
-\caption{\texttt{ab-ba} estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
-\caption{\texttt{ab-by} estimates}
-\end{subfigure}
-\caption{Comparing observed counts and parameter estimates.
-Darker values are larger.
-Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
-\label{fig:parmats}
-\end{figure}
-
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}). 
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
-
-%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-
-%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
-Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
-
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.  
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.  
-We also note that we improve over the situation where we constrain $K=1$.  
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.  
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
-
-Trivially extended to co-clustering applications and for directed data.  
-Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
-
-Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.  
-For example, dynamics might change over time \cite{Vu2011}.  
-Alternatively, if nodes change groups.  
-One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.  
-In some contexts such extensions might be substantively important and warrant future work.
-
-\subsubsection*{Acknowledgements}
-
-\bibliographystyle{unsrt}
-\bibliography{refs}
-
-\appendix{\textbf{Complexity}}
-
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as 
-\begin{align*}
-\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
-\end{align*}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
-
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
-
-\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
-
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-This is a two-stage MH procedure.
-First, pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
-\begin{itemize}
-\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
-\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
-where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
- If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
-\begin{itemize}
-\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
-  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
-\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
-\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
- The Metropolis-Hastings acceptance probabilities require the quantity
-$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
-
-The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
-Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
-The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
-$q(\boldsymbol{\phi}^{s} \rightarrow
-\boldsymbol{\phi}^{m})$ are products of Normal densities.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
-
-Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
-This increases the acceptance probability for splits (and decreases the probability of merges).
-However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
-
-\spewnotes
-
-\end{document}

### 2012-05-30 docs/discussion.tex
Split draft into different tex files.

--- 2012-05-29 docs/submission.tex
@@
-\documentclass{article}
-\usepackage{nips12submit_e,times}
-\usepackage{tikz}
-\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-\usepackage{verbatim}
-\usepackage{url}
-\usepackage{graphicx, color}
-\usepackage{amsmath}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{footnote}
-
-\title{Stochastic Blockmodels for Relational Event Data}
-
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-}
-
-\begin{document}
-
-\maketitle
-
-\savenotes
-
-\begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.  
-For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
-For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
-We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  
-The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
-Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.  
-The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.  
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-\end{abstract}
-
-\section{Introduction}
-
-Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
-% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
-% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
-
-Network data, however, is often collected as a sequence of events occurring over time.   
-Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify how the process depends on the previous history of events.  
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
-
-The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
-In many cases this is unrealistic.  
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
-
-%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
-
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
-The behavior of the model is illustrated with simulated data.  
-We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
-Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
-
-\section{Model}
-
-%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
-We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
-Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.  
-If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
-&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
-\end{align}
-%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
-
-% \begin{align}
-% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
-% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
-% \end{align}
-
-One can extend the above to marked point processes where each event contains additional information. 
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
-The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
-\label{eqn:llk}
-\end{align}
-%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
-
-\begin{figure}
-\centering 
-\begin{minipage}[b]{0.45\linewidth}
-  \begin{tikzpicture}[scale=.8,transform shape,thick]
-    \node[vertex] (n1) at (0,0)  {1};
-    \node[vertex] (n2) at (2,2)  {2};
-    \node[vertex] (n3) at (4,0) {3};
-    \node[vertex] (n4) at (2,-2) {4};
-    \draw[-stealth] (.4,.4) -- (1.6,1.6);
-    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
-    \node at (.7,1.1) {$t_1$};
-    \node at (3.3,1.1) {$t_2$};
-  \end{tikzpicture}
-  \caption{Dynamic network data}
-\end{minipage}
-\begin{minipage}[b]{0.45\linewidth}
-\begin{tikzpicture}[scale=1,transform shape,thick]
-\draw[-stealth] (0,0) -- (0,3);
-\draw[-stealth] (0,0) -- (5,0);
-
-% First timeline
-\draw (0,2) -- (1,2);
-\draw[fill=black] (1,2) circle (0.75mm);
-\draw (1,1.7) circle (0.75mm);
-\draw (1.08,1.7) -- (3,1.7);
-\draw[fill=black] (3,1.7) circle (0.75mm);
-\draw (3,1.7) circle (0.75mm);
-\draw (3,2.2) circle (0.75mm);
-\draw[-stealth] (3.05,2.2) -- (5,2.2);
-
-% second timeline
-\draw[dashed] (0,1) -- (3,1);
-\draw[fill=black] (3,1) circle (0.75mm);
-\draw (3,.75) circle (0.75mm);
-\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
-
-% labels
-\node at (-.5,2) {$\lambda_{1,2}$};
-\node at (-.5,1) {$\lambda_{3,4}$};
-\node at (1,-.3) {$t_1$};
-\node at (3,-.3) {$t_2$};
-\end{tikzpicture}
-\caption{Intensities for two dyads}
-\end{minipage}
-\caption[]{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
-\label{fig:example}
-\end{figure}
-
-\footnotetext{TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-
-% \begin{figure}
-%  \def\svgwidth{6in}
-%   \input{example1.pdf_tex}
-% \caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-% \label{fig:example}
-% %\includegraphics[width=6in]{example}
-% \end{figure}
-
-%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
-
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
-\begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
-%z_i &\sim \mbox{CRP}(\alpha) &
-%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
-\end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-
- In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.  
-This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.  
-In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
-
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.  
-The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
-
-\subsection{Model specification}
-
-\begin{table}[t]
-\footnotesize
-\center
-\begin{tabular}{|l|l|}
-\hline
-Statistic & Formula \\
-\hline
-\hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-\hline
-\end{tabular}
-\label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
-\end{table}
-
-In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$). 
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.  
-For example, a particular node sending often may indicate they will continue to be a sender in the future.  
-We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
-TODO: Explore the theory about these sorts of statistics.
-Aalen has a few suggestions.}
-Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
-
-The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.  
-For example, an AB-BA effect indicates an increased propensity for reciprocity.  
-The final set of effects allow one to model phenomenon such as triadic closure.  
-Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.  
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
-
-\subsection{Relation to other models}
-
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
-In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
-
-The above framework generalizes several important special cases.  
-For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  
-Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  
-As these intensities do not change under this specification, the likelihood simplifies to 
-$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-
-Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
-\begin{align}
-\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
-\label{eqn:multllk}
-\end{align}
-The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
-\section{Inference}
-
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
-
-\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
-
-We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
-\begin{align*}
-p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
-\end{align*} 
-where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.  
-Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.  
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
-This aims for similarly sized groups.}
-
-\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
-
-For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
-\begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
-p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
-\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-\end{align*}
-where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
-This distribution is sampled using slice sampling.
-
-\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
-
-The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
-\begin{align}
-\label{eqn:gibbs.sigma}
-\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
- \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
-\end{align}
-Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$, 
-\begin{align}
-\label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
-\end{align}
-
-\begin{figure}
-\center
-\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
-\label{fig:syncounts}
-\end{figure}
-
-\section{Simulation}
-
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.  
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.  
-This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
-
-Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
-Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
-
-\section{Model fitting and experiments}
-
-\subsection{Data}
-
-Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
-
-\begin{itemize}
-%\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
-\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
-\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
-We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
-Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  
-We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
-\end{itemize}
-
-\subsection{Exploratory analysis}
-
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we
-
-\subsection{Prediction experiments}
-
-We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.  
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
-We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.  
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
-
-In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.  
-At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
-
-\input{../figs/results-llk.tex}
-
-\subsection{Baselines}
-
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
-Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
-
-Our method jointly models \emph{which} dyads occur and \emph{when} they occur.  
-To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
-
-\input{../figs/results-recall.tex}
-
-% \begin{figure}[t]
-% \center
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-% \label{fig:recall}
-% \end{figure}
-
-\subsection{Results}
-
-TODO: Quickly describe main take-aways from the results tables.
-
-% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
-For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
-Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.
-
-%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
-This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
-\footnote{Will need to discuss results for other datasets once they are complete.}
-
 \section{Discussion}
@@
 In some contexts such extensions might be substantively important and warrant future work.
-
-\subsubsection*{Acknowledgements}
-
-\bibliographystyle{unsrt}
-\bibliography{refs}
-
-\appendix{\textbf{Complexity}}
-
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as 
-\begin{align*}
-\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
-\end{align*}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
-
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
-
-\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
-
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-This is a two-stage MH procedure.
-First, pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
-\begin{itemize}
-\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
-\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
-where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
- If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
-\begin{itemize}
-\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
-  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
-\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
-\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
- The Metropolis-Hastings acceptance probabilities require the quantity
-$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
-
-The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
-Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
-The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
-$q(\boldsymbol{\phi}^{s} \rightarrow
-\boldsymbol{\phi}^{m})$ are products of Normal densities.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
-
-Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
-This increases the acceptance probability for splits (and decreases the probability of merges).
-However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
-
-\spewnotes
-
-\end{document}

### 2012-05-30 docs/experiments.tex
Split draft into different tex files.

--- 2012-05-29 docs/submission.tex
@@
-\documentclass{article}
-\usepackage{nips12submit_e,times}
-\usepackage{tikz}
-\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-\usepackage{verbatim}
-\usepackage{url}
-\usepackage{graphicx, color}
-\usepackage{amsmath}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{footnote}
-
-\title{Stochastic Blockmodels for Relational Event Data}
-
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-}
-
-\begin{document}
-
-\maketitle
-
-\savenotes
-
-\begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.  
-For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
-For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
-We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  
-The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
-Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.  
-The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.  
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-\end{abstract}
-
-\section{Introduction}
-
-Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
-% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
-% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
-
-Network data, however, is often collected as a sequence of events occurring over time.   
-Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify how the process depends on the previous history of events.  
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
-
-The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
-In many cases this is unrealistic.  
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
-
-%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
-
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
-The behavior of the model is illustrated with simulated data.  
-We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
-Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
-
-\section{Model}
-
-%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
-We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
-Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.  
-If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
-&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
-\end{align}
-%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
-
-% \begin{align}
-% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
-% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
-% \end{align}
-
-One can extend the above to marked point processes where each event contains additional information. 
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
-The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
-\label{eqn:llk}
-\end{align}
-%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
-
-\begin{figure}
-\centering 
-\begin{minipage}[b]{0.45\linewidth}
-  \begin{tikzpicture}[scale=.8,transform shape,thick]
-    \node[vertex] (n1) at (0,0)  {1};
-    \node[vertex] (n2) at (2,2)  {2};
-    \node[vertex] (n3) at (4,0) {3};
-    \node[vertex] (n4) at (2,-2) {4};
-    \draw[-stealth] (.4,.4) -- (1.6,1.6);
-    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
-    \node at (.7,1.1) {$t_1$};
-    \node at (3.3,1.1) {$t_2$};
-  \end{tikzpicture}
-  \caption{Dynamic network data}
-\end{minipage}
-\begin{minipage}[b]{0.45\linewidth}
-\begin{tikzpicture}[scale=1,transform shape,thick]
-\draw[-stealth] (0,0) -- (0,3);
-\draw[-stealth] (0,0) -- (5,0);
-
-% First timeline
-\draw (0,2) -- (1,2);
-\draw[fill=black] (1,2) circle (0.75mm);
-\draw (1,1.7) circle (0.75mm);
-\draw (1.08,1.7) -- (3,1.7);
-\draw[fill=black] (3,1.7) circle (0.75mm);
-\draw (3,1.7) circle (0.75mm);
-\draw (3,2.2) circle (0.75mm);
-\draw[-stealth] (3.05,2.2) -- (5,2.2);
-
-% second timeline
-\draw[dashed] (0,1) -- (3,1);
-\draw[fill=black] (3,1) circle (0.75mm);
-\draw (3,.75) circle (0.75mm);
-\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
-
-% labels
-\node at (-.5,2) {$\lambda_{1,2}$};
-\node at (-.5,1) {$\lambda_{3,4}$};
-\node at (1,-.3) {$t_1$};
-\node at (3,-.3) {$t_2$};
-\end{tikzpicture}
-\caption{Intensities for two dyads}
-\end{minipage}
-\caption[]{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
-\label{fig:example}
-\end{figure}
-
-\footnotetext{TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-
-% \begin{figure}
-%  \def\svgwidth{6in}
-%   \input{example1.pdf_tex}
-% \caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-% \label{fig:example}
-% %\includegraphics[width=6in]{example}
-% \end{figure}
-
-%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
-
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
-\begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
-%z_i &\sim \mbox{CRP}(\alpha) &
-%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
-\end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-
- In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.  
-This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.  
-In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
-
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.  
-The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
-
-\subsection{Model specification}
-
-\begin{table}[t]
-\footnotesize
-\center
-\begin{tabular}{|l|l|}
-\hline
-Statistic & Formula \\
-\hline
-\hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-\hline
-\end{tabular}
-\label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
-\end{table}
-
-In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$). 
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.  
-For example, a particular node sending often may indicate they will continue to be a sender in the future.  
-We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
-TODO: Explore the theory about these sorts of statistics.
-Aalen has a few suggestions.}
-Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
-
-The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.  
-For example, an AB-BA effect indicates an increased propensity for reciprocity.  
-The final set of effects allow one to model phenomenon such as triadic closure.  
-Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.  
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
-
-\subsection{Relation to other models}
-
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
-In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
-
-The above framework generalizes several important special cases.  
-For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  
-Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  
-As these intensities do not change under this specification, the likelihood simplifies to 
-$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-
-Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
-\begin{align}
-\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
-\label{eqn:multllk}
-\end{align}
-The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
-\section{Inference}
-
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
-
-\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
-
-We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
-\begin{align*}
-p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
-\end{align*} 
-where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.  
-Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.  
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
-This aims for similarly sized groups.}
-
-\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
-
-For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
-\begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
-p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
-\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-\end{align*}
-where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
-This distribution is sampled using slice sampling.
-
-\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
-
-The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
-\begin{align}
-\label{eqn:gibbs.sigma}
-\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
- \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
-\end{align}
-Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$, 
-\begin{align}
-\label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
-\end{align}
-
-\begin{figure}
-\center
-\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
-\label{fig:syncounts}
-\end{figure}
-
-\section{Simulation}
-
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.  
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.  
-This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
-
-Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
-Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
-
 \section{Model fitting and experiments}
@@
 \footnote{Will need to discuss results for other datasets once they are complete.}
-
-\section{Discussion}
-
-%% TODO: Discuss scalability.
-%% TODO: Describe and include posterior predictive checks.
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
-%\caption{Observed counts}
-\end{subfigure}
-\qquad
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
-%\caption{Samples of $\beta_{p,k,l}$}
-\end{subfigure}
-
-\caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions between  members of group 1 and group 3.
-Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-In addition to finding differences in mean activity for these dyads, estimates suggest events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
-\label{fig:posteriorparams}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
-\caption{Observed counts}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
-\caption{Intercept estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
-\caption{\texttt{ab-ba} estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
-\caption{\texttt{ab-by} estimates}
-\end{subfigure}
-\caption{Comparing observed counts and parameter estimates.
-Darker values are larger.
-Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
-\label{fig:parmats}
-\end{figure}
-
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}). 
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
-
-%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-
-%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
-Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
-
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.  
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.  
-We also note that we improve over the situation where we constrain $K=1$.  
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.  
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
-
-Trivially extended to co-clustering applications and for directed data.  
-Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
-
-Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.  
-For example, dynamics might change over time \cite{Vu2011}.  
-Alternatively, if nodes change groups.  
-One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.  
-In some contexts such extensions might be substantively important and warrant future work.
-
-\subsubsection*{Acknowledgements}
-
-\bibliographystyle{unsrt}
-\bibliography{refs}
-
-\appendix{\textbf{Complexity}}
-
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as 
-\begin{align*}
-\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
-\end{align*}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
-
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
-
-\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
-
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-This is a two-stage MH procedure.
-First, pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
-\begin{itemize}
-\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
-\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
-where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
- If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
-\begin{itemize}
-\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
-  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
-\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
-\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
- The Metropolis-Hastings acceptance probabilities require the quantity
-$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
-
-The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
-Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
-The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
-$q(\boldsymbol{\phi}^{s} \rightarrow
-\boldsymbol{\phi}^{m})$ are products of Normal densities.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
-
-Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
-This increases the acceptance probability for splits (and decreases the probability of merges).
-However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
-
-\spewnotes
-
-\end{document}

### 2012-05-30 docs/inference.tex
Split draft into different tex files.

--- 2012-05-29 docs/submission.tex
@@
-\documentclass{article}
-\usepackage{nips12submit_e,times}
-\usepackage{tikz}
-\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-\usepackage{verbatim}
-\usepackage{url}
-\usepackage{graphicx, color}
-\usepackage{amsmath}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{footnote}
-
-\title{Stochastic Blockmodels for Relational Event Data}
-
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-}
-
-\begin{document}
-
-\maketitle
-
-\savenotes
-
-\begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.  
-For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
-For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
-We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  
-The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
-Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.  
-The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.  
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-\end{abstract}
-
-\section{Introduction}
-
-Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
-% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
-% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
-
-Network data, however, is often collected as a sequence of events occurring over time.   
-Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify how the process depends on the previous history of events.  
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
-
-The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
-In many cases this is unrealistic.  
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
-
-%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
-
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
-The behavior of the model is illustrated with simulated data.  
-We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
-Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
-
-\section{Model}
-
-%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
-We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
-Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.  
-If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
-&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
-\end{align}
-%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
-
-% \begin{align}
-% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
-% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
-% \end{align}
-
-One can extend the above to marked point processes where each event contains additional information. 
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
-The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
-\label{eqn:llk}
-\end{align}
-%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
-
-\begin{figure}
-\centering 
-\begin{minipage}[b]{0.45\linewidth}
-  \begin{tikzpicture}[scale=.8,transform shape,thick]
-    \node[vertex] (n1) at (0,0)  {1};
-    \node[vertex] (n2) at (2,2)  {2};
-    \node[vertex] (n3) at (4,0) {3};
-    \node[vertex] (n4) at (2,-2) {4};
-    \draw[-stealth] (.4,.4) -- (1.6,1.6);
-    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
-    \node at (.7,1.1) {$t_1$};
-    \node at (3.3,1.1) {$t_2$};
-  \end{tikzpicture}
-  \caption{Dynamic network data}
-\end{minipage}
-\begin{minipage}[b]{0.45\linewidth}
-\begin{tikzpicture}[scale=1,transform shape,thick]
-\draw[-stealth] (0,0) -- (0,3);
-\draw[-stealth] (0,0) -- (5,0);
-
-% First timeline
-\draw (0,2) -- (1,2);
-\draw[fill=black] (1,2) circle (0.75mm);
-\draw (1,1.7) circle (0.75mm);
-\draw (1.08,1.7) -- (3,1.7);
-\draw[fill=black] (3,1.7) circle (0.75mm);
-\draw (3,1.7) circle (0.75mm);
-\draw (3,2.2) circle (0.75mm);
-\draw[-stealth] (3.05,2.2) -- (5,2.2);
-
-% second timeline
-\draw[dashed] (0,1) -- (3,1);
-\draw[fill=black] (3,1) circle (0.75mm);
-\draw (3,.75) circle (0.75mm);
-\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
-
-% labels
-\node at (-.5,2) {$\lambda_{1,2}$};
-\node at (-.5,1) {$\lambda_{3,4}$};
-\node at (1,-.3) {$t_1$};
-\node at (3,-.3) {$t_2$};
-\end{tikzpicture}
-\caption{Intensities for two dyads}
-\end{minipage}
-\caption[]{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
-\label{fig:example}
-\end{figure}
-
-\footnotetext{TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-
-% \begin{figure}
-%  \def\svgwidth{6in}
-%   \input{example1.pdf_tex}
-% \caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-% \label{fig:example}
-% %\includegraphics[width=6in]{example}
-% \end{figure}
-
-%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
-
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
-\begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
-%z_i &\sim \mbox{CRP}(\alpha) &
-%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
-\end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-
- In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.  
-This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.  
-In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
-
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.  
-The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
-
-\subsection{Model specification}
-
-\begin{table}[t]
-\footnotesize
-\center
-\begin{tabular}{|l|l|}
-\hline
-Statistic & Formula \\
-\hline
-\hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-\hline
-\end{tabular}
-\label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
-\end{table}
-
-In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$). 
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.  
-For example, a particular node sending often may indicate they will continue to be a sender in the future.  
-We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
-TODO: Explore the theory about these sorts of statistics.
-Aalen has a few suggestions.}
-Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
-
-The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.  
-For example, an AB-BA effect indicates an increased propensity for reciprocity.  
-The final set of effects allow one to model phenomenon such as triadic closure.  
-Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.  
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
-
-\subsection{Relation to other models}
-
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
-In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
-
-The above framework generalizes several important special cases.  
-For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  
-Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  
-As these intensities do not change under this specification, the likelihood simplifies to 
-$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-
-Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
-\begin{align}
-\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
-\label{eqn:multllk}
-\end{align}
-The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
 \section{Inference}
@@
 \end{figure}
-
-\section{Simulation}
-
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.  
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.  
-This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
-
-Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
-Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
-
-\section{Model fitting and experiments}
-
-\subsection{Data}
-
-Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
-
-\begin{itemize}
-%\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
-\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
-\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
-We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
-Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  
-We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
-\end{itemize}
-
-\subsection{Exploratory analysis}
-
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we
-
-\subsection{Prediction experiments}
-
-We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.  
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
-We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.  
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
-
-In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.  
-At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
-
-\input{../figs/results-llk.tex}
-
-\subsection{Baselines}
-
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
-Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
-
-Our method jointly models \emph{which} dyads occur and \emph{when} they occur.  
-To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
-
-\input{../figs/results-recall.tex}
-
-% \begin{figure}[t]
-% \center
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-% \label{fig:recall}
-% \end{figure}
-
-\subsection{Results}
-
-TODO: Quickly describe main take-aways from the results tables.
-
-% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
-For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
-Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.
-
-%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
-This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
-\footnote{Will need to discuss results for other datasets once they are complete.}
-
-\section{Discussion}
-
-%% TODO: Discuss scalability.
-%% TODO: Describe and include posterior predictive checks.
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
-%\caption{Observed counts}
-\end{subfigure}
-\qquad
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
-%\caption{Samples of $\beta_{p,k,l}$}
-\end{subfigure}
-
-\caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions from members of group 1 to group 3 (178 total events in total) and from members of group 3 to group 1 (179 events in total).
-Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-In addition to finding differences in mean activity for these dyads, the model learns that events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
-\label{fig:posteriorparams}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
-\caption{Observed counts}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
-\caption{Intercept estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
-\caption{\texttt{ab-ba} estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
-\caption{\texttt{ab-by} estimates}
-\end{subfigure}
-\caption{Comparing observed counts and parameter estimates.
-Darker values are larger.
-Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
-\label{fig:parmats}
-\end{figure}
-
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}). 
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
-
-%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-
-%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
-Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
-
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.  
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.  
-We also note that we improve over the situation where we constrain $K=1$.  
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.  
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
-
-Trivially extended to co-clustering applications and for directed data.  
-Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
-
-Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.  
-For example, dynamics might change over time \cite{Vu2011}.  
-Alternatively, if nodes change groups.  
-One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.  
-In some contexts such extensions might be substantively important and warrant future work.
-
-\subsubsection*{Acknowledgements}
-
-\bibliographystyle{unsrt}
-\bibliography{refs}
-
-\appendix{\textbf{Complexity}}
-
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as 
-\begin{align*}
-\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
-\end{align*}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
-
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
-
-\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
-
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-This is a two-stage MH procedure.
-First, pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
-\begin{itemize}
-\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
-\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
-where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
- If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
-\begin{itemize}
-\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
-  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
-\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
-\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
- The Metropolis-Hastings acceptance probabilities require the quantity
-$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
-
-The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
-Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
-The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
-$q(\boldsymbol{\phi}^{s} \rightarrow
-\boldsymbol{\phi}^{m})$ are products of Normal densities.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
-
-Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
-This increases the acceptance probability for splits (and decreases the probability of merges).
-However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
-
-\spewnotes
-
-\end{document}

### 2012-05-30 docs/intro.tex
Split draft into different tex files.

--- 2012-05-30 docs/submission.tex
@@
-\documentclass{article}
-\usepackage{nips12submit_e,times}
-\usepackage{tikz}
-\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-\usepackage{verbatim}
-\usepackage{url}
-\usepackage{graphicx, color}
-\usepackage{amsmath}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{footnote}
-
-\title{Stochastic Blockmodels for Relational Event Data}
-% Latent Variable Modeling of Relational Event Dynamics
-% Latent Variable Modeling of Relational Event Sequences
-% Latent Variable Modeling for Relational Event Sequences
-
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-}
-
-\begin{document}
-
-\maketitle
-
-\savenotes
-
-\begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.
-For continuous-time network data, a number of approaches have recently been proposed for modeling the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
-For static networks, methods such as stochastic blockmodels have traditionally been used for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
-We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
-The method is illustrated using dyadic interaction data such as email communication and Twitter direct messages.
-Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
-The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-\end{abstract}
-
 \section{Introduction}
@@
 
-%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
+% TODO: Mention dynamic stochastic equivalence
+
+% TODO: Incorporate notes from joint meeting:
+% - used to thinking about actors differentiated by rates
+% - other things in the dynamics than rates, e.g. form of interaction.
+% - Also important aspect of modeling social dynamics.
+% especially important in cases like plot 1(c) where classical methods
+% will fail.
+
+\begin{figure}
+
+\caption{}
+\label{fig:example}
+\end{figure}
 
@@
 Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, allowing identification of particular \emph{roles} in communication.
-
-\section{Model}
-
-%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
-We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
-Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.
-If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
-&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
-\end{align}
-%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
-
-% \begin{align}
-% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
-% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
-% \end{align}
-
-One can extend the above to marked point processes where each event contains additional information.
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
-The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
-\label{eqn:llk}
-\end{align}
-%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
-
-\begin{figure}
-\centering
-\begin{minipage}[b]{0.45\linewidth}
-  \begin{tikzpicture}[scale=.8,transform shape,thick]
-    \node[vertex] (n1) at (0,0)  {1};
-    \node[vertex] (n2) at (2,2)  {2};
-    \node[vertex] (n3) at (4,0) {3};
-    \node[vertex] (n4) at (2,-2) {4};
-    \draw[-stealth] (.4,.4) -- (1.6,1.6);
-    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
-    \node at (.7,1.1) {$t_1$};
-    \node at (3.3,1.1) {$t_2$};
-  \end{tikzpicture}
-  \caption{Dynamic network data}
-\end{minipage}
-\begin{minipage}[b]{0.45\linewidth}
-\begin{tikzpicture}[scale=1,transform shape,thick]
-\draw[-stealth] (0,0) -- (0,3);
-\draw[-stealth] (0,0) -- (5,0);
-
-% First timeline
-\draw (0,2) -- (1,2);
-\draw[fill=black] (1,2) circle (0.75mm);
-\draw (1,1.7) circle (0.75mm);
-\draw (1.08,1.7) -- (3,1.7);
-\draw[fill=black] (3,1.7) circle (0.75mm);
-\draw (3,1.7) circle (0.75mm);
-\draw (3,2.2) circle (0.75mm);
-\draw[-stealth] (3.05,2.2) -- (5,2.2);
-
-% second timeline
-\draw[dashed] (0,1) -- (3,1);
-\draw[fill=black] (3,1) circle (0.75mm);
-\draw (3,.75) circle (0.75mm);
-\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
-
-% labels
-\node at (-.5,2) {$\lambda_{1,2}$};
-\node at (-.5,1) {$\lambda_{3,4}$};
-\node at (1,-.3) {$t_1$};
-\node at (3,-.3) {$t_2$};
-\end{tikzpicture}
-\caption{Intensities for two dyads}
-\end{minipage}
-\caption[]{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
-\label{fig:example}
-\end{figure}
-
-\footnotetext{TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-
-% \begin{figure}
-%  \def\svgwidth{6in}
-%   \input{example1.pdf_tex}
-% \caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-% \label{fig:example}
-% %\includegraphics[width=6in]{example}
-% \end{figure}
-
-%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
-
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
-\begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
-%z_i &\sim \mbox{CRP}(\alpha) &
-%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
-\end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-
- In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.
-This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
-In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
-
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.
-The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
-
-\subsection{Model specification}
-
-\begin{table}[t]
-\footnotesize
-\center
-\begin{tabular}{|l|l|}
-\hline
-Statistic & Formula \\
-\hline
-\hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-\hline
-\end{tabular}
-\label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
-\end{table}
-
-In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$).
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.
-For example, a particular node sending often may indicate they will continue to be a sender in the future.
-We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
-TODO: Explore the theory about these sorts of statistics.
-Aalen has a few suggestions.}
-Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
-
-The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
-For example, an AB-BA effect indicates an increased propensity for reciprocity.
-The final set of effects allow one to model phenomenon such as triadic closure.
-Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
-
-\subsection{Relation to other models}
-
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
-In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
-
-The above framework generalizes several important special cases.
-For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.
-Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
-As these intensities do not change under this specification, the likelihood simplifies to
-$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-
-Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
-\begin{align}
-\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
-\label{eqn:multllk}
-\end{align}
-The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
-\section{Inference}
-
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
-
-\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
-
-We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
-\begin{align*}
-p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-\end{align*}
-where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
-Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
-This aims for similarly sized groups.}
-
-\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
-
-For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
-\begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
-p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]}
-\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-\end{align*}
-where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
-This distribution is sampled using slice sampling.
-
-\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
-
-The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
-\begin{align}
-\label{eqn:gibbs.sigma}
-\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
- \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
-\end{align}
-Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$,
-\begin{align}
-\label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
-\end{align}
-
-\begin{figure}
-\center
-\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
-\label{fig:syncounts}
-\end{figure}
-
-\section{Simulation}
-
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
-This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
-
-Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
-Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
-
-\section{Model fitting and experiments}
-
-\subsection{Data}
-
-Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
-
-\begin{itemize}
-%\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
-\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
-\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
-We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
-Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
-We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
-\end{itemize}
-
-\subsection{Model-based exploratory analysis}
-
-Figure \ref{fig:posteriorparams} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
-We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
-For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
-The two panels on the left show the variation in $s_0(t_M,i,j)$, $s_1(t_M,i,j)$ and $s_3(t_M,i,j)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
-The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
-Similarly, block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
-The model has learned that this structure exists in the data and our inferences from the parameter estimates are sensible given the observed statistics.
-
-\subsection{Prediction experiments}
-
-We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
-We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
-
-In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.
-At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
-
-\input{../figs/results-llk.tex}
-
-\subsection{Baselines}
-
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
-Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
-
-Our method jointly models \emph{which} dyads occur and \emph{when} they occur.
-To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
-
-\input{../figs/results-recall.tex}
-
-% \begin{figure}[t]
-% \center
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-% \label{fig:recall}
-% \end{figure}
-
-\subsection{Results}
-
-TODO: Quickly describe main take-aways from the results tables.
-
-% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
-For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
-Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.
-
-%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
-This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
-\footnote{Will need to discuss results for other datasets once they are complete.}
-
-\section{Discussion}
-
-%% TODO: Discuss scalability.
-%% TODO: Describe and include posterior predictive checks.
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
-%\caption{Observed counts}
-\end{subfigure}
-\qquad
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
-%\caption{Samples of $\beta_{p,k,l}$}
-\end{subfigure}
-
-\caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions between  members of group 1 and group 3.
-Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-In addition to finding differences in mean activity for these dyads, estimates suggest events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
-\label{fig:posteriorparams}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
-\caption{Observed counts}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
-\caption{Intercept estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
-\caption{\texttt{ab-ba} estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
-\caption{\texttt{ab-by} estimates}
-\end{subfigure}
-\caption{Comparing observed counts and parameter estimates.
-Darker values are larger.
-Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
-\label{fig:parmats}
-\end{figure}
-
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}).
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
-
-%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-
-%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
-Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
-
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
-We also note that we improve over the situation where we constrain $K=1$.
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
-
-Trivially extended to co-clustering applications and for directed data.
-Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
-
-Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.
-For example, dynamics might change over time \cite{Vu2011}.
-Alternatively, if nodes change groups.
-One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.
-In some contexts such extensions might be substantively important and warrant future work.
-
-%\subsubsection*{Acknowledgements}
-
-\bibliographystyle{unsrt}
-\bibliography{refs}
-
-\appendix{\textbf{Complexity}}
-
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as
-\begin{align*}
-\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
-\end{align*}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
-
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
-
-\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
-
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-This is a two-stage MH procedure.
-First, pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
-\begin{itemize}
-\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
-\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
-where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
- If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
-\begin{itemize}
-\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
-  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
-\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
-\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
- The Metropolis-Hastings acceptance probabilities require the quantity
-$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
-
-The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
-Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
-The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
-$q(\boldsymbol{\phi}^{s} \rightarrow
-\boldsymbol{\phi}^{m})$ are products of Normal densities.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
-
-Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
-This increases the acceptance probability for splits (and decreases the probability of merges).
-However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
-
-\spewnotes
-
-\end{document}

### 2012-05-30 docs/model.tex
Split draft into different tex files.

--- 2012-05-29 docs/submission.tex
@@
-\documentclass{article}
-\usepackage{nips12submit_e,times}
-\usepackage{tikz}
-\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-\usepackage{verbatim}
-\usepackage{url}
-\usepackage{graphicx, color}
-\usepackage{amsmath}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{footnote}
-
-\title{Stochastic Blockmodels for Relational Event Data}
-
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-}
-
-\begin{document}
-
-\maketitle
-
-\savenotes
-
-\begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.  
-For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
-For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
-We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  
-The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
-Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.  
-The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.  
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-\end{abstract}
-
-\section{Introduction}
-
-Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
-% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
-% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
-
-Network data, however, is often collected as a sequence of events occurring over time.   
-Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify how the process depends on the previous history of events.  
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
-
-The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
-In many cases this is unrealistic.  
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
-
-%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
-
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
-The behavior of the model is illustrated with simulated data.  
-We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
-Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
-
 \section{Model}
@@
 The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
-\section{Inference}
-
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
-
-\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
-
-We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
-\begin{align*}
-p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
-\end{align*} 
-where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.  
-Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.  
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
-This aims for similarly sized groups.}
-
-\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
-
-For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
-\begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
-p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
-\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-\end{align*}
-where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
-This distribution is sampled using slice sampling.
-
-\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
-
-The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
-\begin{align}
-\label{eqn:gibbs.sigma}
-\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
- \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
-\end{align}
-Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$, 
-\begin{align}
-\label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
-\end{align}
-
-\begin{figure}
-\center
-\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
-\label{fig:syncounts}
-\end{figure}
-
-\section{Simulation}
-
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.  
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.  
-This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
-
-Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
-Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
-
-\section{Model fitting and experiments}
-
-\subsection{Data}
-
-Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
-
-\begin{itemize}
-%\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
-\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
-\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
-We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
-Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  
-We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
-\end{itemize}
-
-\subsection{Exploratory analysis}
-
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we
-
-\subsection{Prediction experiments}
-
-We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.  
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
-We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.  
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
-
-In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.  
-At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
-
-\input{../figs/results-llk.tex}
-
-\subsection{Baselines}
-
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
-Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
-
-Our method jointly models \emph{which} dyads occur and \emph{when} they occur.  
-To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
-
-\input{../figs/results-recall.tex}
-
-% \begin{figure}[t]
-% \center
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-% \label{fig:recall}
-% \end{figure}
-
-\subsection{Results}
-
-TODO: Quickly describe main take-aways from the results tables.
-
-% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
-For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
-Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.
-
-%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
-This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
-\footnote{Will need to discuss results for other datasets once they are complete.}
-
-\section{Discussion}
-
-%% TODO: Discuss scalability.
-%% TODO: Describe and include posterior predictive checks.
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
-%\caption{Observed counts}
-\end{subfigure}
-\qquad
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
-%\caption{Samples of $\beta_{p,k,l}$}
-\end{subfigure}
-
-\caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions from members of group 1 to group 3 (178 total events in total) and from members of group 3 to group 1 (179 events in total).
-Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-In addition to finding differences in mean activity for these dyads, the model learns that events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
-\label{fig:posteriorparams}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
-\caption{Observed counts}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
-\caption{Intercept estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
-\caption{\texttt{ab-ba} estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
-\caption{\texttt{ab-by} estimates}
-\end{subfigure}
-\caption{Comparing observed counts and parameter estimates.
-Darker values are larger.
-Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
-\label{fig:parmats}
-\end{figure}
-
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}). 
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
-
-%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-
-%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
-Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
-
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.  
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.  
-We also note that we improve over the situation where we constrain $K=1$.  
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.  
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
-
-Trivially extended to co-clustering applications and for directed data.  
-Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
-
-Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.  
-For example, dynamics might change over time \cite{Vu2011}.  
-Alternatively, if nodes change groups.  
-One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.  
-In some contexts such extensions might be substantively important and warrant future work.
-
-\subsubsection*{Acknowledgements}
-
-\bibliographystyle{unsrt}
-\bibliography{refs}
-
-\appendix{\textbf{Complexity}}
-
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as 
-\begin{align*}
-\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
-\end{align*}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
-
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
-
-\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
-
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-This is a two-stage MH procedure.
-First, pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
-\begin{itemize}
-\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
-\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
-where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
- If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
-\begin{itemize}
-\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
-  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
-\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
-\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
- The Metropolis-Hastings acceptance probabilities require the quantity
-$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
-
-The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
-Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
-The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
-$q(\boldsymbol{\phi}^{s} \rightarrow
-\boldsymbol{\phi}^{m})$ are products of Normal densities.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
-
-Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
-This increases the acceptance probability for splits (and decreases the probability of merges).
-However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
-
-\spewnotes
-
-\end{document}

### 2012-05-30 docs/simulation.tex
Split draft into different tex files.

--- 2012-05-29 docs/submission.tex
@@
-\documentclass{article}
-\usepackage{nips12submit_e,times}
-\usepackage{tikz}
-\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
-\usepackage{verbatim}
-\usepackage{url}
-\usepackage{graphicx, color}
-\usepackage{amsmath}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{footnote}
-
-\title{Stochastic Blockmodels for Relational Event Data}
-
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-}
-
-\begin{document}
-
-\maketitle
-
-\savenotes
-
-\begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.  
-For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
-For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
-We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  
-The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
-Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.  
-The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.  
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-\end{abstract}
-
-\section{Introduction}
-
-Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
-% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
-% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
-
-Network data, however, is often collected as a sequence of events occurring over time.   
-Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify how the process depends on the previous history of events.  
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
-
-The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
-In many cases this is unrealistic.  
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
-
-%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
-
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
-The behavior of the model is illustrated with simulated data.  
-We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
-Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
-
-\section{Model}
-
-%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
-We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
-Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.  
-If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
-&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
-\end{align}
-%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
-
-% \begin{align}
-% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
-% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
-% \end{align}
-
-One can extend the above to marked point processes where each event contains additional information. 
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
-The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
-\label{eqn:llk}
-\end{align}
-%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
-
-\begin{figure}
-\centering 
-\begin{minipage}[b]{0.45\linewidth}
-  \begin{tikzpicture}[scale=.8,transform shape,thick]
-    \node[vertex] (n1) at (0,0)  {1};
-    \node[vertex] (n2) at (2,2)  {2};
-    \node[vertex] (n3) at (4,0) {3};
-    \node[vertex] (n4) at (2,-2) {4};
-    \draw[-stealth] (.4,.4) -- (1.6,1.6);
-    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
-    \node at (.7,1.1) {$t_1$};
-    \node at (3.3,1.1) {$t_2$};
-  \end{tikzpicture}
-  \caption{Dynamic network data}
-\end{minipage}
-\begin{minipage}[b]{0.45\linewidth}
-\begin{tikzpicture}[scale=1,transform shape,thick]
-\draw[-stealth] (0,0) -- (0,3);
-\draw[-stealth] (0,0) -- (5,0);
-
-% First timeline
-\draw (0,2) -- (1,2);
-\draw[fill=black] (1,2) circle (0.75mm);
-\draw (1,1.7) circle (0.75mm);
-\draw (1.08,1.7) -- (3,1.7);
-\draw[fill=black] (3,1.7) circle (0.75mm);
-\draw (3,1.7) circle (0.75mm);
-\draw (3,2.2) circle (0.75mm);
-\draw[-stealth] (3.05,2.2) -- (5,2.2);
-
-% second timeline
-\draw[dashed] (0,1) -- (3,1);
-\draw[fill=black] (3,1) circle (0.75mm);
-\draw (3,.75) circle (0.75mm);
-\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
-
-% labels
-\node at (-.5,2) {$\lambda_{1,2}$};
-\node at (-.5,1) {$\lambda_{3,4}$};
-\node at (1,-.3) {$t_1$};
-\node at (3,-.3) {$t_2$};
-\end{tikzpicture}
-\caption{Intensities for two dyads}
-\end{minipage}
-\caption[]{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
-\label{fig:example}
-\end{figure}
-
-\footnotetext{TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-
-% \begin{figure}
-%  \def\svgwidth{6in}
-%   \input{example1.pdf_tex}
-% \caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-% \label{fig:example}
-% %\includegraphics[width=6in]{example}
-% \end{figure}
-
-%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
-
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
-\begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
-%z_i &\sim \mbox{CRP}(\alpha) &
-%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
-\end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-
- In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.  
-This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.  
-In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
-
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.  
-The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
-
-\subsection{Model specification}
-
-\begin{table}[t]
-\footnotesize
-\center
-\begin{tabular}{|l|l|}
-\hline
-Statistic & Formula \\
-\hline
-\hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-\hline
-\end{tabular}
-\label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
-\end{table}
-
-In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$). 
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.  
-For example, a particular node sending often may indicate they will continue to be a sender in the future.  
-We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
-TODO: Explore the theory about these sorts of statistics.
-Aalen has a few suggestions.}
-Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
-
-The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.  
-For example, an AB-BA effect indicates an increased propensity for reciprocity.  
-The final set of effects allow one to model phenomenon such as triadic closure.  
-Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.  
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
-
-\subsection{Relation to other models}
-
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
-In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
-
-The above framework generalizes several important special cases.  
-For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  
-Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  
-As these intensities do not change under this specification, the likelihood simplifies to 
-$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-
-Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
-\begin{align}
-\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
-\label{eqn:multllk}
-\end{align}
-The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
-\section{Inference}
-
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
-
-\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
-
-We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
-\begin{align*}
-p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
-\end{align*} 
-where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.  
-Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.  
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
-This aims for similarly sized groups.}
-
-\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
-
-For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
-\begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
-p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
-\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-\end{align*}
-where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
-This distribution is sampled using slice sampling.
-
-\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
-
-The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
-\begin{align}
-\label{eqn:gibbs.sigma}
-\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
- \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
-\end{align}
-Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$, 
-\begin{align}
-\label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
-\end{align}
-
-\begin{figure}
-\center
-\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
-\label{fig:syncounts}
-\end{figure}
-
 \section{Simulation}
@@
 Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
-
-\section{Model fitting and experiments}
-
-\subsection{Data}
-
-Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
-
-\begin{itemize}
-%\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
-\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
-\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
-We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
-Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  
-We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
-\end{itemize}
-
-\subsection{Exploratory analysis}
-
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we
-
-\subsection{Prediction experiments}
-
-We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.  
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
-We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.  
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
-
-In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.  
-At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
-
-\input{../figs/results-llk.tex}
-
-\subsection{Baselines}
-
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
-Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
-
-Our method jointly models \emph{which} dyads occur and \emph{when} they occur.  
-To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
-
-\input{../figs/results-recall.tex}
-
-% \begin{figure}[t]
-% \center
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-% \label{fig:recall}
-% \end{figure}
-
-\subsection{Results}
-
-TODO: Quickly describe main take-aways from the results tables.
-
-% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
-For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
-Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.
-
-%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
-This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
-\footnote{Will need to discuss results for other datasets once they are complete.}
-
-\section{Discussion}
-
-%% TODO: Discuss scalability.
-%% TODO: Describe and include posterior predictive checks.
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
-%\caption{Observed counts}
-\end{subfigure}
-\qquad
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
-%\caption{Samples of $\beta_{p,k,l}$}
-\end{subfigure}
-
-\caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions from members of group 1 to group 3 (178 total events in total) and from members of group 3 to group 1 (179 events in total).
-Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-In addition to finding differences in mean activity for these dyads, the model learns that events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
-\label{fig:posteriorparams}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
-\caption{Observed counts}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
-\caption{Intercept estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
-\caption{\texttt{ab-ba} estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
-\caption{\texttt{ab-by} estimates}
-\end{subfigure}
-\caption{Comparing observed counts and parameter estimates.
-Darker values are larger.
-Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
-\label{fig:parmats}
-\end{figure}
-
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}). 
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
-
-%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-
-%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
-Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
-
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.  
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.  
-We also note that we improve over the situation where we constrain $K=1$.  
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.  
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
-
-Trivially extended to co-clustering applications and for directed data.  
-Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
-
-Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.  
-For example, dynamics might change over time \cite{Vu2011}.  
-Alternatively, if nodes change groups.  
-One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.  
-In some contexts such extensions might be substantively important and warrant future work.
-
-\subsubsection*{Acknowledgements}
-
-\bibliographystyle{unsrt}
-\bibliography{refs}
-
-\appendix{\textbf{Complexity}}
-
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as 
-\begin{align*}
-\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
-\end{align*}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
-
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
-
-\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
-
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-This is a two-stage MH procedure.
-First, pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
-\begin{itemize}
-\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
-\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
-where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
- If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
-\begin{itemize}
-\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
-  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
-\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
-\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
- The Metropolis-Hastings acceptance probabilities require the quantity
-$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
-
-The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
-Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
-The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
-$q(\boldsymbol{\phi}^{s} \rightarrow
-\boldsymbol{\phi}^{m})$ are products of Normal densities.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
-
-Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
-This increases the acceptance probability for splits (and decreases the probability of merges).
-However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
-
-\spewnotes
-
-\end{document}

### 2012-05-30 docs/submission.tex
Split draft into different tex files.

--- 2012-05-29 docs/submission.tex
@@
 \begin{document}
-
 \maketitle
-
 \savenotes
-
-\begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.  
-For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
-For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
-We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  
-The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
-Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.  
-The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.  
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
-\end{abstract}
-
-\section{Introduction}
-
-Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
-% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
-% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
-
-Network data, however, is often collected as a sequence of events occurring over time.   
-Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify how the process depends on the previous history of events.  
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
-
-The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
-In many cases this is unrealistic.  
-For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
-
-%, and in many cases emails between one pair of individuals will not affect emailing behavior between a separate pair of individuals.
-
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
-The behavior of the model is illustrated with simulated data.  
-We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
-Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.
-
-\section{Model}
-
-%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
-We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
-Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.  
-If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
-&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
-\end{align}
-%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
-
-% \begin{align}
-% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
-% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
-% \end{align}
-
-One can extend the above to marked point processes where each event contains additional information. 
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
-The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
-\begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
-\label{eqn:llk}
-\end{align}
-%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
-
-\begin{figure}
-\centering 
-\begin{minipage}[b]{0.45\linewidth}
-  \begin{tikzpicture}[scale=.8,transform shape,thick]
-    \node[vertex] (n1) at (0,0)  {1};
-    \node[vertex] (n2) at (2,2)  {2};
-    \node[vertex] (n3) at (4,0) {3};
-    \node[vertex] (n4) at (2,-2) {4};
-    \draw[-stealth] (.4,.4) -- (1.6,1.6);
-    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
-    \node at (.7,1.1) {$t_1$};
-    \node at (3.3,1.1) {$t_2$};
-  \end{tikzpicture}
-  \caption{Dynamic network data}
-\end{minipage}
-\begin{minipage}[b]{0.45\linewidth}
-\begin{tikzpicture}[scale=1,transform shape,thick]
-\draw[-stealth] (0,0) -- (0,3);
-\draw[-stealth] (0,0) -- (5,0);
-
-% First timeline
-\draw (0,2) -- (1,2);
-\draw[fill=black] (1,2) circle (0.75mm);
-\draw (1,1.7) circle (0.75mm);
-\draw (1.08,1.7) -- (3,1.7);
-\draw[fill=black] (3,1.7) circle (0.75mm);
-\draw (3,1.7) circle (0.75mm);
-\draw (3,2.2) circle (0.75mm);
-\draw[-stealth] (3.05,2.2) -- (5,2.2);
-
-% second timeline
-\draw[dashed] (0,1) -- (3,1);
-\draw[fill=black] (3,1) circle (0.75mm);
-\draw (3,.75) circle (0.75mm);
-\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
-
-% labels
-\node at (-.5,2) {$\lambda_{1,2}$};
-\node at (-.5,1) {$\lambda_{3,4}$};
-\node at (1,-.3) {$t_1$};
-\node at (3,-.3) {$t_2$};
-\end{tikzpicture}
-\caption{Intensities for two dyads}
-\end{minipage}
-\caption[]{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
-\label{fig:example}
-\end{figure}
-
-\footnotetext{TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-
-% \begin{figure}
-%  \def\svgwidth{6in}
-%   \input{example1.pdf_tex}
-% \caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-% \label{fig:example}
-% %\includegraphics[width=6in]{example}
-% \end{figure}
-
-%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
-
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
-\begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
-%z_i &\sim \mbox{CRP}(\alpha) &
-%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
-\end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
-
- In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.  
-This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.  
-In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
-
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.  
-The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
-
-\subsection{Model specification}
-
-\begin{table}[t]
-\footnotesize
-\center
-\begin{tabular}{|l|l|}
-\hline
-Statistic & Formula \\
-\hline
-\hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-\hline
-\end{tabular}
-\label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
-\end{table}
-
-In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.  
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$). 
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.  
-For example, a particular node sending often may indicate they will continue to be a sender in the future.  
-We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
-TODO: Explore the theory about these sorts of statistics.
-Aalen has a few suggestions.}
-Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
-
-The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.  
-For example, an AB-BA effect indicates an increased propensity for reciprocity.  
-The final set of effects allow one to model phenomenon such as triadic closure.  
-Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.  
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
-
-\subsection{Relation to other models}
-
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
-In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
-
-The above framework generalizes several important special cases.  
-For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  
-Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  
-As these intensities do not change under this specification, the likelihood simplifies to 
-$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-
-Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
-\begin{align}
-\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
-\label{eqn:multllk}
-\end{align}
-The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
-
-\section{Inference}
-
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
-
-\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
-
-We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
-\begin{align*}
-p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
-\end{align*} 
-where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.  
-Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.  
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
-This aims for similarly sized groups.}
-
-\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
-
-For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
-\begin{align*}
-p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
-p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
-p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
-\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
-\end{align*}
-where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
-This distribution is sampled using slice sampling.
-
-\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
-
-The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
-\begin{align}
-\label{eqn:gibbs.sigma}
-\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
- \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
-\end{align}
-Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$, 
-\begin{align}
-\label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
-\end{align}
-
-\begin{figure}
-\center
-\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
-\label{fig:syncounts}
-\end{figure}
-
-\section{Simulation}
-
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.  
-For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.  
-This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
-
-Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
-Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
-Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
-
-\section{Model fitting and experiments}
-
-\subsection{Data}
-
-Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
-
-\begin{itemize}
-%\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
-\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
-\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
-We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
-Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  
-We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
-\end{itemize}
-
-\subsection{Exploratory analysis}
-
-Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we
-
-\subsection{Prediction experiments}
-
-We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.  
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
-We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.  
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
-
-In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.  
-At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
-
-\input{../figs/results-llk.tex}
-
-\subsection{Baselines}
-
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
-Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
-
-Our method jointly models \emph{which} dyads occur and \emph{when} they occur.  
-To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.
-
-\input{../figs/results-recall.tex}
-
-% \begin{figure}[t]
-% \center
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-% \caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
-% \label{fig:recall}
-% \end{figure}
-
-\subsection{Results}
-
-TODO: Quickly describe main take-aways from the results tables.
-
-% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.
-For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.
-Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.
-
-%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.
-This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).
-\footnote{Will need to discuss results for other datasets once they are complete.}
-
-\section{Discussion}
-
-%% TODO: Discuss scalability.
-%% TODO: Describe and include posterior predictive checks.
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
-%\caption{Observed counts}
-\end{subfigure}
-\qquad
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
-%\caption{Samples of $\beta_{p,k,l}$}
-\end{subfigure}
-
-\caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions from members of group 1 to group 3 (178 total events in total) and from members of group 3 to group 1 (179 events in total).
-Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
-In addition to finding differences in mean activity for these dyads, the model learns that events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
-\label{fig:posteriorparams}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
-\caption{Observed counts}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
-\caption{Intercept estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
-\caption{\texttt{ab-ba} estimates}
-\end{subfigure}
-~
-\begin{subfigure}[b]{0.22\textwidth}
-\centering
-\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
-\caption{\texttt{ab-by} estimates}
-\end{subfigure}
-\caption{Comparing observed counts and parameter estimates.
-Darker values are larger.
-Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
-\label{fig:parmats}
-\end{figure}
-
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}). 
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
-
-%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-
-%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
-Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
-
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.  
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.  
-We also note that we improve over the situation where we constrain $K=1$.  
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.  
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
-
-Trivially extended to co-clustering applications and for directed data.  
-Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
-
-Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.  
-For example, dynamics might change over time \cite{Vu2011}.  
-Alternatively, if nodes change groups.  
-One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.  
-In some contexts such extensions might be substantively important and warrant future work.
-
-\subsubsection*{Acknowledgements}
-
+\input{abstract}
+\input{intro}
+\input{model}
+\input{inference}
+\input{simulation}
+\input{experiments}
+\input{discussion}
+%\subsubsection*{Acknowledgements}
 \bibliographystyle{unsrt}
 \bibliography{refs}
-
-\appendix{\textbf{Complexity}}
-
-We take advantage of our restriction on the types of statistics $s$ to reduce the computational complexity of computing our likelihood Equation \ref{eqn:likelihood} as 
-\begin{align*}
-\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | \cdot) \}
-\end{align*}
-\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs  \cite{Butts2008}.
-\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
-
- By limiting the changepoints to times when either node is involved, computing the likelihood $p(A|z,\beta,\mbox{node} \ r \ \mbox{involved})$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$.
-We precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j)$ for all $m$, $i$, and $j$.
-
-\appendix{\textbf{Split-merge Metropolis-Hastings move for non-conjugate stochastic blockmodels}}
-
-%Motivation: simply Gibbs sampling does not seem to mix well.
-\footnote{This section is planned future work.
-In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.
-For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}
-
-Here we describe a novel split-merge move for non-conjugate blockmodels.\footnote{IRM (modeling Bernoulli data with a conjugate Beta prior) used a split-merge algorithm \cite{Kempe2006}.}  For illustration, we consider the model
-\begin{align*}
-  y_{ij} & \sim \mbox{Poisson}(\Lambda_{z_i,z_j}) & \Lambda_{ij} &\sim \mbox{Gamma}(\alpha_{\Lambda},\beta_{\Lambda})  & z_i &\sim \mbox{CRP}(\gamma)
-\end{align*}
-
-IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).
-A proposed split has REM parameters that are only slightly different than those for the original block.
-Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.
-By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.
-
-This is a two-stage MH procedure.
-First, pick two nodes $i,j$ uniformly.
-If they belong to the same block $k$ (i.e.
-$z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
-\begin{itemize}
-\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$
-\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
-where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
-\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
-\end{itemize}
-
- If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e.
-$z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
-\begin{itemize}
-\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
-  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
-\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
-\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
-\end{itemize}
-
- The Metropolis-Hastings acceptance probabilities require the quantity
-$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$
-
-The first term is the likelihood ratio of the two states.
-Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.
-Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.
-The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and
-$q(\boldsymbol{\phi}^{s} \rightarrow
-\boldsymbol{\phi}^{m})$ are products of Normal densities.
-\footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.
-
-Since the proposed split vectors $\phi^s$ are close the $\phi^m$, this has a higher probability than the split states having arisen from the proposed merge state, which is closer to the midpoint between the two $\phi^s$ vectors.
-This increases the acceptance probability for splits (and decreases the probability of merges).
-However, the probability of getting assigned to each cluster (through the restricted Gibbs scan) might balance things out.
-
+\input{appendices}
 \spewnotes
-
 \end{document}

### 2012-05-30 docs/model.tex
Clean up model section.

@@
 
-%Each interaction then has a set of periods $d \in D_{ij}$ each with a start time $a_d$, an end time $b_d$, and a rate $\lambda_d$.
-%Consider a nonhomogeneous Poisson process with  intensity $\lambda(t)$ that is piecewise constant with respect to a set of knots $\tau$.
-We can then write the likelihood of $M$ events $\mathcal{A} = (t_1, \ldots, t_M)$ with $t_m > 0$ as
 Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.
@@
 \end{align}
-%\noindent where the $m$th event occurs at time $t_m$, the intensity function $\lambda(t)$ is left continuous, and each $\tau_k \in [0,t_M]$.
-
-% \begin{align}
-% \mathcal{L}(A|\theta) &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}}  \exp\left\{ - \int_{0}^{t_M} \lambda_{i}(\tau | \cdot)d\tau \right\} \\
-% &= \prod_{m=1}^M \lambda_{i_m}(t_m|\cdot) \prod_{i \in \mathcal{R}} \prod_{v \in D_{i}} \exp\left\{ - (b_v - a_v) \lambda_{i}(b_v | \cdot ) \right\}
-% \end{align}
 
@@
 \end{align}
-%\prod_{(i,j) \in \mathcal{R}}\exp\{ - (t - t_{M}) \lambda_{ij}(t | \cdot) \}.
 
@@
 
-\footnotetext{TODO?
-Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-
-% \begin{figure}
-%  \def\svgwidth{6in}
-%   \input{example1.pdf_tex}
-% \caption{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.
-Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
-% \label{fig:example}
-% %\includegraphics[width=6in]{example}
-% \end{figure}
-
-%This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.
-
 We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
@@
 In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.
-Several statistics use $v_{mij} \in [0,M]$, the index of the last event where $(i,j)$ occurred.  % (n.b. $t_{v_{mij}} = \tau_{mij}$).
-If $(i,j)$ has never occurred, we set $v_{mij}=0$.  We include degree effects to model possible rich-get-richer phenomenon.
 For example, a particular node sending often may indicate they will continue to be a sender in the future.
-We normalize these counts by the number of events up until a dyad's prior changepoint.\footnote{This seems to avoid ``explosion'' when simulating from the model.
+We normalize these counts by the number of events up until a dyad's prior changepoint.%\footnote{This seems to avoid ``explosion'' when simulating from the model.
 TODO: Explore the theory about these sorts of statistics.
@@
 The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
-For example, an AB-BA effect indicates an increased propensity for reciprocity.
-The final set of effects allow one to model phenomenon such as triadic closure.
-Though we use only the above statistics in our experiments, one may use any quantity or set of covariates about a given dyad that is computed using the previous history of events.
+For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
+Though we use only the above statistics in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
 The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}

### 2012-05-31 process/synthetic.r
Add figure for intro example.

@@ -124,3 +124,103 @@ ds <- sapply(1:niter,function(i) {
 ds <- melt(t(ds))
 qplot(X1,value,data=ds,geom="line",colour=factor(X2)) + theme_bw() + labs(x="iteration",colour="block",y="Euclidean distance to truth")
 ggsave("figs/syn/bias.pdf",width=5,height=4)
+
+
+## Alternative example
+N <- 10
+K <- 2
+beta <- list("intercept"=matrix(c(1.5,0,0,2),K,K),
+             "abba" = matrix(c(2,0,.5,2),K,K),
+             "abby"=matrix(0,K,K),
+             "abxa"=matrix(0,K,K),
+             "abxb"=matrix(0,K,K),
+             "abay"=matrix(0,K,K),
+             "abab"=matrix(0,K,K),
+             "sod"=matrix(0,K,K),
+             "rod"=matrix(0,K,K),
+             "sid"=matrix(0,K,K),
+             "rid"=matrix(0,K,K),
+             "dc" =matrix(0,K,K),
+             "cc" =matrix(0,K,K))
+P <- length(beta)
+beta <- abind(beta,rev.along=3)
+M <- 1000
+set.seed(1)
+z <- c(rep(1,5),rep(2,5))
+sim <- generate.brem(M,N,beta,z,ego=TRUE)
+y1 <- sim$edgelist
+
+beta <- list("intercept"=matrix(0,K,K),
+             "abba" = matrix(c(2,0,.5,2),K,K),
+             "abby"=matrix(0,K,K),
+             "abxa"=matrix(0,K,K),
+             "abxb"=matrix(0,K,K),
+             "abay"=matrix(0,K,K),
+             "abab"=matrix(0,K,K),
+             "sod"=matrix(0,K,K),
+             "rod"=matrix(0,K,K),
+             "sid"=matrix(0,K,K),
+             "rid"=matrix(0,K,K),
+             "dc" =matrix(0,K,K),
+             "cc" =matrix(0,K,K))
+beta <- abind(beta,rev.along=3)
+set.seed(1)
[…]

### 2012-05-31 docs/abstract.tex
Paper edits.  Intro reorg.

@@
 Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
-The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.
-Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
+The fitted models have better predictive accuracy than baselines and
+relational event models without the latent structure and a standard
+stochastic blockmodel approach.
+%Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
 \end{abstract}

### 2012-05-31 docs/discussion.tex
Paper edits.  Intro reorg.

@@
 
- We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
-The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}).
-This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
-
-%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.
-
-%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.
-Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.
+ We propose a hierarchical approach for modeling event-based network data as a continuous-time Markov process.
+Given a specification for the, the proposed method employs latent variables to model the heterogeneity in the event dynamics.
+%is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
+%The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}).
+%This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
 
@@
 In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
-We also note that we improve over the situation where we constrain $K=1$.
+We also note that we can improve over relational event models that lack latent clusters as well as stochastic blockmodels for count data, both special cases 
 Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.

### 2012-05-31 docs/intro.tex
Add figure for intro example.
Paper edits.  Intro reorg.

@@
 Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
-% Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
-% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels
+One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO] that is associated with some other quantity, e.g.
+a covariate such as age or sex, or perhaps shared [TODO].
+Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
 
-Network data, however, is often collected as a sequence of events occurring over time.
-Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-These models allow one to specify how the process depends on the previous history of events.
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
+If we are interested in studying the events in a network, there are other aspects of the dynamics in addition to the mean rate. 
+For example, human conversation often has an increased propensity for reciprocated events.
+Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
+[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
+Figure \ref{fig:introexample} shows how an analysis using a blockmodels will be unable to learn about these differences.
 
-A limitation of all of these approaches is the assumption of a single common behavior for all individuals.
-In contrast, in real-world networks it is reasonable to expect {\it heterogeneity} in dynamic behavior.  
-In an organization, the manner in which an individual communicates with others may be a function of an individual's {\it role} in the organization.
-For example, in a university,  email communication patterns over time between professors, students, and staff will likely be quite different within and between the three groups.
+%A limitation of all of these approaches is the assumption of a single common behavior for all individuals.
+%In contrast, in real-world networks it is reasonable to expect {\it heterogeneity} in dynamic behavior.  
+%In an organization, the manner in which an individual communicates with others may be a function of an individual's {\it role} in the organization.
+%For example, in a university,  email communication patterns over time between professors, students, and staff will likely be quite different within and between the three groups.
+
+% - used to thinking about actors differentiated by rates
+% - other things in the dynamics than rates, e.g. form of interaction.
+% - Also important aspect of modeling social dynamics.
+% especially important in cases like plot 1(c) where classical methods
+% will fail.
+
 %PS: not totally happy with this example since the "roles" are completely determined by one's "job title" and I think we would prefer something
@@
 
-% TODO: Incorporate notes from joint meeting:
-% - used to thinking about actors differentiated by rates
-% - other things in the dynamics than rates, e.g. form of interaction.
-% - Also important aspect of modeling social dynamics.
-% especially important in cases like plot 1(c) where classical methods
-% will fail.
+Borrowing from the intuition of stochastic blockmodels, we propose a continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
+Our approach employs a flexible framework for  specifing how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
+%extend continuous-time models from event history analysis to model sequences of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
+% \cite{AalenOddO.2008}
+%These models allow one to specify how the process depends on the previous history of events. % TODO use 'flexible'?
+In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
+
+We describe parameter estimation and learning of  latent cluster assignments via MCMC, and illustrate the behavior of the model  with simulated data.
+Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
+Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, enabling the study of a dynamic extension to stochastic equivalence.
+%allowing identification of particular \emph{roles} in  that enable
 
 \begin{figure}
-
+\centering
+\includegraphics[scale=.8]{../figs/introexample/all}
 \caption{}
@@
 \end{figure}
-
-Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data to capture node-level heterogeneity, where we assume that latent clusters of nodes share similar patterns of interaction.
-We describe parameter estimation and learning of  latent cluster assignments via MCMC, and illustrate the behavior of the model  with simulated data.
-Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, allowing identification of particular \emph{roles} in communication.

### 2012-05-31 docs/model.tex
Paper edits.  Intro reorg.

@@
 This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
-In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}
+In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
+%\footnote{TODO: This is analogous to a Markov condition for this class of models.}
 

### 2012-05-31 docs/submission.tex
Paper edits.  Intro reorg.

@@
 
-\title{Stochastic Blockmodels for Relational Event Data}
+\title{Stochastic blockmodeling of relational event dynamics}
+% Dynamic stochastic equivalence for relational events
+% Stochastic Blockmodels for Relational Event Data}
 % Latent Variable Modeling of Relational Event Dynamics
@@
 
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-}
+% \author{
+% Christopher DuBois
+% Department of Statistics\\
+% University of California, Irvine\\
+% \texttt{duboisc@ics.uci.edu} \\
+% \And
+% Coauthor \\
+% Affiliation \\
+% Address \\
+
+% \AND
+% Coauthor \\
+% Affiliation \\
+% Address \\
+% }
 

### 2012-05-31 docs/inference.tex
Add caption to intro figure.

@@
 Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
-This aims for similarly sized groups.}
 
@@
 \end{figure}
+
+\subsection{Hyperparameter settings}
+In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
+This aims for similarly sized groups.}
+We set $\alpha_{\sigma}=3$ and $\beta_{sigma}=1$ so that in the presence of little data we encourage shrinkage towards the upper level parameters $\mu$.  
+As with other models, we note that the predictive accuracy of the model can depend on the hyperparameters.

### 2012-05-31 docs/intro.tex
Add caption to intro figure.

@@
 
-% TODO: Mention dynamic stochastic equivalence
-
 Borrowing from the intuition of stochastic blockmodels, we propose a continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
@@
 \includegraphics[scale=.8]{../figs/introexample/all}
-\caption{}
+\caption{Illustration of differentiation by rate of interaction versus by dynamics.
+The left column shows an example where within-block communication is large, and there is a higher tendency for reciprocity within a block.
+In the situation where the two groups are not differentiated by rate, as in the right column, a standard blockmodel is unable to distinguish between groups A and B.
+The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.}
 \label{fig:example}

### 2012-05-31 docs/model.tex
Add caption to intro figure.

@@
 
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \sim N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(3,1)$.
+We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta{\sigma})$.
 The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.

### 2012-05-31 docs/discussion.tex
Small paper edits.

@@
 
-%% TODO: Discuss scalability.
-%% TODO: Describe and include posterior predictive checks.
+% \begin{figure}[t]
+% \centering
+% \begin{subfigure}[b]{0.35\textwidth}
+% \centering
+% \includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
+% %\caption{Observed counts}
+% \end{subfigure}
+% \qquad
+% \begin{subfigure}[b]{0.35\textwidth}
+% \centering
+% \includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
+% %\caption{Samples of $\beta_{p,k,l}$}
+% \end{subfigure}
 
-\begin{figure}[t]
-\centering
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-obs-stats}
-%\caption{Observed counts}
-\end{subfigure}
-\qquad
-\begin{subfigure}[b]{0.35\textwidth}
-\centering
-\includegraphics[scale=.5]{../figs/eckmann-small/example-estimates}
-%\caption{Samples of $\beta_{p,k,l}$}
-\end{subfigure}
-
-\caption{%Observed counts versus model estimates from Eckmann email data.
- Left: Observed statistics across all dyadic email interactions between  members of group 1 and group 3.
+% \caption{%Observed counts versus model estimates from Eckmann email data.
+%  Left: Observed statistics across all dyadic email interactions between  members of group 1 and group 3.
 Right: Posterior samples of the corresponding $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
 In addition to finding differences in mean activity for these dyads, estimates suggest events from 3 to 1 have a higher propensity for \texttt{ab-by} transitions.}
-\label{fig:posteriorparams}
-\end{figure}
+% \label{fig:posteriorparams}
+% \end{figure}
 

### 2012-05-31 docs/abstract.tex
Made some changes to the abstract.

@@
 \begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.
-For continuous-time network data, a number of approaches have recently been proposed for modeling the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
-For static networks, methods such as stochastic blockmodels have traditionally been used for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
-We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
-The method is illustrated using dyadic interaction data such as email communication and Twitter direct messages.
-Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
-The fitted models have better predictive accuracy than baselines and
-relational event models without the latent structure and a standard
-stochastic blockmodel approach.
+Observing network data as a stream of relational events provides an opportunity to understand differences in how individuals interact over time.
+For continuous-time network data, several approaches have recently been proposed for modeling dyadic event rates, conditioned on the observed history of events and nodal or dyadic covariates.
+In many cases, however, interaction propensities -- and even the underlying mechanisms of interaction -- vary systematically across subgroups whose identities are unobserved.
+For static networks, such heterogeneity has been treated via methods such as stochastic blockmodeling, which operate by assuming latent groups of individuals with similar tendencies in their group-wise interactions.
+Here, we combine these two approaches by positing a latent partition on the vertex set, such that event dynamics within and between node sets evolve in potentially distinct ways.
+We illustrate the use of our model family by application to several forms of dyadic interaction data, including email communication and Twitter direct messages.
+Parameter estimates from the fitted models clearly reveal heterogeneity in the dynamics among groups of individuals.
+We also find that the fitted models have better predictive accuracy than either baseline models or relational event models without latent structure.
+Our approach illustrates the utility of latent structure methods based on detailed dynamics, which can succeed even in the absence of differences in marginal interaction rates across groups. 
 %Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.

### 2012-05-31 docs/inference.tex
Small paper edits.
Fix up intro example figure.

@@
 where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
-This distribution is sampled using slice sampling.
+We sample each $\beta_{k,l,p}$ via slice sampling.
 
@@
 \sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
- \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
+ \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{k,l,p} - \mu_p)^2\right).
 \end{align}
@@
 \label{eqn:gibbs.mu}
-\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
+\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{k,l,p},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
 \end{align}
@@
 This aims for similarly sized groups.}
-We set $\alpha_{\sigma}=3$ and $\beta_{sigma}=1$ so that in the presence of little data we encourage shrinkage towards the upper level parameters $\mu$.  
+We set $\alpha_{\sigma}=3$ and $\beta_{\sigma}=1$ so that in the presence of little data we encourage shrinkage towards the upper level parameters $\mu$.  
 As with other models, we note that the predictive accuracy of the model can depend on the hyperparameters.

### 2012-05-31 process/synthetic.r

@@ -1,7 +1,6 @@
 
 library(ggplot2)
 library(brem)
-source("pkg/R/brem.r")
 
 N <- 10
 K <- 2
@@ -168,59 +167,64 @@ set.seed(1)
 sim <- generate.brem(M,N,beta,z,ego=TRUE)
 y2 <- sim$edgelist
 
+s <- 2.5
+stext <- 1.5
 cols <- rev(grey.colors(100,start=0,end=.95,gamma=1))
 tb <- table(factor(y1[,2],1:N),factor(y1[,3],1:N))
-pdf("figs/introexample/mat1.pdf",width=3,height=3)
-par(mar=c(2,2,2,2),mgp=c(1.15,0,0))# oma=c(0,0,0,0),
+pdf("figs/introexample/mat1.pdf",width=s,height=s)
+par(mar=c(2,2,2,2),mgp=c(1.1,0,0))# oma=c(0,0,0,0),
 tb <- as.matrix(tb)
-image(1:N,1:N,tb,xaxt="n",yaxt="n",col=cols,xlab="",ylab="Observed counts",axes=FALSE,cex=1.25)
+image(1:N,1:N,tb,xaxt="n",yaxt="n",col=cols,xlab="",ylab="Observed counts",axes=FALSE,cex=stext,main="Rate Differentiated Case")
 axis(1,at=c(3,8),labels=c("A","B"),tick=FALSE)
 axis(2,at=c(3,8),labels=c("A","B"),tick=FALSE)
 dev.off()
 
 tb <- table(factor(y2[,2],1:N),factor(y2[,3],1:N))
-pdf("figs/introexample/mat2.pdf",width=3,height=3)
+pdf("figs/introexample/mat2.pdf",width=s,height=s)
 par(mar=c(2,2,2,2),mgp=c(1.15,0,0))
-image(1:N,1:N,tb,xaxt="n",yaxt="n",xlab="",ylab="",col=cols,axes=FALSE,cex=1.25)
+image(1:N,1:N,tb,xaxt="n",yaxt="n",xlab="",ylab="",col=cols,axes=FALSE,cex=stext)
 axis(1,at=c(3,8),labels=c("A","B"),tick=FALSE)
 axis(2,at=c(3,8),labels=c("A","B"),tick=FALSE)
 dev.off()
 
[…]

### 2012-05-31 docs/intro.tex
Fix up intro example figure.
edited the intro (look for "PS" in the comments)

@@
 
-Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.
-One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO] that is associated with some other quantity, e.g.
+Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein interactions \cite{Goldenberg2009}.
+One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO], where this variation in rates may be associated with some unobserved quantity, e.g.
 a covariate such as age or sex, or perhaps shared [TODO].
 Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
+%PS: much of this past work is for static network data.
+Do we want to make the observation here that we are thinking about blockmodels for *rate* data here in our discussion and in our example in figure 1 - if we don't some readers/reviewers might get a little confused.
+For example we could say that instead of a binomial model on edges we can think of a Poisson process yielding counts over time, where we can still think about blockmodels.
+%PS: also, we introduce the term "rates" above without really defining it, and then go on to talk about "mean rate" below.
+So I think we definitely need a sentence somewhere above that makes it clear we are talking about edges that instantaneously appear and disappear over time at some rate, rather than static edges.
 
-If we are interested in studying the events in a network, there are other aspects of the dynamics in addition to the mean rate. 
+If we are interested in studying a network of events, there are other important aspects of the dynamics in addition to the mean rate. 
 For example, human conversation often has an increased propensity for reciprocated events.
+%PS: wording of next sentence is a little odd - not sure I understand what it means...
 Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
 [TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
-Figure \ref{fig:introexample} shows how an analysis using a blockmodels will be unable to learn about these differences.
+Figure \ref{fig:introexample} shows a simple example of how traditional blockmodels are unable to make inferences these types of dynamic phenomena.
 
@@
 Borrowing from the intuition of stochastic blockmodels, we propose a continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
-Our approach employs a flexible framework for  specifing how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
+Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
 %extend continuous-time models from event history analysis to model sequences of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
@@
 %These models allow one to specify how the process depends on the previous history of events. % TODO use 'flexible'?
-In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.
+In this way one can investigate theories about the underlying processes and make predictions about future data conditioned on the past.
 
@@
 
-\begin{figure}
+\begin{SCfigure}
 \centering
-\includegraphics[scale=.8]{../figs/introexample/all}
+\includegraphics[scale=.45]{../figs/introexample/all}
 \caption{Illustration of differentiation by rate of interaction versus by dynamics.
@@
 \label{fig:example}
-\end{figure}
+\end{SCfigure}

### 2012-05-31 docs/discussion.tex
edited the discussion section

@@
 
+%PS: I would prefer to say: In this paper, we propose (or we proposed)….
 We propose a hierarchical approach for modeling event-based network data as a continuous-time Markov process. 
-Given a specification for the, the proposed method employs latent variables to model the heterogeneity in the event dynamics.
+Our approach generalizes traditional static notions of stochastic equivalence on nodes (such as stochastic blockmodels) to a dynamic formulation. 
+The proposed dynamic model posits the existence of groups of nodes, such that all members of a group are governed by the same dynamic process, and groups are differentiated by having different dynamic processes.
+%PS: sentence below is incomplete (after the first "the")…..I decided to take it out anyway..
+% Given a specification for the, the proposed method employs latent variables to model the heterogeneity in the event dynamics.
 %is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
@@
 
-This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.
+%This analysis suggests 
+Our approach has the ability to uncover systematic differences in dynamic behaviors or roles among subsets of nodes.
+% across a variety of data sets.
 In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
-We also note that we can improve over relational event models that lack latent clusters as well as stochastic blockmodels for count data, both special cases 
+In particular, we show that our proposed approach leads to improved predictive accuracy when compared to both (a) relational event models that lack latent clusters, and (b) stochastic blockmodels for count data, both of which are both special cases of our model.
 Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
 
+%PS: not sure where this fits…seems out of place here…and needs to be rewritten to be clearer - its not clear why we are comparing to this model other than the fact that it appeared at NIPS last year and also talks about relational event data.
+One way to make its inclusion here more natural would to set this up by first stating that there are other modeling approaches for relational events (such as the Guna… work), and comment on whether our "block approach" could be applied there too (not sure this makes sense since the models are so different…).
 Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
@@
 
+%PS: these last two paragraphs need to be finished….
 Trivially extended to co-clustering applications and for directed data.

### 2012-05-31 docs/intro.tex
Outline new intro.

@@
 Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein interactions \cite{Goldenberg2009}.
+- observe network data over time
+- interested how nodes differ in the rate of activity with other parts of the network
+  - why?
+  - based on some other covariate
+- methods for static networks can be applied to the marginal rate
+  - model this heterogeneity by assuming subgroups of nodes have similar tendencies in their group-wise interactions.
+
+- Some of the questions of interest for dynamic data may not be answerable from the marginal rate
+- Example of what we'll call a type of dynamics
+- Example where the question cannot be answered (referring to Figure1 )
+- Drawing parallels with the blockmodels above, for these questions what we need is a dynamic version of stochastic equivalence
+
+When such .
 One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO], where this variation in rates may be associated with some unobserved quantity, e.g.
 a covariate such as age or sex, or perhaps shared [TODO].
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
+Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by .
+Such techniques are readily applied to count data, [TODO: These let us study how shared behavior is associated with other covariates.]
+
+%1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of a tie between members
+
 %PS: much of this past work is for static network data.

### 2012-05-31 docs/intro.tex
Small attempts at intro.

@@
 Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
-- observe network data over time
-- interested how nodes differ in the rate of activity with other parts of the network
-  - why?
-  - based on some other covariate
-- methods for static networks can be applied to the marginal rate
-  - model this heterogeneity by assuming subgroups of nodes have similar tendencies in their group-wise interactions.
+We often  observe events occurring between nodes over time, and a common goal is to understand the differences in the total number of interactions --- perhaps with respect to information known about the nodes.
+%In this case, we have a count for the total activity of each dyad, and one may analyze it has a static network. 
+%Methods for static networks such as a 
+Stochastic blockmodels \cite{} are a class of techniques which can model this heterogeneity by assuming latent subgroups of nodes have similar tendencies in their group-wise rate of interaction.
 
+Network-based event data is powerful, however, because we may study processes occurring at 
+However, other questions of int
 - Some of the questions of interest for dynamic data may not be answerable from the marginal rate

### 2012-05-31 docs/intro.tex
First few paragraphs in intro.

--- 2012-05-31 docs/intro.tex
@@
 Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
+We often  observe events occurring between nodes over time, and a common goal is to understand the differences in the total number of interactions --- perhaps with respect to information known about the nodes.
+%In this case, we have a count for the total activity of each dyad, and one may analyze it has a static network. 
+%Methods for static networks such as a 
+Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of techniques which can model this heterogeneity by assuming latent subgroups of nodes have similar tendencies in their group-wise rate of interaction.
+
+Network-based event data are powerful, however, because we enable the study of the processes governing these events.
+Theories about the dynamics among nodes will be often be difficult to study when only using the marginal rate of events as illustrated in Figure \ref{fig:introexample}
+For example, a common aspect of human communication is \emph{reciprocity} --- after person A speaks to person B, B subsequently has a higher propensity to speak to A.  
+These sorts of dynamics may be studied with event data, but again, our focus may be on the differences in these dynamics among the nodes in the network.  
+In the context of studying interactions among students, one might be interested particular conversational tendencies are associated with covariates about the student or the classroom.  
+A standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.
+
+%When such .
 One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO], where this variation in rates may be associated with some unobserved quantity, e.g.
 a covariate such as age or sex, or perhaps shared [TODO].
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.
+%Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by .
+Such techniques are readily applied to count data, [TODO: These let us study how shared behavior is associated with other covariates.]
+
+%1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of a tie between members
+
 %PS: much of this past work is for static network data.
@@
 
-If we are interested in studying a network of events, there are other important aspects of the dynamics in addition to the mean rate. 
-For example, human conversation often has an increased propensity for reciprocated events.
+%If we are interested in studying a network of events, there are other important aspects of the dynamics in addition to the mean rate. 
+%For example, human conversation often has an increased propensity for reciprocated events.
 %PS: wording of next sentence is a little odd - not sure I understand what it means...
-Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
-[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
-Figure \ref{fig:introexample} shows a simple example of how traditional blockmodels are unable to make inferences these types of dynamic phenomena.
+%Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
+%[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
+%Figure \ref{fig:introexample} shows a simple example of how traditional blockmodels are unable to make inferences these types of dynamic phenomena.
 

### 2012-05-31 docs/intro.tex
More intro edits.

@@
 Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
-We often  observe events occurring between nodes over time, and a common goal is to understand the differences in the total number of interactions --- perhaps with respect to information known about the nodes.
+In such settings we often observe events occurring between nodes over time, and a common goal is to understand the differences in the total number of interactions --- perhaps with respect to known information about the nodes.
 %In this case, we have a count for the total activity of each dyad, and one may analyze it has a static network. 
@@
 
-Network-based event data are powerful, however, because we enable the study of the processes governing these events.
-Theories about the dynamics among nodes will be often be difficult to study when only using the marginal rate of events as illustrated in Figure \ref{fig:introexample}
-For example, a common aspect of human communication is \emph{reciprocity} --- after person A speaks to person B, B subsequently has a higher propensity to speak to A.  
-These sorts of dynamics may be studied with event data, but again, our focus may be on the differences in these dynamics among the nodes in the network.  
-In the context of studying interactions among students, one might be interested particular conversational tendencies are associated with covariates about the student or the classroom.  
-A standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.
+Network-based event data are powerful, however, because they enable the study of the processes governing these events.
+Theories about the dynamics among nodes will often be difficult to study when only using the marginal rate of events, as illustrated in Figure \ref{fig:introexample}.  
+For example, a common aspect of human communication is \emph{reciprocity}: person A speaks to person B, then B has a higher propensity to speak to A.  
+Through an event-based analysis we can study the similarity --- and differences --- in the dynamics of nodes in the network. 
+When studying interactions among students, for example, one might be interested in conversational tendencies are associated with covariates about the student or classroom, or based on unobserved groupings.
+In this context, a standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.
 

### 2012-05-31 docs/intro.tex
Clean up second paragraph.

@@
 Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
-In such settings we often observe events occurring between nodes over time, and a common goal is to understand the differences in the total number of interactions --- perhaps with respect to known information about the nodes.
-%In this case, we have a count for the total activity of each dyad, and one may analyze it has a static network. 
-%Methods for static networks such as a 
+In such settings we often observe events occurring between nodes over time, and a common goal is to understand systematic differences in how often various interactions occur --- perhaps with respect to known information about the nodes.
 Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of techniques which can model this heterogeneity by assuming latent subgroups of nodes have similar tendencies in their group-wise rate of interaction.
@@
 For example, a common aspect of human communication is \emph{reciprocity}: person A speaks to person B, then B has a higher propensity to speak to A.  
-Through an event-based analysis we can study the similarity --- and differences --- in the dynamics of nodes in the network. 
-When studying interactions among students, for example, one might be interested in conversational tendencies are associated with covariates about the student or classroom, or based on unobserved groupings.
-In this context, a standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.
+Through an event-based analysis we can study how nodes vary in their tendencies to behave in this manner.  %the similarity --- and differences --- in the dynamics of nodes in the network. 
+This is particularly relevant when studying interactions among students in a classroom, where differences in conversational tendencies might be associated with covariates about the student or classroom, or based on unobserved groupings, and may have ramifications for education.
+In such contexts, a standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.
 

### 2012-05-31 docs/discussion.tex
Edit discussion and reorganize a few things.

@@
 
+Though prediction is not the main focus of the model, the results provide some evidence that having latent structure (i.e.
+K>1) can lead to improvements in predictive power.
+Our prediction results show that it is easy to overfit with this model.
+If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model (TODO: confirm this).
+However, as K increases beyond 2 (e.g., for the unrestrained CRP prior), the model is prone to overfit.
+This is not surprising, given that the number of fitted parameters scales as the square of the number of components K in the model.
+A natural future direction worth exploring for this model is a class of priors that are more resistant to overfitting.%^ (e.g., somehow have the priors  be sensitive to K^2 rather than K, as you mentioned yesterday).
+
 %PS: not sure where this fits…seems out of place here…and needs to be rewritten to be clearer - its not clear why we are comparing to this model other than the fact that it appeared at NIPS last year and also talks about relational event data.
 One way to make its inclusion here more natural would to set this up by first stating that there are other modeling approaches for relational events (such as the Guna… work), and comment on whether our "block approach" could be applied there too (not sure this makes sense since the models are so different…).
-Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
 

### 2012-05-31 docs/model.tex
Edit discussion and reorganize a few things.

@@
 The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
+
+\subsection{Related work}
+The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
+% \cite{AalenOddO.2008}
+Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
+By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.

### 2012-05-31 process/synthetic.r

@@ -167,39 +167,6 @@ set.seed(1)
 sim <- generate.brem(M,N,beta,z,ego=TRUE)
 y2 <- sim$edgelist
 
-s <- 2.5
-stext <- 1.5
-cols <- rev(grey.colors(100,start=0,end=.95,gamma=1))
-tb <- table(factor(y1[,2],1:N),factor(y1[,3],1:N))
-pdf("figs/introexample/mat1.pdf",width=s,height=s)
-par(mar=c(2,2,2,2),mgp=c(1.1,0,0))# oma=c(0,0,0,0),
-tb <- as.matrix(tb)
-image(1:N,1:N,tb,xaxt="n",yaxt="n",col=cols,xlab="",ylab="Observed counts",axes=FALSE,cex=stext,main="Rate Differentiated Case")
-axis(1,at=c(3,8),labels=c("A","B"),tick=FALSE)
-axis(2,at=c(3,8),labels=c("A","B"),tick=FALSE)
-dev.off()
-
-tb <- table(factor(y2[,2],1:N),factor(y2[,3],1:N))
-pdf("figs/introexample/mat2.pdf",width=s,height=s)
-par(mar=c(2,2,2,2),mgp=c(1.15,0,0))
-image(1:N,1:N,tb,xaxt="n",yaxt="n",xlab="",ylab="",col=cols,axes=FALSE,cex=stext)
-axis(1,at=c(3,8),labels=c("A","B"),tick=FALSE)
-axis(2,at=c(3,8),labels=c("A","B"),tick=FALSE)
-dev.off()
-
-pdf("figs/introexample/dynamics1.pdf",width=s,height=s)
-par(mar=c(3,3,0,0),mgp=c(2,1.15,0))
-x1 <- c(2,0,.5,2)
-plot(1:4,x1,xlim=c(0.5,4.5),ylim=c(-1,3),xlab="",xaxt="n",ylab="Propensity for reciprocal events",pch=19,col="red",cex=stext)
-axis(1,at=1:4,labels=c("(A,A)","(A,B)","(B,A)","(B,B)"),tick=TRUE)
-dev.off()
-
-pdf("figs/introexample/dynamics2.pdf",width=s,height=s)
-par(mar=c(3,3,0,0),mgp=c(2,1.15,0))
[…]

### 2012-05-31 docs/intro.tex
Edited the intro section, and added a reference.  (BTW, did you really intend W+F to be 1992?  You seem to have been citing their book proposal, instead of the 1994 book itself (albeit not in this paper).)

--- 2012-05-31 docs/intro.tex
@@
 Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
-We often  observe events occurring between nodes over time, and a common goal is to understand the differences in the total number of interactions --- perhaps with respect to information known about the nodes.
-%In this case, we have a count for the total activity of each dyad, and one may analyze it has a static network. 
-%Methods for static networks such as a 
-Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of techniques which can model this heterogeneity by assuming latent subgroups of nodes have similar tendencies in their group-wise rate of interaction.
+Recent work has expanded to include settings in which we observe events occurring between nodes over time (i.e., \emph{relational events}, as opposed to static edge structures or ongoing relationships), with the common goal of modeling interaction dynamics in terms of both endogenous mechanisms and exogenous covariates.
+A key concern in this regard is the ability to detect differential behavioral tendencies on the part of subsets of nodes, the dynamic analog of \emph{role structure} in classical network analysis \cite{Wasserman1994}.
 
-Network-based event data are powerful, however, because we enable the study of the processes governing these events.
-Theories about the dynamics among nodes will be often be difficult to study when only using the marginal rate of events as illustrated in Figure \ref{fig:introexample}
-For example, a common aspect of human communication is \emph{reciprocity} --- after person A speaks to person B, B subsequently has a higher propensity to speak to A.  
-These sorts of dynamics may be studied with event data, but again, our focus may be on the differences in these dynamics among the nodes in the network.  
-In the context of studying interactions among students, one might be interested particular conversational tendencies are associated with covariates about the student or the classroom.  
-A standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.
+In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.
+While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.
+Consider, e.g., Figure~\ref{fig:introexample}.
+In the top left panel, we depict the time-marginalized frequencies of simulated interactions between members of two groups (A and B), with darker cells indicating higher interaction frequencies.
+As can be seen, a clear role structure is present, with members of each subgroup interacting at higher rates with co-members than out-group members; such a structure is easily detectable via conventional blockmodeling techniques.
+By contrast, consider the interaction patterns shown in the top right panel.
+Here, there is no systematic difference in marginal rates between the two groups, rendering them invisible under a standard blockmodeling approach.
+
+There is, however, a difference to be detected.
+The bottom panels of Figure~\ref{}fig:introexample} show, for each respective simulation, parameters (as discussed in Section~\ref{}) governing the tendency towards reciprocation for members of each group vis a vis communications coming from in- or out-group members.
+For both simulated cases, the log-hazard for an event in which a member of group A or group B immediately responds to an incoming communication from an in-group member is increased by an increment of 2 (relative to a non-reciprocating event), and the log-hazard for an event in which a member of group B immediately responds to an incoming communication from a member of group A is increased by a corresponding increment of 0.5.
+Groups A and B are thus distinctive in terms of their dynamic behavior, even in the absence of marginal differences in propensity to communicate.
+A blockmodeling approach that classifies nodes based on shared dynamics (rather than merely shared marginal communication rates) can potentially identify such subtle distinctions; in this paper, we introduce such an approach.
+
+%For example, a common aspect of human communication is \emph{reciprocity}: person A speaks to person B, then B has a higher propensity to speak to A.  
+%Through an event-based analysis we can study how nodes vary in their tendencies to behave in this manner.  %the similarity --- and differences --- in the dynamics of nodes in the network. 
+%This is particularly relevant when studying interactions among students in a classroom, where differences in conversational tendencies might be associated with covariates about the student or classroom, or based on unobserved groupings, and may have ramifications for education.
+%In such contexts, a standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.
 
@@
 \includegraphics[scale=.45]{../figs/introexample/all}
-\caption{Illustration of differentiation by rate of interaction versus by dynamics.
+\caption{Illustration of differentiation by rate of interaction versus by dynamic behavior.
 The left column shows an example where within-block communication is large, and there is a higher tendency for reciprocity within a block.
 In the situation where the two groups are not differentiated by rate, as in the right column, a standard blockmodel is unable to distinguish between groups A and B.
-The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.}
+The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.
+{\color{red} (Suggest renaming "Dynamics parameters" to "Behavioral parameters" or something.
+"Dynamics parameters" is not gramatically sound.)}}
 \label{fig:example}

### 2012-06-01 docs/abstract.tex
Small edits.

@@
 For static networks, such heterogeneity has been treated via methods such as stochastic blockmodeling, which operate by assuming latent groups of individuals with similar tendencies in their group-wise interactions.
-Here, we combine these two approaches by positing a latent partition on the vertex set, such that event dynamics within and between node sets evolve in potentially distinct ways.
+Here, we combine these two approaches by positing a latent partition of the vertex set such that event dynamics within and between node sets evolve in potentially distinct ways.
 We illustrate the use of our model family by application to several forms of dyadic interaction data, including email communication and Twitter direct messages.
@@
 Our approach illustrates the utility of latent structure methods based on detailed dynamics, which can succeed even in the absence of differences in marginal interaction rates across groups. 
-%Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
+% TODO: This last sentence is a bit unclear for a new reader.
 \end{abstract}

### 2012-06-01 docs/intro.tex
Small edits.

@@
 
-Borrowing from the intuition of stochastic blockmodels, we propose a continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
+Borrowing from the intuition of stochastic blockmodels, we propose a novel continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
 Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
-%extend continuous-time models from event history analysis to model sequences of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-% \cite{AalenOddO.2008}
-%These models allow one to specify how the process depends on the previous history of events. % TODO use 'flexible'?
-In this way one can investigate theories about the underlying processes and make predictions about future data conditioned on the past.
+In this way one can compare theories about the underlying processes and make predictions about future data conditioned on the past.
 
-We describe parameter estimation and learning of  latent cluster assignments via MCMC, and illustrate the behavior of the model with simulated data.
+We describe how we learn the latent cluster assignments and model parameters via MCMC, and illustrate the behavior of the model with simulated data.
 Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
-Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, enabling the study of a dynamic extension to stochastic equivalence.
-%allowing identification of particular \emph{roles} in  that enable
+We conclude by showing that the parameter estimates reveal interpretable structure in the event dynamics, enabling the study of a dynamic extension to stochastic equivalence.
 

### 2012-06-01 docs/model.tex
Small edits.

@@
 \end{align}
+%{\color{red} (You've left out the right-censoring factor here (and also below).)}
 
@@
 In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
+We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
 The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
@@
 \caption[]{Illustration of event data and the assumptions of the model.
-Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
+Left: A sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
 Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
@@
 
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ is assigned a latent cluster $z_i$ and use a log linear model for the intensity functions
+We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
 \begin{align*}
@@
 \end{align*}
-where for each pair of clusters $(k,l)$ we have a vector of parameters $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the vector of $P$ statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.
+where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
+Thus, our model for the rate of $(i,j)$ events has the same parameters as for other events occurring between group $z_i$ and $z_j$.
 
  In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
-As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.
+As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{ij}(t)$ to change following an event where $i$ was the sender or the recipient.
 This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
 In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
+% TODO: Lead off with computational convenience, then justify?
 %\footnote{TODO: This is analogous to a Markov condition for this class of models.}
@@
 We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta{\sigma})$.
-The cluster assignments are given a non-parameteric priors $z_i \sim \mbox{CRP}(\alpha)$.
+The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
+% TODO: Expand on this?
 
@@
 \hline
-Statistic & Formula \\
+Statistic & Formula\\
 \hline
@@
 \label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
+\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{ij}$ previous to event $m$.}
 \end{table}
@@
 For example, a particular node sending often may indicate they will continue to be a sender in the future.
-We normalize these counts by the number of events up until a dyad's prior changepoint.%\footnote{This seems to avoid ``explosion'' when simulating from the model.
+We normalize these counts by the number of events up until a dyad's prior changepoint.
+%\footnote{This seems to avoid ``explosion'' when simulating from the model.
 TODO: Explore the theory about these sorts of statistics.
@@
 
+% TODO: Awkward next sentence
 The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
 For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
-Though we use only the above statistics in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
+Though we use only the statistics in Table \ref{tab:stats} in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
 The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
+% TODO: Also, need to be restricted to ego conditions if we want it fast
 
@@
 
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
+Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} for static networkswhich models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
 In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
 
-The above framework generalizes several important special cases.
+The proposed family of models generalizes several important special cases.
 For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.
@@
 Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
-By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions, while here we use latent variables to allow for heterogeneity in intensities across possible events.
+By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions.  
+We instead use latent variables to allow for heterogeneity in intensities across possible events.

### 2012-06-01 docs/model.tex
Minor edits to the modeling section.

--- 2012-05-31 docs/model.tex
@@
 \end{align}
+{\color{red} (You've left out the right-censoring factor here (and also below).)}
 
@@
 Though we use only the above statistics in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
+The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.
+However, such effects can be approximated by introducting exogenous ``clock events'' not discussed here.}
 

### 2012-06-01 docs/intro.tex
Fix a few figure references.

@@
 There is, however, a difference to be detected.
-The bottom panels of Figure~\ref{}fig:introexample} show, for each respective simulation, parameters (as discussed in Section~\ref{}) governing the tendency towards reciprocation for members of each group vis a vis communications coming from in- or out-group members.
+The bottom panels of Figure~\ref{fig:introexample} show, for each respective simulation, parameters (as discussed in Section~\ref{sec:specification}) governing the tendency towards reciprocation for members of each group vis a vis communications coming from in- or out-group members.
 For both simulated cases, the log-hazard for an event in which a member of group A or group B immediately responds to an incoming communication from an in-group member is increased by an increment of 2 (relative to a non-reciprocating event), and the log-hazard for an event in which a member of group B immediately responds to an incoming communication from a member of group A is increased by a corresponding increment of 0.5.

### 2012-06-01 docs/discussion.tex
Small reorg of discussion.

@@
 Our approach generalizes traditional static notions of stochastic equivalence on nodes (such as stochastic blockmodels) to a dynamic formulation. 
-The proposed dynamic model posits the existence of groups of nodes, such that all members of a group are governed by the same dynamic process, and groups are differentiated by having different dynamic processes.
+The proposed model posits the existence of groups of nodes, such that all members of a group are governed by the same dynamic process, and groups are differentiated by having different dynamic processes.
 %PS: sentence below is incomplete (after the first "the")…..I decided to take it out anyway..
@@
 In particular, we show that our proposed approach leads to improved predictive accuracy when compared to both (a) relational event models that lack latent clusters, and (b) stochastic blockmodels for count data, both of which are both special cases of our model.
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-
 Though prediction is not the main focus of the model, the results provide some evidence that having latent structure (i.e.
@@
 Our prediction results show that it is easy to overfit with this model.
-If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model (TODO: confirm this).
+If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model.
+% (TODO: confirm this).
 However, as K increases beyond 2 (e.g., for the unrestrained CRP prior), the model is prone to overfit.
 This is not surprising, given that the number of fitted parameters scales as the square of the number of components K in the model.
-A natural future direction worth exploring for this model is a class of priors that are more resistant to overfitting.%^ (e.g., somehow have the priors  be sensitive to K^2 rather than K, as you mentioned yesterday).
+A natural future direction worth exploring for this model is a class of priors that are more resistant to overfitting.
+
+Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
+One interesting extension, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.  
+We leave this to future work.
+
+%Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.  
+%For example, dynamics might change over time \cite{Vu2011}.
+
+%^ (e.g., somehow have the priors  be sensitive to K^2 rather than K, as you mentioned yesterday).
 
@@
 %PS: these last two paragraphs need to be finished….
-Trivially extended to co-clustering applications and for directed data.
-Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
+% Trivially extended to co-clustering applications and for directed data.
+% Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
 
-Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.
-For example, dynamics might change over time \cite{Vu2011}.
 Alternatively, if nodes change groups.
-One approach, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.
 In some contexts such extensions might be substantively important and warrant future work.

### 2012-06-01 docs/discussion.tex
Small edits to discussion.  Mention observation time for llk.
Add zoomed-in portion to exploratory analysis plot.

@@
 \caption{\texttt{ab-by} estimates}
+\end{subfigure} \\
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed-zoom}
+\caption{Observed counts}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\end{subfigure}
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1-zoom}
+\caption{Intercept estimates}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2-zoom}
+\caption{\texttt{ab-ba} estimates}
+\end{subfigure}
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3-zoom}
+\caption{\texttt{ab-by} estimates}
 \end{subfigure}
@@
 Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
-Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).}
+Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).
+Figures (e-h) give a zoomed-in view of groups 1 and 3, showing that while the overall rate of (1,3) and (3,1) events is similar, the tendency for \texttt{ab-by} differs.}
 \label{fig:parmats}
@@
 We leave this to future work.
+%Alternatively, if nodes change groups.
+%In some contexts such extensions might be substantively important and warrant future work.
 
@@
 % Also useful for our model when nodes belong to one latent class as a sender but another latent class as a recipient.
-
-Alternatively, if nodes change groups.
-In some contexts such extensions might be substantively important and warrant future work.

### 2012-06-01 figs/eckmann-stats.r

@@ -110,3 +110,18 @@ for (p in px) {
   image(mat,xaxt="n",yaxt="n",col=cols)
   dev.off()
 }
+
+# Zoom in plots
+ix <- o[which(z[o] %in% c(1,3))]  # members of 1 or 3
+tb <- table(factor(train[,2],1:N),factor(train[,3],1:N))
+pdf("figs/eckmann-small/parmat/observed-zoom.pdf",height=4,width=4)
+par(mar=c(0,0,0,0))
+image(log(tb[ix,ix]+1),xaxt="n",yaxt="n",col=cols)
+dev.off()
+for (p in px) {
+  mat <- (mats[[p]] - mu.hat[p]) / sigma.hat[p]
+  pdf(paste("figs/eckmann-small/parmat/",p,"-zoom.pdf",sep=""),height=4,width=4)
+  par(mar=c(0,0,0,0))
+  image(mat[ix,ix],xaxt="n",yaxt="n",col=cols)
+  dev.off()
+}

### 2012-06-01 docs/intro.tex
Change y-axis on example in Intro.

@@
 In the situation where the two groups are not differentiated by rate, as in the right column, a standard blockmodel is unable to distinguish between groups A and B.
-The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.
-{\color{red} (Suggest renaming "Dynamics parameters" to "Behavioral parameters" or something.
-"Dynamics parameters" is not gramatically sound.)}}
+The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.}
 \label{fig:example}
 \end{SCfigure}
+% {\color{red} (Suggest renaming "Dynamics parameters" to "Behavioral parameters" or something.  "Dynamics parameters" is not gramatically sound.)

### 2012-06-01 process/synthetic.r

@@ -187,7 +187,7 @@ axis(2,at=c(3,8),labels=c("A","B"),tick=FALSE)
 
 par(mar=c(3,3,1,1),mgp=c(2,1.15,0))
 x1 <- c(2,0,.5,2)
-plot(1:4,x1,xlim=c(0.5,4.5),ylim=c(-1,3),xlab="",xaxt="n",ylab="Dynamics parameters",pch=19,col="red",cex=stext)
+plot(1:4,x1,xlim=c(0.5,4.5),ylim=c(-1,3),xlab="",xaxt="n",ylab="Behavioral parameters",pch=19,col="red",cex=stext)
 axis(1,at=1:4,labels=c("(A,A)","(A,B)","(B,A)","(B,B)"),tick=TRUE,cex.axis=1)
 
 par(mar=c(3,2,1,2),mgp=c(2,1.15,0))

### 2012-06-01 docs/discussion.tex
Fix up some references.

@@
 Darker values are larger.
-Estimates are rescaled posterior means $(\hat{\beta}_{p,z_i,z_j} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
+Estimates are rescaled posterior means $(\hat{\beta}_{z_i,z_j,p} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
 Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).
-Figures (e-h) give a zoomed-in view of groups 1 and 3, showing that while the overall rate of (1,3) and (3,1) events is similar, the tendency for \texttt{ab-by} differs.}
+Figures (e-h) give a zoomed-in view of groups 1 and 3, showing that while the overall rate of (1,3) and (3,1) events is similar, the tendency for \texttt{ab-by} transitions differs.}
 \label{fig:parmats}

### 2012-06-01 docs/experiments.tex
Edit discussion and reorganize a few things.
Small edits.
Fix up some references.

@@
 %\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2003}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Enron}
-\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
+\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2001}
+\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann2004}
+\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Klimt2004}
+%\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
 \item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
@@
 We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{MITreality}
+\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{Eagle2009}
 \end{itemize}
@@
 
-\input{../figs/results-llk.tex}
+\input{../figs/results-llk2.tex}
 
-\subsection{Baselines}
+%\subsection{Baselines}
 
@@
 
-\subsection{Results}
-
-TODO: Quickly describe main take-aways from the results tables.
-
 % Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.

### 2012-06-01 docs/intro.tex
Fix up some references.

@@
 
-In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.
+In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez2011} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.
 While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.
@@
 The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.}
-\label{fig:example}
+\label{fig:introexample}
 \end{SCfigure}

### 2012-06-01 docs/experiments.tex
Fix up references.

@@
 \section{Model fitting and experiments}
+\label{sec:experiments}
 
@@
 
-Figure \ref{fig:posteriorparams} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
+Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
 We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
@@
 
-\input{../figs/results-recall.tex}
+%\input{../figs/results-recall.tex}
 

### 2012-06-01 docs/inference.tex
Fix up references.

@@
 \subsection{Hyperparameter settings}
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
+In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2000} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
 This aims for similarly sized groups.}

### 2012-06-01 docs/model.tex
Fix a few figure references.
Small edits to discussion.  Mention observation time for llk.
Fix up references.

--- 2012-06-01 docs/model.tex
@@
 We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
-The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
+%The likelihood of an observed event history $\mathcal{A}$ extending from time 0 to the time of the final event, $t_M$, is 
+The likelihood of an observed event history $\mathcal{A}_{t_M}$ (extending from time 0 to the time of the final event and denoted $\mathcal{A}$ for convenience) is
+%The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
 \begin{align}
@@
 \subsection{Model specification}
+\label{sec:specification}
 
@@
 \end{tabular}
+\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{ij}$ previous to event $m$.}
 \label{tab:stats}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{ij}$ previous to event $m$.}
 \end{table}
@@
 
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp} for static networkswhich models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
+Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp2006} for static networks which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
 In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

### 2012-06-01 docs/discussion.tex
Incorporate Padhraic's change in the end of the discussion.

@@
 
-Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
-One interesting extension, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.  
+Another direction to explore is that of including different statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and using our approach to study how the roles of these statistics vary across nodes.
+One potentially interesting extension, analogous to \cite{Airoldi2008}, would be to allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point---we leave this to future work.
+
+%Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
+%One interesting extension, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.
 We leave this to future work.

### 2012-06-01 getcounts.r
Directory clean up.

@@ -1,21 +1,0 @@
-#!/usr/bin/env Rscript
-suppressPackageStartupMessages(library(optparse))
-suppressPackageStartupMessages(library(brem))
-
-
-option_list <- list(
-  make_option(c("-d", "--dataset"), 
-              help="Name of dataset with data at /data/[dataset].rdata 
-                    and results at /results/[dataset]/."),
-  make_option(c("-m", "--model"), default=NULL,
-              help=""))
-parser <- OptionParser(usage = "%prog [options]", option_list=option_list)
-opts   <- parse_args(OptionParser(option_list=option_list))
-
-load(paste("data/",opts$dataset,".rdata",sep=""))
-
-# Get indices for events that had dyads which were previously unobserved
-ctrain <- counts.of.observed(train,N)
-ctest <- counts.of.observed(test,N)
-
-save(ctrain,ctest,opts,file=paste("results/",opts$dataset,"/counts.rdata",sep=""))

### 2012-06-01 llk.r

@@ -74,8 +74,8 @@ tmp[,-c(1:4)] <- round(tmp[,-c(1:4)],3)
 chosen <- c(1:4,8,12,15,16,23)
 subset(tmp,alpha %in% c(1,5))[,chosen]
 
-res <- subset(tmp,alpha %in% c(5) & type=="test")[,c("dataset","2_1_1_1","2_10_0_0","2_2_1_1","2_3_1_1","2_10_1_1")]
-colnames(res) <- c("Dataset","K=1","SBM","K=2","K=3","K=10")
+res <- subset(tmp,alpha %in% c(5) & type=="test")[,c("dataset","2_10_0_0","2_1_1_1","2_2_1_1","2_3_1_1","2_10_1_1")]
+colnames(res) <- c("Dataset","SBM","K=1","K=2","K=3","K=10")
 
 library(xtable)
 xr <- xtable(res,caption="Comparing mean loglikelihood under different methods for each event in a given test set.  Larger values are better.  See text for details.",label="tab:results",digits=3)

### 2012-06-01 results.r

@@ -26,6 +26,16 @@ for (i in 1:length(r)) {
   }
 }
 r <- melt(r)
+colnames(r)[2:5] <- c("type","metric","model","dataset")
+ma <- lapply(unique(r$model),function(m) data.frame(model=m,modelatts(m)))
+ma <- do.call(rbind,ma)
+x <- merge(r,ma,by="model")
+tmp <- dcast(x,xsigalpha +xsigbeta+metric+type+ dataset ~ kinit + kmax + pshift + deg, fun.aggregate=mean)
+tmp$xsigalpha <- tmp$xsigalpha/1000
+tmp$xsigbeta  <- tmp$xsigbeta/1000
+colnames(tmp)[1:2] <- c("alpha","beta")
+
+subset(tmp,metric=="llk" & alpha==5 & type=="test")[,c("dataset","2_10_0_0","2_1_1_1","2_2_1_1","2_3_1_1","2_10_1_1")]
 
 #datasets <- c("synthetic-1","eckmann-small","realitymining-small",
  datasets <- c("synthetic-1","eckmann-small","classroom-16", "classroom-17", "classroom-27","realitymining-small","enron-small","twitter")

### 2012-06-01 run.sh

@@ -6,6 +6,7 @@
 
 ./parallel --sshlogin 8/d5,8/d7,8/d8,4/d9,4/d10,8/d11,4/d12 'cd /extra/duboisc0/blockrem;./brem.r --dataset {1} --kinit 2 --kmax {2} --numiterations 500 --splitmerge FALSE --pshifts {3} --degrees {4} --negbinom FALSE --collapse TRUE --numextra 5 --force TRUE --sigma.hyper.alpha {5} --sigma.hyper.beta {6} --crp.hyper 1' :::  "synthetic-1" "eckmann-small" "classroom-16" "classroom-17" "classroom-27" "enron-small" "realitymining-small" "twitter-small" ::: 1 2 3 10 ::: TRUE FALSE ::: TRUE FALSE ::: 5 1 ::: 1
 
+./parallel --sshlogin  4/d5,4/d8,4/d11,4/d12 'cd /extra/duboisc0/blockrem;./predict.r --dataset {1} --force {2} --baselines {3} --niters {4}' :::  "synthetic-1" "eckmann-small" "classroom-16" "classroom-17" "classroom-27" "enron-small" "realitymining-small" "twitter-small" ::: TRUE ::: TRUE FALSE ::: 10
 
 
 

### 2012-06-01 synthetic.example.r

@@ -1,150 +1,0 @@
-opts <- list(dataset="synthetic-1",numclusters=2,model.type="full",numiterations=20)
-
-load(paste("data/",opts$dataset,".rdata",sep=""))
-library(brem)
-source("pkg/R/brem.r")
-source("pkg/R/splitmerge.r")
-source("pkg/R/brem.alt.r")
-
-# Precompute data structures
-N <- max(c(train[,2],train[,3]))
-M <- nrow(train)
-P <- 13
-ego <- 1
-K <- opts$numclusters
-s <- new(RemStat,train[,1],as.integer(train[,2])-1,as.integer(train[,3])-1,N,M,P,ego)
-s$precompute()
-
-## Set priors
-priors <- list(alpha=1,sigma=.1,phi=list(mu=0,sigma=1))
-px <- rep(1,13)
-px[8:13] <- 0
-px[7] <- 0
-
-## Test basic likelihood functions
-K <- 2
-k1 <- k2 <- 1
-current <- array(rnorm(P*K^2,priors$phi$mu,priors$phi$sigma),c(P,K,K))
-olp <- brem.lpost.fast(A,N,K,z,s,beta)
-
-priors <- list(alpha=1,sigma=.1,phi=list(mu=0,sigma=1))
-llk <- sum(RemLogLikelihoodPc(beta,z-1,s$ptr(),K))
-llk <- sum(RemLogLikelihoodPcSubset(beta,z-1,s$ptr(),K,1:5))
-llk <- sum(RemLogLikelihoodActorPc(3,beta,z-1,s$ptr(),K))
-k1nodes <- which(z==k1)
-k2nodes <- which(z==k2)
-llk <- sum(RemLogLikelihoodBlockPc(k1-1,k2-1,k1nodes-1,k2nodes-1,beta,z-1,s$ptr(),K))
-
-# Fit model
-source("pkg/R/brem.alt.r")
-lp(beta,z,priors)
-px <- c(1,2,3,4,5,6)
-K <- 5; P <- 13
-fit <- mcmc.blockmodel(lp,llk_node,priors,N,P,K,do.sm=FALSE,num.extra=10,niter=15,verbose=TRUE)
-lp(fit$samples[[5]]$phi,fit$samples[[5]]$z,priors)
-
-fit <- mcmc.blockmodel(lp,llk_node,priors,N,P,10,do.sm=TRUE,num.extra=10,niter=20,verbose=TRUE)
-
[…]

### 2012-06-01 twitter.example.r

@@ -1,83 +1,0 @@
-# Get estimates for full 2 when we manually set the groups
-library(brem)
-load("data/twitter-small.rdata")
-source("pkg/R/brem.r")
-
-tb <- table(factor(c(A[,2],A[,3]),1:N))
-chosen <- names(tb)[which(tb > 20)]
-chosen <- as.numeric(chosen)
-z <- rep(1,N)
-z[chosen] <- 2
-
-opts <- list(slice=FALSE,gibbs=FALSE,model.type="full",dataset="twitter-small",numiterations=500,numclusters=2,initialize=TRUE)
-
-# Precompute data structures
-M <- nrow(train)
-P <- 13
-K <- opts$numclusters
-s <- new(RemStat,train[,1],as.integer(train[,2])-1,as.integer(train[,3])-1,N,M,P)
-s$precompute()
-
-px <- rep(1,13)
-px[13] <- 0
-px[7]  <- 0
-f <- paste("results/",opts$dataset,"/full.1.rdata",sep="")
-if (K > 1 & file.exists(f) & opts$initialize) {
-  load(f)
-  beta <- array(res$beta,c(P,K,K))
-} else {
-  beta <- NULL
-}
-
-# fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice, mcmc.sd=.05,m=5,
-#                  niter=opts$numiterations, gibbs=opts$gibbs,  beta=beta, px=px, z=z,   
-#                  outfile="tmp.slice.m5.rdata")
-# 
-
-opts$gibbs <- "fast"
-opts$slice <- TRUE
-# fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice, mcmc.sd=.05,m=20,
-#                  niter=opts$numiterations, gibbs=opts$gibbs,  beta=beta, px=px, z=z,   
-#                  outfile="tmp.slice.gibbs.rdata")
-
-
-fit <- brem.mcmc(train,N,K,s,model.type=opts$model.type,mh=!opts$slice, mcmc.sd=.05,m=20,
-                 niter=opts$numiterations, gibbs=opts$gibbs,  beta=beta, px=px,  
-                 outfile="tmp.slice.gibbs.learn2.rdata")
-
-# Visualzie model paramters
-
-load("tmp.slice.rdata")
-r <- melt(res$param)
-r <- subset(r,X1<10)
[…]

### 2012-06-01 docs/inference.tex
Move figures around and use subfigures more.

@@
 
-\begin{figure}
-\center
-\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
-\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
-\caption{Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
-The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
-\label{fig:syncounts}
-\end{figure}
-
 \subsection{Hyperparameter settings}

### 2012-06-01 docs/simulation.tex
Move figures around and use subfigures more.

@@
+\begin{figure}
+%\centering
+\begin{subfigure}[b]{0.25\textwidth}
+\centering
+\includegraphics[width=1.5in]{../figs/synthetic/mat.pdf}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.3\textwidth}
+\centering
+\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.3\textwidth}
+\centering
+\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
+\end{subfigure}
+\caption{Illustration of 2000 simulated events, as described in text.
+Left: Counts of each dyad.
+Center: Boxplot of distribution of participation counts across dyads.
+The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
+Right: Parameters (in red) and posterior credible intervals (in black).}
+\label{fig:syncounts}
+\end{figure}
+
 \section{Simulation}

### 2012-06-01 docs/model.tex
Move figures around and use subfigures more.
Small notation fixes.

@@
 \centering
-\begin{minipage}[b]{0.45\linewidth}
+\begin{subfigure}[b]{0.45\linewidth}
   \begin{tikzpicture}[scale=.8,transform shape,thick]
@@
   \caption{Dynamic network data}
-\end{minipage}
-\begin{minipage}[b]{0.45\linewidth}
+\end{subfigure}
+\begin{subfigure}[b]{0.45\linewidth}
 \begin{tikzpicture}[scale=1,transform shape,thick]
@@
 \caption{Intensities for two dyads}
-\end{minipage}
-\caption[]{Illustration of event data and the assumptions of the model.
-Left: A sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
-Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).
-We assume $\lambda_{34}$ can only change after events sent or received by node 3.\footnotemark }
+\end{subfigure}
+\caption[]{Illustration of relational event data and the assumptions of the model.
+(a) A sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
+(b) The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{3,4}(t)$ (dotted).
+We assume $\lambda_{3,4}$ can only change after events sent or received by node 3.
+}
 \label{fig:example}
@@
 \label{sec:specification}
-
-\begin{table}[t]
-\footnotesize
-\center
-\begin{tabular}{|l|l|}
-\hline
-Statistic & Formula\\
-\hline
-\hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
-%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
-%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
-%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
-%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
-\hline
-\end{tabular}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{ij}$ previous to event $m$.}
-\label{tab:stats}
-\end{table}
 
@@
 
+\begin{table}[t]
+\footnotesize
+\center
+\begin{tabular}{|l|l|}
+\hline
+Statistic & Formula\\
+\hline
+\hline
+Intercept& $s_{0}(t,i,j) = 1$\\
+Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
+Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
+Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
+%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
+%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
+Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
+Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
+Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
+%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
+%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
+\hline
+\end{tabular}
+\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{ij}$ previous to event $m$.}
+\label{tab:stats}
+\end{table}
+
 \subsection{Related work}

### 2012-06-01 docs/intro.tex
Small paper edits.  A few TODOs inserted.

@@
 
-In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez2011} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.
+In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp2006, Ishiguro2010, Rodriguez2011} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.
 While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.

### 2012-06-01 docs/model.tex
Small paper edits.  A few TODOs inserted.

@@
 
-We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
+{\color{red} (Need a better transition here.)}  We aim to learn about the dynamics between subsets of nodes.
+To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
 \begin{align*}
@@
 where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
-Thus, our model for the rate of $(i,j)$ events has the same parameters as for other events occurring between group $z_i$ and $z_j$.
+Thus, the rate of $(i,j)$ events has the same parameters as other events occurring between group $z_i$ and $z_j$.
 
- In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
+% In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
 As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{ij}(t)$ to change following an event where $i$ was the sender or the recipient.
@@
 In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
-% TODO: Lead off with computational convenience, then justify?
-%\footnote{TODO: This is analogous to a Markov condition for this class of models.}
 
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta{\sigma})$.
+We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta_{\sigma})$.
 The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
@@
 
-In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.
+Table \ref{tab:stats}  lists the statistics  $\mathbf{s}(t,i,j,\mathcal{A}_t)$ used in Section \ref{sec:experiments}.
 For example, a particular node sending often may indicate they will continue to be a sender in the future.
@@
 For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
-Though we use only the statistics in Table \ref{tab:stats} in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
+Though only the statistics in Table \ref{tab:stats} are used in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
 The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}

### 2012-06-01 docs/simulation.tex
Small notation fixes.
Small paper edits.  A few TODOs inserted.

@@
 \end{subfigure}
-\caption{Illustration of 2000 simulated events, as described in text.
+\caption{[TODO: Remove AB-XA, etc from the plot.
+Rotate them.
+Use subfigure labels.] Illustration of 2000 simulated events, as described in text.
 Left: Counts of each dyad.
@@
 
-We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
-The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.
+We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 groups where 1) within group communication is more likely, 2) events among members in the first group are more likely to be reciprocated  (i.e.
+a positive \texttt{ab-ba} effect), and 3) events among members of the second group are more likely to be followed by an event with the same sender (i.e.
+a positive \texttt{ab-ay} effect).
+The specification of $\textbf{s}$ is therefore $\textbf{s}(t,i,j,\mathcal{A}_t) = [s_0, s_{1}(t,i,j,\mathcal{A}_t), s_{2}(t,i,j,\mathcal{A}_t)]$.
 For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
-This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
+Data is generated by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
 

### 2012-06-01 docs/discussion.tex
Minor paper edits.  A few TODOs inserted.

@@
 %This analysis suggests 
-Our approach has the ability to uncover systematic differences in dynamic behaviors or roles among subsets of nodes.
+Our approach has the ability to uncover systematic differences in dynamic behaviors among subsets of nodes.
 % across a variety of data sets.
@@
 If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model.
-% (TODO: confirm this).
+{\red(TODO: confirm this)}
   However, as K increases beyond 2 (e.g., for the unrestrained CRP prior), the model is prone to overfit.

### 2012-06-01 docs/experiments.tex
Add TODOs in results section.
Add another placeholder and update TODO file.
Minor paper edits.  A few TODOs inserted.

@@
 
+A variety of real world datasets are used to explore the efficacy of the model.
 Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
@@
 The two panels on the left show the variation in $s_0(t_M,i,j)$, $s_1(t_M,i,j)$ and $s_3(t_M,i,j)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
-The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
+The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
 Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
@@
 
-We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.
+We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood and recall for held-out data.
 Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
@@
 
+[TODO: Recall table.]
+
+[TODO: Mention that we cap the value of $K$]
+
+[TODO: Discuss results.]
 %\input{../figs/results-recall.tex}

### 2012-06-01 docs/simulation.tex
Minor paper edits.  A few TODOs inserted.

@@
 Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
-Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
+Intensities for reciprocal actions among nodes in block 1 are $e^3 \approx 20$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
 Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).

### 2012-06-01 docs/discussion.tex
Work on prediction

@@
 Our prediction results show that it is easy to overfit with this model.
-If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model.
-{\red(TODO: confirm this)}
-  However, as K increases beyond 2 (e.g., for the unrestrained CRP prior), the model is prone to overfit.
+%If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model. 
+%{\red (TODO: confirm this)}
+ As K increases beyond 2 (e.g., for the unrestrained CRP prior), the model is prone to overfit.
 This is not surprising, given that the number of fitted parameters scales as the square of the number of components K in the model.

### 2012-06-01 dashboard.r

@@ -29,25 +29,13 @@ opts   <- parse_args(OptionParser(option_list=option_list))
 options(verbose=FALSE)
 library(brem)
 library(ggplot2)
-opts <- list(dataset="realitymining-small",predictions=TRUE)
+opts <- list(dataset="classroom-16",predictions=TRUE)
 dataset <- opts$dataset
 
 # Pull data from each saved file and grab the name of the fit
 folder <- paste("results/",opts$dataset,"/fits",sep="")
 load(paste("data/",opts$dataset,".rdata",sep=""))
-modelnames <- function(folder) {
-  as.vector(sapply(dir(folder),function(x) strsplit(x,".rdata")[[1]][1]))
-}
-
-modelatts <- function(model) {
-  atts <- strsplit(model,"\\.")[[1]]
-  g <- function(x,y) as.numeric(strsplit(x,y)[[1]][2])
-  xs <- c("kinit","kmax","sm","nb","pshift","deg","trans","collapse","xsigalpha","xsigbeta")
-  a <- lapply(1:length(xs),function(i) g(atts[i],xs[i]))
-  names(a) <- xs
-  return(a)
-}
-
+source("utils.r")
 fs <- list.files(folder,full.names=TRUE)
 models <- modelnames(folder)
 atts <- lapply(models,function(m) as.data.frame(modelatts(m)))
@@ -80,7 +68,7 @@ q1
 cat("Trace plot of beta.\n")
 library(coda)
 t(t(names(fits)))
-fit <- fits[[24]]
+fit <- fits[[35]]
 z <- fit$params$z
 table(z)
 table(z[train[,2]],z[train[,3]])
@@ -90,7 +78,7 @@ mus <- lapply(fit$samples[1:fit$iter],function(s) {
   as.matrix(s$mu)
 })
 r <- melt(mus)
-q1c <- qplot(L1,value,data=subset(r, Var1 %in% fit$priors$px), colour=factor(Var1),geom="line") + labs(colour="parameters for\n each block",x="iteration") + theme_bw()# + facet_grid(Var2~Var3,scales="free")
+q1c <- qplot(L1,value,data=subset(r, L1 > 20 & Var1 %in% fit$priors$px), colour=factor(Var1),geom="line") + labs(colour="parameters for\n each block",x="iteration") + theme_bw()# + facet_grid(Var2~Var3,scales="free")
 q1c
 
 # Trace plot of upper level sigmas

### 2012-06-01 predict.r

@@ -1,21 +1,19 @@
 #!/usr/bin/env Rscript
 suppressPackageStartupMessages(library(optparse))
 suppressPackageStartupMessages(library(brem))
-suppressPackageStartupMessages(library(ggplot2))
 suppressPackageStartupMessages(library(reshape2))
+suppressPackageStartupMessages(library(multicore))
 
 option_list <- list(
   make_option(c("-d", "--dataset"), 
               help="Name of dataset with data at /data/[dataset].rdata 
                     and results at /results/[dataset]/."),
-  make_option("--force", default=FALSE),
-  make_option("--baselines", default=FALSE),
   make_option("--niters", type="integer", default=10)  )
 parser <- OptionParser(usage = "%prog [options]", option_list=option_list)
 opts   <- parse_args(OptionParser(option_list=option_list))
 
 #
-library(brem);library(reshape2);opts <- list(dataset="eckmann-small",baselines=FALSE,force=TRUE,niters=1)
+#library(brem);library(reshape2);library(multicore);opts <- list(dataset="twitter-small",niters=10)
 source("utils.r")
 
 options(verbose=FALSE)
@@ -24,26 +22,45 @@ dataset <- opts$dataset
 load(paste("data/",dataset,".rdata",sep=""))
 train.ix <- 1:nrow(train)
 test.ix <- (1:nrow(A))[-train.ix]
-file.remove("progress.rdata")
-dir.create(paste("results/",dataset,"/preds/",sep=""),showWarnings=FALSE)
 
-for (model in c("online","uniform","marg")) {
-  progress(dataset,model,"started")
-  pred <- evaluate.baseline(A,N,train.ix,test.ix,model)
-  save(pred,opts,file=paste("results/",dataset,"/preds/",model,".rdata",sep=""))
-  progress(dataset,model,"completed")  
-}
+dir.create(paste("results/",dataset,"/preds/",sep=""),showWarnings=FALSE)
+pf <- "progress.rdata"
+
 
 folder <- paste("results/",opts$dataset,"/fits",sep="")
 models <- modelnames(folder)
-for (model in models) {
-  progress(dataset,model,"started")
+options(cores=8)
+
+for (model in models) progress(pf,c(dataset,model,"started",""))
+
+x <- mclapply(models,function(model) {
+#for (model in models) {
+  st <- proc.time()
[…]

### 2012-06-01 results.r

@@ -1,6 +1,6 @@
 # Go through available pred objects.
-
-datasets <- c("synthetic-1","eckmann-small","classroom-16", "classroom-17", "classroom-27","realitymining-small","enron-small","twitter")
+source("utils.r")
+datasets <- c("synthetic-1","classroom-16", "classroom-17", "classroom-27","eckmann-small","realitymining-small","enron-small","twitter")
 
 res <- lapply(datasets,function(dataset) {
   folder <- paste("results/",dataset,"/preds/",sep="")
@@ -13,6 +13,9 @@ res <- lapply(datasets,function(dataset) {
   return(r)
 })
 names(res) <- datasets
+allpred <- res
+
+save(allpred,file="results/allpred.rdata")
 
 r <- res
 ix <- which(sapply(r,length) == 0)
@@ -25,7 +28,20 @@ for (i in 1:length(r)) {
     } 
   }
 }
+
+
+## rks <- list()
+## for (i in 1:length(r)) {
+##   for (j in 1:length(r[[i]])) {
+##     rks[[i]][[j]] <- list(train=recall(r[[i]][[j]][[k]]$train),
+2##                           test =recall(r[[i]][[j]][[k]]$test))
+##   }
+## }
+
+
+
 # recall(r$test,top=1:30)
+library(reshape2)
 r <- melt(r)
 colnames(r)[2:5] <- c("type","metric","model","dataset")
 ma <- lapply(unique(r$model),function(m) data.frame(model=m,modelatts(m)))
@@ -35,8 +51,21 @@ tmp <- dcast(x,xsigalpha +xsigbeta+metric+type+ dataset ~ kinit + kmax + pshift 
 tmp$xsigalpha <- tmp$xsigalpha/1000
 tmp$xsigbeta  <- tmp$xsigbeta/1000
 colnames(tmp)[1:2] <- c("alpha","beta")
+subset(tmp,metric=="mllk" & alpha==5 & type=="test")[,c("dataset","2_10_0_0","2_1_1_1","2_2_1_1","2_3_1_1","2_10_1_1")]
 
-subset(tmp,metric=="llk" & alpha==5 & type=="test")[,c("dataset","2_10_0_0","2_1_1_1","2_2_1_1","2_3_1_1","2_10_1_1")]
+
[…]

### 2012-06-01 utils.r

@@ -10,40 +10,58 @@ modelnames <- function(folder) {
   as.vector(sapply(dir(folder),function(x) strsplit(x,".rdata")[[1]][1]))
 }
 
-progress <- function(dataset,model,action,f="progress.rdata") {
+# TODO: Just input a tuple...?
+progress <- function(f,x,num.match=2) {
   if (!file.exists(f)) {
-    current <- list()
-    save(current,file=f)
-  } 
-  load(f)
-  if (length(current[[dataset]]) == 0) {
-    current[[dataset]] <- list()
+    current <- matrix(x,nr=1)
+  } else {
+    load(f)
+    ix <- which(current[,1] == x[1] & current[,2] == x[2])
+    if (length(ix) == 0) {
+      current <- rbind(current,x)
+    } else {
+      current[ix,] <- x
+    }
   }
-  current[[dataset]][[model]] <- action
   save(current,file=f)
 }
 
-get_progress <- function(f="progress.rdata") {
+get_progress <- function(f) {
+  if (!file.exists(f)) stop("progress file does not exist")
   load(f)
-  x <- melt(current)
-  colnames(x) <- c("action","model","dataset")
+  x <- data.frame(dataset=current[,1],model=current[,2],action=current[,3],time=as.numeric(current[,4]))
+  rownames(x) <- NULL
+  x$model <- as.character(x$model)
   ma <- lapply(unique(x$model),function(m) data.frame(model=m,modelatts(m)))
   ma <- do.call(rbind,ma)
   merge(x,ma,by="model")
 }
 
-library(testthat)
-test_that("view progress makes correct list", {
-  f <- "tmp.rdata"
-  file.remove(f)
-  progress("a","b","start",f=f)
-  load(f)
-  expect_that(current, equals(list(a = list(b = "start"))))
-  progress("a","c","start",f=f)
-  load(f)
-  expect_that(current, equals(list(a = list(b = "start", c = "start"))))
-  progress("a","b","end",f=f)
-  load(f)
-  expect_that(current, equals(list(a = list(b = "end", c = "start"))))  
-  file.remove(f)
-})
+## progress <- function(dataset,model,action,notes=NULL,f="progress.rdata") {
+##   if (!file.exists(f)) {
+##     current <- list()
+##     save(current,file=f)
+##   } 
+##   load(f)
+##   if (length(current[[dataset]]) == 0) {
+##     current[[dataset]] <- list()
+##   }
[…]

### 2012-06-01 docs/experiments.tex
Work on prediction
Add latest results tables.

@@
 
-\input{../figs/results-llk2.tex}
+\input{../figs/results-llk3.tex}
+\input{../figs/results-recall-5.tex}
+\input{../figs/results-recall-20.tex}
 
@@
 
-[TODO: Recall table.]
-
-[TODO: Mention that we cap the value of $K$]
-
-[TODO: Discuss results.]
-%\input{../figs/results-recall.tex}
+In order to investigate the role of the number of clusters, $K$, we introduce an upper bound on the number of clusters during the fitting process.
+In Table ~\ref{tab:results} we compare the loglikelihood on held-out test data while varying the maximum allowed $K$.
+A relational event model (i.e.
+our model with $K=1$) outperforms the baselines on these data sets.
+For the synthetic data example introduced in Section \ref{sec:simulation} the fitted models with $K>1$ have predictive performance comparable to the true model; a standard relational event model does not perform as well because it does not have the flexibility to model the dynamics of each block separately.
+For the classroom data set we see only a slight improvement with $K>1$, while with the mobile phone calls the $K=3$ model performs best.
+In Table~\ref{tab:recall5} and Table~\ref{tab:recall20} we include the corresponding results for the recall experiment at a cutoff of 5 and 20, respectively.  The results there are similar.
 

### 2012-06-01 debug/llktable.r

@@ -1,0 +1,37 @@
+library(brem)
+#load("results/llk.rdata")  # using last draw
+load("results/allpred.rdata") # averaged over draws
+
+dataset <- "classroom-16"
+models <- c("kinit2.kmax1.sm0.nb0.pshift1.deg1.trans1.collapse1.xsigalpha1000.xsigbeta1000",
+            "kinit2.kmax2.sm0.nb0.pshift1.deg1.trans1.collapse1.xsigalpha1000.xsigbeta1000")
+
+  load(paste("data/",dataset,".rdata",sep=""))
+  M <- nrow(A)
+  P <- 13
+  ego <- 1
+  train.ix <- 1:nrow(train)
+  test.ix  <- (1:nrow(A))[-train.ix]
+  s <- new(RemStat,A[,1],as.integer(A[,2])-1,as.integer(A[,3])-1,N,M,P,ego)
+  s$precompute()
+  s$transform()
+
+folder <- paste("results/",dataset,"/fits/",sep="")
+llks <- lapply(models,function(model) {
+  load(paste(folder,model,".rdata",sep=""))
+  K <- dim(fit$params$beta)[2]
+  list(train = RemLogLikelihoodPcSubset(fit$params$beta, fit$params$z - 1, s$ptr(),  K, train.ix - 1)[train.ix],
+       test  = RemLogLikelihoodPcSubset(fit$params$beta, fit$params$z - 1, s$ptr(),  K, test.ix  - 1)[test.ix])
+})
+names(llks) <- models
+
+# Averaged over samples
+a <- allpred[[dataset]][[models[1]]]
+b <- allpred[[dataset]][[models[2]]]
+mean(a$llk$test)
+mean(b$llk$test)
+
+# Using last sample
+mean(llks[[models[1]]]$test)
+mean(llks[[models[2]]]$test)
+

### 2012-06-01 results.r

@@ -30,15 +30,6 @@ for (i in 1:length(r)) {
 }
 
 
-## rks <- list()
-## for (i in 1:length(r)) {
-##   for (j in 1:length(r[[i]])) {
-##     rks[[i]][[j]] <- list(train=recall(r[[i]][[j]][[k]]$train),
-2##                           test =recall(r[[i]][[j]][[k]]$test))
-##   }
-## }
-
-
 
 # recall(r$test,top=1:30)
 library(reshape2)
@@ -67,6 +58,51 @@ xr <- xtable(final,caption="Comparing mean loglikelihood for each event across m
 print(xr,include.rownames=FALSE,file=paste("figs/results-llk3.tex",sep=""),NA.string="",table.placement="t",size="footnotesize",sanitize.text.function=identity)
 
 
+r <- allpred
+ix <- which(sapply(r,length) == 0)
+if (length(ix) > 0) r <- r[-ix]
+rks <- list()
+for (i in 1:length(r)) {
+  rks[[i]] <- list()
+  for (j in 1:length(r[[i]])) {
+    tr <- recall(r[[i]][[j]]$rks$train)$recall#[c(5,20)]
+    te <- recall(r[[i]][[j]]$rks$test)$recall#[c(5,20)]
+    rks[[i]][[j]] <- list(train=list("5"=tr[1],"20"=tr[2]),
+                          test =list("5"=te[1],"20"=te[2]))
+  }
+  names(rks[[i]]) <- names(r[[i]])
+}
+names(rks) <- names(r)
+
+
+r <- melt(rks)
+colnames(r)[2:5] <- c("metric","type","model","dataset")
+ma <- lapply(unique(r$model),function(m) data.frame(model=m,modelatts(m)))
+ma <- do.call(rbind,ma)
+x <- merge(r,ma,by="model")
+tmp <- dcast(x,xsigalpha +xsigbeta+metric+type+ dataset ~ kinit + kmax + pshift + deg, fun.aggregate=mean)
+tmp$xsigalpha <- tmp$xsigalpha/1000
+tmp$xsigbeta  <- tmp$xsigbeta/1000
+colnames(tmp)[1:2] <- c("alpha","beta")
+subset(tmp,metric=="5" & alpha==5 & type=="test")[,c("dataset","2_10_0_0","2_1_1_1","2_2_1_1","2_3_1_1","2_10_1_1")]
[…]

### 2012-06-01 docs/experiments.tex
Small edits and TODOs in Experiments/Model sections.

@@
 For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
-The two panels on the left show the variation in $s_0(t_M,i,j)$, $s_1(t_M,i,j)$ and $s_3(t_M,i,j)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
+The two panels on the left show the variation in $s_0(t_M,i,j,\mathcal{A}_t)$, $s_1(t_M,i,j,\mathcal{A}_t)$ and $s_3(t_M,i,j,\mathcal{A}_t)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
 The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
@@
 We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood and recall for held-out data.
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.
-I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
-We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
-The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
+Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\hat{\lambda}_{i,j}(t_m) = \frac{1}{S}\sum_k \lambda_{i,j}(t_m | \boldsymbol{\beta}^{(k)}, \mathbf{z}^{(k)})$ is averaged using $K$ posterior samples given the training data.
+% TODO: We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
+% TODO: The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
 
@@
 
-In order to investigate the role of the number of clusters, $K$, we introduce an upper bound on the number of clusters during the fitting process.
-In Table ~\ref{tab:results} we compare the loglikelihood on held-out test data while varying the maximum allowed $K$.
+In order to investigate the role of the number of clusters, we introduce an upper bound $K^*$ on the number of clusters during the fitting process.
+In Table ~\ref{tab:results} we compare the loglikelihood on held-out test data while varying the maximum allowed $K^*$.
 A relational event model (i.e.

### 2012-06-01 docs/model.tex
Small edits and TODOs in Experiments/Model sections.

@@
 The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
-% TODO: Expand on this?
+{\color{red} TODO: Expand on this?}
 
@@
 For example, a particular node sending often may indicate they will continue to be a sender in the future.
-We normalize these counts by the number of events up until a dyad's prior changepoint.
+We normalize these counts by the number of events up until a dyad's prior changepoint by using $f(x) = \log \frac{x+1}{m + N(N-1)}$.
 %\footnote{This seems to avoid ``explosion'' when simulating from the model.
@@
 Aalen has a few suggestions.}
-Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}
+Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.
 
@@
 The proposed family of models generalizes several important special cases.
-For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.
+For example,  using only the intercept statistic $s_0(t,i,j,\mathcal{A}_t) = 1$ is analogous to the stochastic block model for static networks.
 Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
@@
 \end{align}
-The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.
+The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1973}, though here the possible choices are all the dyads in $\mathcal{R}$.
 
@@
 \hline
-Intercept& $s_{0}(t,i,j) = 1$\\
-Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
-Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
-Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
+Intercept& $s_{0}(t,i,j,\mathcal{A}_t) = 1$\\
+Reciprocity (AB-BA)& $s_{1}(t,i,j,\mathcal{A}_t) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
+Turn-continuing (AB-AY)& $s_{2}(t,i,j,\mathcal{A}_t) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
+Turn-taking (AB-BY)&$s_{3}(t,i,j,\mathcal{A}_t) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
 %Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
 %Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
-Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
-Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
-Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
+Sender out-degree& $s_{4}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
+Sender in-degree& $s_{5}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
+Dyad count& $s_{6}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
 %Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
@@
 \end{tabular}
-\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{ij}$ previous to event $m$.}
+\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{i,j}(t)$ previous to event $m$.}
 \label{tab:stats}

### 2012-06-01 docs/intro.tex
Move some citations around.

--- 2012-06-01 docs/intro.tex
@@
 
-In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez2011} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.
+In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.
 While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.
@@
 In this way one can compare theories about the underlying processes and make predictions about future data conditioned on the past.
+{\color{red} ...and simultaneously adjust for heterogeneity in these
+  processes?}
 

### 2012-06-01 docs/model.tex
Move some citations around.

@@
 (b) The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{3,4}(t)$ (dotted).
-We assume $\lambda_{3,4}$ can only change after events sent or received by node 3.
+We assume $\lambda_{3,4}(t)$ can only change after events sent or received by node 3.
 }
@@
 \subsection{Related work}
-The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
+Stochastic blockmodels have been extended to longitudinal network data
+involving a sequence of networks occurring at discrete times
+\cite{Ishiguro2010, Rodriguez2011}, while we focus on methods for
+analyzing a stream of relational events. 
+The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
 % \cite{AalenOddO.2008}
@@
 By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions.  
-We instead use latent variables to allow for heterogeneity in intensities across possible events.
+We instead use latent variables to allow for heterogeneity in intensities across the set of possible events.

### 2012-06-01 docs/experiments.tex
Experiments section.

@@
 
-A variety of real world datasets are used to explore the efficacy of the model.
-Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
+A variety of real world data sets are used to explore the efficacy of the model.
+The following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
+For each data set, we hold out the final $M_{test}$ events for evaluation of the model.
 
@@
 %\item Kiel email: \cite{Ebel2002}
-\item Classrooms: TODO describe datasets. Each have about 20 people and 400 events.  \cite{McFarland2001}
-\item Eckmann email: dyadic emails.  2000 training events among 88 people and 1300 test events.  \cite{Eckmann2004}
-\item Enron email: dyadic emails between months [TODO], 3000 training and 1000 test events among 141 actors \cite{Klimt2004}
+\item Classroom \cite{McFarland2001}: 445 directed communication among 27 people a high
+  school classroom collected via participant observation ($M_{test}= 145$).
+\item University email \cite{Eckmann2004}: 3300 dyadic emails among 88 people with at least 30
+  emails each ($M_{test} = 1300$)
+\item Enron email \cite{Klimt2004}: 4000 dyadic emails among 141
+  individuals between TODO months ($M_{test} = 1000$)
 %\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
-\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
+\item Twitter direct messages: Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.
+This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
 We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
 Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
-We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
-\item MIT Reality Mining: phone calls among the 89 recipients between the dates , using of 2000 training events followed by 1000 test events. \cite{Eagle2009}
+We use a subset 4330 events among 487 users who participated in more than one event ($M_{test} = 1330$).\footnote{This dataset will be made publicly available.}
+\item MIT Reality Mining \cite{Eagle2009}: 2000 phone calls among the 89 recipients between the dates TODO ($M_{test}= 1000$).
 \end{itemize}
@@
 We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood and recall for held-out data.
-Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\hat{\lambda}_{i,j}(t_m) = \frac{1}{S}\sum_k \lambda_{i,j}(t_m | \boldsymbol{\beta}^{(k)}, \mathbf{z}^{(k)})$ is averaged using $K$ posterior samples given the training data.
+%Each data set is first split into a training set and a test set, and 
+The loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\hat{\lambda}_{i,j}(t_m) = \frac{1}{L}\sum_l \lambda_{i,j}(t_m | \boldsymbol{\beta}^{(l)}, \mathbf{z}^{(l)},\mathcal{A}_t)$ is averaged using $L$ posterior samples given the training data.
 % TODO: We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
@@
 
-\input{../figs/results-llk3.tex}
+{ \footnotesize
+\input{../figs/results-llk4.tex}
 \input{../figs/results-recall-5.tex}
 \input{../figs/results-recall-20.tex}
-
+}
 %\subsection{Baselines}

### 2012-06-01 docs/inference.tex
Small notation fixes.
Minor edits and clean up.

@@
 p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
-\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
+\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{U}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
 \end{align*}
-where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
+where $\mathcal{U}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
 Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
@@
 \subsection{Hyperparameter settings}
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2000} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
+In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2000} with 5 extra clusters drawn from the prior.%\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
 This aims for similarly sized groups.}
-We set $\alpha_{\sigma}=3$ and $\beta_{\sigma}=1$ so that in the presence of little data we encourage shrinkage towards the upper level parameters $\mu$.  
+We set $\alpha_{\sigma}=5$ and $\beta_{\sigma}=1$ so that in the presence of little data we encourage shrinkage towards the upper level parameters $\mu$.  
 As with other models, we note that the predictive accuracy of the model can depend on the hyperparameters.

### 2012-06-01 docs/model.tex
Minor edits and clean up.

@@
 
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp2006} for static networks which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
+Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001} for static networks which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
 In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
@@
 As these intensities do not change under this specification, the likelihood simplifies to
-$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
-
+\begin{align}
+\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m}
+\prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}
+\end{align}
 Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
@@
 analyzing a stream of relational events. 
-The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
-% \cite{AalenOddO.2008}
+The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2011,Vu2011,Vu2011a}.
+% \cite{AalenOddO.2008} Opsahl2011,Stadtfeld2010
 Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.

### 2012-06-01 docs/experiments.tex
Update results tables.

@@
 \item Enron email \cite{Klimt2004}: 4000 dyadic emails among 141
-  individuals between TODO months ($M_{test} = 1000$)
+  individuals between July 2001 and August 2001 ($M_{test} = 1000$)
 %\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
@@
 We use a subset 4330 events among 487 users who participated in more than one event ($M_{test} = 1330$).\footnote{This dataset will be made publicly available.}
-\item MIT Reality Mining \cite{Eagle2009}: 2000 phone calls among the 89 recipients between the dates TODO ($M_{test}= 1000$).
+\item MIT Reality Mining \cite{Eagle2009}: 2000 phone calls among the
+  89 recipients between October 2001 and February 2002 ($M_{test}= 1000$).
 \end{itemize}
@@
 
-{ \footnotesize
+{ \scriptsize
 \input{../figs/results-llk4.tex}
-\input{../figs/results-recall-5.tex}
-\input{../figs/results-recall-20.tex}
+\input{../figs/results-recall-all.tex}
 }

### 2012-06-01 results.r

@@ -44,14 +44,13 @@ tmp$xsigbeta  <- tmp$xsigbeta/1000
 colnames(tmp)[1:2] <- c("alpha","beta")
 subset(tmp,metric=="mllk" & alpha==5 & type=="test")[,c("dataset","2_10_0_0","2_1_1_1","2_2_1_1","2_3_1_1","2_10_1_1")]
 
-
 a <- subset(tmp,metric=="llk" & alpha==5 & type=="test")[,c("dataset","2_10_0_0","2_1_1_1","2_2_1_1","2_3_1_1","2_10_1_1")]
 b <- subset(x,model %in% c("online","marg","uniform"))
 b <- dcast(b,metric + type + dataset ~ model)
 b <- b[1:length(datasets),c(3,6,4,5)]
 a <- a[,-1]
 final <- cbind(b,a)[c(7,3,4,5,6,8),]
-colnames(final)[1:9] <- c("Dataset","\\texttt{unif}","\\texttt{marg}","\\texttt{online}","\\texttt{BM}","$K=1$","$K=2$","$K=3$","$K=10$")
+colnames(final)[1:9] <- c("Dataset","\\texttt{unif}","\\texttt{marg}","\\texttt{online}","\\texttt{BM}","$K^*=1$","$K^*=2$","$K^*=3$","$K^*=10$")
 final[,1] <- c("Synthetic","Classroom","University Email","Enron Email","Mobile Phone Calls","Twitter Dir. Messages")
 library(xtable)
 xr <- xtable(final,caption="Comparing mean loglikelihood for each event across methods for each dataset.  Larger values are better.  See text for details.",label="tab:results",digits=3)
@@ -93,26 +92,28 @@ colnames(r)[2:5] <- c("metric","type","model","dataset")
 ma <- lapply(unique(r$model),function(m) data.frame(model=m,modelatts(m)))
 ma <- do.call(rbind,ma)
 x <- merge(r,ma,by="model")
+x$metric <- factor(x$metric,c("5","20"))
 tmp <- dcast(x,xsigalpha +xsigbeta+metric+type+ dataset ~ kinit + kmax + pshift + deg, fun.aggregate=mean)
 tmp$xsigalpha <- tmp$xsigalpha/1000
 tmp$xsigbeta  <- tmp$xsigbeta/1000
 colnames(tmp)[1:2] <- c("alpha","beta")
[…]

### 2012-06-01 docs/abstract.tex
Small paper edits.  A few TODOs inserted.
Move some citations around.
Minor edits and clean up.
Tweaked the abstract.  Doubt I'm going to find a better way to write it.

@@
 \begin{abstract}
-Observing network data as a stream of relational events provides an opportunity to understand differences in how individuals interact over time.
-For continuous-time network data, several approaches have recently been proposed for modeling dyadic event rates, conditioned on the observed history of events and nodal or dyadic covariates.
+%Observing network data as a sequence of relational events provides an opportunity to understand differences in how individuals interact over time.
+For continuous-time network data, several approaches have recently been proposed for modeling dyadic event rates conditioned on the observed history of events and nodal or dyadic covariates.
 In many cases, however, interaction propensities -- and even the underlying mechanisms of interaction -- vary systematically across subgroups whose identities are unobserved.
 For static networks, such heterogeneity has been treated via methods such as stochastic blockmodeling, which operate by assuming latent groups of individuals with similar tendencies in their group-wise interactions.
-Here, we combine these two approaches by positing a latent partition of the vertex set such that event dynamics within and between node sets evolve in potentially distinct ways.
+Here, we combine these two approaches by positing a latent partition of the node set such that event dynamics within and between subsets evolve in potentially distinct ways.
 We illustrate the use of our model family by application to several forms of dyadic interaction data, including email communication and Twitter direct messages.
@@
 Our approach illustrates the utility of latent structure methods based on detailed dynamics, which can succeed even in the absence of differences in marginal interaction rates across groups. 
-% TODO: This last sentence is a bit unclear for a new reader.
+%{ \color{red}  TODO: This last sentence is a bit unclear for a new reader. TODO: A few of theses sentences might be too long. TODO: stream vs sequence in first sentence?}
 \end{abstract}

### 2012-06-01 docs/discussion.tex
Tweaked the discussion section.

@@
 %PS: I would prefer to say: In this paper, we propose (or we proposed)….
-We propose a hierarchical approach for modeling event-based network data as a continuous-time Markov process. 
-Our approach generalizes traditional static notions of stochastic equivalence on nodes (such as stochastic blockmodels) to a dynamic formulation. 
-The proposed model posits the existence of groups of nodes, such that all members of a group are governed by the same dynamic process, and groups are differentiated by having different dynamic processes.
+%We propose a hierarchical approach for modeling event-based network data as a continuous-time Markov process. 
+We have here introduced a family of relational event models that can flexibly capture heterogeneity in the underlying interaction dynamics.
+Our approach generalizes traditional, static notions of stochastic equivalence on nodes (such as stochastic blockmodels) to the dynamic context. 
+The proposed model family posits the existence of groups of nodes, such that all members of a group are governed by the same dynamic process, and groups are differentiated by having different dynamic processes.
 %PS: sentence below is incomplete (after the first "the")…..I decided to take it out anyway..
@@
 %This analysis suggests 
-Our approach has the ability to uncover systematic differences in dynamic behaviors among subsets of nodes.
+Our approach has the ability to uncover systematic differences in dynamic behaviors among subsets of nodes, even in the absence of differences in marginal interaction rates.
 % across a variety of data sets.
-In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
-In particular, we show that our proposed approach leads to improved predictive accuracy when compared to both (a) relational event models that lack latent clusters, and (b) stochastic blockmodels for count data, both of which are both special cases of our model.
-Though prediction is not the main focus of the model, the results provide some evidence that having latent structure (i.e.
+The analyses of Section \ref{sec:experiments} show that this model family has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
+In particular, our proposed approach leads to improved predictive accuracy when compared to both (a) relational event models that lack latent clusters, and (b) stochastic blockmodels for count data, both of which are special cases of our model family.
+Though prediction is not the main focus of our effort, these results provide evidence that having latent structure (i.e.
 K>1) can lead to improvements in predictive power.
-Our prediction results show that it is easy to overfit with this model.
+However, our prediction results also show that it is easy to overfit with this model family.
 %If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model. 
 %{\red (TODO: confirm this)}
- As K increases beyond 2 (e.g., for the unrestrained CRP prior), the model is prone to overfit.
+ As K increases beyond 2 (e.g., for the unrestrained CRP prior), these models are prone to overfit.
 This is not surprising, given that the number of fitted parameters scales as the square of the number of components K in the model.
-A natural future direction worth exploring for this model is a class of priors that are more resistant to overfitting.
+A natural direction worth exploring for this family is a class of priors that are more resistant to overfitting (e.g., by imposing more structure on the parameters within each group).
 
 Another direction to explore is that of including different statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and using our approach to study how the roles of these statistics vary across nodes.
-One potentially interesting extension, analogous to \cite{Airoldi2008}, would be to allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point---we leave this to future work.
+One potentially interesting extension, analogous to \cite{Airoldi2008}, would be to allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.
+Allowing underlying dynamics to be governed by a latent membership structure opens the door to a wide range of possibilities, with many opportunities for further elaboration.
 

### 2012-06-01 docs/simulation.tex
Minor edits and clean up.

@@
 \centering
-\includegraphics[width=1.5in]{../figs/synthetic/mat.pdf}
+\includegraphics[width=1.5in]{../figs/synthetic/mat.pdf} %counts2.pdf
+\vspace{.25cm}
+\caption{}
 \end{subfigure}
@@
 \includegraphics[width=2in]{../figs/synthetic/counts.pdf}
+\caption{}
 \end{subfigure}
@@
 \centering
-\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
+\includegraphics[width=2.1in]{../figs/synthetic/params-estimates.pdf}
+\vspace{-.5cm}
+\caption{}
 \end{subfigure}
-\caption{[TODO: Remove AB-XA, etc from the plot.
-Rotate them.
-Use subfigure labels.] Illustration of 2000 simulated events, as described in text.
-Left: Counts of each dyad.
-Center: Boxplot of distribution of participation counts across dyads.
+\caption{Illustration of 2000 simulated events, as described in text.
+(a) Counts of each dyad.
+(b) Boxplot of distribution of participation counts across dyads.
 The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
-Right: Parameters (in red) and posterior credible intervals (in black).}
+(c) Parameters (in red) and posterior credible intervals (in black).}
 \label{fig:syncounts}
@@
 \section{Simulation}
+\label{sec:simulation}
 

### 2012-06-01 docs/intro.tex
Some small tweaks to the intro and model sections.

@@
 
-Borrowing from the intuition of stochastic blockmodels, we propose a novel continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
+Borrowing from the intuition of stochastic blockmodels, we propose a novel continuous-time model family for network-based event sequences where latent clusters of nodes share similar patterns of interaction.
 Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
-In this way one can compare theories about the underlying processes and make predictions about future data conditioned on the past.
-{\color{red} ...and simultaneously adjust for heterogeneity in these
-  processes?}
+In this way one can compare theories about underlying processes and make predictions about future data conditioned on the past, simultaneously adjusting for unobserved heterogeneity.
 

### 2012-06-01 docs/inference.tex
Commenting out footnotes.

@@
 
-We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
+We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.%\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
 

### 2012-06-01 docs/model.tex
Some small tweaks to the intro and model sections.
Commenting out footnotes.

@@
 
-{\color{red} (Need a better transition here.)}  We aim to learn about the dynamics between subsets of nodes.
+Here, we aim to learn about the dynamics within and between subsets of nodes.
 To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
@@
 Though only the statistics in Table \ref{tab:stats} are used in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
-The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
+The only restriction is that the statistic may not change in value between each observed event.%\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
 % TODO: Also, need to be restricted to ego conditions if we want it fast

### 2012-06-01 docs/experiments.tex
Space-saving stuff.

@@
 
-\subsection{Data}
+%\subsection{Data}
 
@@
 For each data set, we hold out the final $M_{test}$ events for evaluation of the model.
-
-\begin{itemize}
+\begin{list}{\labelitemi}{\leftmargin=1em}
+%\begin{itemize}
 %\item Kiel email: \cite{Ebel2002}
@@
   school classroom collected via participant observation ($M_{test}= 145$).
-\item University email \cite{Eckmann2004}: 3300 dyadic emails among 88 people with at least 30
-  emails each ($M_{test} = 1300$)
+\item University email \cite{Eckmann2004}: 3300 dyadic emails among 88
+  users with at least 30 emails ($M_{test} = 1300$).
 \item Enron email \cite{Klimt2004}: 4000 dyadic emails among 141
-  individuals between July 2001 and August 2001 ($M_{test} = 1000$)
+  individuals between July 2001 and August 2001 ($M_{test} = 1000$).
 %\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
@@
   89 recipients between October 2001 and February 2002 ($M_{test}= 1000$).
-\end{itemize}
+%\end{itemize}
+\end{list}
 
@@
 For the classroom data set we see only a slight improvement with $K>1$, while with the mobile phone calls the $K=3$ model performs best.
-In Table~\ref{tab:recall5} and Table~\ref{tab:recall20} we include the corresponding results for the recall experiment at a cutoff of 5 and 20, respectively.  The results there are similar.
+In Table~\ref{tab:recall20} we include the corresponding results for the recall experiment at a cutoff of 5 and 20, respectively.  The results there are similar.
 

### 2012-06-01 docs/experiments.tex
Minor edits.

@@
 Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.
-Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.
+Finally, \texttt{BM} is a stochastic blockmodel (i.e. our model with only an intercept term).
+Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} and \texttt{BM} can capture individual heterogeneity and group level heterogeneity in overall activity.
 
@@
 In order to investigate the role of the number of clusters, we introduce an upper bound $K^*$ on the number of clusters during the fitting process.
-In Table ~\ref{tab:results} we compare the loglikelihood on held-out test data while varying the maximum allowed $K^*$.
+In Table ~\ref{tab:results} we compare the loglikelihood on held-out test data while varying  $K^*$.
 A relational event model (i.e.

### 2012-06-01 docs/inference.tex
Minor edits.

@@
 where $\mathcal{U}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
-Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
+Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = K+1 |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
 
@@
 \subsection{Hyperparameter settings}
-In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2000} with 5 extra clusters drawn from the prior.%\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
+In our experiments we use $\alpha=1$ and use Algorithm 8 from \cite{Neal2000} with 5 extra clusters drawn from the prior.%\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
 This aims for similarly sized groups.}

### 2012-06-01 docs/model.tex
Space-saving stuff.
Take out a TODO item.

@@
 \begin{subfigure}[b]{0.45\linewidth}
-  \begin{tikzpicture}[scale=.8,transform shape,thick]
+  \begin{tikzpicture}[scale=.65,transform shape,thick]
     \node[vertex] (n1) at (0,0)  {1};
@@
 \begin{subfigure}[b]{0.45\linewidth}
-\begin{tikzpicture}[scale=1,transform shape,thick]
+\begin{tikzpicture}[scale=.8,transform shape,thick]
 \draw[-stealth] (0,0) -- (0,3);
@@
 
-We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta_{\sigma})$.
-The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
-{\color{red} TODO: Expand on this?}
+We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta_{\sigma})$.
+The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$, allowing for a flexible number of clusters.
+% TODO: expand on this
 

### 2012-06-01 docs/discussion.tex
Update results tables to save some space.

@@
 Though prediction is not the main focus of our effort, these results provide evidence that having latent structure (i.e.
-K>1) can lead to improvements in predictive power.
+$K > 1$) can lead to improvements in predictive power.
 However, our prediction results also show that it is easy to overfit with this model family.
@@
 %{\red (TODO: confirm this)}
- As K increases beyond 2 (e.g., for the unrestrained CRP prior), these models are prone to overfit.
-This is not surprising, given that the number of fitted parameters scales as the square of the number of components K in the model.
+ As $K$ increases beyond 2 (e.g., for the unrestrained CRP prior), these models are prone to overfit.
+This is not surprising, given that the number of fitted parameters scales as the square of the number of components $K$ in the model.
 A natural direction worth exploring for this family is a class of priors that are more resistant to overfitting (e.g., by imposing more structure on the parameters within each group).

### 2012-06-01 docs/experiments.tex
Update results tables to save some space.

@@
 In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.
-At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean rank across the $M$ events.
+At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean number of events ranking above cutoffs $\kappa=5$ and $\kappa=20$.
 
-{ \scriptsize
 \input{../figs/results-llk4.tex}
 \input{../figs/results-recall-all.tex}
-}
-%\subsection{Baselines}
 

### 2012-06-01 docs/experiments.tex
Final NIPS submission.

@@
 Similarly, block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
-The model has learned that this structure exists in the data and our inferences from the parameter estimates are sensible given the observed statistics.
+The model has learned that this structure exists in the data, and furthermore our inferences from the parameter estimates are sensible given the observed statistics.
 

### 2012-10-30 docs/submission.tex
Fix up intro example figure.
Edit discussion and reorganize a few things.
Removing supplemental material from main doc.
Commenting out footnotes.
Add AISTATS style files.

@@
-\documentclass{article}
-\usepackage{nips12submit_e,times}
+\documentclass[twoside]{article}
+\usepackage{aistats2e}
+% \documentclass{article}
+% \usepackage{nips12submit_e,times}
 \usepackage{tikz}
@@
 \usepackage{footnote}
+\usepackage{sidecap}
 
-\title{Stochastic blockmodeling of relational event dynamics}
-% Dynamic stochastic equivalence for relational events
-% Stochastic Blockmodels for Relational Event Data}
-% Latent Variable Modeling of Relational Event Dynamics
-% Latent Variable Modeling of Relational Event Sequences
-% Latent Variable Modeling for Relational Event Sequences
-
-% \author{
-% Christopher DuBois
-% Department of Statistics\\
-% University of California, Irvine\\
-% \texttt{duboisc@ics.uci.edu} \\
-% \And
-% Coauthor \\
-% Affiliation \\
-% Address \\
-
-% \AND
-% Coauthor \\
-% Affiliation \\
-% Address \\
-% }
+%\usepackage[accepted]{aistats2e}
 
 \begin{document}
-\maketitle
-\savenotes
+
+\twocolumn[
+
+\aistatstitle{Stochastic blockmodeling of relational event dynamics}
+
+\aistatsauthor{ Anonymous Author 1 \And Anonymous Author 2 \And Anonymous Author 3 }
+
+\aistatsaddress{ Unknown Institution 1 \And Unknown Institution 2 \And Unknown Institution 3 } ]
+
+%\maketitle
 \input{abstract}
@@
 \bibliography{refs}
-\input{appendices}
-\spewnotes
+%\newpage
+%\input{appendices}
 \end{document}

### 2012-10-30 continuous/sample.r

@@ -1,3 +1,63 @@
+#' At each time t:
+#' for each (i,j) 
+#'   compute lambda_ij(t|.) = sum_kl exp{ phi_ik + phi_jl + beta_kl s_ij(t|.)}
+#' draw (i,j) ~ lambda_ij/sum_ij lambda_ij
+#' draw next t ~ Exp(sum_ij lambda_ij)
+
+CLogLambda <- function(i,j,beta,phi,s) {
+  lam <- 0
+  for (k in 1:K) {
+    for (l in 1:K) {
+      lam <- lam + 
+    }
+  }
+}
+
+simulate.cbrem.cond <- function(beta,phi,train,M,N,ego=FALSE) {
+  K <- dim(beta)[2]
+  P <- dim(beta)[1]
+  edgelist <- train[nrow(train),,drop=FALSE]
+  time <- edgelist[1]
+  s <- InitializeStatisticsArray(N,P);
+  for (m in 1:nrow(train)) {
+    i <- train[m,2]
+    j <- train[m,3]
+    s <- UpdateStatisticsArray(s,m-1,i-1,j-1,N,P)
+  }
+  lambda <- matrix(0,N,N)
+  for (i in 1:N) {
+    for (j in 1:N) {
+      lambda[i,j] <- LogLambda(i-1,j-1,z[i]-1,z[j]-1,s,beta,N,K,P)
+    }
+  }
+  for (m in 1:M) {
+    for (r in 1:N) {
+      lambda[j,r] <- LogLambda(j-1,r-1,z[j]-1,z[r]-1,s,beta,N,K,P)
+      lambda[i,r] <- LogLambda(i-1,r-1,z[i]-1,z[r]-1,s,beta,N,K,P)
+      if (!ego) {
+        lambda[r,j] <- LogLambda(r-1,j-1,z[r]-1,z[j]-1,s,beta,N,K,P)
+        lambda[r,i] <- LogLambda(r-1,i-1,z[r]-1,z[i]-1,s,beta,N,K,P)
+      }
+    }
+    diag(lambda) <- -Inf
+    cells <- cbind(as.vector(row(lambda)), as.vector(col(lambda)), exp(as.vector(lambda)))
+    if (any(is.nan(cells[,3]))) {
+      stop("Intensity explosion: infinite lambda_ij present")
+    }
+    drawcell <- sample(1:NROW(cells),1,prob=cells[,3])
+    i <- cells[drawcell,1]
+    j <- cells[drawcell,2]
[…]

### 2012-10-30 pkg/src/brem.cpp

@@ -400,6 +400,14 @@ private:
   vector< vector<int> > u;
 };
 
+double LogLambdaContinuous(int i, int j, vector<double> s, Rcpp::NumericVector beta, Rcpp::NumericVector phi, int N, int K, int P) {
+  double lam = 0;
+  //  double lam = beta[threeDIndex(0,zi,zj,P,K,K)]; // intercept
+  for (int p = 1; p < 12; p++) {
+    //    lam += s[p] * beta[threeDIndex(p,zi,zj,P,K,K)];
+  }
+  return lam;
+}
 
 // Use the precomputed set of statistics to copmute lambda
 

### 2012-10-31 docs/submission.tex
Update submission with a few bits of reformatting.
Also included some of the material from the reviewer responses.

@@
 \usepackage{sidecap}
+\DeclareCaptionType{copyrightbox}
 
@@
 %\maketitle
-\input{abstract}
-\input{intro}
-\input{model}
-\input{inference}
-\input{simulation}
-\input{experiments}
-\input{discussion}
+\begin{abstract}
+%Observing network data as a sequence of relational events provides an opportunity to understand differences in how individuals interact over time.
+For continuous-time network data, several approaches have recently been proposed for modeling dyadic event rates conditioned on the observed history of events and nodal or dyadic covariates.
+In many cases, however, interaction propensities -- and even the underlying mechanisms of interaction -- vary systematically across subgroups whose identities are unobserved.
+For static networks, such heterogeneity has been treated via methods such as stochastic blockmodeling, which operate by assuming latent groups of individuals with similar tendencies in their group-wise interactions.
+Here, we combine these two approaches by positing a latent partition of the node set such that event dynamics within and between subsets evolve in potentially distinct ways.
+We illustrate the use of our model family by application to several forms of dyadic interaction data, including email communication and Twitter direct messages.
+Parameter estimates from the fitted models clearly reveal heterogeneity in the dynamics among groups of individuals.
+We also find that the fitted models have better predictive accuracy than either baseline models or relational event models without latent structure.
+Our approach illustrates the utility of latent structure methods based on detailed dynamics, which can succeed even in the absence of differences in marginal interaction rates across groups. 
+%{ \color{red}  TODO: This last sentence is a bit unclear for a new reader. TODO: A few of theses sentences might be too long. TODO: stream vs sequence in first sentence?}
+\end{abstract}
+
+\section{Introduction}
+
+Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
+Recent work has expanded to include settings in which we observe events occurring between nodes over time (i.e., \emph{relational events}, as opposed to static edge structures or ongoing relationships), with the common goal of modeling interaction dynamics in terms of both endogenous mechanisms and exogenous covariates.
+A key concern in this regard is the ability to detect differential behavioral tendencies on the part of subsets of nodes, the dynamic analog of \emph{role structure} in classical network analysis \cite{Wasserman1994}.
+
+In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.
+While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.
+Consider, e.g., Figure~\ref{fig:introexample}.
+In the top left panel, we depict the time-marginalized frequencies of simulated interactions between members of two groups (A and B), with darker cells indicating higher interaction frequencies.
+As can be seen, a clear role structure is present, with members of each subgroup interacting at higher rates with co-members than out-group members; such a structure is easily detectable via conventional blockmodeling techniques.
+By contrast, consider the interaction patterns shown in the top right panel.
+Here, there is no systematic difference in marginal rates between the two groups, rendering them invisible under a standard blockmodeling approach.
+
+There is, however, a difference to be detected.
+The bottom panels of Figure~\ref{fig:introexample} show, for each respective simulation, parameters (as discussed in Section~\ref{sec:specification}) governing the tendency towards reciprocation for members of each group vis a vis communications coming from in- or out-group members.
+For both simulated cases, the log-hazard for an event in which a member of group A or group B immediately responds to an incoming communication from an in-group member is increased by an increment of 2 (relative to a non-reciprocating event), and the log-hazard for an event in which a member of group B immediately responds to an incoming communication from a member of group A is increased by a corresponding increment of 0.5.
+Groups A and B are thus distinctive in terms of their dynamic behavior, even in the absence of marginal differences in propensity to communicate.
+A blockmodeling approach that classifies nodes based on shared dynamics (rather than merely shared marginal communication rates) can potentially identify such subtle distinctions; in this paper, we introduce such an approach.
+
+%For example, a common aspect of human communication is \emph{reciprocity}: person A speaks to person B, then B has a higher propensity to speak to A.  
+%Through an event-based analysis we can study how nodes vary in their tendencies to behave in this manner.  %the similarity --- and differences --- in the dynamics of nodes in the network. 
+%This is particularly relevant when studying interactions among students in a classroom, where differences in conversational tendencies might be associated with covariates about the student or classroom, or based on unobserved groupings, and may have ramifications for education.
+%In such contexts, a standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.
+
+%When such .
+One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO], where this variation in rates may be associated with some unobserved quantity, e.g.
+a covariate such as age or sex, or perhaps shared [TODO].
+%Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by .
+Such techniques are readily applied to count data, [TODO: These let us study how shared behavior is associated with other covariates.]
+
+%1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of a tie between members
+
+%PS: much of this past work is for static network data.
+Do we want to make the observation here that we are thinking about blockmodels for *rate* data here in our discussion and in our example in figure 1 - if we don't some readers/reviewers might get a little confused.
+For example we could say that instead of a binomial model on edges we can think of a Poisson process yielding counts over time, where we can still think about blockmodels.
+%PS: also, we introduce the term "rates" above without really defining it, and then go on to talk about "mean rate" below.
+So I think we definitely need a sentence somewhere above that makes it clear we are talking about edges that instantaneously appear and disappear over time at some rate, rather than static edges.
+
+%If we are interested in studying a network of events, there are other important aspects of the dynamics in addition to the mean rate. 
+%For example, human conversation often has an increased propensity for reciprocated events.
+%PS: wording of next sentence is a little odd - not sure I understand what it means...
+%Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
+%[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
+%Figure \ref{fig:introexample} shows a simple example of how traditional blockmodels are unable to make inferences these types of dynamic phenomena.
+
+%A limitation of all of these approaches is the assumption of a single common behavior for all individuals.
+%In contrast, in real-world networks it is reasonable to expect {\it heterogeneity} in dynamic behavior.  
+%In an organization, the manner in which an individual communicates with others may be a function of an individual's {\it role} in the organization.
+%For example, in a university,  email communication patterns over time between professors, students, and staff will likely be quite different within and between the three groups.
+
+% - used to thinking about actors differentiated by rates
+% - other things in the dynamics than rates, e.g. form of interaction.
+% - Also important aspect of modeling social dynamics.
+% especially important in cases like plot 1(c) where classical methods
+% will fail.
+
+%PS: not totally happy with this example since the "roles" are completely determined by one's "job title" and I think we would prefer something
+% less explicit, e.g., that there are "leaders", "followers", etc, that go beyond just job titles. So feel free to edit this further!
+
+%
+%comprised of several teams, each team may uncover different patterns for collaborating via email---individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.
+
+Borrowing from the intuition of stochastic blockmodels, we propose a novel continuous-time model family for network-based event sequences where latent clusters of nodes share similar patterns of interaction.
+Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
+In this way one can compare theories about underlying processes and make predictions about future data conditioned on the past, simultaneously adjusting for unobserved heterogeneity.
+
+We describe how we learn the latent cluster assignments and model parameters via MCMC, and illustrate the behavior of the model with simulated data.
+Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
+We conclude by showing that the parameter estimates reveal interpretable structure in the event dynamics, enabling the study of a dynamic extension to stochastic equivalence.
+
+\begin{figure}
+\centering
+\includegraphics[scale=.45]{../figs/introexample/all}
+\caption{Illustration of differentiation by rate of interaction versus by dynamic behavior.
+The left column shows an example where within-block communication is large, and there is a higher tendency for reciprocity within a block.
+In the situation where the two groups are not differentiated by rate, as in the right column, a standard blockmodel is unable to distinguish between groups A and B.
+The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.}
+\label{fig:introexample}
+\end{figure}
+% {\color{red} (Suggest renaming "Dynamics parameters" to "Behavioral parameters" or something.  "Dynamics parameters" is not gramatically sound.)
+
+\section{Model}
+
+Consider a sequence of events $\mathcal{A} = (0,t_1, \ldots, t_M)$ arising from a nonhomogeneous Poisson process with  intensity $\lambda(t)$.
+If the intensity is left continuous and piecewise constant with respect to a set of knots $\tau$ then the likelihood can be written
+\begin{align}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda(t_m) \exp\left\{ - \int_{0}^{t_M} \lambda(s)ds \right\} \notag  \\
+&= \prod_{m=1}^M \lambda(t_m) \prod_{k=1}^{|\tau|} \exp\left\{ - (\tau_{k} - \tau_{k-1}) \lambda(\tau_k) \right\}%  \exp\left\{ - (t - \tau_{M}) \lambda(t) \right\}
+\end{align}
+%{\color{red} (You've left out the right-censoring factor here (and also below).)}
+
+One can extend the above to marked point processes where each event contains additional information.
+In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
+We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
+%The likelihood of an observed event history $\mathcal{A}$ extending from time 0 to the time of the final event, $t_M$, is 
+The likelihood of an observed event history $\mathcal{A}_{t_M}$ (extending from time 0 to the time of the final event and denoted $\mathcal{A}$ for convenience) is
+%The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
+\begin{align}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
+\label{eqn:llk}
+\end{align}
+
+\begin{figure*}
+\begin{subfigure}[b]{0.45\linewidth}
+\centering
+  \begin{tikzpicture}[scale=.65,transform shape,thick]
+    \node[vertex] (n1) at (0,0)  {1};
+    \node[vertex] (n2) at (2,2)  {2};
+    \node[vertex] (n3) at (4,0) {3};
+    \node[vertex] (n4) at (2,-2) {4};
+    \draw[-stealth] (.4,.4) -- (1.6,1.6);
+    \draw[-stealth] (2.4,1.6) -- (3.6,.4);
+    \node at (.7,1.1) {$t_1$};
+    \node at (3.3,1.1) {$t_2$};
+  \end{tikzpicture}
+  \caption{Dynamic network data}
+\end{subfigure}
+\begin{subfigure}[b]{0.45\linewidth}
+\centering
+\begin{tikzpicture}[scale=.8,transform shape,thick]
+\draw[-stealth] (0,0) -- (0,3);
+\draw[-stealth] (0,0) -- (5,0);
+
+% First timeline
+\draw (0,2) -- (1,2);
+\draw[fill=black] (1,2) circle (0.75mm);
+\draw (1,1.7) circle (0.75mm);
+\draw (1.08,1.7) -- (3,1.7);
+\draw[fill=black] (3,1.7) circle (0.75mm);
+\draw (3,1.7) circle (0.75mm);
+\draw (3,2.2) circle (0.75mm);
+\draw[-stealth] (3.05,2.2) -- (5,2.2);
+
+% second timeline
+\draw[dashed] (0,1) -- (3,1);
+\draw[fill=black] (3,1) circle (0.75mm);
+\draw (3,.75) circle (0.75mm);
+\draw[dashed, -stealth] (3.08,.75) -- (5,.75);
+
+% labels
+\node at (-.5,2) {$\lambda_{1,2}$};
+\node at (-.5,1) {$\lambda_{3,4}$};
+\node at (1,-.3) {$t_1$};
+\node at (3,-.3) {$t_2$};
+\end{tikzpicture}
+\caption{Intensities for two dyads}
+\end{subfigure}
+\caption[]{Illustration of relational event data and the assumptions of the model.
+(a) A sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.
+(b) The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{3,4}(t)$ (dashed).
+We assume $\lambda_{3,4}(t)$ can only change after events sent or received by node 3.
+}
+\label{fig:example}
+\end{figure*}
+
+Here, we aim to learn about the dynamics within and between subsets of nodes.
+To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
+\begin{align*}
+\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
+%z_i &\sim \mbox{CRP}(\alpha) &
+%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
+\end{align*}
+where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
+Thus, the rate of $(i,j)$ events has the same parameters as other events occurring between group $z_i$ and $z_j$.
+
+% In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
+As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{ij}(t)$ to change following an event where $i$ was the sender or the recipient.
+This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
+In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
+
+We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto N(0,1)$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta_{\sigma})$.
+The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$, allowing for a flexible number of clusters.
+% TODO: expand on this
+
+\subsection{Model specification}
+\label{sec:specification}
+
+Table ~\ref{tab:stats}  lists the statistics  $\mathbf{s}(t,i,j,\mathcal{A}_t)$ used in Section \ref{sec:experiments}.
+For example, a particular node sending often may indicate they will continue to be a sender in the future.
+We use $\mathbf{I}$ to denote the indicator function where $\mathbf{I}(A) = 1$ if $A$ is true, and 0 otherwise.
+We normalize these counts by the number of events up until a dyad's prior changepoint by using $f(x) = \log \frac{x+1}{m + N(N-1)}$.
+%\footnote{This seems to avoid ``explosion'' when simulating from the model.
+TODO: Explore the theory about these sorts of statistics.
+Aalen has a few suggestions.}
+Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.
+
+% TODO: Awkward next sentence
+The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
+For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
+Though only the statistics in Table \ref{tab:stats} are used in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
+The only restriction is that the statistic may not change in value between each observed event.%\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
+% TODO: Also, need to be restricted to ego conditions if we want it fast
+
+\subsection{Relation to other models}
+
+Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001} for static networks which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
+In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
+
+The proposed family of models generalizes several important special cases.
+For example,  using only the intercept statistic $s_0(t,i,j,\mathcal{A}_t) = 1$ is analogous to the stochastic block model for static networks.
+Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
+As these intensities do not change under this specification, the likelihood simplifies to
+\begin{align}
+\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m}
+\prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}
+\end{align}
+Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
+\begin{align}
+\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
+\label{eqn:multllk}
+\end{align}
+The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1973}, though here the possible choices are all the dyads in $\mathcal{R}$.
+
+\begin{table*}[t]
+\footnotesize
+\center
+\begin{tabular}{|l|l|}
+\hline
+Statistic & Formula\\
+\hline
+\hline
+Intercept& $s_{0}(t,i,j,\mathcal{A}_t) = 1$\\
+Reciprocity (AB-BA)& $s_{1}(t,i,j,\mathcal{A}_t) = \mathbf{I}(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
+Turn-continuing (AB-AY)& $s_{2}(t,i,j,\mathcal{A}_t) =  \mathbf{I}(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
+Turn-taking (AB-BY)&$s_{3}(t,i,j,\mathcal{A}_t) = \mathbf{I}(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
+%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = \mathbf{I}(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
+%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = \mathbf{I}(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
+Sender out-degree& $s_{4}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} \mathbf{I}(i_m=i) )$\\
+Sender in-degree& $s_{5}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} \mathbf{I}(j_m=i) )$\\
+Dyad count& $s_{6}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} \mathbf{I}(i_m=i,j_m=j) )$\\
+%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} \mathbf{I}(i_m=j))$\\
+%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} \mathbf{I}(j_m=j))$\\
+\hline
+\end{tabular}
+\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{i,j}(t)$ previous to event $m$. }
+\label{tab:stats}
+\end{table*}
+
+\subsection{Related work}
+Stochastic blockmodels have been extended to longitudinal network data
+involving a sequence of networks occurring at discrete times
+\cite{Ishiguro2010, Rodriguez2011}, while we focus on methods for
+analyzing a stream of relational events. 
+The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2011,Vu2011,Vu2011a}.
+% \cite{AalenOddO.2008} Opsahl2011,Stadtfeld2010
+Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
+By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions.  
+We instead use latent variables to allow for heterogeneity in intensities across the set of possible events.
+
+\section{Inference}
+
+We use Markov chain Monte Carlo to sample from the posterior distribution of our parameters.%\footnote{TODO: Describe why EM is hard.} % Note that approaches such as EM are difficult since the form of the likelihood does not admit an analytic expression for the expected complete data log likelihood.
+
+\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }
+
+We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
+\begin{align*}
+p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
+p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
+\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]} \ \   \times \\
+&\prod_{(i,j) \in \mathcal{U}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
+\end{align*}
+where $\mathcal{U}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
+Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = K+1 |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.
+
+\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }
+
+For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
+\begin{align*}
+p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
+p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
+p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} \ \times \\
+& \ \ \prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
+\end{align*}
+where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.
+We sample each $\beta_{k,l,p}$ via slice sampling.
+
+\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }
+
+The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
+\begin{align}
+\label{eqn:gibbs.sigma}
+\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim \\
+& \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{k,l,p} - \mu_p)^2\right).
+\end{align}
+Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$,
+\begin{align}
+\label{eqn:gibbs.mu}
+\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{k,l,p},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
+\end{align}
+
+\subsection{Hyperparameter settings}
+In our experiments we use $\alpha=1$ and use Algorithm 8 from \cite{Neal2000} with 5 extra clusters drawn from the prior.%\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.
+This aims for similarly sized groups.}
+We set $\alpha_{\sigma}=5$ and $\beta_{\sigma}=1$ so that in the presence of little data we encourage shrinkage towards the upper level parameters $\mu$.  
+As with other models, we note that the predictive accuracy of the model can depend on the hyperparameters.
+
+\begin{figure*}
+%\centering
+\begin{subfigure}[b]{0.25\textwidth}
+\centering
+\includegraphics[width=1.5in]{../figs/synthetic/mat.pdf} %counts2.pdf
+\vspace{.25cm}
+\caption{}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.3\textwidth}
+\centering
+\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
+\caption{}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.3\textwidth}
+\centering
+\includegraphics[width=2.1in]{../figs/synthetic/params-estimates.pdf}
+\vspace{-.5cm}
+\caption{}
+\end{subfigure}
+\caption{Illustration of 2000 simulated events, as described in text.
+(a) Counts of each dyad.
+(b) Boxplot of distribution of participation counts across dyads.
+The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.
+(c) Parameters (in red) and posterior credible intervals (in black).}
+\label{fig:syncounts}
+\end{figure*}
+
+\section{Simulation}
+\label{sec:simulation}
+
+We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 groups where 1) within group communication is more likely, 2) events among members in the first group are more likely to be reciprocated  (i.e.
+a positive \texttt{ab-ba} effect), and 3) events among members of the second group are more likely to be followed by an event with the same sender (i.e.
+a positive \texttt{ab-ay} effect).
+The specification of $\textbf{s}$ is therefore $\textbf{s}(t,i,j,\mathcal{A}_t) = [s_0, s_{1}(t,i,j,\mathcal{A}_t), s_{2}(t,i,j,\mathcal{A}_t)]$.
+For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
+Data is generated by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
+
+Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
+Intensities for reciprocal actions among nodes in block 1 are $e^3 \approx 20$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
+Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
+
+\section{Model fitting and experiments}
+\label{sec:experiments}
+
+%\subsection{Data}
+
+A variety of real world data sets are used to explore the efficacy of the model.
+The following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.
+For each data set, we hold out the final $M_{test}$ events for evaluation of the model.
+\begin{list}{\labelitemi}{\leftmargin=1em}
+%\begin{itemize}
+%\item Kiel email: \cite{Ebel2002}
+\item Classroom \cite{McFarland2001}: 445 directed communication among 27 people a high
+  school classroom collected via participant observation ($M_{test}= 145$).
+\item University email \cite{Eckmann2004}: 3300 dyadic emails among 88
+  users with at least 30 emails ($M_{test} = 1300$).
+\item Enron email \cite{Klimt2004}: 4000 dyadic emails among 141
+  individuals between July 2001 and August 2001 ($M_{test} = 1000$).
+%\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
+\item Twitter direct messages: Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.
+This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.
+We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
+Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
+We use a subset 4330 events among 487 users who participated in more than one event ($M_{test} = 1330$).\footnote{This dataset will be made publicly available.}
+\item MIT Reality Mining \cite{Eagle2009}: 2000 phone calls among the
+  89 recipients between October 2001 and February 2002 ($M_{test}= 1000$).
+%\end{itemize}
+\end{list}
+
+\subsection{Model-based exploratory analysis}
+
+Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
+We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
+For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
+% The two panels on the left show the variation in $s_0(t_M,i,j,\mathcal{A}_t)$, $s_1(t_M,i,j,\mathcal{A}_t)$ and $s_3(t_M,i,j,\mathcal{A}_t)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
+%The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
+% Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
+ block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
+The model has learned that this structure exists in the data, and furthermore our inferences from the parameter estimates are sensible given the observed statistics.
+{\color{red} TODO: Elaborate on what this figure shows.}
+
+\subsection{Prediction experiments}
+
+We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood and recall for held-out data.
+Throughout the experiments we use all of the statistics found in Table ~\ref{tab:stats}.
+%Each data set is first split into a training set and a test set, and 
+The loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $$\hat{\lambda}_{i,j}(t_m) = \frac{1}{L}\sum_l \lambda_{i,j}(t_m | \boldsymbol{\beta}^{(l)}, \mathbf{z}^{(l)},\mathcal{A}_t)$$ is averaged using $L$ posterior samples given the training data.
+% TODO: We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
+% TODO: The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.
+
+In addition, we compute recall to evaluate whether the next observed event is among the most likely according to the model.
+At each event $m$ we sort the predicted intensities of all possible events in decreasing order, find the rank of the observed event in the list of predicted intensities, and compute the mean number of events ranking above cutoffs $\kappa=5$ and $\kappa=20$.
+
+\input{../figs/results-llk4.tex}
+\input{../figs/results-recall-all.tex}
+
+Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i) \sum_{m:t_m < t} \mathbf{I}(j_m=j)$.
+Finally, \texttt{BM} is a stochastic blockmodel (i.e. our model with only an intercept term).
+Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} and \texttt{BM} can capture individual heterogeneity and group level heterogeneity in overall activity.
+
+\begin{figure*}[t]
+\centering
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed}
+\caption{Observed counts}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1}
+\caption{Intercept estimates}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2}
+\caption{\texttt{ab-ba} estimates}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3}
+\caption{\texttt{ab-by} estimates}
+\end{subfigure} \\
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/observed-zoom}
+\caption{Observed counts}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\end{subfigure}
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/1-zoom}
+\caption{Intercept estimates}
+\end{subfigure}
+~
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/2-zoom}
+\caption{\texttt{ab-ba} estimates}
+\end{subfigure}
+\begin{subfigure}[b]{0.22\textwidth}
+\centering
+\includegraphics[scale=.25]{../figs/eckmann-small/parmat/3-zoom}
+\caption{\texttt{ab-by} estimates}
+\end{subfigure}
+\caption{Comparing observed counts and parameter estimates.
+Darker values are larger.
+Estimates are rescaled posterior means $(\hat{\beta}_{z_i,z_j,p} - \hat{\mu}_p)/\hat{\sigma}_p$ for each dyad $(i,j)$.
+Learned parameters suggest heterogeneity exists in both total activity (b) as well as dynamics, as seen in (c) and (d).
+Figures (e-h) give a zoomed-in view of groups 1 and 3, showing that while the overall rate of (1,3) and (3,1) events is similar, the tendency for \texttt{ab-by} transitions differs.}
+\label{fig:parmats}
+\end{figure*}
+
+We additionally perform experiments that evaluate the likelihood of future data under our model.
+As our method jointly models \emph{which} dyads occur and \emph{when} they occur, any alternative method needs to also model these two aspects.
+In this vein, we extend the above baselines by assuming each dyad is a Poisson process with estimated  rate $$\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$$, where $r_b(i,j)$ is the statistic for a given baseline (described above) and $\xi=1$ is a smoothing parameter.
+In Table ~\ref{tab:results} we compare the loglikelihood of held-out test data under our proposed model to each of these baselines.
+In order to investigate the role of the number of clusters, we introduce an upper bound $K^*$ on the number of clusters during the fitting process.
+Note that the model with $K*=1$ is an important baseline -- it is simply a relational event model with no groups.
+There are a large number of other possible baselines that one might use, but most are not directly applicable to Table 2 as they do not work with events in continuous time.
+
+For the synthetic data example introduced in Section \ref{sec:simulation} the fitted models with $K>1$ have predictive performance comparable to the true model; a standard relational event model does not perform as well because it does not have the flexibility to model the dynamics of each block separately.
+For the classroom data set we see only a slight improvement with $K>1$, while with the mobile phone calls the $K=3$ model performs best.
+
+In Table~\ref{tab:recall20} we include the corresponding results for the recall experiment at a cutoff of 5 and 20, respectively.
+The results there are similar.
+As an example, using the Mobile data we are making a prediction of the next event among $89^2$ possible events: e.g., the next event is among the 5 highest-ranked edges 16.3 percent of the time by the $K*$=2 model.
+This is 1291 times more often than predicting events with our uniform baseline.
+Note that the absolute numbers for the baselines (e.g., zero in the 1st 3 decimal places for uniform) are so low because of the very large number of outcomes (e.g, $89^2$).
+For ranking, relative to all baselines including K*=1, on the synthetic data set and Enron there are small systematic improvements, while on the phone and University data there are large systematic improvements.
+Note that we used ranking performance (Table 3) to investigate the predictive power of the proposed models, as a function of $K*$, rather than only evaluating test log-likelihood; the latter is what the model was optimized for and thus the optimal $K*$ for Tables 2 and 3 need not agree.
+
+\section{Discussion}
+
+We have here introduced a family of relational event models that can flexibly capture heterogeneity in the underlying interaction dynamics.
+Our approach generalizes traditional, static notions of stochastic equivalence on nodes (such as stochastic blockmodels) to the dynamic context. 
+The proposed model family posits the existence of groups of nodes, such that all members of a group are governed by the same dynamic process, and groups are differentiated by having different dynamic processes.
+%PS: sentence below is incomplete (after the first "the")É..I decided to take it out anyway..
+% Given a specification for the, the proposed method employs latent variables to model the heterogeneity in the event dynamics.
+%is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
+%The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}).
+%This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.
+
+%This analysis suggests 
+Our approach has the ability to uncover systematic differences in dynamic behaviors among subsets of nodes, even in the absence of differences in marginal interaction rates.
+% across a variety of data sets.
+The analyses of Section \ref{sec:experiments} show that this model family has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
+In particular, our proposed approach leads to improved predictive accuracy when compared to both (a) relational event models that lack latent clusters, and (b) stochastic blockmodels for count data, both of which are special cases of our model family.
+Though prediction is not the main focus of our effort, these results provide evidence that having latent structure (i.e.
+$K > 1$) can lead to improvements in predictive power.
+However, our prediction results also show that it is easy to overfit with this model family.
+%If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model. 
+%{\red (TODO: confirm this)}
+ As $K$ increases beyond 2 (e.g., for the unrestrained CRP prior), these models are prone to overfit.
+This is not surprising, given that the number of fitted parameters scales as the square of the number of components $K$ in the model.
+A natural direction worth exploring for this family is a class of priors that are more resistant to overfitting (e.g., by imposing more structure on the parameters within each group).
+
+Another direction to explore is that of including different statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and using our approach to study how the roles of these statistics vary across nodes.
+One potentially interesting extension, analogous to \cite{Airoldi2008}, would be to allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.
+Allowing underlying dynamics to be governed by a latent membership structure opens the door to a wide range of possibilities, with many opportunities for further elaboration.
+
+%Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
+%One interesting extension, analogous to \cite{Airoldi2008}, would allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.
+We leave this to future work.
+%Alternatively, if nodes change groups.
+%In some contexts such extensions might be substantively important and warrant future work.
+
+%Our method learns about event dynamics within latent groups of individuals, but other types of heterogeneity likely exist in some data sets.  
+%For example, dynamics might change over time \cite{Vu2011}.
+
 %\subsubsection*{Acknowledgements}
+\newpage
+\newpage
 \bibliographystyle{unsrt}
 \bibliography{refs}
-%\newpage
-%\input{appendices}
+\newpage
+\textbf{Supplementary material}
+\input{appendices}
 \end{document}

### 2012-10-31 docs/submission.tex
Make novelty a bit more clear.

@@
 One can extend the above to marked point processes where each event contains additional information.
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
-We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
+In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} is defined as the set of all possible dyadic events.
+We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{i,j}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
 %The likelihood of an observed event history $\mathcal{A}$ extending from time 0 to the time of the final event, $t_M$, is 
@@
 \begin{align}
-\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
+\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{i,j}(t_m | \cdot) \}
 \label{eqn:llk}
@@
 Here, we aim to learn about the dynamics within and between subsets of nodes.
-To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
+To facilitate this, we propose a novel extension which assumes each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
 \begin{align*}
-\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
+\log \lambda_{i,j}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
 %z_i &\sim \mbox{CRP}(\alpha) &
@@
 \end{align*}
-where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
+where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{Normal}(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
 Thus, the rate of $(i,j)$ events has the same parameters as other events occurring between group $z_i$ and $z_j$.
@@
 % In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
-As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{ij}(t)$ to change following an event where $i$ was the sender or the recipient.
+As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{i,j}(t)$ to change following an event where $i$ was the sender or the recipient.
 This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
@@
 
-% TODO: Awkward next sentence
+{\color{red} TODO: Awkward next sentence...}
 The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
@@
 
-Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001} for static networks which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
+Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001} for static networks which models the probability of a dyad as $p(y_{i,j}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
 In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.
@@
 \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]} \ \   \times \\
-&\prod_{(i,j) \in \mathcal{U}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
+&\prod_{(i,j) \in \mathcal{U}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{i,j}(t_m|\cdot)\}
 \end{align*}
@@
 p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} \ \times \\
-& \ \ \prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
+& \ \ \prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{i,j}(t_m|\cdot)\}
 \end{align*}
@@
 As with other models, we note that the predictive accuracy of the model can depend on the hyperparameters.
+
+\subsection{Scalability}
+
+Likelihood computation essentially depends on the number of changepoints (i.e. knots) in all of the intensity functions being modeled.   
+We take advantage of our restriction on the types of statistics $\mathbf{s}$ to reduce the computational complexity of computing our likelihood  as
+\begin{align*}
+\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \times \\
+&  \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{m,i,j}) \lambda_{i,j}(t_m | \cdot) \}
+\end{align*}
+\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{m,i,j}$ is the time  of the changepoint for $\lambda_{i,j}(t|\cdot)$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs.%  \cite{Butts2008}.
+%\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}
+
+ By limiting the number changepoints, computing the likelihood  $p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta})$ for Gibbs sampling $z_r$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$, avoiding a factor of $N^2$.
+In practice, we precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j,\mathcal{A}_t)$ for all $m$, $i$, and $j$.
+This can be done in one pass through the data set.
+
+% The Supplementary Material mentions one attempt at speeding this up, for larger graphs, but additional assumptions/approximations are likely needed for scalability.
 
@@
 For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
-Data is generated by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
+Data is generated by sequentially computing $\lambda_{i,j}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{i,j} \lambda_{i,j}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{i,j}(t_m|\cdot) / \sum_{i,j}\lambda_{i,j}(t_m|\cdot))$.
 
@@
 Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
-We use a subset 4330 events among 487 users who participated in more than one event ($M_{test} = 1330$).\footnote{This dataset will be made publicly available.}
+We use a subset 4330 events among 487 users who participated in more than one event ($M_{test} = 1330$).\footnote{This data set will be made publicly available.}
 \item MIT Reality Mining \cite{Eagle2009}: 2000 phone calls among the
@@
 Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
+In this particular example we used the University data set and fit the proposed model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
+
 We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
-For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
 % The two panels on the left show the variation in $s_0(t_M,i,j,\mathcal{A}_t)$, $s_1(t_M,i,j,\mathcal{A}_t)$ and $s_3(t_M,i,j,\mathcal{A}_t)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
@@
 
-Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i) \sum_{m:t_m < t} \mathbf{I}(j_m=j)$.
-Finally, \texttt{BM} is a stochastic blockmodel (i.e. our model with only an intercept term).
+Several baselines are included for comparison: 
+\begin{enumerate}
+\item \texttt{uniform} places uniform probability on all possible dyads, 
+\item \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i,j_m=j)$, 
+\item \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i) \sum_{m:t_m < t} \mathbf{I}(j_m=j)$.  
+\item \texttt{BM} is a stochastic blockmodel (i.e. our model with only an intercept term).
+\end{enumerate}
 Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} and \texttt{BM} can capture individual heterogeneity and group level heterogeneity in overall activity.
@@
 As our method jointly models \emph{which} dyads occur and \emph{when} they occur, any alternative method needs to also model these two aspects.
-In this vein, we extend the above baselines by assuming each dyad is a Poisson process with estimated  rate $$\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$$, where $r_b(i,j)$ is the statistic for a given baseline (described above) and $\xi=1$ is a smoothing parameter.
+In this vein, we extend the above baselines by assuming each dyad is a Poisson process with estimated  rate $$\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{i,j} r_{b}(i,j) + \xi}$$, where $r_b(i,j)$ is the statistic for a given baseline (described above) and $\xi=1$ is a smoothing parameter.
 In Table ~\ref{tab:results} we compare the loglikelihood of held-out test data under our proposed model to each of these baselines.
@@
 \bibliography{refs}
-\newpage
-\textbf{Supplementary material}
-\input{appendices}
 \end{document}

### 2012-10-31 docs/submission.tex
Add a few extra sentences regarding Figure 2.

@@
 One can extend the above to marked point processes where each event contains additional information.
-In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} is defined as the set of all possible dyadic events.
-We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{i,j}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
+{\color{red} In Figure \ref{fig:example}a we illustrate the case of \emph{relational events}  occurring among $N$ nodes}, where each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} is defined as the set of all possible dyadic events.
+{\color{red} The study of human communication often involves such data (e.g.
+phone calls, online interaction, etc.) where the nodes represent people and each event represents one person speaking to another.}
+
+As in \cite{Butts} in the proposed approach we assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{i,j}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.
 %The likelihood of an observed event history $\mathcal{A}$ extending from time 0 to the time of the final event, $t_M$, is 
@@
 This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
-In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
+% In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
 
@@
 In order to investigate the role of the number of clusters, we introduce an upper bound $K^*$ on the number of clusters during the fitting process.
-Note that the model with $K*=1$ is an important baseline -- it is simply a relational event model with no groups.
+Note that the model with $K^*=1$ is an important baseline -- it is simply a relational event model with no groups.
 There are a large number of other possible baselines that one might use, but most are not directly applicable to Table 2 as they do not work with events in continuous time.
@@
 %\subsubsection*{Acknowledgements}
-\newpage
-\newpage
+
+\clearpage
 \bibliographystyle{unsrt}

### 2012-11-01 docs/submission.tex
Add some preliminary discussion regarding Figure 4.

@@
 \centering
-\includegraphics[scale=.45]{../figs/introexample/all}
+\includegraphics[scale=.5]{../figs/introexample/all}
 \caption{Illustration of differentiation by rate of interaction versus by dynamic behavior.
@@
 \centering
-  \begin{tikzpicture}[scale=.65,transform shape,thick]
+  \begin{tikzpicture}[scale=.8,transform shape,thick]
     \node[vertex] (n1) at (0,0)  {1};
@@
 Here, we aim to learn about the dynamics within and between subsets of nodes.
-To facilitate this, we propose a novel extension which assumes each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
+To facilitate this, we propose a novel extension which assumes each node $i$ belongs to latent cluster $z_i$ while still using a log linear model for the intensity functions
 \begin{align*}
@@
 
-{\color{red} TODO: Awkward next sentence...}
-The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
+The next set of statistics aim to capture various types of \emph{participation shift} that play a role conversational norms \cite{Gibson2003}.
 For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
+{\color{red} Another example is a \emph{turn-taking} effect where an event $(a,b)$ is followed by $b$ initiating an event with an individual other than $a$, denoted \texttt{ab-by}.
+These statistics are binary and simply indicate whether or not the event in question can be classified as an example of that particular transition. }
 Though only the statistics in Table \ref{tab:stats} are used in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
-The only restriction is that the statistic may not change in value between each observed event.%\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
+The only restriction within the proposed framework is that the statistic may not change in value between each observed event.%\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
 % TODO: Also, need to be restricted to ego conditions if we want it fast
@@
 \centering
-\includegraphics[width=1.5in]{../figs/synthetic/mat.pdf} %counts2.pdf
+\includegraphics[width=1.8in]{../figs/synthetic/mat.pdf} %counts2.pdf
 \vspace{.25cm}
@@
 \centering
-\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
+\includegraphics[width=2.4in]{../figs/synthetic/counts.pdf}
 \caption{}
@@
 \centering
-\includegraphics[width=2.1in]{../figs/synthetic/params-estimates.pdf}
+\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
 \vspace{-.5cm}
@@
 In this particular example we used the University data set and fit the proposed model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
+{\color{red} In Figure \ref{fig:parmats}a we show the observed number of times each dyadic event occurred over the course of the dataset.
+The rows and columns have been sorted according to block membership.
+In Figure \ref{fig:parmats}b-d we show estimates of the parameters corresponding to each of the statistics included in our model of the dyadic intensities $\lambda_{i,j}$.
+The intercept estimates, for example, reveal the general propensity at which two blocks tend to communicate.
+One salient feature of the parameter estimates is the symmetry in the block structure of the overall communication rates: if a member of group $A$ sends an email to a member of group $B$, the reverse occurs at a similar rate.
 
-We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
-% The two panels on the left show the variation in $s_0(t_M,i,j,\mathcal{A}_t)$, $s_1(t_M,i,j,\mathcal{A}_t)$ and $s_3(t_M,i,j,\mathcal{A}_t)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
-%The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
-% Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
- block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
+However, the asymmetries that the model reveals may often be of particular interest.
+An assymetry in intercept estimates between two blocks reveals that for a particular pair of groups that one tends to be the sender and the other tends to be the recipient.
+Assymetries for other effects are similarly interpretable.
+Consider interactions initiated by a member of block $B_1$ and directed to a member of block $B_2$.
+If the estimate for the $\texttt{ab-ba}$ parameter is larger for the set of $(B_1,B_2)$ interactions than for $(B_2,B_1)$ interactions, then the model suggests that those in block $B_2$ tend to respond to emails from those in $B_1$ than vice versa.
+
+Such asymmetric behavior exists in the University data between block 1 and block 3.
+Figure \ref{fig:parmats}e-h focuses on the interactions between members of block 1 to members of block 3.
+In this example the between-group events are for a similar number of events to the 179 events occurred originating from a member of group 3 and 178 originated from group 1.
+Under the model, the block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
+Though such assymetries are unsurprising from a sociological point of view, our model has allowed us to examine this tendency while accounting for the overall propensity for interaction.
 The model has learned that this structure exists in the data, and furthermore our inferences from the parameter estimates are sensible given the observed statistics.
-{\color{red} TODO: Elaborate on what this figure shows.}
-
+}
 \subsection{Prediction experiments}
@@
 Note that the absolute numbers for the baselines (e.g., zero in the 1st 3 decimal places for uniform) are so low because of the very large number of outcomes (e.g, $89^2$).
-For ranking, relative to all baselines including K*=1, on the synthetic data set and Enron there are small systematic improvements, while on the phone and University data there are large systematic improvements.
-Note that we used ranking performance (Table 3) to investigate the predictive power of the proposed models, as a function of $K*$, rather than only evaluating test log-likelihood; the latter is what the model was optimized for and thus the optimal $K*$ for Tables 2 and 3 need not agree.
+For ranking, relative to all baselines including $K^*$=1, on the synthetic data set and Enron there are small systematic improvements, while on the phone and University data there are large systematic improvements.
+Note that we used ranking performance (Table 3) to investigate the predictive power of the proposed models, as a function of $K^*$, rather than only evaluating test log-likelihood; the latter is what the model was optimized for and thus the optimal $K^*$ for Tables 2 and 3 need not agree.
 
@@
 A natural direction worth exploring for this family is a class of priors that are more resistant to overfitting (e.g., by imposing more structure on the parameters within each group).
+%  {\color{red} The choice of hierarchical structure would depend on the problem of interest, but such an approach could be of use when modeling multiple sequences simultaneously \cite{Arxiv?}.}
 
-Another direction to explore is that of including different statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and using our approach to study how the roles of these statistics vary across nodes.
+Another direction to explore is that of including different statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, modifying on the prior on each block's $\beta$ that induces some sparsity in the parameter estimates, and using our approach to study how the roles of these statistics vary across nodes.
 One potentially interesting extension, analogous to \cite{Airoldi2008}, would be to allow the latent class $z_i$ to be drawn from node-specific membership vectors $\pi_i$  after each change point.

### 2012-11-03 docs/submission.tex
Add cutom version of results tables and add new multicol headers.
Modify Figure 4 to focus on different clusters.
Add outline around clusters that we use for zoom in.

@@
 The intercept estimates, for example, reveal the general propensity at which two blocks tend to communicate.
-One salient feature of the parameter estimates is the symmetry in the block structure of the overall communication rates: if a member of group $A$ sends an email to a member of group $B$, the reverse occurs at a similar rate.
 
+One salient feature of the parameter estimates is the presence of both symmetries and assymmetries in the block structure of the overall communication rates.
+For example, if a member of group $A$ sends an email to a member of group $B$, the reverse occurs at a similar rate.
 However, the asymmetries that the model reveals may often be of particular interest.
@@
 
-\input{../figs/results-llk4.tex}
-\input{../figs/results-recall-all.tex}
+% Tables created by results.r, then hand edited a bit:
+% - Added verticle and horizontal lines to make it easier
+% - Removed synthetic results
+% - Added headers
+\input{results-llk4.tex}
+\input{results-recall-all.tex}
 
